---
title: "Sample_indexing"
author: "Fabian Roger"
date: "`r Sys.Date()`"
output: html_notebook
---

Loading libraries

```{r, message=FALSE}

library(dplyr)
library(tidyr)
library(readxl)
library(readr)
library(jsonlite)
library(Biostrings)
library(here)
library(clipr)

```

```{r}
sessionInfo()
```

# demultiplexing by index

I use the [pheniqs](https://biosails.github.io/pheniqs/) software for demultiplexing. 

> Galanti L, Shasha D, Gunsalus KC. 2021 Pheniqs 2.0: accurate, high-performance Bayesian decoding and confidence estimation for combinatorial barcode indexing. BMC Bioinformatics 22, 359. (doi:10.1186/s12859-021-04267-5)

It uses Phred adjusted maximum likelihood decoding
> A maximum likelihood decoder that directly estimates the decoding error probability from the base calling error probabilities provided by the sequencing platform. Abbreviated PAMLD.

We need to provide it with a configuration file that links the samples to the dual unique indices. 

## indices

We used the Nextera Indecies from Shinis Group for barcoding. However, different markes were barcoded with different sample arrangements, so that we first demultiplex by barcode and then assign samples. 

```{r}
shini_barcodes <- read_xlsx(here("Documents", "Shini_barcodes_4_plates.xlsx"))

shini_barcodes <- 
  shini_barcodes %>% 
  select(1,2) %>% 
  mutate(P5_P7 = gsub(".+_.+_(.+)", "\\1", `DNA Oligo Name`),
         Plate = rep(rep(1:4, each = 96),2)) %>% 
  mutate(Barcode = gsub("CAAGCAGAAGACGGCATACGAGAT(.+)GTCTCGTGGGCTCGG", "\\1", `DNA Sequence (5' -> 3')`)) %>% 
  mutate(Barcode = gsub("AATGATACGGCGACCACCGAGATCTACAC(.+)TCGTCGGCAGCGTC", "\\1", Barcode)) %>% 
  mutate(name = gsub("(.+_.+)_.+", "\\1", `DNA Oligo Name`)) %>% 
  select(-`DNA Sequence (5' -> 3')`, -`DNA Oligo Name`) %>%
  pivot_wider(names_from = P5_P7, values_from = Barcode) %>% 
  select(Plate, name, i5, i7) %>% 
  mutate(i7_rc = as.character(reverseComplement(DNAStringSet(i7)))) %>% 
  mutate(i5_rc = as.character(reverseComplement(DNAStringSet(i5)))) %>% 
  filter(Plate %in% c(1,2))
```



## configuration file 

phenix takes a json formatted configuration file. An example is given [here](https://github.com/biosails/pheniqs/blob/master/example/illumina_vignette/H7LT2DSXX_l01_sample_static.json) which we take as basis for modification 

```{r}
phenix_config <- read_json(here("Data", "iSeq", "20231128_FS10000965_18_BSB09423-0723", "phenix_config_file.json"))
```

## base configurations

Here we give the
+ input directory (where the files are)
+ output directory (where the files should be written to)
+ the 4 filenames (reads (R) and indices (I) for forward and reverse)
+ and the filename of the sample report

note that $output will be ignored if we write the reads to demultiplexed fastq files
```{r}
phenix_config$PM <- "unkown"

phenix_config$`base input url` <- here("Data", "iSeq", "20231128_FS10000965_18_BSB09423-0723")

phenix_config$`base output url` <- here("Data", "iSeq", "20231128_FS10000965_18_BSB09423-0723", "demultiplexed")

phenix_config$`flowcell id` <- "unkown"

#forward reads
phenix_config$input[[1]] <- "Undetermined_S0_L001_R1_001.fastq.gz"

#forward index
phenix_config$input[[2]] <- "Undetermined_S0_L001_I1_001.fastq.gz"

#reverse index
phenix_config$input[[3]] <- "Undetermined_S0_L001_I2_001.fastq.gz"

#reverse reads
phenix_config$input[[4]] <- "Undetermined_S0_L001_R2_001.fastq.gz"

#output
phenix_config$output[[1]] <- "Undetermined_S0_L001.bam"

#report
phenix_config$`report url` <- "sample_report.json"

```


#barcodes

Here we format the core of the configuration file, associating each sample with the unique barcode configuration

shini_barcodes is a dataframe containing the sample indicies, the P5 barcode, the P7 barcode and the reverseComplement of the barcode (in my case)

```{r}
my_config <- 
phenix_config$sample$codec[1:nrow(shini_barcodes)]

names(my_config) <- paste("@", shini_barcodes$name, sep ="")

for(i in names(my_config)){
  
  sample <- gsub("@", "", i)
  my_config[[i]]$LB <- sample
  my_config[[i]]$barcode[[1]] <- shini_barcodes[shini_barcodes$name == sample,]$i7_rc
  my_config[[i]]$barcode[[2]] <- shini_barcodes[shini_barcodes$name == sample,]$i5_rc
  
  my_config[[i]]$output[[1]] <- paste(sample, "_R1.fastq.gz", sep ="")
  my_config[[i]]$output[[2]] <- paste(sample, "_R2.fastq.gz", sep ="")
  
}

phenix_config$sample$codec <- my_config

phenix_config$sample$undetermined$output[[1]] <- "unclear_barcode_R1.fastq.gz"
phenix_config$sample$undetermined$output[[2]] <- "unclear_barcode_R2.fastq.gz"

#barcode length 10 bp
phenix_config$sample$transform$token[[1]] <- "1::10"
phenix_config$sample$transform$token[[2]] <- "2::10"
```

## Save configuration file
```{r}
path_to_file <- here("Data", "iSeq", "20231128_FS10000965_18_BSB09423-0723", "phenix_config_iSeq.json")

write_json(phenix_config, path_to_file, pretty = T, auto_unbox = TRUE)
```


##create output folder
```{r}
dir.create(phenix_config$`base output url`)
```


## run pheniqs
```{r}
system(paste("/Users/fabian/pheniqs_head_static/bin/static-HEAD/install/bin/pheniqs mux --config ",path_to_file, sep = ""))
```

## check report
```{r}
report <- read_json(here("Data", "iSeq", "20231128_FS10000965_18_BSB09423-0723", "demultiplexed", "sample_report.json"))
```

total number of reads
```{r}
report$outgoing$count
```


fraction of reads with valid indices
```{r}
report$outgoing$`pf fraction`
```


```{r}
report$sample$classified %>% 
  lapply(., as.data.frame) %>% 
  bind_rows() %>% 
  select(LB, BC, count) %>% 
  mutate(prct = signif(count/sum(count)*100, 2)) %>% 
  arrange(desc(count))
```

find barcodes
```{r}
I1 <- ShortRead::readFastq(here("Data", "iSeq", "20231128_FS10000965_18_BSB09423-0723", "Undetermined_S0_L001_I1_001.fastq.gz")) %>% ShortRead::sread()

I2 <- ShortRead::readFastq(here("Data", "iSeq", "20231128_FS10000965_18_BSB09423-0723", "Undetermined_S0_L001_I2_001.fastq.gz")) %>% ShortRead::sread()

BC <- tibble(I1 = as.character(I1),
             I2 = as.character(I2))
BC <- 
BC %>% 
  group_by(I1,I2) %>% 
  summarise(n = n()) %>% 
  arrange(desc(n)) %>% 
  ungroup() %>% 
  mutate(n_frac = round(n/sum(n) * 100,2))

BC %>% 
  left_join(shini_barcodes, by = c("I1" = "i5_rc", "I2" = "i7_rc"))
```

I1 --> i7_rc
```{r}
sum(shini_barcodes$i7_rc %in% BC$I1)
```

I2 --> i5_rc
```{r}
sum(shini_barcodes$i5_rc %in% BC$I2)
```


# demultiplexing by primer

We pooled four different primer pairs

1) COI Primer for Insects (30%)
2) ITS Primer for Fungi (30%)
3) 16S Primer for Bacteria (30%)
4) 16S Primer for Vertebrates (10%)

I will use cutadapt to sort the reads by primer into separate files


I use [cutadapt](https://cutadapt.readthedocs.io/en/stable/guide.html) to sort the reads by primer into different files.

setting up variables:
```{r}
dm_files <- list.files(here("Data", "iSeq", "20231128_FS10000965_18_BSB09423-0723","demultiplexed"),
                        full.names = F)

dm_files <- dm_files[!grepl("unclear_barcode", dm_files)]

forward_raw <- dm_files[grepl("_R1.fastq.gz", dm_files)]
reverse_raw <- dm_files[grepl("_R2.fastq.gz", dm_files)]

#path to sequences
path_demulti <- here("Data", "iSeq", "20231128_FS10000965_18_BSB09423-0723","demultiplexed")

#minimum length for sequences after trimming (shorter sequences are discarded)
minlength <- 100

#number of cores to use 
Cores <- 8
```



-g specifies that the primer adaptor is at the 5' end of the read.

> 5’ adapters preceed the sequence of interest


### Insects

COI-ArtF12_P5 : GCHCCHGAYATRGCHTTYCCNC
COI-ArtR22xls_P7: GRTCAAARAADGAWGTATTWARATTWCG


```{r}
Primer <- list()
Primer$Frwd <- "GCHCCHGAYATRGCHTTYCCNC" 
Primer$Rev <- "GRTCAAARAADGAWGTATTWARATTWCG" 

Primer$Frwd_rc <- as.character(reverseComplement(DNAStringSet(Primer$Frwd)))
Primer$Rev_rc <- as.character(reverseComplement(DNAStringSet(Primer$Rev)))
```

```{r}
#create directories for primer pair
Primer_path <- here("Data", "iSeq", "20231128_FS10000965_18_BSB09423-0723", "Insects")
dir.create(Primer_path)
```


```{r}

cutadapt_out <- vector(mode = "list", length = length(forward_raw))
names(cutadapt_out) <- gsub("(.+?)_R\\d.+", "\\1", forward_raw)

for (i in seq_along(forward_raw)) {
  
  message(paste("processing file", i, "out of", length(forward_raw), sep=" "))

# will sort in separate file below
temp <- 
system(
paste('cutadapt -a ', #replace by your path to cutadapt
      '"', Primer$Frwd,';required;min_overlap=12','"','...', Primer$Rev_rc,
      ' -A ', '"',Primer$Rev,';required;min_overlap=12','"', '...', Primer$Rev_rc,
      ' --discard-untrimmed',
      ' --minimum-length=', minlength, #minimum length for sequences after trimming
      ' --cores=', Cores, # Number of cores to use
      ' -o ', paste(Primer_path, '/', forward_raw[i], sep = ''),
      ' -p ', paste(Primer_path, '/', reverse_raw[i], sep = ''),
      paste(' ',path_demulti , '/', forward_raw[i], sep = ''),
      paste(' ', path_demulti, '/', reverse_raw[i], sep = ''),
      ' --report=minimal',
      sep = ''
      ), intern = TRUE
)

df <- read.table(text = temp)[2,]
colnames(df) <- read.table(text = temp)[1,]
cutadapt_out[[i]] <- df
}

cutadapt_out_df <- bind_rows(cutadapt_out, .id = "Index")

cutadapt_out_df$Marker <- "Insects"

cutadapt_out_df %>% 
  write_tsv(here("Data", "iSeq", "cutadapt_stats_inscects.txt"))

cutadapt_out_df
```



### Vertebrates

For vertebrates I used primers targeting the mitochondrial vertebrate 16S gene
The forward Primer has two versions, one targeting birds specifically

mam1 :   GGTTGGGGTGACCTCGGA
mam1_b : GGTTGGGGCGACCTTGGA
                 Y     Y
ve16S : AATCGTTGAACAAACGAAC


```{r}
Primer <- list()
Primer$Frwd <- "GGTTGGGGYGACCTYGGA" 
Primer$Rev <- "AATCGTTGAACAAACGAAC" 

Primer$Frwd_rc <- as.character(reverseComplement(DNAStringSet(Primer$Frwd)))
Primer$Rev_rc <- as.character(reverseComplement(DNAStringSet(Primer$Rev)))
```

```{r}
#create directories for primer pair
Primer_path <- here("Data", "iSeq", "20231128_FS10000965_18_BSB09423-0723", "Vertebrates")
dir.create(Primer_path)
```


```{r}

cutadapt_out <- vector(mode = "list", length = length(forward_raw))
names(cutadapt_out) <- gsub("(.+?)_R\\d.+", "\\1", forward_raw)

for (i in seq_along(forward_raw)) {
  
  message(paste("processing file", i, "out of", length(forward_raw), sep=" "))

# will sort in separate file below
temp <- 
system(
paste('cutadapt -a ', #replace by your path to cutadapt
      '"', Primer$Frwd,';required;min_overlap=12','"','...', Primer$Rev_rc,
      ' -A ', '"',Primer$Rev,';required;min_overlap=12','"', '...', Primer$Rev_rc,
      ' --discard-untrimmed',
      ' --minimum-length=', minlength, #minimum length for sequences after trimming
      ' --cores=', Cores, # Number of cores to use
      ' -o ', paste(Primer_path, '/', forward_raw[i], sep = ''),
      ' -p ', paste(Primer_path, '/', reverse_raw[i], sep = ''),
      paste(' ',path_demulti , '/', forward_raw[i], sep = ''),
      paste(' ', path_demulti, '/', reverse_raw[i], sep = ''),
      ' --report=minimal',
      sep = ''
      ), intern = TRUE
)

df <- read.table(text = temp)[2,]
colnames(df) <- read.table(text = temp)[1,]
cutadapt_out[[i]] <- df
}

cutadapt_out_df <- bind_rows(cutadapt_out, .id = "Index")

cutadapt_out_df$Marker <- "Vertebrates"

cutadapt_out_df %>% 
  write_tsv(here("Data", "iSeq", "cutadapt_stats_vertebrates.txt"))

cutadapt_out_df
```

### Bacteria

515F : GTGYCAGCMGCCGCGGTAA
806R : GGACTACNVGGGTWTCTAAT


```{r}
Primer <- list()
Primer$Frwd <- "GTGYCAGCMGCCGCGGTAA" 
Primer$Rev <- "GGACTACNVGGGTWTCTAAT" 

Primer$Frwd_rc <- as.character(reverseComplement(DNAStringSet(Primer$Frwd)))
Primer$Rev_rc <- as.character(reverseComplement(DNAStringSet(Primer$Rev)))
```

```{r}
#create directories for primer pair
Primer_path <- here("Data", "iSeq", "20231128_FS10000965_18_BSB09423-0723", "Bacteria")
dir.create(Primer_path)
```


```{r}

cutadapt_out <- vector(mode = "list", length = length(forward_raw))
names(cutadapt_out) <- gsub("(.+?)_R\\d.+", "\\1", forward_raw)

for (i in seq_along(forward_raw)) {
  
  message(paste("processing file", i, "out of", length(forward_raw), sep=" "))

# will sort in separate file below
temp <- 
system(
paste('cutadapt -a ', #replace by your path to cutadapt
      '"', Primer$Frwd,';required;min_overlap=12','"','...', Primer$Rev_rc,
      ' -A ', '"',Primer$Rev,';required;min_overlap=12','"', '...', Primer$Rev_rc,
      ' --discard-untrimmed',
      ' --minimum-length=', minlength, #minimum length for sequences after trimming
      ' --cores=', Cores, # Number of cores to use
      ' -o ', paste(Primer_path, '/', forward_raw[i], sep = ''),
      ' -p ', paste(Primer_path, '/', reverse_raw[i], sep = ''),
      paste(' ',path_demulti , '/', forward_raw[i], sep = ''),
      paste(' ', path_demulti, '/', reverse_raw[i], sep = ''),
      ' --report=minimal',
      sep = ''
      ), intern = TRUE
)

df <- read.table(text = temp)[2,]
colnames(df) <- read.table(text = temp)[1,]
cutadapt_out[[i]] <- df
}

cutadapt_out_df <- bind_rows(cutadapt_out, .id = "Index")

cutadapt_out_df$Marker <- "Bacteria"

cutadapt_out_df %>% 
  write_tsv(here("Data", "iSeq", "cutadapt_stats_bacteria.txt"))

cutadapt_out_df
```

### Fungi

```{r}
ITS_primer <- read_xlsx(here("Data", "Fungal_primer_Schlegel.xlsx"))
```

we sequence wit a mix of primers, we specify the adapters in separate files:

```{r}
ITS_fwrd <- 
ITS_primer %>% 
  filter(Primer_dir == "Fwrd") %>% 
  pull(`primer sequence`) %>% 
  unique() %>% 
  DNAStringSet()

fwrd_search <- 
paste(ITS_fwrd, ";required;min_overlap=12","...",reverseComplement(ITS_fwrd), sep = "")

paste(">primer", 1:length(fwrd_search), sep = "_") %>% 
  rbind(., fwrd_search) %>% 
  c() %>% 
  write_lines(here("Data", "ITS_Primer_fwrd.fasta"))


ITS_rev <- 
ITS_primer %>% 
  filter(Primer_dir == "Rev") %>% 
  pull(`primer sequence`) %>% 
  unique() %>% 
  DNAStringSet()

rev_search <- 
paste(ITS_rev, ";required;min_overlap=12","...",reverseComplement(ITS_rev), sep = "")

paste(">primer", 1:length(rev_search), sep = "_") %>% 
  rbind(., rev_search) %>% 
  c() %>% 
  write_lines(here("Data", "ITS_Primer_rev.fasta"))
```

```{r}
#create directories for primer pair
Primer_path <- here("Data", "iSeq", "20231128_FS10000965_18_BSB09423-0723", "Fungi")
dir.create(Primer_path)
```


```{r}

cutadapt_out <- vector(mode = "list", length = length(forward_raw))
names(cutadapt_out) <- gsub("(.+?)_R\\d.+", "\\1", forward_raw)

for (i in seq_along(forward_raw)) {
  
  message(paste("processing file", i, "out of", length(forward_raw), sep=" "))

# will sort in separate file below
temp <- 
system(
paste('cutadapt -a ', #replace by your path to cutadapt
      '"', 
      'file:',here("Data", "ITS_Primer_fwrd.fasta"),
      '"',
      ' -A ', 
      '"',
      'file:', here("Data", "ITS_Primer_rev.fasta"),
      '"', 
      ' --discard-untrimmed',
      ' --minimum-length=', minlength, #minimum length for sequences after trimming
      ' --cores=', Cores, # Number of cores to use
      ' -o ', paste(Primer_path, '/', forward_raw[i], sep = ''),
      ' -p ', paste(Primer_path, '/', reverse_raw[i], sep = ''),
      paste(' ',path_demulti , '/', forward_raw[i], sep = ''),
      paste(' ', path_demulti, '/', reverse_raw[i], sep = ''),
      ' --report=minimal',
      sep = ''
      ), intern = TRUE
)

df <- read.table(text = temp)[2,]
colnames(df) <- read.table(text = temp)[1,]
cutadapt_out[[i]] <- df
}

cutadapt_out_df <- bind_rows(cutadapt_out, .id = "Index")

cutadapt_out_df$Marker <- "Fungi"

cutadapt_out_df %>% 
  write_tsv(here("Data", "iSeq", "cutadapt_stats_fungi.txt"))

cutadapt_out_df
```

### Plants

ITS2F : ATGCGATACTTGGTGTGAAT
ITS2R : GACGCTTCTCCAGACTACAAT


```{r}
Primer <- list()
Primer$Frwd <- "ATGCGATACTTGGTGTGAAT" 
Primer$Rev <- "GACGCTTCTCCAGACTACAAT" 

Primer$Frwd_rc <- as.character(reverseComplement(DNAStringSet(Primer$Frwd)))
Primer$Rev_rc <- as.character(reverseComplement(DNAStringSet(Primer$Rev)))
```

```{r}
#create directories for primer pair
Primer_path <- here("Data", "iSeq", "20231128_FS10000965_18_BSB09423-0723", "Plants")
dir.create(Primer_path)
```


```{r}

cutadapt_out <- vector(mode = "list", length = length(forward_raw))
names(cutadapt_out) <- gsub("(.+?)_R\\d.+", "\\1", forward_raw)

for (i in seq_along(forward_raw)) {
  
  message(paste("processing file", i, "out of", length(forward_raw), sep=" "))

# will sort in separate file below
temp <- 
system(
paste('cutadapt -a ', #replace by your path to cutadapt
      '"', Primer$Frwd,';required;min_overlap=12','"','...', Primer$Rev_rc,
      ' -A ', '"',Primer$Rev,';required;min_overlap=12','"', '...', Primer$Rev_rc,
      ' --discard-untrimmed',
      ' --minimum-length=', minlength, #minimum length for sequences after trimming
      ' --cores=', Cores, # Number of cores to use
      ' -o ', paste(Primer_path, '/', forward_raw[i], sep = ''),
      ' -p ', paste(Primer_path, '/', reverse_raw[i], sep = ''),
      paste(' ',path_demulti , '/', forward_raw[i], sep = ''),
      paste(' ', path_demulti, '/', reverse_raw[i], sep = ''),
      ' --report=minimal',
      sep = ''
      ), intern = TRUE
)

df <- read.table(text = temp)[2,]
colnames(df) <- read.table(text = temp)[1,]
cutadapt_out[[i]] <- df
}

cutadapt_out_df <- bind_rows(cutadapt_out, .id = "Index")

cutadapt_out_df$Marker <- "Plants"

cutadapt_out_df %>% 
  write_tsv(here("Data", "iSeq", "cutadapt_stats_plants.txt"))

cutadapt_out_df
```

# count PhiX

```{r}
dir.create(here("Data", "iSeq", "bbduk_stats"))

lapply(seq_along(forward_raw), function(x)
{

  system(paste("/Applications/bbmap/bbduk.sh in1=",
               paste(path_demulti , '/', forward_raw[x], sep = ''),
               " in2=",
               paste(path_demulti , '/', reverse_raw[x], sep = ''),
               " stats=",
               paste(here("Data", "iSeq", "bbduk_stats"),
                     paste("bbduk", x, ".txt", sep = ""), sep = "/"),
               " maxns=0",
               " ref=/Applications/bbmap/resources/phix174_ill.ref.fa.gz k=20 hdist=2",
               " threads=6",
                sep = ""
               ), intern = TRUE)

})
```


```{r}

bbduk_files <- list.files(here("Data", "iSeq", "bbduk_stats"), full.names = TRUE)

phix = lapply(bbduk_files, function(x){
  temp <- read_lines(x)
  strsplit(temp[3], "\t")[[1]][2]
}) %>% unlist() %>% as.numeric

bbduk_stat <- tibble(Index = gsub("(.+?)_R\\d.+", "\\1", forward_raw),
                     phiX = phix)

```

#size distribution

```{r}

Adapter_cont <- 
lapply(forward_raw, function(x){
  
  ShortRead::readFastq(paste(path_demulti, x, sep = "/")) %>% 
    ShortRead::sread(test) %>% 
    as.character() %>% 
    grepl("GGGGG+$", .) %>% 
    sum()
  
})

names(Adapter_cont) <- gsub("(.+?)_R\\d.+", "\\1", forward_raw)

Adapter_cont_df <- 
  Adapter_cont %>% bind_rows(.id = "Index") %>% 
  pivot_longer(cols = everything(), names_to = "Index", values_to = "Adapter_reads")

Adapter_cont_df %>% 
write_tsv(here("Data", "iSeq", "Adapter_stats.txt"))
```

# distribution of reads

```{r}
ca_insect <- read_tsv(here("Data", "iSeq", "cutadapt_stats_inscects.txt"))

ca_vert <- read_tsv(here("Data", "iSeq", "cutadapt_stats_vertebrates.txt"))

ca_bac <- read_tsv(here("Data", "iSeq", "cutadapt_stats_bacteria.txt")) 

ca_fungi <- read_tsv(here("Data", "iSeq", "cutadapt_stats_fungi.txt"))

ca_plants <- read_tsv(here("Data", "iSeq", "cutadapt_stats_plants.txt"))

Adapter_cont_df <- read_tsv(here("Data", "iSeq", "Adapter_stats.txt")) %>% 
  mutate(Marker = "Adapter") %>% 
  dplyr::rename(out_reads = Adapter_reads)

read_stat <- 
list(ca_insect, ca_vert, ca_bac, ca_fungi, ca_plants, Adapter_cont_df) %>% 
  bind_rows() %>% 
  select(Index, out_reads, Marker) %>% 
  pivot_wider(names_from = Marker, values_from = out_reads) %>% 
  left_join(select(ca_bac, Index, in_reads, too_short)) %>% 
  left_join(bbduk_stat) %>% 
  mutate(accounted = (Insects + Vertebrates + Bacteria + Fungi + Plants + phix + Adapter)) %>% 
  mutate(missing_prct = 100-round(accounted/in_reads*100,1))

write_tsv(read_stat, here("Data", "iSeq", "readStat.txt"))
```

```{r}
read_stat <- read_tsv(here("Data", "iSeq", "readStat.txt"))
```

```{r}
read_stat %>% 
  mutate(unaccounted = in_reads - accounted) %>% 
  select(-accounted, -missing_prct) %>% 
  pivot_longer(!one_of(c("Index", "in_reads")),
               names_to = "Read_type", values_to = "reads") %>% 
  group_by(Read_type) %>% 
  summarise(reads = sum(reads, na.rm = TRUE)) %>% 
  mutate(reads = reads/sum(reads)*100) %>% 
  mutate(Read_type = factor(Read_type, levels = c("Bacteria", "Fungi", 
                                                  "Insects", "Vertebrates", "Plants", "phiX", "Adapter", "too_short", "unaccounted"))) %>% 
  ggplot(aes(x = 1, y = reads, fill = Read_type))+
  geom_bar(position = "stack", stat = "identity")+
  scale_fill_brewer(palette = "Set1")+
  theme_bw()
```
```{r}

read_stat %>% 
  mutate(unaccounted = in_reads - accounted) %>% 
  select("Insects", "Vertebrates", "Bacteria", "Fungi", 
         "Plants", "Adapter", "too_short", "unaccounted", "phiX") %>% 
  pivot_longer(everything(), values_to = "reads") %>% 
  group_by(name) %>% 
  summarise(reads = sum(reads, na.rm = TRUE)) %>% 
  mutate(prct = round(reads/sum(reads)*100,2)) %>% 
  arrange(desc(prct)) %>% write_clip()

```
# read lengths

```{r}
readsR1 <-  ShortRead::readFastq(paste(path_demulti, forward_raw[1], sep = "/")) %>% 
  ShortRead::sread()

readsR2 <-  ShortRead::readFastq(paste(path_demulti, forward_raw[2], sep = "/")) %>% 
  ShortRead::sread()

readsR2 %>% as.character() %>% gsub("(.+?)[GA]{5}[GA]+$","\\1", .) %>% nchar() %>% 
table() %>% 
  as.data.frame() %>%
  dplyr::rename(Var1 = 1) %>% 
  mutate(across(one_of("Var1", "Freq"), function(x) as.numeric(as.character(x)))) %>% 
  arrange(desc(Freq)) %>% 
  mutate(prct = Freq / sum(Freq)*100) %>% 
  mutate(rank = 1:n()) %>% 
  mutate(Label = case_when( rank < 6 ~ paste(Var1, "(",round(prct,1), " %",")", sep = ""),
                   TRUE ~ NA_character_)) %>% 
    ggplot(aes(x = Var1, y = Freq))+
  geom_line(colour = "darkgrey", alpha = 0.5)+
  geom_text_repel(aes(label = Label))+
  geom_point(data = ~filter(.x, !is.na(Label)), colour = "red", size = 0.5)+
  theme_bw()+
  scale_x_continuous(limits = c(0,500))
```

```{r}
temp <- 
system(
paste('cutadapt -g ', #replace by your path to cutadapt
      '"','GGGGGGGGGG',';min_overlap=10','"',
      ' --discard-untrimmed',
      ' --minimum-length=', 0, #minimum length for sequences after trimming
      ' --cores=', Cores, # Number of cores to use
      ' -o ', paste(path_demulti, '/', "test1.fastq", sep = ''),
      ' -p ', paste(path_demulti, '/', "test2.fastq", sep = ''),
      paste(' ',path_demulti , '/', forward_raw[1], sep = ''),
      paste(' ', path_demulti, '/', reverse_raw[1], sep = ''),
      ' --report=minimal',
      sep = ''
      ), intern = TRUE
)


df <- read.table(text = temp)[2,]
colnames(df) <- read.table(text = temp)[1,]

df

readsR1 <-  ShortRead::readFastq(paste(path_demulti, '/', "test1.fastq", sep = '')) %>% 
  ShortRead::sread()

{150-width(readsR1)} %>% 
table() %>% 
  as.data.frame() %>%
  dplyr::rename(Var1 = 1) %>% 
  mutate(across(one_of("Var1", "Freq"), function(x) as.numeric(as.character(x)))) %>% 
  arrange(desc(Freq)) %>% 
  mutate(prct = Freq / sum(Freq)*100) %>% 
  mutate(rank = 1:n()) %>% 
  mutate(Label = case_when( rank < 6 ~ paste(Var1, "(",round(prct,1), " %",")", sep = ""),
                   TRUE ~ NA_character_)) %>% 
    ggplot(aes(x = Var1, y = Freq))+
  geom_line(colour = "darkgrey", alpha = 0.5)+
  geom_text_repel(aes(label = Label))+
  geom_point(data = ~filter(.x, !is.na(Label)), colour = "red", size = 0.5)+
  theme_bw()+
  scale_x_continuous(limits = c(0,500))
```

# unassigned reads

```{r}


system(paste("/Applications/bbmap/bbduk.sh in1=",
               paste(path_demulti , '/', "unclear_barcode_R1.fastq.gz", sep = ''),
               " in2=",
               paste(path_demulti , '/', "unclear_barcode_R2.fastq.gz", sep = ''),
               " stats=",
               paste(here("Data", "iSeq", "bbduk_stats"),
                     paste("bbduk", "_unassigned_phix", ".txt", sep = ""), sep = "/"),
               " maxns=0",
               " ref=/Applications/bbmap/resources/phix174_ill.ref.fa.gz k=20 hdist=2",
               " threads=6",
                sep = ""
               ), intern = TRUE)
 paste(here("Data", "iSeq", "bbduk_stats"),
                     paste("bbduk", "_unassigned_phix", ".txt", sep = ""), sep = "/")

read_tsv(here("Data","iSeq","bbduk_stats","bbduk_unassigned_phix.txt"), col_names = FALSE) %>% 
  dplyr::slice(2:3) %>% 
  select(1:2) %>% 
  mutate(X1 = gsub("#", "", X1)) %>% 
  mutate(X2 = as.numeric(X2)) %>% 
  pivot_wider(names_from = 1, values_from = 2) %>% 
  mutate(phix_prct = round(Matched / Total * 100,2))


```
```{r}
UA_R1 <-  ShortRead::readFastq(paste(path_demulti, "unclear_barcode_R1.fastq.gz", sep = "/")) %>% 
  ShortRead::sread()

Adapter_cont_UA <- 
  UA_R1 %>% 
    as.character() %>% 
    grepl("GGGGGGG+$", .) %>% 
    sum()
  

```

#read distribution samples

```{r}

gURL <- "https://docs.google.com/spreadsheets/d/1GKWzD0q683oH3I_i5ueulbf4P4Q-9-a9YPbdyUZyF2A/edit#gid=847966271"

Meta <- read_sheet(gURL, sheet = "Sample_data")

shini_barcodes <- 
  read_xlsx(here("Documents", "Shini_barcodes_4_plates.xlsx")) %>% 
  mutate(Plate = rep(rep(paste("Plate", 1:4, sep = "_"), each = 96),2)) %>% 
  filter(Plate %in% c("Plate_1", "Plate_2")) %>% 
  dplyr::rename(Index = 1) %>% 
  select(Index, Plate) %>% 
  filter(grepl("_i5$", Index)) %>% 
  mutate(Index = gsub("(.+?)_i5", "\\1", Index)) %>% 
  group_by(Plate) %>% 
  mutate(row = rep(LETTERS[1:8], each = 12)) %>% 
  mutate(col = as.character(rep(1:12, 8)))
```

## Insects

```{r}
Plates_I <- read_sheet(gURL, sheet = "Plates", range = "A1:M21", col_types = "c") %>% 
  dplyr::rename(row = 1) %>% 
  filter(row %in% LETTERS[1:8]) %>% 
  mutate(Plate = rep(c("Plate_1", "Plate_2"), each = 8)) %>% 
  pivot_longer(!one_of(c("row", "Plate")), names_to = "col", values_to = "Sample_names") %>% 
  select(Plate, row, col, Sample_names)
```

valid reads
```{r}
ASV_sp <- read_tsv(here("Data","iSeq", "COI_ASV_glom.txt"))
COI_tax_glom <- read_tsv(here("Data","iSeq", "COI_taxa_80_glom.txt"))

Arth_ASV <- COI_tax_glom %>% filter(phylum == "Arthropoda") %>% pull(seq)

Valid_arth <- 
ASV_sp %>% 
  select(1, one_of(Arth_ASV)) %>% 
  pivot_longer(-Sample) %>% 
  group_by(Sample) %>% 
  summarise(reads_arth = sum(value)) %>% 
  dplyr::rename(Index = Sample)
```


```{r}
insect_reads <- 
shini_barcodes %>% 
  left_join(Plates_I) %>% 
  left_join(select(ca_insect, Index, out_reads)) %>% 
  left_join(Valid_arth)

write_tsv(insect_reads, here("Data", "iSeq", "readsN_insects.txt"))

clipr::write_clip(insect_reads)

ggplot(insect_reads, aes(x = out_reads, y = reads_arth))+
  geom_point()
```


## Vertebrates

(same plate layout as insects)

valid reads
```{r}
ASV_sp <- read_tsv(here("Data",  "iSeq", "20231128_FS10000965_18_BSB09423-0723", "Vertebrates", "Vert_ASV_sp.text"))

Vert_tax_glom <- read_tsv(here("Data",  "iSeq", "20231128_FS10000965_18_BSB09423-0723", "Vertebrates", "Vert_tax_sp.txt"))

Vert_ASV <- Vert_tax_glom %>% filter(phylum == "Chordata") %>% pull(seq)

Valid_vert <- 
ASV_sp %>% 
  select(1, one_of(Vert_ASV)) %>% 
  pivot_longer(-Sample) %>% 
  group_by(Sample) %>% 
  summarise(reads_vert = sum(value)) %>% 
  dplyr::rename(Index = Sample)
```

```{r}
vert_reads <- 
shini_barcodes %>% 
  left_join(Plates_I) %>% 
  left_join(select(ca_vert, Index, out_reads)) %>% 
  left_join(Valid_vert)

write_tsv(vert_reads, here("Data", "iSeq", "readsN_vert.txt"))

clipr::write_clip(vert_reads)

ggplot(vert_reads, aes(x = out_reads, y = reads_vert))+
  geom_point()
```





## Plants


it looks like the short forward read alone of the plant marker is not diagnostic? 


valid reads
```{r}
ASV_sp <- read_tsv(here("Data","iSeq", "Plants_ASW_glom.txt"))
Plants_tax <- read_tsv(here("Data","iSeq", run_name, primer, "Plants_Sintax_80.txt"))

Plants_tax$phylum %>% table()
```

all sequences are identified as plants

```{r}
Valid_plants <- 
ASV_sp %>% 
  pivot_longer(-Sample) %>% 
  group_by(Sample) %>% 
  summarise(reads_plants = sum(value)) %>% 
  dplyr::rename(Index = Sample)
```

```{r}
plant_reads <- 
shini_barcodes %>% 
  left_join(Plates_I) %>% 
  left_join(select(ca_plants, Index, out_reads)) %>% 
  left_join(Valid_plants)

write_tsv(plant_reads, here("Data", "iSeq", "readsN_plants.txt"))

clipr::write_clip(plant_reads)
```

```{r}
ggplot(plant_reads, aes(x = out_reads, y = reads_plants))+
  geom_point()+
  geom_abline(slope = 1)
```


## Bacteria

```{r}
Plates_B <- read_sheet(gURL, sheet = "Plates", range = "A48:M66", col_types = "c") %>% 
  dplyr::rename(row = 1) %>% 
  filter(row %in% LETTERS[1:8]) %>% 
  mutate(Plate = rep(c("Plate_1", "Plate_2"), each = 8)) %>% 
  pivot_longer(!one_of(c("row", "Plate")), names_to = "col", values_to = "Sample_names") %>% 
  select(Plate, row, col, Sample_names)
```



valid reads
```{r}
ASV_sp <- read_tsv(here("Data","iSeq", "Bac_ASW_glom.txt"))
Bac_tax_glom <- read_tsv(here("Data","iSeq", "Bac_taxa_glom.txt"))

Bac_ASV <- Bac_tax_glom %>% filter(Kingdom == "Bacteria") %>% pull(seq)

Valid_bac <- 
ASV_sp %>% 
  select(1, one_of(Bac_ASV)) %>% 
  pivot_longer(-Sample) %>% 
  group_by(Sample) %>% 
  summarise(reads_bac = sum(value)) %>% 
  dplyr::rename(Index = Sample)
```

```{r}
bac_reads <- 
shini_barcodes %>% 
  left_join(Plates_B) %>% 
  left_join(select(ca_bac, Index, out_reads)) %>% 
  left_join(Valid_bac)

write_tsv(bac_reads, here("Data", "iSeq", "readsN_bacteria.txt"))

clipr::write_clip(bac_reads)

ggplot(bac_reads, aes(x = out_reads, y = reads_bac))+
  geom_point()
```

## Fungi

```{r}
Plates_F <- read_sheet(gURL, sheet = "Plates", range = "A26:M44", col_types = "c") %>% 
  dplyr::rename(row = 1) %>% 
  filter(row %in% LETTERS[1:8]) %>% 
  mutate(Plate = rep(c("Plate_1", "Plate_2"), each = 8)) %>% 
  pivot_longer(!one_of(c("row", "Plate")), names_to = "col", values_to = "Sample_names") %>% 
  select(Plate, row, col, Sample_names) 
```

for Fungi, the library was teh same on both sequencing runs. We can compare the results

```{r}
fungi_iSeq1 <- read_tsv(here("Data", "iSeq",
                             "20230913_FS10000965_17_BSB09423-1124",
                             "Fungi",
                             "ee_filter_stats.txt")) %>% 
  dplyr::rename(Index  = Sample_name, iSeq1 = reads.in ) %>% 
  select(Index, iSeq1)
```


valid reads
```{r}
ASV_sp <- read_tsv(here("Data","iSeq", "COI_ASV_glom.txt"))
COI_tax_glom <- read_tsv(here("Data","iSeq", "COI_taxa_80_glom.txt"))

Arth_ASV <- COI_tax_glom %>% filter(phylum == "Arthropoda") %>% pull(seq)

Valid_arth <- 
ASV_sp %>% 
  select(1, one_of(Arth_ASV)) %>% 
  pivot_longer(-Sample) %>% 
  group_by(Sample) %>% 
  summarise(reads_bac = sum(value)) %>% 
  dplyr::rename(Index = Sample)
```

```{r}
fungi_reads <- 
shini_barcodes %>% 
  left_join(Plates_F) %>% 
  left_join(select(ca_fungi, Index, out_reads)) %>% 
  left_join(fungi_iSeq1)

write_tsv(fungi_reads, here("Data", "iSeq", "readsN_fungi.txt"))

clipr::write_clip(fungi_reads)

```


```{r}
fungi_reads %>% 
  left_join(fungi_iSeq1) %>% 
  ggplot(aes(x = out_reads,y = iSeq1))+
  geom_point()+
  geom_smooth(method = "lm")+
  theme_minimal()+
  labs(x = "reads in first run", y = "reads in second run", title = "Fungi")
```

```{r}
fungi_reads %>% 
  left_join(fungi_iSeq1) %>% 
  filter(out_reads < 100)
```
Fungal reads and taxa run 1

```{r}
ASV_sp <- read.table(here("Data","iSeq", "20230913_FS10000965_17_BSB09423-1124", "Fungi", "ASV_swarm.text"))

Fungi_tax <- read_tsv(here("Data","iSeq", "20230913_FS10000965_17_BSB09423-1124", "Fungi", "ITS_Sintax_80.txt"))

ASV_sp %>% 
  pivot_longer(-Sample, names_to = "seq", values_to = "reads") %>% 
  left_join(Fungi_tax) %>% 
  select(1:4) %>% 
  mutate(kingdom = case_when(kingdom != "Fungi" | is.na(kingdom) ~ "other",
                             TRUE ~ kingdom)) %>% 
  group_by(Sample, kingdom) %>% 
  summarise(reads = sum(reads)) %>% 
  pivot_wider(names_from = kingdom, values_from = reads) %>%
  mutate(total = sum(Fungi + other)) %>% 
  dplyr::rename(Index = Sample) %>% 
  left_join(fungi_reads,.) %>% 
  clipr::write_clip() 
  ggplot(aes(x = total, y = Fungi))+
  geom_point()+
  geom_abline(slope = 1)

```

