---
title: "Bead-beating_sim"
author: "Fabian Roger"
date: '2022-05-06'
output: html_document
---
#libraries
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
```

#Simulations
here I simulate how relative abundance change with bead beating, both 

1) within a sample
2) across samples

For samples of 

1) constant composition
2) variable composition

The background for the simulation is that we will extract DNA from filters for which we would like to analyse different target groups which differ in their 'sturdiness' (Bacteria, Fungi, Plants, Metazoan cells)

For the simulations I assume we have 5 species in equal abundances but different extraction efficiencies - depending on the amount of bead beating

```{r}

n <- 5

extract_bump = function(eff, fact){
  res <-  eff*fact
  res[res>=100] <- 100
  return(res)
}


Sample_df <- 
  tibble(Spec = LETTERS[1:n],
         abund = rep(10000/n, n),
         bead_0_eff = exp(seq(1,log(100), length.out = n)),
         bead_1_eff = extract_bump(bead_0_eff,2),
         bead_2_eff = extract_bump(bead_0_eff,5),
         bead_3_eff = extract_bump(bead_0_eff,10),
         bead_4_eff = extract_bump(bead_0_eff,100)
         )

```


## efficiency
bead beating efficiency per species

```{r}
Sample_df %>% 
  pivot_longer(!one_of(c("Spec", "abund"))) %>% 
  ggplot(aes(x = name, y = value, colour = Spec))+
  geom_point()+
  geom_line(aes(group = Spec))+
  theme_bw()+
  scale_colour_viridis_d()
```

## abs abundance
absolute abundance of the DNA after extraction
```{r}
Sample_df %>% 
  mutate(org = abund) %>% 
  mutate(across(starts_with("bead"), ~.x*(abund/100))) %>% 
  pivot_longer(!one_of("Spec", "abund")) %>% 
  ggplot(aes(x = name, y = value, fill = Spec))+
  geom_bar(stat = "identity", position = "stack")+
  theme_bw()+
  scale_fill_viridis_d()
  
```

## rel abundance
relative abundance of the DNA after extraction
```{r}
Sample_df %>% 
  mutate(org = abund) %>% 
  mutate(across(starts_with("bead"), ~.x*(abund/100))) %>% 
  pivot_longer(!one_of("Spec", "abund")) %>%
  group_by(name) %>% 
  mutate(value = value / (sum(value))) %>% 
  ggplot(aes(x = name, y = value, fill = Spec))+
  geom_bar(stat = "identity", position = "stack")+
  theme_bw()+
  scale_fill_brewer(palette = "Dark2")+
  scale_fill_viridis_d()
  
  
```

# varying composition
For absol abundance across sample with varying composition
```{r}

Sample_df %>% 
  mutate(abund_1 = 10000/2^(0:(n-1)),
         abund_2 = abund_1[n():1]) %>% 
  pivot_longer(starts_with("abund"), names_to = "Sample", values_to = "abund") %>% 
  pivot_longer(starts_with("bead"), names_to = "Treatment", values_to = "efficiency") %>% 
  mutate(value = abund * efficiency/100) %>% 
  group_by(Treatment, Sample) %>% 
  mutate(value = value / sum(value)) %>% 
  ggplot(aes(x = Treatment, y = value, fill = Spec))+
  geom_bar(stat = "identity", position = "stack")+
  facet_wrap(~Sample)+
  theme_bw()+
  scale_fill_brewer(palette = "Dark2")+
  scale_fill_viridis_d()+
  theme(axis.text.x = element_text(angle = 90))

```

```{r}
Sample_df %>% 
  mutate(abund_1 = 10000/2^(0:(n-1)),
         abund_2 = abund_1[n():1]) %>% 
  pivot_longer(starts_with("abund"), names_to = "Sample", values_to = "abund") %>% 
  pivot_longer(starts_with("bead"), names_to = "Treatment", values_to = "efficiency") %>% 
  mutate(value = abund * efficiency/100) %>% 
  group_by(Treatment, Sample) %>% 
  mutate(value = value / sum(value)) %>% 
  ggplot(aes(x = Sample, y = value, fill = Spec))+
  geom_bar(stat = "identity", position = "stack")+
  facet_wrap(~Treatment)+
  theme_bw()+
  scale_fill_brewer(palette = "Dark2")+
  scale_fill_viridis_d()+
  theme(axis.text.x = element_text(angle = 90))
```

