---
title: "Fungi data analysis"
output: html_notebook
---

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(forcats)
library(here)
library(readr)
library(ggplot2)
library(vegan)
library(readxl)
library(corrplot)
library(Biostrings) 
library(decontam)
library(microDecon)
#library(fungaltraits)
```

#Set-up

##define Marker

```{r}
primer <- "Fungi"
```

##load meta-data

```{r}
#load metadata
Meta <- read_tsv(here("Data", "Meta_data.tsv"))
```

#load data
```{r}

Fungi_ASV <- read_tsv(here("Data", primer, "out_final", "Fungi_ASV_glom.txt")) %>% 
  mutate(Sample = gsub(".+?(NGS.+)", "\\1", Sample)) %>% 
  dplyr::rename(Index = Sample)

Fungi_tax <- read_tsv(here("Data", primer, "out_final", "Fungi_taxa_80_glom.txt"))

Fungi_seq <- readDNAStringSet(here("Data", primer, "out_final", paste(primer, "Seq_glom", sep = "_")))

DNAC <- read_tsv(here("Data", "DNA_conc_air_sampler_comp.txt"))
```

#fragment lenth distribution

```{r}
Fungi_len <- 
Fungi_seq %>% 
  width() %>% 
  tibble(len = .) %>% 
  mutate(seq = names(Fungi_seq)) 

Fungi_len %>% 
  group_by(len) %>% 
  summarise(n = n()) %>% 
  ggplot(aes(x = len, y = n))+
  geom_bar(stat = "identity")


#Fungi_len_reads <- 
Fungi_ASV %>% 
  pivot_longer(-Index, names_to = "seq") %>% 
  group_by(seq) %>% 
  summarise(prev = sum(value > 0), 
            reads = sum(value)) %>%
  na.omit() %>% 
  filter(reads < quantile(reads, 0.9)) %>% 
  left_join(Fungi_len) %>% 
  mutate(reds_o_prev = reads/prev) %>% 
  ggplot(aes(x = len, y = reds_o_prev))+
  geom_point(shape = 21, alpha = 0.5, size = 3)



Fungi_len_reads %>% 
  pivot_wider(names_from = var, values_from = value) %>% 
  ggplot(aes(x = count_seq, y = reads))+
  geom_point()+
  coord_equal(ratio = 1)+
  geom_abline(slope = 1)

```



# rarefaction

```{r}
Fungi_tax %>% 
  filter(phylum %in% c("Ascomycota", "Basidiomycota", "Tracheophyta"))
```


```{r}
# Define the target phyla
major_phyla <- c("Ascomycota", "Basidiomycota", "Tracheophyta", "Other")

# Build one tidy data frame of rarefaction curves per phylum
rc_df <- map_df(major_phyla, function(ph) {
 
   # 1) Pull ASV IDs for this phylum
  
  if(ph  == "Other"){
    
    seqs <- Fungi_tax %>%
    filter(!phylum %in%  major_phyla) %>%
    pull(seq)
    
  } else{
    
    seqs <- Fungi_tax %>%
    filter(phylum  == ph) %>%
    pull(seq)
    
  }
  
  
  # 2) Subset & filter ASV table
  
  asv_sub <- Fungi_ASV %>%
    select(Index, any_of(seqs)) %>%                  # only ASVs from this phylum
    #select(c(1, which(colSums(.[-1]) > 0)+1)) %>%           # drop zero‐sum ASVs
    data.frame()
  
  # 3) Convert to matrix with rownames
  rownames(asv_sub) <- asv_sub$Index
  asv_mat <- asv_sub[, -1] %>% as.matrix()
  
  # 4) Compute rarefaction curve (tidy = TRUE gives a tibble)
  rarecurve(
    asv_mat,
    step      = 1000,
    label     = FALSE,
    tidy      = TRUE
  ) %>%
    dplyr::rename(Index = Site) %>%                   # match metadata key
    mutate(phylum = ph)                               # tag the phylum
})

# 5) Join metadata
rc_plot_df <- rc_df %>%
  left_join(Meta, by = "Index")

# 6) Plot: one facet per phylum
ggplot(rc_plot_df, aes(x = Sample, y = Species, color = duration, group = Index)) +
  geom_line() +
  #geom_vline(xintercept = 50000, colour = "darkred") +
  facet_wrap(~phylum, scales = "free_y", ncol = 2) +
  theme_minimal()+
  theme(legend.position = "right")+
  scale_color_brewer(palette = "Set1")

ggplot(rc_plot_df, aes(x = Sample, y = Species, color = phylum, group = paste(Index,phylum))) +
  geom_line() +
  #geom_vline(xintercept = 50000, colour = "darkred") +
  facet_wrap(~Sampler, scales = "free_y") +
  theme_minimal()+
  theme(legend.position = "right")+
  scale_color_brewer(palette = "Set1")

```

Sequencing depth varied between samples but reached saturation in all cases for all target taxa. 

# clean data

9F2control2 got contaminated during DNA extraction --> could not be used as Blank anymore. Sample removed
1DK5 → Filter dropped on the floor. Only kept buffer (not filter paper) but most particles seemed to be in buffer (turbid)
1DD30 → No liquid left after 1st Lysis. No supernatent could be transferred
1CD_ → No liquid left after 1st Lysis. No supernatent could be transferred
1DB30 → Only small amount of liquid → ½ transferred.
8RH5 quite some buffer missing. Added less beads (ca ½) as I was afraid that the full amount would absorb the remaining supernatent. 

Plate1:
Pipetted some of column 2 in column 1?

Well E3 only very little liquid in

## index hopping

We demultiplexed with all indicies (full Plate) but not all were used. Unused indices should not have been found but could be present in the data due to random mutations and Maximum likelihood index-assignment (should be rare as we use dual indices)

```{r}
index_hop <- 
Fungi_ASV %>%
  filter(!Index %in% Meta$Index) %>%
  pivot_longer(-Index) %>% 
  filter(value > 0)

index_hop
```

5 unused indices have any reads, with the maximum being 2 reads

```{r}
Fungi_ASV %>% 
  select(unique(index_hop$name)) %>% 
  colSums()
```

all reads are from highly abundant ASVs, indicating a low but existing level of index-hopping

we remove the unused indices from the ASV table

```{r}
Fungi_ASV <-
  Fungi_ASV %>%
  filter(Index %in%  
           {filter(Meta, Sample_names != "BLANK8") %>%  #this blank got massivley contaminated 
             pull(Index)
             } 
         )
```


##check controls

### decontam

```{r}
# define which samples are blanks

# we set field controls to NA (which excludes them) as they have been exposed and contain genuine sequences. We look at this in the next step. 

Blanks <- 
Meta %>% 
  filter(Index %in% Fungi_ASV$Index) %>% #samples with zero reads are missing from the ASV table
  select(Index, Sample_names, Type) %>% 
  filter(Sample_names != "BLANK8") %>% #this blank got massivley contaminated (see below)
  mutate(is_neg = case_when(Type %in% c("active", "passive") ~ FALSE,
                            Type %in% c("extr_control", "PCR_control") ~ TRUE,
                            Type == "field_control" ~ NA)) %>% 
  arrange(Index) 


#check that sample order is matching 
all(Fungi_ASV$Index == Blanks$Index)

#run decontam
contamdf.prev <- isContaminant(as.matrix(Fungi_ASV[,-1]), 
                               neg = Blanks$is_neg, 
                               method="prevalence")

p_thresholds = seq(1,0,-0.001)

# Compute the number of contaminants for each threshold
contam_counts <- purrr::map_df(p_thresholds, function(x) {
  tibble(
    p_threshold = x,
    num_contaminants = sum(contamdf.prev$p < x, na.rm = TRUE)
  )
})

# Plot the results
ggplot(contam_counts, aes(x = p_threshold, y = num_contaminants)) +
  geom_point(size = 0.1)+
  geom_smooth(linewidth = 0.5, color = "red") +
  geom_vline(xintercept = 0.1, colour = "blue")+
  labs(
    title = "Prevalence-based contaminant detection",
    x = "P-value Threshold",
    y = "Number of Contaminants"
  ) +
  theme_minimal()+
  scale_x_reverse()
```


```{r}
contam_seqs <- 
contamdf.prev %>% 
  filter(p < 0.1) %>% 
  tibble::rownames_to_column("seq")

```


decontam identifies no contamninants.


visualize contaminant sequences as abundance ratio between blanks and sample

```{r}
cont_ratio <- 
Fungi_ASV %>% 
  pivot_longer(-Index, names_to = "seq") %>% 
  left_join(Meta) %>% 
  select(Index, seq, Sample_names, Type, value) %>% 
  mutate(Type = case_when(Type %in% c("active", "passive") ~ "Sample",
                            Type %in% c("extr_control", "PCR_control") ~ "Lab",
                            Type == "field_control" ~ "Field")) %>% 
  group_by(Type, seq) %>% 
  summarise(mean_val = round(mean(value),2),
            max_val = max(value)) %>% 
  pivot_longer(one_of(c("mean_val", "max_val"))) %>% 
  pivot_wider(names_from = Type, values_from = value) %>% 
  filter(Lab > 0 | Field > 0) %>% 
  pivot_longer(!one_of("seq", "Sample", "name"), names_to= "Type") %>%
  mutate(ratio = round(value/Sample,3))

Field_only <- 
cont_ratio %>% 
  filter(Type == "Field") %>% 
  filter(value > 0)

Lab_only <- 
cont_ratio %>% 
  filter(Type == "Lab") %>% 
  filter(value > 0)
```


###lab controls

one blank (blank 8) got contaminated by a sample. it cannot serve as a blank and is excluded

```{r}
Lab_only$seq %>% unique %>% length()
```

```{r}
Lab_only %>% 
  filter(name == "mean_val") %>% 
  filter(ratio > 1) %>% 
  nrow()
```


```{r}
Fungi_ASV %>% 
  pivot_longer(-Index, names_to = "seq") %>% 
  filter(seq %in% unique(Lab_only$seq)) %>% 
  left_join(Meta) %>% 
  filter(Type %in% c("extr_control", "PCR_control")) %>% 
  group_by(seq) %>% 
  filter(sum(value) > 0) %>% 
  left_join(
        {filter(cont_ratio, 
                Type == "Lab",
                name == "mean_val") %>% 
            select(seq, ratio)
            }
  ) %>% 
  mutate(ratio = ifelse(value == 0, NA, ratio)) %>% 
  left_join(Fungi_tax) %>% 
  mutate(spec_name = 
           case_when(!is.na(species) ~ paste(species, seq, sep = " "),
                     !is.na(genus) ~ paste(genus, "sp.", seq, sep = " "),
                     !is.na(family) ~ paste(family, "(fam)", seq, sep = " "),
                     !is.na(order) ~ paste(order, "(ord)", seq, sep = " "),
                     !is.na(class) ~ paste(class, "(cl)", seq, sep = " "),
                     !is.na(phylum) ~ paste(phylum, "(ph)", seq, sep = " "),
                     TRUE ~ seq)) %>%
  ggplot(aes(x = Sample_names, y = spec_name, fill = ratio))+
  geom_tile()+
  scale_fill_viridis_c(na.value = "white")+
  theme(axis.text.x = element_text(angle = 90, hjust = 0))
```



There are 28 sequences present across extraction blanks and pcr negatives. 6 are present in higher average abundance in the PCR negatives than in the samples - these are clear contaminants. 

The remaining are more likely the result cross-contamination (i.e. true OTUs that contaminated a blank)

###field controls

```{r}
Field_only$seq %>% unique %>% length()
```


```{r}
Field_only %>% 
  filter(name == "mean_val") %>% 
  filter(ratio > 1) %>% 
  nrow()
```

```{r}
Field_only %>% 
  filter(name == "mean_val") %>% 
  filter(ratio > 10) %>% 
  nrow()
```

there are a total of 542 sequences present across the 14 field blanks. 

There are also 223 sequences present with higher abundance in the controls than in any sample, in 119 of which the sequence is 10X more abundant in the field control than in any sample

```{r}
Field_cont <- 
Fungi_ASV %>% 
  pivot_longer(-Index, names_to = "seq") %>% 
  filter(seq %in% unique(Field_only$seq)) %>% 
  left_join(Meta) %>% 
  filter(Type == "field_control") %>% 
  filter(value > 0) %>% 
  left_join(
        {filter(cont_ratio, Type == "Field",
                name == "mean_val") %>% 
            select(seq, ratio)
            }
  ) 

Field_cont %>% 
  mutate(ratio_cat = ifelse(ratio > 1, "> 1", "< 1")) %>% 
  ggplot(aes(x = Sampler, y = value, colour = ratio_cat))+
  geom_point(position = position_jitterdodge(jitter.width =  0.2, dodge.width = 0.6))+
  scale_fill_viridis_c()+
  scale_y_log10()+
  scale_color_brewer(palette = "Set1")+
  theme(axis.text.x = element_text(angle = 90, hjust = 0))+
  theme_minimal()+
  coord_flip()
```

```{r}
Field_cont %>% 
  mutate(ratio_cat = ifelse(ratio > 1, "> 1", "< 1")) %>% 
  group_by(ratio_cat, Sampler) %>% 
  summarise(n = n()) %>% 
  ggplot(aes(x = Sampler, y = n, colour = ratio_cat))+
  geom_point(position = position_dodge(width = 0.4), size = 3, shape = 21)+
  scale_color_brewer(palette = "Set1")+
  theme_minimal()+
  coord_flip()
```


```{r}
Field_cont %>% 
  group_by(Sampler) %>% 
  summarise(n = n())
```

```{r}
Field_cont %>% 
  mutate(ratio_cat = case_when(
    ratio < 1 ~ "< 1",
    ratio > 1 & ratio < 10 ~ "> 01",
    ratio > 10 ~ "> 10")) %>% 
  group_by(ratio_cat, Sampler) %>% 
  summarise(n = n()) %>% 
  pivot_wider(names_from = ratio_cat, values_from = n, values_fill = 0)
```

There is significant differences between the number of sequences detected in the field blanks across samplers. 

The two passive methods have very low number of detentions (6 & 7 OTUs respectively) - but for these two the field samples were 'carried-along' but not opened in the field. 

The Burkhard, the Coriolis and the Sass have 4, 13 and 19 OTUS in the field control - respectively. 

The Kärcher, the Electrostatic and the HEPA sampler (as well as the Drone sampler) have between 89 (Drone) and 310 (HEPA) OTUs as contaminants. 

Notably, for the Kärcher - and to a lesser extend for the HEPA - abundance in the control is below abundance in the samples (ratio < 1). For the Kärcher, only five sequences are more abundant in the control than in the sample. 

The Drone, the Electrostatic amd the HEPA have a significant number of OTUs with much higher abundances in the control than in the sample. These are likely contaminants and need to be excluded. 


#exclude contaminats

I follow the following rules for excluding contaminants 

1) For every OTU present in any laboratory negative control we remove twice the number of the maximum reads from all samples. 

2) For every OTU present in a field negative control at higher average abundance than in a the samples  we remove twice the max number of reads from all samples. 

3) OTUs present in the field negative control at lower average abundance than in the sample are kept untouched. 

```{r}
Remove_reads_lab <- 
  Lab_only %>% 
  filter(name == "max_val") %>% 
  select(seq, value) 

Remove_reads_field <- 
Field_only %>% 
  group_by(seq) %>% 
  filter(min(ratio) > 1) %>% 
  ungroup() %>% 
  filter(name == "max_val") %>% 
  select(seq, value)

Remove_reads <- 
  bind_rows(Remove_reads_lab, 
            Remove_reads_field) %>% 
  group_by(seq) %>% 
  filter(value == max(value)) %>% #in case an OTU occured in both lab and field controls
  mutate(seq_remove = value * 2) %>% #remove twice the average number of reads in the control
  mutate(seq_remove = ceiling(seq_remove)) #round up to next higher integer

Fungi_mat <- as.matrix(Fungi_ASV[,-1]) # matrix of reads
Remove_mat <- Fungi_mat # matrix of same size
Remove_mat[] <- 0 # filled with 0s
Remove_mat[,Remove_reads$seq] <- rep(Remove_reads$seq_remove, each = nrow(Remove_mat)) #fill with reads to remove (colwise)

Fungi_mat_clean <- Fungi_mat - Remove_mat #remove reads

Fungi_mat_clean[Fungi_mat_clean < 0] <- 0
```

remove empty OTUS
```{r}
Fungi_mat_clean <- Fungi_mat_clean[, colSums(Fungi_mat_clean) > 0] # remove empty sequences
```

```{r}
ncol(Fungi_mat) -ncol(Fungi_mat_clean)
```

208 sequences were removed from all samples

back-transform to tibble and add back sample index

```{r}

Fungi_ASV <- 
as_tibble(Fungi_mat_clean) %>% #transform back to tibble
  bind_cols(select(Fungi_ASV, "Index"), .) #add Sample Index back

```

remove controls from ASV data

```{r}
blank_index <- 
  Meta %>% 
  filter(!Type %in% c("active", "passive")) %>% 
  pull(Index)

Fungi_ASV <- 
Fungi_ASV %>% 
  filter(!Index %in% blank_index)
```


check for empty samples

```{r}

Sample_reads <- 
Fungi_ASV %>% 
  pivot_longer(-Index) %>% 
  group_by(Index) %>% 
  summarise(value = sum(value)) %>% 
  arrange(value) %>% 
  left_join(Meta)

Sample_reads

```
two samples have no (or < 1000) reads left after clean-up: 

1) one Drone Sample (1DD30)
2) a split Burkhard sample 1DB30_1

I remove both sample for downstream analysis

all other samples have >29'000 reads 

```{r}
empty_samples <- 
  Sample_reads %>% 
  filter(value < 1000) %>% 
  pull(Index)
  
Fungi_ASV <- 
Fungi_ASV %>%
  filter(!Index %in% empty_samples)
```

```{r}
Fungi_ASV_long <- 
  pivot_longer(Fungi_ASV, -Index, names_to = "seq", values_to = "reads")
```


##subset for Fungi only

```{r}
Fungi_tax <- 
  Fungi_tax %>% 
  filter(seq %in% colnames(Fungi_ASV))
```


```{r}
Fungi_ASV_long %>% 
  group_by(seq) %>% 
  summarise(reads = sum(reads)) %>% 
  left_join(Fungi_tax) %>% 
  group_by(kingdom, phylum) %>% 
  summarise(n_OTUs = n(), 
            reads = sum(reads)) %>% 
  pivot_longer(one_of(c("n_OTUs", "reads"))) %>% 
  ggplot(aes(x = phylum, y = value)) +
  facet_grid(kingdom~name, scales = "free", space = "free")+
  geom_bar(stat = "identity")+
  coord_flip()+
  scale_y_log10()+
  theme(axis.text.x = element_text(angle = 90))
```

The primers pick up more than only Fungi. Notably they pick up also a lot of plants (mostly Tracheophyta). 

For this comparison we keep only sequences that are assigned to the kingdom "Fungi" and are assigned at phylum level 

```{r}
Fungi_seq <- 
Fungi_tax %>% 
  filter(kingdom == "Fungi") %>% 
  filter(!is.na(phylum)) %>% 
  pull(seq)
```

```{r}
ncol(Fungi_ASV[,-1]) - length(Fungi_seq)
```
this excludes 6000 OTUs

```{r}
Fungi_ASV_long %>% 
  group_by(seq) %>% 
  summarise(reads = sum(reads)) %>% 
  left_join(Fungi_tax) %>% 
  mutate(cat = case_when(kingdom == "Fungi" & !is.na(phylum) ~ "Fungi",
                         kingdom == "Fungi" & is.na(phylum) ~ "na_Fungi",
                         kingdom != "Fungi" ~ kingdom)) %>% 
  group_by(cat) %>% 
  summarise(reads = sum(reads),
            n_seq = n()) %>% 
  mutate(reads_prct = signif(reads/sum(reads)*100,2),
         seq_prct = signif(n_seq/sum(n_seq)*100,2),)
```
this keeps 59% of the sequences accounting for 78% of the reads. The largest group of excluded sequences is Plantea with 9.5% of the sequences corresponding to 20% of the reads.

```{r}
Fungi_ASV_clean <- 
  Fungi_ASV %>% 
  select(one_of("Index", Fungi_seq))
```


```{r}
Fungi_ASV_clean_long <- 
  pivot_longer(Fungi_ASV_clean, -Index, names_to = "seq", values_to = "reads")
```


#prevalence
```{r}
Fungi_ASV_clean_long %>% 
  filter(reads > 0) %>% 
  group_by(seq) %>%
  summarise(prev = n()) %>% 
  ggplot(aes(x = prev))+
  geom_histogram(binwidth = 5, fill = NA, color = "darkred")+
 # scale_y_log10()+
  theme_minimal()
```



## Species richness

```{r}
Fungi_ASV_clean_long %>%
  left_join(Fungi_tax) %>% 
  select(seq, phylum) %>% 
  distinct() %>% 
  group_by(phylum) %>% 
  summarise(n = n())
  
```



```{r}

rar_curve <- 
Fungi_ASV_clean %>% 
  select(-Index) %>% 
  as.matrix %>% 
  `rownames<-`(Fungi_ASV_clean$Index) %>% 
  rarecurve(x = .,step = 1000, tidy = TRUE)

rar_curve %>% 
  left_join(Meta, by = c("Site" = "Index")) %>% 
  filter(grepl("active|passive", Type)) %>% 
  ggplot(aes(x = Sample, y = Species, group = Site, colour = duration)) +
  geom_line()+
  facet_wrap(~Sampler)

```

```{r}

S_df <- 
Fungi_ASV_clean_long %>% 
  group_by(Index) %>% 
  filter(reads > 0) %>% 
  summarise(S = n()) %>% 
  left_join(Meta) %>% 
  filter(Sampler != "Drone") %>% 
  mutate(Sampler = factor(Sampler, levels= unique(.$Sampler)[c(8,1,3,2,5,6,4,7)])) 

S_df %>% 
 mutate(Sampler = factor(Sampler, levels =  c("Kärcher","Coriolis", "Sass", "Hepa",
                                             "Burkhart", "Electrostatic", "MWAC",
                                            "WSL_Filter"))) %>%
  ggplot(aes(y = S, x = duration,  colour = Sites))+
  facet_grid(~Sampler, scales = "free_x")+
  geom_line(aes(group = Sites), size = 0.2, colour = "grey")+
  geom_point()+
  theme_minimal()+
  scale_color_brewer(palette = "Set1")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
        legend.position = "bottom")
```

#comparision with Burkhard

```{r}
S_df %>% 
  filter(duration != "30 min") %>% 
  select(Sites, Sampler, S) %>% 
  pivot_wider(names_from = Sampler, values_from = S) %>% 
  mutate(across(!one_of("Sites"), ~.x/Burkhart)) %>% 
  pivot_longer(-Sites) %>% 
  group_by(name) %>% 
  summarise(mean = round(mean(value),1), sd = round(sd(value),1))
```


export richness
```{r}
# S_df %>% 
#   mutate(taxon = "Fungi") %>% 
#   write_tsv(here("Data", "Fungi_richness.txt"))
```

### richness vs air volume

```{r}
flow_rate <- 
  Meta %>% 
  filter(Type == "active" & Sampler != "Drone") %>% 
  select(Sampler, duration) %>% 
  distinct() %>% 
  mutate(flow = case_when(
    Sampler == "Hepa" ~ 60,
    Sampler == "Kärcher" ~ 3000,
    Sampler == "Sass" ~ 300,
    Sampler == "Coriolis" ~ 300,
    Sampler == "Electrostatic" ~ 10,
    Sampler == "Burkhart" ~ 16.5,
  )) %>% 
  mutate(air_vol = case_when(
    duration == "30 min" ~ flow*30,
    duration == "5 hours" ~ flow*5*60
  ))
```


```{r}
S_df %>% 
  left_join(flow_rate) %>% 
  filter(Type == "active" & Sampler != "Drone") %>% 
  ggplot(aes(x = air_vol, y = S, colour = Sampler, shape = duration))+
  #geom_line(aes(group = Sites), size = 0.2, colour = "grey")+
  geom_point()+
 # facet_grid(~duration)+
  geom_smooth(aes(group = 1), method = "lm", colour = "darkblue", size = 0.2)+
#  geom_smooth(aes(group = duration), method = "lm", colour = "darkred", size = 0.2, se = F)+
  scale_x_log10()+
  scale_color_brewer(palette = "Set1")+
  theme_bw()
  

```

### richness corr

```{r}
Cor_mat <- 
S_df %>% 
  left_join(flow_rate) %>% 
  #filter(Type %in% c("active", "passive")) %>% 
  mutate(Sampler = paste(Sampler, duration)) %>% 
  arrange(Type, duration, desc(flow)) %>% 
  #filter(duration == "5 hours") %>% 
  select(Sites, Sampler, S) %>% 
  pivot_wider(names_from = Sampler, values_from = S) %>% 
  select(-Sites) %>% 
  cor(method = "pearson") 



{  
Cor_mat %>% 
  corrplot(type = "lower",diag = F,tl.col = "black", tl.cex = 0.8, method = "ellipse",
           na.label = "square", na.label.col = "white")
  
rect(xleft = 0.5, ybottom = 8.5, xright = 5.5, ytop = 13.5, density = NULL, angle = 45,
     col = NULL, border = "darkred", lty = par("lty"), lwd = par("lwd"))

rect(xleft = 6.5, ybottom = 2.5, xright = 11.5, ytop = 7.5, density = NULL, angle = 45,
     col = NULL, border = "darkred", lty = par("lty"), lwd = par("lwd"))

}
         
```


### sample completeness
```{r}
ASV_sample_based <- 
  Fungi_ASV_clean_long %>% 
  left_join(Meta) %>% 
  filter(Sampler != "Drone") %>% 
  mutate(reads = as.numeric(reads> 0)) %>% 
  select(Index, Sampler, duration, seq, reads) %>% 
  pivot_wider(names_from = seq, values_from = reads)
  
site_rar_df <- 
ASV_sample_based %>% 
  split(~Sampler+duration) 

site_rar_df <- 
site_rar_df[unlist(lapply(site_rar_df, nrow))>0] %>% 
  lapply(., function(x){
    
  if(nrow(x) > 0){res <- vegan::specaccum(x[,-c(1:3)])}
    
    resdf = tibble(sites = res$sites,
                   Richness = res$richness,
                   sd = res$sd)
    
    return(resdf)
  })

site_rar_df %>% 
  bind_rows(.id = "var") %>% 
  tidyr::separate_wider_delim(col = var, names = c("Sampler", "duration"), delim = ".") %>% 
  ggplot(aes(x = sites, y = Richness, fill = duration))+
  geom_line()+
  geom_ribbon(aes(ymin = Richness - sd, ymax = Richness + sd), colouor = NULL, alpha = 0.5)+
  facet_wrap(.~Sampler)+
  scale_x_continuous(breaks = seq(1,8,1))
```


## Species comp

```{r}

ASV_dist <- vegdist(Fungi_ASV_clean[,-1], method = "bray", binary = T)

NMDS <- metaMDS(ASV_dist) 

NMDS$points %>% 
  as_tibble() %>% 
  mutate(Index = Fungi_ASV_clean$Index) %>% 
  left_join(Meta) %>% 
#  filter(MDS1 < 100) %>% 
  ggplot(aes(x = MDS1, y = MDS2, colour = Sampler))+
  geom_point()+
  facet_wrap(~duration)
  
```
```{r}
adonis2(ASV_dist~ Sampler + Sites + duration, 
        data = Meta[Meta$Index %in% Fungi_ASV_clean$Index,],
        by = "margin")

betadisper(ASV_dist, group = Meta[Meta$Index %in% Fungi_ASV_clean$Index,]$Sites) %>% 
  anova()

betadisper(ASV_dist, group = Meta[Meta$Index %in% Fungi_ASV_clean$Index,]$Sampler) %>% 
  anova()

betadisper(ASV_dist, group = Meta[Meta$Index %in% Fungi_ASV_clean$Index,]$duration) %>% 
  anova()
```

### similarity ranking

```{r}
ASV_dist_M <- as.matrix(ASV_dist)

dimnames(ASV_dist_M) <- 
  list(Fungi_ASV_clean$Index, Fungi_ASV_clean$Index) 

diag(ASV_dist_M) <- NA

Meta_comb1 <- 
  Meta %>% 
  mutate(comb1 = paste(Sampler, duration, Sites)) %>% 
  select(Index, comb1)

Meta_comb2 <- 
  Meta %>% 
  mutate(comb2 = paste(Sampler, duration, Sites)) %>% 
  select(Index, comb2) %>% 
  dplyr::rename(Index2 = Index)

ASV_dist_df <-   
ASV_dist_M %>% 
  data.frame() %>% 
  tibble::rownames_to_column(var = "Index") %>% 
  pivot_longer(-Index, names_to = "Index2") %>% 
  filter(!is.na(value)) %>% 
  left_join(Meta_comb1) %>% 
  left_join(Meta_comb2) 

ASV_dist_cor <- 
ASV_dist_df %>% 
  filter(!grepl("Drone", comb1) & !grepl("Drone", comb2)) %>%
  mutate(site_comb = paste(gsub(".+\\s(\\w+)$", "\\1", comb1),
                           gsub(".+\\s(\\w+)$", "\\1", comb2),
                           sep = "_")) %>% 
  mutate(Sampler1 = gsub("(.+)\\s\\w+$", "\\1", comb1),
         Sampler2 = gsub("(.+)\\s\\w+$", "\\1", comb2)) %>% 
  filter(Sampler1 == Sampler2) %>% 
  select(value, site_comb, Sampler1) %>% 
  #group_by(site_comb, Sampler1) %>% 
  #filter(n() > 1)
  pivot_wider(names_from = Sampler1, values_from = value) %>% 
  select(-site_comb) %>% 
  distinct() %>% 
  select(rownames(Cor_mat)) %>% 
  cor(method = "pearson") %>% 
  #cor(method = "kendall") %>% 
  {.}
                              
ASV_dist_cor %>% 
  corrplot(type = "lower",diag = F,tl.col = "black", tl.cex = 0.8, method = "ellipse",
           na.label = "square", na.label.col = "white")

```


## Venn diagramm
```{r}
sampler_venn <- 
Fungi_ASV_clean_long %>%
  left_join(Fungi_tax) %>% 
  left_join(Meta) %>% 
  filter(Sampler != "Drone") %>% 
  filter(duration %in% c("30 min", "5 hours")) %>% 
 # filter(Sampler %in% c("Kärcher", "Sass", "Coriolis", "Hepa")) %>% 
  group_by(Sampler, duration, seq) %>% 
  summarise(reads = sum(reads)) %>% 
  mutate(reads = reads > 0) %>% 
  pivot_wider(names_from = duration, values_from = reads) %>% 
  select(-seq) %>% 
  euler(shape = "ellipse", by = list(Sampler)) 

#sampler_venn
eulerr_options(quantities = list(fontsize= 8, col = "darkred", cex = 1.2, fontfamily = "sans"))

plot(sampler_venn, legend = TRUE, quantities = TRUE)
```


```{r}
active <- 
Fungi_ASV_clean_long %>%
  left_join(Fungi_tax) %>% 
  left_join(Meta) %>% 
  filter(Sampler != "Drone") %>% 
  filter(duration %in% c("5 hours")) %>% 
  filter(Sampler %in% c("Kärcher", "Sass", "Coriolis", "Hepa")) %>% 
  group_by(Type, Sampler, seq) %>% 
  summarise(reads = sum(reads))%>% 
  mutate(reads = reads > 0) 

passive <- 
  Fungi_ASV_clean_long %>%
  left_join(Fungi_tax) %>% 
  left_join(Meta) %>% 
  filter(Sampler != "Drone") %>% 
  filter(duration %in% c("2 weeks")) %>% 
  group_by(Type, Sampler, seq) %>% 
  summarise(reads = sum(reads)) %>% 
  mutate(reads = reads > 0) 

comp_df <- active %>% mutate(comp = "") %>% dplyr::slice(0)

for(i in unique(active$Sampler)){
  
  comp_df <- 
  active %>% 
    filter(Sampler == i) %>% 
    bind_rows(passive) %>% 
    mutate(comp = i) %>% 
    bind_rows(comp_df, .)

  
}

sampler_venn <- 
comp_df %>% 
  group_by(comp) %>% 
  mutate(Sampler = case_when(Type == "active" ~ "active",
                             TRUE ~ Sampler)) %>% 
  select(-Type) %>% 
  pivot_wider(names_from = Sampler, values_from = reads) %>% 
  select(-seq) %>% 
  euler(shape = "ellipse", by = list(comp)) 

sampler_venn

eulerr_options(quantities = list(fontsize= 8))

plot(sampler_venn, legend = TRUE, quantities = TRUE)
```


```{r}
sampler_venn <- 
Fungi_ASV_clean_long %>%
  left_join(Fungi_tax) %>% 
  left_join(Meta) %>% 
  filter(Sampler != "Drone") %>% 
  filter(duration == "5 hours") %>% 
  filter(Sampler %in% c("Kärcher", "Sass", "Coriolis", "Hepa")) %>% 
  group_by(Sampler, seq) %>% 
  summarise(reads = sum(reads)) %>% 
  mutate(reads = reads > 0) %>% 
  pivot_wider(names_from = Sampler, values_from = reads) %>% 
  select(-seq) %>% 
  euler(shape = "ellipse") 

eulerr_options(quantities = list(fontsize= 10, col = "black"))

plot(sampler_venn, legend = TRUE, quantities = TRUE)
```


# Subgroup - Large fungi

red list of swiss fungi from  https://www.doi.org/10.16904/envidat.434 (-> download Version 2). 

```{r}
swiss_redlist <- 
  read_tsv(here("Data","Fungi" ,"in_files","redlist_taxonomy.tsv")) 

STATUS <- c("LC", "NT", "VU", "EN", "CR", "EW", "EX", "DD")
Full_Name <- c("Least Concern", "Near Threatened", "Vulnerable", "Endangered", "Critically Endangered", "Extinct in the Wild", "Extinct", "Data Deficient")
Rank <- c(1, 2, 3, 4, 5, 6, 7, 0)

IUCN_Red_List_Status <- data.frame(STATUS, Full_Name, Rank)
```

How much did we detect across all samples and sites?
(in % of swiss macrofungi)

Species 
```{r}
sum(unique(Fungi_tax$species) %in% swiss_redlist$accepted_name) / length(unique(swiss_redlist$accepted_name)) * 100
```

Genus
```{r}
sum(unique(Fungi_tax$genus) %in% swiss_redlist$genus) / length(unique(swiss_redlist$genus)) * 100
```

Family
```{r}
sum(unique(Fungi_tax$family) %in% swiss_redlist$family) / length(unique(swiss_redlist$family)) * 100
```


we match by accepted name (from GBIF) but some accepted names correspond to multiple species in the redlist with partly different threat status

I keep the lower threat status for each group of synonymes (following the logic that if two species are actually one, then the lower threat level can be assumed for the species group)

I ignore data deficient species in this step

```{r}

swiss_redlist_filt <- 
swiss_redlist %>% 
  select(accepted_name, SPECIES, STATUS) %>% 
  filter(!is.na(accepted_name)) %>% 
  left_join(IUCN_Red_List_Status) %>% 
  group_by(accepted_name) %>% 
  filter(Rank == min(Rank)) %>% 
  select(-SPECIES) %>% 
  dplyr::rename(species = accepted_name) %>% 
  group_by(species) %>% 
  dplyr::slice(1)
  

```

```{r}
  Fungi_ASV_clean_long_rl <- 
  Fungi_ASV_clean_long %>% 
  left_join(Fungi_tax) %>% 
  left_join(swiss_redlist_filt) 
```


```{r}
Fungi_ASV_clean_long_rl %>% 
  group_by(Index) %>% 
  mutate(reads_prct = reads/sum(reads)*100) %>% 
  filter(Rank >= 3) %>% # c("VU", "EN", "CR", "EW", "EX")) %>% 
  #filter(STATUS %in% c("LC")) %>% 
  left_join(Meta) %>% 
  filter(duration != "30 min") %>% 
  filter(Sampler != "Drone") %>% 
  mutate(Sampler = factor(Sampler, levels=unique(.$Sampler)[c(8,1,3,2,5,6,4,7)])) %>% 
  mutate(reads = ifelse(reads == 0, NA, reads)) %>% 
  mutate(reads_prct = ifelse(reads_prct == 0, NA, reads_prct)) %>% 
  arrange(Type, Sites, duration) %>% 
 # filter(Sampler == "MWAC") %>% 
  mutate(Sample_names = factor(Sample_names, levels = unique(.$Sample_names))) %>% #filter(is.na(reads)) %>% filter(STATUS == "CR") %>% pull(species) %>% unique()
  ggplot(aes(y = species, x = Sampler, size = reads_prct, colour = STATUS))+
  geom_point(alpha = 0.6)+
  facet_grid(STATUS~Sites, scales = "free_y", space = "free")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))+
  scale_color_brewer(palette = "Set1")+
  ggtitle("Species composition by site")
```


```{r}

vul_spec <- Fungi_ASV_clean_long_rl %>% 
 filter(STATUS %in% c("VU", "EN", "CR", "EW", "EX")) %>% 
  #filter(STATUS %in% c("LC")) %>% 
  pull(species)

Tax_vul <- 
  Fungi_tax %>% 
  filter(species %in% vul_spec) %>% 
  pull(seq)
  
Fungi_ASV_v <- 
Fungi_ASV %>% 
  select("Sample", all_of(Tax_vul)) 
  
```

```{r}

rar_curve <- 
Fungi_ASV_v %>% 
  select(-"Sample") %>% 
  as.matrix %>% 
  `rownames<-`(Fungi_ASV$Sample) %>% 
  `[`(rowSums(.) > 0,) %>% 
  rarecurve(x = .,step = 1000, tidy = TRUE)

rar_curve %>% left_join(Meta, by = c("Site" = "Index")) %>% 
  filter(grepl("active|passive", Type)) %>% 
  ggplot(aes(x = Sample, y = Species, group = Site, colour = duration)) +
  geom_line()+
  facet_wrap(~Sampler)

```


```{r}

S_df_vul <- 
Fungi_ASV_tax_long_rl %>% 
  filter(STATUS %in% c("VU", "EN", "CR", "EW", "EX")) %>% 
 # filter(STATUS %in% c("LC")) %>% 
  filter(Type %in% c("active", "passive")) %>% 
  filter(Sampler != "Drone") %>%
  group_by(duration, Sampler, Sites) %>%
  summarise(S = sum(reads >0)) %>% 
  mutate(Sampler = factor(Sampler, levels= levels(S_df$Sampler))) 


  
S_df_vul %>% 
  ggplot(aes(y = S, x = duration,  colour = Sites))+
  facet_grid(~Sampler, scales = "free_x")+
  geom_line(aes(group = Sites), size = 0.2, colour = "grey")+
  geom_point()+
  theme_minimal()+
  scale_color_brewer(palette = "Set1")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
        legend.position = "bottom")+
  ggtitle("rarefied richness - Vulnerable")
```


```{r}
S_df_vul %>% 
  left_join(Meta) %>% 
  left_join(flow_rate) %>% 
  filter(Type == "active" & Sampler != "Drone") %>% 
  ggplot(aes(x = air_vol, y = S, colour = Sampler, shape = duration))+
 # geom_line(aes(group = Sites), size = 0.2, colour = "grey")+
  geom_point(position = position_jitter(width = 0.1))+
  geom_smooth(aes(group = 1), method = "lm", se = TRUE, colour = "black", size = 0.2)+
  #facet_grid(~duration)+
  scale_x_log10()+
  scale_color_brewer(palette = "Set1")+
  theme_minimal()+
  theme(legend.position = "bottom")+
  labs(x = "sampled air volume (l)", title = "sampled air volume")
  

```


### richness corr

```{r}
Cor_mat <- 
S_df_vul %>% 
  left_join(Meta) %>% 
  filter(Sample_names != "1DB30_1") %>% 
  left_join(flow_rate) %>% 
  filter(Type %in% c("active", "passive")) %>% 
  mutate(Sampler = paste(Sampler, duration)) %>% 
  arrange(Type, duration, desc(flow)) %>% 
  #filter(duration == "5 hours") %>% 
  ungroup() %>% 
  select(Sites, Sampler, S) %>% 
  pivot_wider(names_from = Sampler, values_from = S) %>% 
  select(-Sites) %>% 
  cor(method = "pearson") 



{  
Cor_mat %>% 
  corrplot(type = "lower",diag = F,tl.col = "black", tl.cex = 0.8, method = "ellipse",
           na.label = "square", na.label.col = "white")
  
rect(xleft = 0.5, ybottom = 8.5, xright = 5.5, ytop = 13.5, density = NULL, angle = 45,
     col = NULL, border = "darkred", lty = par("lty"), lwd = par("lwd"))

rect(xleft = 6.5, ybottom = 2.5, xright = 11.5, ytop = 7.5, density = NULL, angle = 45,
     col = NULL, border = "darkred", lty = par("lty"), lwd = par("lwd"))

}
         
```
## Species comp

```{r}

#1DB30_1 = NGS_UDP0094

ASV_sample_vu <- 
Fungi_ASV_v %>% 
  filter(Sample != Meta[Meta$Sample_names == "1DB30_1",]$Index) %>% 
  filter(Sample %in% {filter(Meta, Type %in% c("active", "passive")) %>% pull(Index)}) %>% 
  filter(rowSums(.[,-1])>0) %>% 
  select(where(function(x){is.character(x) || sum(x) > 1}))

ASV_dist_vu <- vegdist(ASV_sample_vu[,-1], method = "bray", binary = T)
ASV_dist_vu <- raupcrick(ASV_sample_vu[,-1])

NMDS_vu <- metaMDS(ASV_dist_vu) 

NMDS_vu$points %>% 
  as_tibble() %>% 
  mutate(Index = ASV_sample_vu$Sample) %>% 
  left_join(Meta) %>% 
#  filter(MDS1 < 100) %>% 
  ggplot(aes(x = MDS1, y = MDS2, colour = Sites))+
  geom_point()+
  facet_wrap(~duration)
  
```

### similarity ranking

```{r}
ASV_dist_M <- as.matrix(ASV_dist_vu)

dimnames(ASV_dist_M) <- 
  list(ASV_sample_vu$Sample, ASV_sample_vu$Sample) 

diag(ASV_dist_M) <- NA

Meta_comb1 <- 
  Meta %>% 
  mutate(comb1 = paste(Sampler, duration, Sites)) %>% 
  select(Index, comb1)

Meta_comb2 <- 
  Meta %>% 
  mutate(comb2 = paste(Sampler, duration, Sites)) %>% 
  select(Index, comb2) %>% 
  dplyr::rename(Index2 = Index)

ASV_dist_df <-   
ASV_dist_M %>% 
  data.frame() %>% 
  tibble::rownames_to_column(var = "Index") %>% 
  pivot_longer(-Index, names_to = "Index2") %>% 
  filter(!is.na(value)) %>% 
  left_join(Meta_comb1) %>% 
  left_join(Meta_comb2) 

ASV_dist_cor <- 
ASV_dist_df %>% 
  filter(!grepl("Drone", comb1) & !grepl("Drone", comb2)) %>%
  mutate(site_comb = paste(gsub(".+\\s(\\w+)$", "\\1", comb1),
                           gsub(".+\\s(\\w+)$", "\\1", comb2),
                           sep = "_")) %>% 
  mutate(Sampler1 = gsub("(.+)\\s\\w+$", "\\1", comb1),
         Sampler2 = gsub("(.+)\\s\\w+$", "\\1", comb2)) %>% 
  filter(Sampler1 == Sampler2) %>% 
  select(value, site_comb, Sampler1) %>% 
#  group_by(site_comb, Sampler1) %>% 
#  filter(n() > 1)
  pivot_wider(names_from = Sampler1, values_from = value) %>% 
  select(-site_comb) %>% 
  distinct() %>% 
  select(rownames(Cor_mat)) %>% 
  cor(method = "pearson") %>% 
  #cor(method = "kendall") %>% 
  {.}
                              
ASV_dist_cor %>% 
  corrplot(type = "lower",diag = F,tl.col = "black", tl.cex = 0.8, method = "ellipse",
           na.label = "square", na.label.col = "lightgrey")

```

## Venn diagramm
```{r}
sampler_venn <- 
Fungi_ASV_tax_long_rl %>% 
  filter(Type %in% c("active")) %>% 
  filter(Sampler != "Drone") %>% 
  filter(STATUS %in% c("VU", "EN", "CR", "EW", "EX")) %>% 
 # filter(Sampler %in% c("Kärcher", "Sass", "Coriolis", "Hepa")) %>% 
  group_by(Sampler, duration, seq) %>% 
  summarise(reads = sum(reads)) %>% 
  mutate(reads = reads > 0) %>% 
  pivot_wider(names_from = duration, values_from = reads) %>% 
  select(-seq) %>% 
  euler(shape = "ellipse", by = list(Sampler)) 

sampler_venn

plot(sampler_venn, legend = TRUE, quantities = FALSE)
```

```{r}
sampler_venn <- 
Fungi_ASV_tax_long_rl %>% 
  filter(is.na(STATUS)) %>% 
#  filter(Type %in% c("passive")) %>% 
  filter(Sampler != "Drone") %>% 
#  filter(STATUS %in% c("VU", "EN", "CR", "EW", "EX")) %>% 
  filter(Sampler %in% c("Kärcher", "MWAC", "WSL_Filter")) %>% 
  group_by(Sampler, seq) %>% 
  summarise(reads = sum(reads)) %>% 
  mutate(reads = reads > 0) %>% 
  pivot_wider(names_from = Sampler, values_from = reads) %>% 
  select(-seq) %>% 
  euler(shape = "circle") 

plot(sampler_venn, legend = TRUE, quantities = TRUE)
```


#Aquatic fungi

```{r}
Fungi_tax %>% filter(phylum %in% c("Blastocladiomycota", "Chytridiomycota", "Neocallimastigomycota"))

Plant_pathogen <- read_csv(here("Data", "Fungus-Host-Data_20211105.csv"))

Plant_pathogen_gbif <- 
gsub("(\\w+)\\s.+", "\\1", unique(Plant_pathogen$sciName)) %>% unique() %>% 
  rgbif::name_backbone_checklist(rank = "genus", kingdom = "Fungi")

Plant_pathogen_gbif_detected <- 
  filter(Plant_pathogen_gbif, genus %in% Fungi_tax$genus) %>% 
  select(verbatim_name, kingdom, phylum, class, order, family, genus)

Plant_pathogen %>% 
  mutate(verbatim_name = gsub("(\\w+)\\s.+", "\\1", sciName)) %>% 
  right_join(Plant_pathogen_gbif_detected) %>% 
  select(sciName) %>% 
  unique()

Fungi_tax %>% 
  filter(species %in% Plant_pathogen$sciName) %>% 
  left_join(distinct(select(Plant_pathogen, sciName, host)), by = join_by(species == sciName)) %>% 
  filter(grepl("Castanea", host))
```

