---
title: "Fungi data analysis"
output: html_notebook
---

# Setup

Load required libraries for analysis.


```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(forcats)
library(here)
library(readr)
library(ggplot2)
library(vegan)
library(corrplot)
library(Biostrings)
library(decontam)
library(eulerr)
library(betapart)
library(ape)
library(picante)
library(iNEXT)
library(glmmTMB)
library(DHARMa)
```

Define the taxonomic group for this analysis.

```{r}
primer <- "Fungi"
```

## Visualization settings

Define a single global visualization configuration (factor ordering, color palettes, and reusable theme).


```{r}
# ===== VISUALIZATION SETTINGS (DEFINE ONCE) =====
sampler_order <- c(
  "Burkhart",
  "Electrostatic",
  "Hepa",
  "Sass",
  "Coriolis",
  "Kärcher",
  "Drone",
  "MWAC",
  "WSL_Filter"
)

duration_order <- c("30 min", "5 hours", "2 weeks")

sampler_colors <- RColorBrewer::brewer.pal(9, "Set1")
site_colors <- RColorBrewer::brewer.pal(8, "Paired")
duration_colors <- RColorBrewer::brewer.pal(3, "Dark2")

theme_airsample <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      axis.line = element_line(color = "black", linewidth = 0.3),
      strip.background = element_rect(fill = "grey90", color = NA),
      strip.text = element_text(face = "bold", size = rel(1.0))
    )
}
```

Load metadata from the standardized repository file.

```{r}
Meta <- read_tsv(here("Data", "Meta_data.tsv")) %>%
  mutate(
    Sampler = factor(Sampler, levels = sampler_order),
    duration = factor(duration, levels = duration_order)
  )
```

# Load data

Load ASV count table, taxonomy, representative sequences, and DNA concentrations.


```{r}
Fungi_ASV <- read_tsv(here("Data", primer, "out_final", "Fungi_ASV_glom.txt")) %>%
  mutate(Sample = gsub(".+?(NGS.+)", "\\1", Sample)) %>%
  dplyr::rename(Index = Sample)

Fungi_tax <- read_tsv(here("Data", primer, "out_final", "Fungi_taxa_80_glom.txt"))

Fungi_seq <- readDNAStringSet(here("Data", primer, "out_final", paste(primer, "Seq_glom", sep = "_")))

DNAC <- read_tsv(here("Data", "DNA_conc_air_sampler_comp.txt"))
```

Load phylogenetic tree (placeholder until trees are built).


```{r}
# Fungi_tree <- ape::read.tree(here("Data", "Fungi", "out_final", "Fungi_tree.nwk"))
```

# Fragment length distribution

Quantify sequence-length distributions and read-length relationships before filtering.

```{r}
Fungi_len <-
  Fungi_seq %>%
  width() %>%
  tibble(len = .) %>%
  mutate(seq = names(Fungi_seq))
```

Plot histogram of sequence lengths.


```{r}
Fungi_len %>%
  group_by(len) %>%
  summarise(n = n()) %>%
  ggplot(aes(x = len, y = n)) +
  geom_bar(stat = "identity") +
  theme_airsample() +
  labs(x = "Sequence length (bp)", y = "Number of ASVs")
```

Compute reads per ASV and plot against sequence length.


```{r}
Fungi_len_reads <-
  Fungi_ASV %>%
  pivot_longer(-Index, names_to = "seq") %>%
  group_by(seq) %>%
  summarise(prev = sum(value > 0),
            reads = sum(value)) %>%
  na.omit() %>%
  filter(reads < quantile(reads, 0.9)) %>%
  left_join(Fungi_len) %>%
  mutate(reads_o_prev = reads/prev)

Fungi_len_reads %>%
  ggplot(aes(x = len, y = reads_o_prev)) +
  geom_point(shape = 21, alpha = 0.5, size = 3) +
  theme_airsample() +
  labs(x = "Sequence length (bp)", y = "Reads per prevalence")
```

# Rarefaction (pre-cleaning)

Evaluate pre-cleaning rarefaction behavior by major taxonomic groups.
This establishes whether sequencing depth is limiting diversity estimates; this is foundational for using observed richness in **Q1** and **Q2**.
Plateauing curves indicate sequencing saturation, so sampler capture (not read depth) is the main bottleneck.

```{r}
major_phyla <- c("Ascomycota", "Basidiomycota", "Tracheophyta", "Other")
```


```{r}
rc_df <- map_df(major_phyla, function(ph) {

  if(ph == "Other"){
    seqs <- Fungi_tax %>%
      filter(!phylum %in% major_phyla) %>%
      pull(seq)
  } else{
    seqs <- Fungi_tax %>%
      filter(phylum == ph) %>%
      pull(seq)
  }

  asv_sub <- Fungi_ASV %>%
    select(Index, any_of(seqs)) %>%
    data.frame()

  rownames(asv_sub) <- asv_sub$Index
  asv_mat <- asv_sub[, -1] %>% as.matrix()

  rarecurve(
    asv_mat,
    step = 1000,
    label = FALSE,
    tidy = TRUE
  ) %>%
    dplyr::rename(Index = Site) %>%
    mutate(phylum = ph)
})

rc_plot_df <- rc_df %>%
  left_join(Meta, by = "Index")
```

Plot rarefaction curves faceted by phylum.

```{r}
ggplot(rc_plot_df, aes(x = Sample, y = Species, color = duration, group = Index)) +
  geom_line() +
  facet_wrap(~phylum, scales = "free_y", ncol = 2) +
  theme_airsample() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "Sequencing depth", y = "ASV richness")
```

Plot rarefaction curves faceted by sampler.


```{r}
ggplot(rc_plot_df, aes(x = Sample, y = Species, color = phylum, group = paste(Index, phylum))) +
  geom_line() +
  facet_wrap(~Sampler, scales = "free_y") +
  theme_airsample() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Set1") +
  labs(x = "Sequencing depth", y = "ASV richness")
```

Sequencing depth varied between samples but reached saturation in all cases for all target taxa.

# Clean data

## Known sample issues

Sample notes from lab records:
- 9F2control2 got contaminated during DNA extraction (cannot be used as blank)
- 1DK5: Filter dropped on floor, only buffer kept
- 1DD30: No liquid left after 1st lysis
- 1CD_: No liquid left after 1st lysis
- 1DB30: Only small amount of liquid, half transferred
- 8RH5: Some buffer missing, reduced beads
- Plate1: Possible pipetting error column 1/2
- Well E3: Very little liquid

## Index hopping

Check for reads in unused indices (should be absent but can occur due to index hopping).


```{r}
index_hop <-
  Fungi_ASV %>%
  filter(!Index %in% Meta$Index) %>%
  pivot_longer(-Index) %>%
  filter(value > 0)

index_hop
```

Five unused indices have reads (maximum 2 reads per index).


```{r}
Fungi_ASV %>%
  select(unique(index_hop$name)) %>%
  colSums()
```

All reads are from highly abundant ASVs, indicating low-level index hopping. Remove unused indices from ASV table.


```{r}
Fungi_ASV <-
  Fungi_ASV %>%
  filter(Index %in%
           {filter(Meta, Sample_names != "BLANK8") %>%
             pull(Index)
             }
         )
```

## Check controls

### Decontam (prevalence-based)

Define lab negatives for prevalence-based decontam while excluding field controls from the negative class.

Field controls can contain real environmental taxa, so treating them as strict negatives would bias contamination calls.


```{r}
Blanks <-
  Meta %>%
  filter(Index %in% Fungi_ASV$Index) %>%
  select(Index, Sample_names, Type) %>%
  filter(Sample_names != "BLANK8") %>%
  mutate(is_neg = case_when(Type %in% c("active", "passive") ~ FALSE,
                            Type %in% c("extr_control", "PCR_control") ~ TRUE,
                            Type == "field_control" ~ NA)) %>%
  arrange(Index)

all(Fungi_ASV$Index == Blanks$Index)
```

Run decontam with prevalence method.


```{r}
contamdf.prev <- isContaminant(as.matrix(Fungi_ASV[,-1]),
                               neg = Blanks$is_neg,
                               method = "prevalence")

n_contaminants <- sum(contamdf.prev$p < 0.1, na.rm = TRUE)
```

Decontam (prevalence method, p < 0.1) identifies `r n_contaminants` contaminants. This low sensitivity is expected for air eDNA because many potential contaminants are also genuine airborne taxa and therefore do not show strong prevalence separation between negatives and true samples.

### Visualize control contamination

Compute per-OTU control/sample ratios using both mean and max counts for lab and field controls.

This provides an explicit, transparent contamination screen that complements decontam and directly supports read-subtraction filtering.
OTUs with ratio > 1 are candidate contaminants; max values are used for conservative subtraction thresholds.

```{r}
cont_ratio <-
  Fungi_ASV %>%
  pivot_longer(-Index, names_to = "seq") %>%
  left_join(Meta) %>%
  select(Index, seq, Sample_names, Type, value) %>%
  mutate(Type = case_when(Type %in% c("active", "passive") ~ "Sample",
                          Type %in% c("extr_control", "PCR_control") ~ "Lab",
                          Type == "field_control" ~ "Field")) %>%
  group_by(Type, seq) %>%
  summarise(mean_val = round(mean(value), 2),
            max_val = max(value)) %>%
  pivot_longer(one_of(c("mean_val", "max_val"))) %>%
  pivot_wider(names_from = Type, values_from = value) %>%
  filter(Lab > 0 | Field > 0) %>%
  pivot_longer(!one_of("seq", "Sample", "name"), names_to = "Type") %>%
  mutate(ratio = round(value/Sample, 3))

Field_only <-
  cont_ratio %>%
  filter(Type == "Field") %>%
  filter(value > 0)

Lab_only <-
  cont_ratio %>%
  filter(Type == "Lab") %>%
  filter(value > 0)
```

### Lab controls

Blank8 got contaminated by a sample and is excluded. Count unique sequences in lab controls.


```{r}
Lab_only$seq %>% unique() %>% length()
```

Count sequences with ratio > 1 (more abundant in controls than samples).


```{r}
Lab_only %>%
  filter(name == "mean_val") %>%
  filter(ratio > 1) %>%
  nrow()
```

Summarize lab contaminant taxonomy.


```{r}
Lab_only %>%
  filter(name == "mean_val") %>%
  left_join(Fungi_tax, by = "seq") %>%
  mutate(contaminant = ratio > 1) %>%
  group_by(phylum, class, contaminant) %>%
  summarise(
    n_OTUs = n(),
    total_reads = sum(value),
    .groups = "drop"
  ) %>%
  arrange(desc(contaminant), desc(n_OTUs))
```

There are 28 sequences present across extraction blanks and PCR negatives. Six are more abundant in controls than samples (clear contaminants). The remaining are likely cross-contamination from true OTUs. Expected contaminant taxa include molds (Aspergillus, Penicillium, Cladosporium) which are omnipresent lab and air contaminants, skin-associated fungi (Malassezia) which are classical human contaminants, and soil fungi (Mortierella) which may indicate sample handling contamination.

### Field controls

Count unique sequences in field controls.

```{r}
Field_only$seq %>% unique() %>% length()
```

Count sequences with ratio > 1.

```{r}
Field_only %>%
  filter(name == "mean_val") %>%
  filter(ratio > 1) %>%
  nrow()
```

Count sequences with ratio > 10.


```{r}
Field_only %>%
  filter(name == "mean_val") %>%
  filter(ratio > 10) %>%
  nrow()
```

There are 542 sequences present across 14 field blanks, with 223 more abundant in controls than samples (119 with ratio > 10).

Create field control dataset for visualization.


```{r}
Field_cont <-
  Fungi_ASV %>%
  pivot_longer(-Index, names_to = "seq") %>%
  filter(seq %in% unique(Field_only$seq)) %>%
  left_join(Meta) %>%
  filter(Type == "field_control") %>%
  filter(value > 0) %>%
  left_join(
    {filter(cont_ratio, Type == "Field",
            name == "mean_val") %>%
        select(seq, ratio)
        }
  )
```

Plot contamination levels by sampler.


```{r}
Field_cont %>%
  mutate(ratio_cat = ifelse(ratio > 1, "> 1", "< 1")) %>%
  ggplot(aes(x = Sampler, y = value, colour = ratio_cat)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.6)) +
  scale_y_log10() +
  scale_color_brewer(palette = "Set1") +
  theme_airsample() +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 0)) +
  labs(x = "Sampler", y = "Reads in field control", colour = "Ratio")
```

Plot number of contaminant OTUs by sampler.

```{r}
Field_cont %>%
  mutate(ratio_cat = ifelse(ratio > 1, "> 1", "< 1")) %>%
  group_by(ratio_cat, Sampler) %>%
  summarise(n = n()) %>%
  ggplot(aes(x = Sampler, y = n, colour = ratio_cat)) +
  geom_point(position = position_dodge(width = 0.4), size = 3) +
  scale_color_brewer(palette = "Set1") +
  theme_airsample() +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 0)) +
  labs(x = "Sampler", y = "Number of OTUs", colour = "Ratio")
```

Count total OTUs per sampler in field controls.


```{r}
Field_cont %>%
  group_by(Sampler) %>%
  summarise(n = n())
```

Summarize contamination by ratio categories.

```{r}
Field_cont %>%
  mutate(ratio_cat = case_when(
    ratio < 1 ~ "< 1",
    ratio > 1 & ratio < 10 ~ "> 1",
    ratio > 10 ~ "> 10")) %>%
  group_by(ratio_cat, Sampler) %>%
  summarise(n = n()) %>%
  pivot_wider(names_from = ratio_cat, values_from = n, values_fill = 0)
```

There are significant differences in contamination levels across samplers. Passive methods have very low detection (6-7 OTUs) but this is because they are not true filed controls (not opened in the filed). Burkhart, Coriolis, and Sass have 4-19 OTUs. Kärcher, Electrostatic, HEPA, and Drone have 89-310 OTUs. For Kärcher, most OTUs have ratio < 1 (likely environmental not contaminant). Drone, Electrostatic, and HEPA show high ratios indicating true contamination.

Summarize field contaminant taxonomy by sampler.


```{r}
Field_cont %>%
  filter(ratio > 1) %>%
  left_join(Fungi_tax, by = "seq") %>%
  group_by(Sampler, phylum, class) %>%
  summarise(n_OTUs = n(), .groups = "drop") %>%
  arrange(Sampler, desc(n_OTUs))
```

Field control contaminants show sampler-specific patterns. High-flow active samplers (Kärcher, HEPA, Electrostatic) show contamination dominated by common air and surface fungi. The taxonomic composition reflects handling contamination and ambient laboratory air rather than the target environmental signal.

### Export contamination summary

Create contamination summary table for cross-group comparison.

```{r}
cont_summary_lab <-
  Lab_only %>%
  filter(name == "mean_val") %>%
  summarise(
    n_OTUs_lab = n(),
    n_ratio_gt1_lab = sum(ratio > 1),
    total_reads_lab = sum(value)
  )

cont_summary_field <-
  Field_cont %>%
  group_by(Sampler) %>%
  summarise(
    n_OTUs_field = n(),
    n_ratio_gt1 = sum(ratio > 1),
    n_ratio_gt10 = sum(ratio > 10),
    total_reads_field = sum(value)
  ) %>%
  mutate(taxon = "Fungi")

cont_summary <-
  cont_summary_field %>%
  mutate(
    n_OTUs_lab = cont_summary_lab$n_OTUs_lab,
    n_ratio_gt1_lab = cont_summary_lab$n_ratio_gt1_lab,
    total_reads_lab = cont_summary_lab$total_reads_lab
  ) %>%
  select(taxon, Sampler, n_OTUs_field, n_ratio_gt1, n_ratio_gt10,
         total_reads_field, n_OTUs_lab, n_ratio_gt1_lab)
```

Export contamination summary to intermediate output folder.


```{r}
write_tsv(cont_summary, here("Data", "Fungi", "out_intermediate", "Fungi_contamination.tsv"))
```

## Exclude contaminants

Apply contamination correction by subtracting `2 x max(control reads)` per flagged OTU.

This conservative rule removes likely contaminant signal while preserving true detections when controls have low relative abundance. After subtraction, negative values are clipped to zero and high-confidence contaminant carryover should be strongly reduced.

```{r}
Remove_reads_lab <-
  Lab_only %>%
  filter(name == "max_val") %>%
  select(seq, value)

Remove_reads_field <-
  Field_only %>%
  group_by(seq) %>%
  filter(min(ratio) > 1) %>%
  ungroup() %>%
  filter(name == "max_val") %>%
  select(seq, value)

Remove_reads <-
  bind_rows(Remove_reads_lab,
            Remove_reads_field) %>%
  group_by(seq) %>%
  filter(value == max(value)) %>%
  mutate(seq_remove = ceiling(value * 2))

Fungi_mat <- as.matrix(Fungi_ASV[,-1])
Remove_mat <- Fungi_mat
Remove_mat[] <- 0
Remove_mat[, Remove_reads$seq] <- rep(Remove_reads$seq_remove, each = nrow(Remove_mat))

Fungi_mat_clean <- Fungi_mat - Remove_mat
Fungi_mat_clean[Fungi_mat_clean < 0] <- 0
```

Remove empty OTUs after contamination subtraction.

```{r}
Fungi_mat_clean <- Fungi_mat_clean[, colSums(Fungi_mat_clean) > 0]
```

Count OTUs removed.

```{r}
ncol(Fungi_mat) - ncol(Fungi_mat_clean)
```

208 sequences were removed from all samples.

Convert back to tibble and add sample index.

```{r}
Fungi_ASV <-
  as_tibble(Fungi_mat_clean) %>%
  bind_cols(select(Fungi_ASV, "Index"), .)
```

Remove control samples from ASV table.

```{r}
blank_index <-
  Meta %>%
  filter(!Type %in% c("active", "passive")) %>%
  pull(Index)

Fungi_ASV <-
  Fungi_ASV %>%
  filter(!Index %in% blank_index)
```

## Check for empty samples

Count reads per sample after cleaning.

```{r}
Sample_reads <-
  Fungi_ASV %>%
  pivot_longer(-Index) %>%
  group_by(Index) %>%
  summarise(value = sum(value)) %>%
  arrange(value) %>%
  left_join(Meta)

Sample_reads
```

Two samples have < 1000 reads: 1DD30 (Drone) and 1DB30_1 (Burkhart split). Remove these samples. All others have > 29000 reads.

```{r}
empty_samples <-
  Sample_reads %>%
  filter(value < 1000) %>%
  pull(Index)

Fungi_ASV <-
  Fungi_ASV %>%
  filter(!Index %in% empty_samples)
```

Create long format ASV table.


```{r}
Fungi_ASV_long <-
  pivot_longer(Fungi_ASV, -Index, names_to = "seq", values_to = "reads")
```

# Subset for target group

Restrict the cleaned table to fungal OTUs with phylum-level assignment.

The ITS2 amplicon includes off-target taxa (notably plants), and fungal-focused inference for **Q1-Q4** requires a taxonomically consistent target set.


```{r}
Fungi_tax <-
  Fungi_tax %>%
  filter(seq %in% colnames(Fungi_ASV))
```

Visualize reads and OTUs by kingdom and phylum.

```{r}
Fungi_ASV_long %>%
  group_by(seq) %>%
  summarise(reads = sum(reads)) %>%
  left_join(Fungi_tax) %>%
  group_by(kingdom, phylum) %>%
  summarise(n_OTUs = n(),
            reads = sum(reads)) %>%
  pivot_longer(one_of(c("n_OTUs", "reads"))) %>%
  ggplot(aes(x = phylum, y = value)) +
  facet_grid(kingdom~name, scales = "free", space = "free") +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_y_log10() +
  theme_airsample() +
  theme(axis.text.x = element_text(angle = 0))
```

The primers amplify more than fungi. Notably, many plant sequences (Tracheophyta) are present. For this analysis, retain only sequences assigned to kingdom Fungi with phylum-level identification.


```{r}
Fungi_seq_keep <-
  Fungi_tax %>%
  filter(kingdom == "Fungi") %>%
  filter(!is.na(phylum)) %>%
  pull(seq)
```

Count excluded OTUs.

```{r}
ncol(Fungi_ASV[,-1]) - length(Fungi_seq_keep)
```

This excludes 6013 OTUs.

Compute proportion of reads and sequences retained.

```{r}
Fungi_ASV_long %>%
  group_by(seq) %>%
  summarise(reads = sum(reads)) %>%
  left_join(Fungi_tax) %>%
  mutate(cat = case_when(kingdom == "Fungi" & !is.na(phylum) ~ "Fungi",
                         kingdom == "Fungi" & is.na(phylum) ~ "na_Fungi",
                         kingdom != "Fungi" ~ kingdom)) %>%
  group_by(cat) %>%
  summarise(reads = sum(reads),
            n_seq = n()) %>%
  mutate(reads_prct = signif(reads/sum(reads)*100, 2),
         seq_prct = signif(n_seq/sum(n_seq)*100, 2))
```

This keeps 59% of sequences accounting for 78% of reads. The largest excluded group is Plantae (9.5% of sequences, 20% of reads).

Create cleaned ASV table with fungi only.

```{r}
Fungi_ASV_clean <-
  Fungi_ASV %>%
  select(one_of("Index", Fungi_seq_keep))

Fungi_ASV_clean_long <-
  pivot_longer(Fungi_ASV_clean, -Index, names_to = "seq", values_to = "reads")
```

# Prevalence

Summarize OTU prevalence across samples.

Prevalence profiles contextualize whether richness patterns are driven by many rare OTUs or a stable core community.


```{r}
Fungi_ASV_clean_long %>%
  filter(reads > 0) %>%
  group_by(seq) %>%
  summarise(prev = n()) %>%
  ggplot(aes(x = prev)) +
  geom_histogram(binwidth = 5, fill = NA, color = "darkred") +
  theme_airsample() +
  labs(x = "Prevalence (number of samples)", y = "Number of OTUs")
```

# Species richness

Quantify and model observed richness by sampler, duration, and site.

This is the primary **Q1** analysis (how many species each sampler detects) and establishes the main performance ranking.


```{r}
Fungi_ASV_clean_long %>%
  left_join(Fungi_tax) %>%
  select(seq, phylum) %>%
  distinct() %>%
  group_by(phylum) %>%
  summarise(n = n())
```

Compute rarefaction curves on cleaned dataset.

```{r}
rar_curve <-
  Fungi_ASV_clean %>%
  select(-Index) %>%
  as.matrix() %>%
  `rownames<-`(Fungi_ASV_clean$Index) %>%
  rarecurve(x = ., step = 1000, tidy = TRUE)

rar_curve %>%
  left_join(Meta, by = c("Site" = "Index")) %>%
  filter(grepl("active|passive", Type)) %>%
  ggplot(aes(x = Sample, y = Species, group = Site, colour = duration)) +
  geom_line() +
  facet_wrap(~Sampler) +
  theme_airsample() +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "Sequencing depth", y = "Observed richness")
```

Compute observed richness per sample.


```{r}
S_df <-
  Fungi_ASV_clean_long %>%
  group_by(Index) %>%
  filter(reads > 0) %>%
  summarise(S = n()) %>%
  left_join(Meta) %>%
  filter(Sampler != "Drone") %>%
  mutate(taxon = "Fungi")
```

Plot richness by sampler, duration, and site.

```{r}
S_df %>%
  ggplot(aes(y = S, x = duration, colour = Sites)) +
  facet_grid(~Sampler, scales = "free_x") +
  geom_line(aes(group = Sites), linewidth = 0.2, colour = "grey") +
  geom_point() +
  theme_airsample() +
  scale_color_brewer(palette = "Paired") +
  labs(x = "Duration", y = "Observed richness (S)")
```

## Formal statistical model

Fit a negative-binomial GLMM with sampler and duration as fixed effects and site as random intercept.

Sites are blocking units, and richness is overdispersed in metabarcoding; this model is the formal inferential test for **Q1**. Fixed-effect contrasts provide sampler and duration effects after accounting for between-site heterogeneity.

```{r}
S_df_model <-
  S_df %>%
  filter(!is.na(S))

richness_glmm <- glmmTMB(S ~ Sampler + duration + (1|Sites),
                         data = S_df_model,
                         family = nbinom2)

summary(richness_glmm)
```

Check residuals to verify model adequacy.

```{r}
sim_res <- simulateResiduals(richness_glmm, n = 1000)
plot(sim_res)
```


## Comparison with Burkhart

Compute fold-differences relative to Burkhart at 5 hours.


```{r}
S_df %>%
  filter(duration != "30 min") %>%
  select(Sites, Sampler, S) %>%
  pivot_wider(names_from = Sampler, values_from = S) %>%
  mutate(across(!one_of("Sites"), ~.x/Burkhart)) %>%
  pivot_longer(-Sites) %>%
  group_by(name) %>%
  summarise(mean = round(mean(value), 1), sd = round(sd(value), 1))
```

## Export richness data

Export richness for cross-group figures.


```{r}
write_tsv(S_df, here("Data", "Fungi", "out_intermediate", "Fungi_richness.tsv"))
```

# Faith's phylogenetic diversity

Compute Faith's PD in parallel to observed richness when a tree is available.
PD is the phylogenetic complement to **Q1**, testing whether sampler rankings hold for branch-length diversity as well as OTU counts.


```{r}
if(exists("Fungi_tree")) {

  # Prepare community matrix
  comm_mat <-
    Fungi_ASV_clean %>%
    data.frame() %>%
    `rownames<-`(.$Index) %>%
    select(-Index) %>%
    as.matrix()

  # Compute Faith's PD
  pd_result <- picante::pd(comm_mat, Fungi_tree, include.root = FALSE)

  # Add to S_df
  S_df <-
    S_df %>%
    left_join(
      pd_result %>%
        tibble::rownames_to_column("Index") %>%
        select(Index, PD)
    )

  # Plot PD by sampler, duration, and site
  S_df %>%
    ggplot(aes(y = PD, x = duration, colour = Sites)) +
    facet_grid(~Sampler, scales = "free_x") +
    geom_line(aes(group = Sites), linewidth = 0.2, colour = "grey") +
    geom_point() +
    theme_airsample() +
    scale_color_brewer(palette = "Paired") +
    labs(x = "Duration", y = "Faith's PD")

  # Export updated richness with PD
  write_tsv(S_df, here("Data", "Fungi", "out_intermediate", "Fungi_richness.tsv"))

} else {
  message("Phylogenetic tree not available. Skipping Faith's PD analysis.")
}
```

# Richness vs air volume

Relate observed richness to sampled air volume for active samplers.
This directly addresses **Q2** (does more sampled air yield more detected species?).


```{r}
flow_rate <-
  Meta %>%
  filter(Type == "active" & Sampler != "Drone") %>%
  select(Sampler, duration) %>%
  distinct() %>%
  mutate(flow = case_when(
    Sampler == "Hepa" ~ 60,
    Sampler == "Kärcher" ~ 3000,
    Sampler == "Sass" ~ 300,
    Sampler == "Coriolis" ~ 300,
    Sampler == "Electrostatic" ~ 10,
    Sampler == "Burkhart" ~ 16.5
  )) %>%
  mutate(air_vol = case_when(
    duration == "30 min" ~ flow * 30,
    duration == "5 hours" ~ flow * 5 * 60
  ))
```

Plot richness vs air volume with linear model.

```{r}
S_df %>%
  left_join(flow_rate) %>%
  filter(Type == "active" & Sampler != "Drone") %>%
  ggplot(aes(x = air_vol, y = S, colour = Sampler, shape = duration)) +
  geom_point() +
  geom_smooth(aes(group = 1), method = "lm", colour = "darkblue", linewidth = 0.5) +
  scale_x_log10() +
  scale_color_brewer(palette = "Set1") +
  theme_airsample() +
  labs(x = "Air volume (L)", y = "Observed richness (S)")
```

Fit linear model for richness vs log air volume.

```{r}
S_vol_df <-
  S_df %>%
  left_join(flow_rate) %>%
  filter(Type == "active" & Sampler != "Drone")

vol_lm <- lm(S ~ log10(air_vol), data = S_vol_df)
summary(vol_lm)
```

# Sampling completeness

Estimate per-site completeness as the fraction of each site's pooled reference community detected by each sampler-duration combination.
This complements **Q1/Q2** by putting richness in relative terms against the site-level "known" community.
OTUs are classified by site prevalence as: widespread (>60% of sites), intermediate (20-60%), and rare (<20%).

Build the site-level reference community and classify OTUs by site prevalence.

```{r}
analysis_grid <-
  Meta %>%
  filter(Index %in% Fungi_ASV_clean$Index, Sampler != "Drone") %>%
  distinct(Sites, Sampler, duration)

n_sites_total <- analysis_grid %>% distinct(Sites) %>% nrow()

otu_prevalence <-
  Fungi_ASV_clean_long %>%
  left_join(Meta, by = "Index") %>%
  filter(Sampler != "Drone", reads > 0) %>%
  distinct(seq, Sites) %>%
  count(seq, name = "n_sites_detected") %>%
  mutate(
    site_prevalence = n_sites_detected / n_sites_total,
    prevalence_class = case_when(
      site_prevalence > 0.60 ~ "widespread",
      site_prevalence >= 0.20 ~ "intermediate",
      TRUE ~ "rare"
    ),
    prevalence_class = factor(
      prevalence_class,
      levels = c("widespread", "intermediate", "rare")
    )
  )

ref_community_species <-
  Fungi_ASV_clean_long %>%
  left_join(Meta, by = "Index") %>%
  filter(Sampler != "Drone", reads > 0) %>%
  distinct(Sites, seq) %>%
  left_join(otu_prevalence %>% select(seq, prevalence_class), by = "seq")

ref_community_richness <-
  ref_community_species %>%
  group_by(Sites, prevalence_class) %>%
  summarise(S_reference = n_distinct(seq), .groups = "drop")
```

Compute class-specific completeness (fraction of each site's class-specific community detected).

```{r}
sampler_class_richness <-
  Fungi_ASV_clean_long %>%
  left_join(Meta, by = "Index") %>%
  filter(Sampler != "Drone", reads > 0) %>%
  distinct(Sites, Sampler, duration, seq) %>%
  left_join(otu_prevalence %>% select(seq, prevalence_class), by = "seq") %>%
  group_by(Sites, Sampler, duration, prevalence_class) %>%
  summarise(S_sampler = n_distinct(seq), .groups = "drop")

completeness_df <-
  analysis_grid %>%
  crossing(prevalence_class = c("widespread", "intermediate", "rare")) %>%
  mutate(
    prevalence_class = factor(
      prevalence_class,
      levels = c("widespread", "intermediate", "rare")
    )
  ) %>%
  left_join(
    sampler_class_richness,
    by = c("Sites", "Sampler", "duration", "prevalence_class")
  ) %>%
  left_join(ref_community_richness, by = c("Sites", "prevalence_class")) %>%
  mutate(
    S_sampler = replace_na(S_sampler, 0L),
    S_reference = replace_na(S_reference, 0L),
    fraction_detected = ifelse(S_reference > 0, S_sampler / S_reference, NA_real_),
    taxon = "Fungi"
  )
```

Plot class-specific completeness by sampler and duration.


```{r}
completeness_df %>%
  ggplot(aes(x = Sampler, y = fraction_detected, color = duration)) +
  geom_point(alpha = 0.6, size = 2, position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.4)) +
  stat_summary(
    fun.data = function(x) {
      m <- mean(x, na.rm = TRUE)
      se <- sd(x, na.rm = TRUE) / sqrt(sum(!is.na(x)))
      data.frame(y = m, ymin = m - 1.96 * se, ymax = m + 1.96 * se)
    },
    geom = "errorbar",
    width = 0.25,
    linewidth = 0.8,
    position = position_dodge(width = 0.4)
  ) +
  stat_summary(
    fun = mean,
    geom = "point",
    size = 3,
    position = position_dodge(width = 0.4)
  ) +
  facet_grid(.~prevalence_class) +
  coord_flip() +
  theme_airsample() +
  theme(axis.text.x = element_text(angle = 0)) +
  scale_color_brewer(palette = "Dark2") +
  labs(
    x = "Sampler",
    y = "Fraction of site community detected (within prevalence class)",
    color = "Duration"
  ) +
  ylim(0, 1)
```

Add unstratified "all" rows alongside the prevalence-class rows, then export the combined completeness table for use in cross-group figures (`Joined_Figures.Rmd`).

```{r}
completeness_all <-
  Fungi_ASV_clean_long %>%
  left_join(Meta, by = "Index") %>%
  filter(Sampler != "Drone", reads > 0) %>%
  group_by(Sites, Sampler, duration) %>%
  summarise(S_sampler = n_distinct(seq), .groups = "drop") %>%
  left_join(
    ref_community_species %>%
      group_by(Sites) %>%
      summarise(S_reference = n_distinct(seq), .groups = "drop"),
    by = "Sites"
  ) %>%
  mutate(
    prevalence_class = "all",
    fraction_detected = S_sampler / S_reference,
    taxon = "Fungi"
  )

completeness_df <- bind_rows(completeness_df, completeness_all)

write_tsv(completeness_df, here("Data", "Fungi", "out_intermediate", "Fungi_completeness.tsv"))
```

# Richness correlation across sites

Correlate site-level richness rankings across sampler-duration combinations. This addresses **Q4** by testing whether samplers agree on which sites are species-rich versus species-poor. High positive correlations imply robust ecological site gradients detectable across technologies.

```{r}
Cor_mat <-
  S_df %>%
  left_join(flow_rate) %>%
  mutate(Sampler_dur = paste(Sampler, duration)) %>%
  arrange(Type, duration, desc(flow)) %>%
  select(Sites, Sampler_dur, S) %>%
  pivot_wider(names_from = Sampler_dur, values_from = S) %>%
  select(-Sites) %>%
  cor(method = "pearson", use = "pairwise.complete.obs")
```

We restrict the correlation display to biologically meaningful contrasts: same sampler across durations, and same duration across samplers. Mixed sampler+duration contrasts are hard to interpret and are masked. For **Q4**, consistent positive correlations within these focused comparisons indicate that samplers detect the same site-level richness gradients.

```{r}
cor_meta <-
  tibble(Sampler_dur = colnames(Cor_mat)) %>%
  mutate(
    duration = sub(".* (30 min|5 hours|2 weeks)$", "\\1", Sampler_dur),
    Sampler = sub(" (30 min|5 hours|2 weeks)$", "", Sampler_dur),
    duration = factor(duration, levels = duration_order),
    Sampler = factor(Sampler, levels = sampler_order)
  ) %>%
  arrange(duration, Sampler)

ord <- match(cor_meta$Sampler_dur, colnames(Cor_mat))
Cor_mat_ord <- Cor_mat[ord, ord]

same_sampler_diff_time <-
  outer(as.character(cor_meta$Sampler), as.character(cor_meta$Sampler), "==") &
  outer(as.character(cor_meta$duration), as.character(cor_meta$duration), "!=")

same_time_diff_sampler <-
  outer(as.character(cor_meta$duration), as.character(cor_meta$duration), "==") &
  outer(as.character(cor_meta$Sampler), as.character(cor_meta$Sampler), "!=")

keep_focus <- same_sampler_diff_time | same_time_diff_sampler
diag(keep_focus) <- FALSE

Cor_focus_mat <- Cor_mat_ord
Cor_focus_mat[!keep_focus] <- NA

corrplot(
  Cor_focus_mat,
  type = "lower",
  diag = FALSE,
  tl.col = "black",
  tl.cex = 0.8,
  method = "ellipse",
  na.label = "square",
  na.label.col = "darkgrey"
)
```

# Richness by taxonomic subgroup

Repeat richness analyses within major fungal clades (Ascomycota, Basidiomycota, and Agaricomycetes). This extends **Q1** by testing whether sampler rankings depend on fungal subgroup biology, including spore morphology differences. If sampler rankings stay similar across clades, sampler performance is broad and not strongly clade-specific.

```{r}
S_subgroup_df <-
  Fungi_ASV_clean_long %>%
  left_join(Fungi_tax) %>%
  filter(phylum %in% c("Ascomycota", "Basidiomycota")) %>%
  left_join(Meta) %>%
  filter(Sampler != "Drone") %>%
  group_by(Index, phylum) %>%
  filter(reads > 0) %>%
  summarise(S = n()) %>%
  left_join(Meta)
```

Visualize richness separately for Ascomycota and Basidiomycota. If sampler rankings are consistent across phyla, this indicates broad performance rather than spore-type bias. Differences would suggest that spore morphology (e.g., unicellular conidia vs multicellular basidiospores) influences sampler capture efficiency.

```{r}
S_subgroup_df %>%
  ggplot(aes(y = S, x = duration, colour = Sites)) +
  facet_grid(phylum ~ Sampler, scales = "free_y") +
  geom_line(aes(group = Sites), linewidth = 0.2, colour = "grey") +
  geom_point() +
  theme_airsample() +
  scale_color_brewer(palette = "Paired") +
  labs(x = "Duration", y = "Observed richness (S)")
```

Fit GLMM for each phylum separately (main effects only, negative binomial family).

```{r}
ascomycota_glmm <- glmmTMB(S ~ Sampler + duration + (1|Sites),
                           data = filter(S_subgroup_df, phylum == "Ascomycota"),
                           family = nbinom2)
summary(ascomycota_glmm)

basidiomycota_glmm <- glmmTMB(S ~ Sampler + duration + (1|Sites),
                              data = filter(S_subgroup_df, phylum == "Basidiomycota"),
                              family = nbinom2)
summary(basidiomycota_glmm)
```

Ascomycota and Basidiomycota show remarkably similar sampler patterns, with consistent ranking of high- and low-performing technologies across both phyla. This indicates that the main sampler effects are robust across fungal groups rather than driven by one lineage. Interpreted biologically, the lack of strong phylum-specific differences suggests little evidence for sampler bias tied to broad spore-type differences (e.g., typical ascomycete conidia vs basidiomycete spores).


# Nestedness vs turnover

Decompose compositional dissimilarity into turnover and nestedness using one pooled community profile per sampler.
This addresses **Q3** by testing whether low-performing samplers detect subsets of Kärcher (nestedness) or systematically different species (turnover). Kärcher is the reference because it has the highest flow rate and highest overall capture.
High nestedness implies volume limitation without strong technological bias; high turnover implies technology-specific capture bias; mixed values imply both effects.

```{r}
comm_pooled <-
  Fungi_ASV_clean_long %>%
  left_join(Meta, by = "Index") %>%
  filter(Sampler != "Drone") %>%
  group_by(Sampler, seq) %>%
  summarise(reads = sum(reads), .groups = "drop") %>%
  pivot_wider(names_from = seq, values_from = reads, values_fill = 0) %>%
  arrange(Sampler) %>%
  data.frame() %>%
  `rownames<-`(.$Sampler) %>%
  select(-Sampler) %>%
  as.matrix()

comm_pooled_pa <- ifelse(comm_pooled > 0, 1, 0)

beta_result <- betapart::beta.pair(comm_pooled_pa, index.family = "sorensen")

beta_sor_mat <- as.matrix(beta_result$beta.sor)
beta_sim_mat <- as.matrix(beta_result$beta.sim)
beta_sne_mat <- as.matrix(beta_result$beta.sne)

karcher_idx <- which(rownames(beta_sor_mat) == "Kärcher")

beta_vs_karcher <-
  tibble(
    Sampler = rownames(beta_sor_mat)[-karcher_idx],
    beta_sor = beta_sor_mat[-karcher_idx, karcher_idx],
    beta_sim = beta_sim_mat[-karcher_idx, karcher_idx],
    beta_sne = beta_sne_mat[-karcher_idx, karcher_idx]
  ) %>%
  mutate(
    turnover_fraction = ifelse(beta_sor > 0, beta_sim / beta_sor, NA_real_),
    nestedness_fraction = ifelse(beta_sor > 0, beta_sne / beta_sor, NA_real_),
    sampler_type = case_when(
      Sampler %in% c("MWAC", "WSL_Filter") ~ "Passive",
      TRUE ~ "Active"
    ),
    Sampler = factor(Sampler, levels = sampler_order[sampler_order != "Kärcher"])
  )
```

Visualize the fraction of dissimilarity attributable to turnover vs nestedness for each sampler relative to Kärcher.
Fraction-based decomposition (`beta_sim / beta_sor`, `beta_sne / beta_sor`) gives interpretable components that sum to 1 and are directly comparable across samplers.
Bars dominated by nestedness indicate subset detection; bars dominated by turnover indicate different captured communities.

```{r}
beta_vs_karcher %>%
  select(Sampler, turnover_fraction, nestedness_fraction, sampler_type) %>%
  pivot_longer(c(turnover_fraction, nestedness_fraction),
               names_to = "component", values_to = "fraction") %>%
  mutate(component = recode(
    component,
    turnover_fraction = "Turnover (beta_sim)",
    nestedness_fraction = "Nestedness (beta_sne)"
  )) %>%
  ggplot(aes(x = Sampler, y = fraction, fill = component)) +
  geom_col(position = "stack") +
  facet_grid(sampler_type~., scales = "free", space = "free") +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  theme_airsample() +
  theme(axis.text.x = element_text(angle = 0)) +
  labs(
    x = "Sampler",
    y = "Fraction of dissimilarity (vs Kärcher)",
    fill = "Component",
    title = "Compositional dissimilarity decomposition"
  )
```

Nestedness-dominated profiles indicate that lower-capture samplers mostly recover subsets of Kärcher detections (volume limitation). Turnover-dominated profiles indicate technology-driven differences in which species are captured (potential sampler bias). Mixed profiles indicate both missing-species effects and technology-specific capture effects are present.


```{r}
write_tsv(
  beta_vs_karcher %>% mutate(taxon = "Fungi"),
  here("Data", "Fungi", "out_intermediate", "Fungi_nestedness_turnover.tsv")
)
```

# Rank-abundance correlation

Build sample-level relative-abundance matrices for non-drone samples and pair each sample with its metadata.

This addresses **Q3** from an abundance perspective by testing whether samplers agree on which taxa are dominant, not only which taxa are present.
If dominant taxa are consistently recovered across samplers, pairwise Spearman rho should be high within sites.

```{r}
non_drone_idx <-
  Meta %>%
  filter(Index %in% Fungi_ASV_clean$Index, Sampler != "Drone") %>%
  pull(Index)

# Build relative-abundance matrix and sample key
rank_mat <-
  Fungi_ASV_clean %>%
  filter(Index %in% non_drone_idx) %>%
  data.frame() %>%
  `rownames<-`(.$Index) %>%
  select(-Index) %>%
  as.matrix()

rank_mat <- sweep(rank_mat, 1, rowSums(rank_mat), "/")

sample_key <-
  Meta %>%
  filter(Index %in% rownames(rank_mat), Sampler != "Drone") %>%
  select(Index, Sites, Sampler, duration) %>%
  distinct() %>%
  mutate(Sampler_dur = paste(Sampler, duration))

combo_lookup <-
  sample_key %>%
  distinct(Sampler, duration, Sampler_dur)

pair_lookup <-
  tidyr::crossing(
    combo_lookup %>%
      dplyr::rename(Sampler1 = Sampler, duration1 = duration, Sampler_dur_1 = Sampler_dur),
    combo_lookup %>%
      dplyr::rename(Sampler2 = Sampler, duration2 = duration, Sampler_dur_2 = Sampler_dur)
  ) %>%
  filter(Sampler_dur_1 < Sampler_dur_2)

idx_tbl <- sample_key %>% select(Sites, Sampler_dur, Index)

site_pair_df <-
  tidyr::crossing(Sites = unique(sample_key$Sites), pair_lookup) %>%
  left_join(
    idx_tbl %>% dplyr::rename(Sampler_dur_1 = Sampler_dur, Index1 = Index),
    by = c("Sites", "Sampler_dur_1")
  ) %>%
  left_join(
    idx_tbl %>% dplyr::rename(Sampler_dur_2 = Sampler_dur, Index2 = Index),
    by = c("Sites", "Sampler_dur_2")
  ) %>%
  filter(!is.na(Index1), !is.na(Index2))

cor_res <-
  purrr::pmap(
    site_pair_df %>% select(Index1, Index2),
    function(Index1, Index2) {
      v1 <- rank_mat[Index1, ]
      v2 <- rank_mat[Index2, ]
      shared <- (v1 > 0) & (v2 > 0)

      if (sum(shared) <= 2) {
        return(list(rho = NA_real_, p_value = NA_real_, n_shared = sum(shared)))
      }

      ct <- suppressWarnings(cor.test(v1[shared], v2[shared], method = "spearman", exact = FALSE))
      list(rho = unname(ct$estimate), p_value = ct$p.value, n_shared = sum(shared))
    }
  )

rank_cor_df <-
  site_pair_df %>%
  mutate(
    rho = purrr::map_dbl(cor_res, "rho"),
    p_value = purrr::map_dbl(cor_res, "p_value"),
    n_shared = purrr::map_int(cor_res, "n_shared")
  ) %>%
  select(Sites, Sampler1, Sampler2, duration1, duration2, rho, p_value, n_shared)

rank_cor_summary <-
  rank_cor_df %>%
  filter(!is.na(rho)) %>%
  group_by(Sampler1, Sampler2, duration1, duration2) %>%
  summarise(
    rho = mean(rho, na.rm = TRUE),
    p_value = mean(p_value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(taxon = "Fungi")
```


```{r}
labels <-
  combo_lookup %>%
  arrange(duration, Sampler) %>%
  pull(Sampler_dur)

# Long pair table
pair_long <- rank_cor_summary %>%
  transmute(
    Sampler_dur1 = paste(Sampler1, duration1),
    Sampler_dur2 = paste(Sampler2, duration2),
    rho
  )

# Wide (upper triangle entries where available)
rank_cor_mat <-
  pair_long %>% 
  pivot_wider(
    names_from = Sampler_dur2,
    values_from = rho,
    values_fill = NA_real_
  ) %>%
  tibble::column_to_rownames("Sampler_dur1") %>%
  as.matrix()

# Symmetrize + diagonal
rank_cor_mat[lower.tri(rank_cor_mat)] <- t(rank_cor_mat)[lower.tri(rank_cor_mat)]
diag(rank_cor_mat) <- 1
```

Display the full pairwise rank-abundance correlation matrix. High values (warm colors) indicate that two sampler × duration combinations agree on which species are dominant. Clusters of high correlation within a duration block support **Q3** — samplers recover similar dominance hierarchies.

```{r}
rank_cor_mat %>%
  corrplot(type = "lower", diag = TRUE, tl.col = "black", tl.cex = 0.8,
           method = "color", na.label = "square", na.label.col = "white")
```

For reporting, we summarize the full pairwise correlation matrix by comparison type. The key **Q3** statistic is the mean rho for same-duration comparisons — this tells us how well samplers agree on species dominance within comparable sampling effort.

```{r}
rank_cor_grouped <-
  rank_cor_df %>%
  filter(!is.na(rho)) %>%
  mutate(comparison_type = case_when(
    duration1 == duration2 & Sampler1 != Sampler2 ~ "Same duration, different sampler",
    Sampler1 == Sampler2 & duration1 != duration2 ~ "Same sampler, different duration",
    TRUE ~ "Different sampler & duration"
  )) %>%
  group_by(comparison_type) %>%
  summarise(
    mean_rho = round(mean(rho), 3),
    sd_rho = round(sd(rho), 3),
    n_comparisons = n(),
    .groups = "drop"
  )

rank_cor_grouped
```

Most pairwise correlations are positive, indicating that shared species tend to maintain their rank ordering regardless of sampler. The same-duration/different-sampler comparisons are the most relevant for **Q3**: a high mean rho here means samplers agree on which species dominate the community.

```{r}
rank_cor_df %>%
  filter(!is.na(rho)) %>%
  mutate(comparison_type = case_when(
    duration1 == duration2 & Sampler1 != Sampler2 ~ "Same duration\ndifferent sampler",
    Sampler1 == Sampler2 & duration1 != duration2 ~ "Same sampler\ndifferent duration",
    TRUE ~ "Different sampler\n& duration"
  )) %>%
  ggplot(aes(x = comparison_type, y = rho)) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 1.5) +
  stat_summary(fun.data = mean_cl_boot, geom = "pointrange", color = "darkblue", size = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  theme_airsample() +
  theme(axis.text.x = element_text(angle = 0)) +
  labs(x = "Comparison type", y = "Spearman rho (site-level)")
```

Export rank correlation results.

```{r}
write_tsv(rank_cor_summary, here("Data", "Fungi", "out_intermediate", "Fungi_rank_correlation.tsv"))
```

# Venn diagram

Visualize shared and unique detections among the four main active samplers at 5 hours. This is a compact complement to **Q3**, showing overlap structure among the most comparable high-effort active methods. Large shared regions indicate consensus detection; large unique regions suggest sampler-specific capture.

```{r}
sampler_venn <-
  Fungi_ASV_clean_long %>%
  left_join(Meta) %>%
  filter(Sampler %in% c("Kärcher", "Sass", "Coriolis", "Hepa")) %>%
  filter(duration == "5 hours") %>%
  group_by(Sampler, seq) %>%
  summarise(reads = sum(reads)) %>%
  mutate(present = reads > 0) %>%
  select(-reads) %>%
  pivot_wider(names_from = Sampler, values_from = present) %>%
  select(-seq) %>%
  euler(shape = "ellipse")

eulerr_options(quantities = list(fontsize = 10, col = "black"))

plot(sampler_venn, legend = TRUE, quantities = TRUE)
```


# Red list detection

Quantify red-listed macrofungi detection per sampler and site. This provides a conservation-facing extension of **Q1** focused on societally relevant taxa, with sites as biological replicates for uncertainty estimation.

```{r}
swiss_redlist <-
  read_tsv(here("Data", "Fungi", "in_files", "redlist_taxonomy.tsv"))

STATUS <- c("LC", "NT", "VU", "EN", "CR", "EW", "EX", "DD")
Full_Name <- c("Least Concern", "Near Threatened", "Vulnerable", "Endangered",
               "Critically Endangered", "Extinct in the Wild", "Extinct", "Data Deficient")
Rank <- c(1, 2, 3, 4, 5, 6, 7, 0)

IUCN_Red_List_Status <- data.frame(STATUS, Full_Name, Rank)
```

Match detected species to red list.


```{r}
swiss_redlist_filt <-
  swiss_redlist %>%
  select(accepted_name, SPECIES, STATUS) %>%
  filter(!is.na(accepted_name)) %>%
  left_join(IUCN_Red_List_Status) %>%
  group_by(accepted_name) %>%
  filter(Rank == min(Rank)) %>%
  dplyr::rename(species = accepted_name) %>%
  group_by(species) %>%
  dplyr::slice(1)

Fungi_ASV_clean_long_rl <-
  Fungi_ASV_clean_long %>%
  left_join(Fungi_tax) %>%
  left_join(swiss_redlist_filt)
```

Count red-listed macrofungi detected per sampler at each site, treating sites as biological replicates. This avoids comparison against the total national checklist (most species do not occur at these 8 sites) and provides real uncertainty estimates. A deeper analysis of which specific red-listed species are detected would be a natural follow-up but is out of scope here.

```{r}
redlist_per_site <-
  Fungi_ASV_clean_long_rl %>%
  left_join(Meta) %>%
  filter(Sampler != "Drone") %>%
  filter(duration == "5 hours" | duration == "2 weeks") %>%
  filter(!is.na(STATUS)) %>%
  filter(reads > 0) %>%
  group_by(Sites, Sampler) %>%
  summarise(n_redlist = n_distinct(species), .groups = "drop")

redlist_per_site
```

Plot red-listed species counts per sampler with site-level replication (jitter + mean ± SE).

```{r}
redlist_per_site %>%
  ggplot(aes(x = Sampler, y = n_redlist)) +
  geom_jitter(width = 0.2, alpha = 0.5, size = 2, color = "steelblue") +
  stat_summary(fun.data = mean_se, geom = "pointrange", size = 0.8) +
  coord_flip() +
  theme_airsample() +
  theme(axis.text.x = element_text(angle = 0)) +
  labs(x = "Sampler", y = "Red-listed species detected per site")
```


# Session info

Record R session information for reproducibility.

```{r}
sessionInfo()
```
