---
title: "Fungi data analysis"
output: html_notebook
---

# Setup

Load required libraries for analysis.

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(forcats)
library(here)
library(readr)
library(ggplot2)
library(vegan)
library(corrplot)
library(Biostrings)
library(decontam)
library(eulerr)
library(betapart)
library(ape)
library(picante)
library(iNEXT)
```

Define the taxonomic group for this analysis.

```{r}
primer <- "Fungi"
```

Load metadata from the standardized file.

```{r}
Meta <- read_tsv(here("Data", "Meta_data.tsv"))
```

# Load data

Load ASV count table, taxonomy, representative sequences, and DNA concentrations.

```{r}
Fungi_ASV <- read_tsv(here("Data", primer, "out_final", "Fungi_ASV_glom.txt")) %>%
  mutate(Sample = gsub(".+?(NGS.+)", "\\1", Sample)) %>%
  dplyr::rename(Index = Sample)

Fungi_tax <- read_tsv(here("Data", primer, "out_final", "Fungi_taxa_80_glom.txt"))

Fungi_seq <- readDNAStringSet(here("Data", primer, "out_final", paste(primer, "Seq_glom", sep = "_")))

DNAC <- read_tsv(here("Data", "DNA_conc_air_sampler_comp.txt"))
```

Load phylogenetic tree (placeholder until trees are built).

```{r}
# Fungi_tree <- ape::read.tree(here("Data", "Fungi", "out_final", "Fungi_tree.nwk"))
```

# Fragment length distribution

Compute sequence lengths for all representative sequences.

```{r}
Fungi_len <-
  Fungi_seq %>%
  width() %>%
  tibble(len = .) %>%
  mutate(seq = names(Fungi_seq))
```

Plot histogram of sequence lengths.

```{r}
Fungi_len %>%
  group_by(len) %>%
  summarise(n = n()) %>%
  ggplot(aes(x = len, y = n)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(x = "Sequence length (bp)", y = "Number of ASVs")
```

Compute reads per ASV and plot against sequence length.

```{r}
Fungi_len_reads <-
  Fungi_ASV %>%
  pivot_longer(-Index, names_to = "seq") %>%
  group_by(seq) %>%
  summarise(prev = sum(value > 0),
            reads = sum(value)) %>%
  na.omit() %>%
  filter(reads < quantile(reads, 0.9)) %>%
  left_join(Fungi_len) %>%
  mutate(reads_o_prev = reads/prev)

Fungi_len_reads %>%
  ggplot(aes(x = len, y = reads_o_prev)) +
  geom_point(shape = 21, alpha = 0.5, size = 3) +
  theme_minimal() +
  labs(x = "Sequence length (bp)", y = "Reads per prevalence")
```

# Rarefaction (pre-cleaning)

Define major phyla to examine rarefaction curves by taxonomic group.

```{r}
major_phyla <- c("Ascomycota", "Basidiomycota", "Tracheophyta", "Other")
```

Compute rarefaction curves for each major phylum.

```{r}
rc_df <- map_df(major_phyla, function(ph) {

  if(ph == "Other"){
    seqs <- Fungi_tax %>%
      filter(!phylum %in% major_phyla) %>%
      pull(seq)
  } else{
    seqs <- Fungi_tax %>%
      filter(phylum == ph) %>%
      pull(seq)
  }

  asv_sub <- Fungi_ASV %>%
    select(Index, any_of(seqs)) %>%
    data.frame()

  rownames(asv_sub) <- asv_sub$Index
  asv_mat <- asv_sub[, -1] %>% as.matrix()

  rarecurve(
    asv_mat,
    step = 1000,
    label = FALSE,
    tidy = TRUE
  ) %>%
    dplyr::rename(Index = Site) %>%
    mutate(phylum = ph)
})

rc_plot_df <- rc_df %>%
  left_join(Meta, by = "Index")
```

Plot rarefaction curves faceted by phylum.

```{r}
ggplot(rc_plot_df, aes(x = Sample, y = Species, color = duration, group = Index)) +
  geom_line() +
  facet_wrap(~phylum, scales = "free_y", ncol = 2) +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Set1") +
  labs(x = "Sequencing depth", y = "ASV richness")
```

Plot rarefaction curves faceted by sampler.

```{r}
ggplot(rc_plot_df, aes(x = Sample, y = Species, color = phylum, group = paste(Index, phylum))) +
  geom_line() +
  facet_wrap(~Sampler, scales = "free_y") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Set1") +
  labs(x = "Sequencing depth", y = "ASV richness")
```

Sequencing depth varied between samples but reached saturation in all cases for all target taxa.

# Clean data

## Known sample issues

Sample notes from lab records:
- 9F2control2 got contaminated during DNA extraction (cannot be used as blank)
- 1DK5: Filter dropped on floor, only buffer kept
- 1DD30: No liquid left after 1st lysis
- 1CD_: No liquid left after 1st lysis
- 1DB30: Only small amount of liquid, half transferred
- 8RH5: Some buffer missing, reduced beads
- Plate1: Possible pipetting error column 1/2
- Well E3: Very little liquid

## Index hopping

Check for reads in unused indices (should be absent but can occur due to index hopping).

```{r}
index_hop <-
  Fungi_ASV %>%
  filter(!Index %in% Meta$Index) %>%
  pivot_longer(-Index) %>%
  filter(value > 0)

index_hop
```

Five unused indices have reads (maximum 2 reads per index).

```{r}
Fungi_ASV %>%
  select(unique(index_hop$name)) %>%
  colSums()
```

All reads are from highly abundant ASVs, indicating low-level index hopping. Remove unused indices from ASV table.

```{r}
Fungi_ASV <-
  Fungi_ASV %>%
  filter(Index %in%
           {filter(Meta, Sample_names != "BLANK8") %>%
             pull(Index)
             }
         )
```

## Check controls

### Decontam (prevalence-based)

Define which samples are blanks. Field controls are set to NA (excluded) as they contain genuine environmental sequences.

```{r}
Blanks <-
  Meta %>%
  filter(Index %in% Fungi_ASV$Index) %>%
  select(Index, Sample_names, Type) %>%
  filter(Sample_names != "BLANK8") %>%
  mutate(is_neg = case_when(Type %in% c("active", "passive") ~ FALSE,
                            Type %in% c("extr_control", "PCR_control") ~ TRUE,
                            Type == "field_control" ~ NA)) %>%
  arrange(Index)

all(Fungi_ASV$Index == Blanks$Index)
```

Run decontam with prevalence method.

```{r}
contamdf.prev <- isContaminant(as.matrix(Fungi_ASV[,-1]),
                               neg = Blanks$is_neg,
                               method = "prevalence")

p_thresholds <- seq(1, 0, -0.001)

contam_counts <- purrr::map_df(p_thresholds, function(x) {
  tibble(
    p_threshold = x,
    num_contaminants = sum(contamdf.prev$p < x, na.rm = TRUE)
  )
})
```

Plot number of contaminants identified across p-value thresholds.

```{r}
ggplot(contam_counts, aes(x = p_threshold, y = num_contaminants)) +
  geom_point(size = 0.1) +
  geom_smooth(linewidth = 0.5, color = "red") +
  geom_vline(xintercept = 0.1, colour = "blue") +
  labs(
    title = "Prevalence-based contaminant detection",
    x = "P-value Threshold",
    y = "Number of Contaminants"
  ) +
  theme_minimal() +
  scale_x_reverse()
```

Extract contaminant sequences at p < 0.1 threshold.

```{r}
contam_seqs <-
  contamdf.prev %>%
  filter(p < 0.1) %>%
  tibble::rownames_to_column("seq")
```

Decontam identifies no contaminants at p < 0.1.

### Visualize control contamination

Compute mean and max abundance ratios between controls and samples.

```{r}
cont_ratio <-
  Fungi_ASV %>%
  pivot_longer(-Index, names_to = "seq") %>%
  left_join(Meta) %>%
  select(Index, seq, Sample_names, Type, value) %>%
  mutate(Type = case_when(Type %in% c("active", "passive") ~ "Sample",
                          Type %in% c("extr_control", "PCR_control") ~ "Lab",
                          Type == "field_control" ~ "Field")) %>%
  group_by(Type, seq) %>%
  summarise(mean_val = round(mean(value), 2),
            max_val = max(value)) %>%
  pivot_longer(one_of(c("mean_val", "max_val"))) %>%
  pivot_wider(names_from = Type, values_from = value) %>%
  filter(Lab > 0 | Field > 0) %>%
  pivot_longer(!one_of("seq", "Sample", "name"), names_to = "Type") %>%
  mutate(ratio = round(value/Sample, 3))

Field_only <-
  cont_ratio %>%
  filter(Type == "Field") %>%
  filter(value > 0)

Lab_only <-
  cont_ratio %>%
  filter(Type == "Lab") %>%
  filter(value > 0)
```

### Lab controls

Blank8 got contaminated by a sample and is excluded. Count unique sequences in lab controls.

```{r}
Lab_only$seq %>% unique() %>% length()
```

Count sequences with ratio > 1 (more abundant in controls than samples).

```{r}
Lab_only %>%
  filter(name == "mean_val") %>%
  filter(ratio > 1) %>%
  nrow()
```

Plot heatmap of contaminant sequences in lab controls.

```{r}
Fungi_ASV %>%
  pivot_longer(-Index, names_to = "seq") %>%
  filter(seq %in% unique(Lab_only$seq)) %>%
  left_join(Meta) %>%
  filter(Type %in% c("extr_control", "PCR_control")) %>%
  group_by(seq) %>%
  filter(sum(value) > 0) %>%
  left_join(
    {filter(cont_ratio,
            Type == "Lab",
            name == "mean_val") %>%
        select(seq, ratio)
        }
  ) %>%
  mutate(ratio = ifelse(value == 0, NA, ratio)) %>%
  left_join(Fungi_tax) %>%
  mutate(spec_name =
           case_when(!is.na(species) ~ paste(species, seq, sep = " "),
                     !is.na(genus) ~ paste(genus, "sp.", seq, sep = " "),
                     !is.na(family) ~ paste(family, "(fam)", seq, sep = " "),
                     !is.na(order) ~ paste(order, "(ord)", seq, sep = " "),
                     !is.na(class) ~ paste(class, "(cl)", seq, sep = " "),
                     !is.na(phylum) ~ paste(phylum, "(ph)", seq, sep = " "),
                     TRUE ~ seq)) %>%
  ggplot(aes(x = Sample_names, y = spec_name, fill = ratio)) +
  geom_tile() +
  scale_fill_viridis_c(na.value = "white") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Sample", y = "Taxon", fill = "Ratio\n(Control/Sample)")
```

There are 28 sequences present across extraction blanks and PCR negatives. Six are more abundant in controls than samples (clear contaminants). The remaining are likely cross-contamination from true OTUs.

### Field controls

Count unique sequences in field controls.

```{r}
Field_only$seq %>% unique() %>% length()
```

Count sequences with ratio > 1.

```{r}
Field_only %>%
  filter(name == "mean_val") %>%
  filter(ratio > 1) %>%
  nrow()
```

Count sequences with ratio > 10.

```{r}
Field_only %>%
  filter(name == "mean_val") %>%
  filter(ratio > 10) %>%
  nrow()
```

There are 542 sequences present across 14 field blanks, with 223 more abundant in controls than samples (119 with ratio > 10).

Create field control dataset for visualization.

```{r}
Field_cont <-
  Fungi_ASV %>%
  pivot_longer(-Index, names_to = "seq") %>%
  filter(seq %in% unique(Field_only$seq)) %>%
  left_join(Meta) %>%
  filter(Type == "field_control") %>%
  filter(value > 0) %>%
  left_join(
    {filter(cont_ratio, Type == "Field",
            name = "mean_val") %>%
        select(seq, ratio)
        }
  )
```

Plot contamination levels by sampler.

```{r}
Field_cont %>%
  mutate(ratio_cat = ifelse(ratio > 1, "> 1", "< 1")) %>%
  ggplot(aes(x = Sampler, y = value, colour = ratio_cat)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.6)) +
  scale_y_log10() +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  coord_flip() +
  labs(x = "Sampler", y = "Reads in field control", colour = "Ratio")
```

Plot number of contaminant OTUs by sampler.

```{r}
Field_cont %>%
  mutate(ratio_cat = ifelse(ratio > 1, "> 1", "< 1")) %>%
  group_by(ratio_cat, Sampler) %>%
  summarise(n = n()) %>%
  ggplot(aes(x = Sampler, y = n, colour = ratio_cat)) +
  geom_point(position = position_dodge(width = 0.4), size = 3) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  coord_flip() +
  labs(x = "Sampler", y = "Number of OTUs", colour = "Ratio")
```

Count total OTUs per sampler in field controls.

```{r}
Field_cont %>%
  group_by(Sampler) %>%
  summarise(n = n())
```

Summarize contamination by ratio categories.

```{r}
Field_cont %>%
  mutate(ratio_cat = case_when(
    ratio < 1 ~ "< 1",
    ratio > 1 & ratio < 10 ~ "> 1",
    ratio > 10 ~ "> 10")) %>%
  group_by(ratio_cat, Sampler) %>%
  summarise(n = n()) %>%
  pivot_wider(names_from = ratio_cat, values_from = n, values_fill = 0)
```

There are significant differences in contamination levels across samplers. Passive methods have very low detection (6-7 OTUs). Burkhart, Coriolis, and Sass have 4-19 OTUs. Kärcher, Electrostatic, HEPA, and Drone have 89-310 OTUs. For Kärcher, most OTUs have ratio < 1 (likely environmental not contaminant). Drone, Electrostatic, and HEPA show high ratios indicating true contamination.

### Export contamination summary

Create contamination summary table for cross-group comparison.

```{r}
cont_summary_lab <-
  Lab_only %>%
  filter(name == "mean_val") %>%
  summarise(
    n_OTUs_lab = n(),
    n_ratio_gt1_lab = sum(ratio > 1),
    total_reads_lab = sum(value)
  )

cont_summary_field <-
  Field_cont %>%
  group_by(Sampler) %>%
  summarise(
    n_OTUs_field = n(),
    n_ratio_gt1 = sum(ratio > 1),
    n_ratio_gt10 = sum(ratio > 10),
    total_reads_field = sum(value)
  ) %>%
  mutate(taxon = "Fungi")

cont_summary <-
  cont_summary_field %>%
  bind_cols(cont_summary_lab) %>%
  select(taxon, Sampler, n_OTUs_field, n_ratio_gt1, n_ratio_gt10,
         total_reads_field, n_OTUs_lab, n_ratio_gt1_lab)
```

Export contamination summary to intermediate output folder.

```{r}
write_tsv(cont_summary, here("Data", "Fungi", "out_intermediate", "Fungi_contamination.tsv"))
```

## Exclude contaminants

Remove reads from contaminant OTUs using 2× max subtraction method. For lab controls: subtract from all samples. For field controls: only subtract if ratio > 1.

```{r}
Remove_reads_lab <-
  Lab_only %>%
  filter(name == "max_val") %>%
  select(seq, value)

Remove_reads_field <-
  Field_only %>%
  group_by(seq) %>%
  filter(min(ratio) > 1) %>%
  ungroup() %>%
  filter(name == "max_val") %>%
  select(seq, value)

Remove_reads <-
  bind_rows(Remove_reads_lab,
            Remove_reads_field) %>%
  group_by(seq) %>%
  filter(value == max(value)) %>%
  mutate(seq_remove = ceiling(value * 2))

Fungi_mat <- as.matrix(Fungi_ASV[,-1])
Remove_mat <- Fungi_mat
Remove_mat[] <- 0
Remove_mat[, Remove_reads$seq] <- rep(Remove_reads$seq_remove, each = nrow(Remove_mat))

Fungi_mat_clean <- Fungi_mat - Remove_mat
Fungi_mat_clean[Fungi_mat_clean < 0] <- 0
```

Remove empty OTUs after contamination subtraction.

```{r}
Fungi_mat_clean <- Fungi_mat_clean[, colSums(Fungi_mat_clean) > 0]
```

Count OTUs removed.

```{r}
ncol(Fungi_mat) - ncol(Fungi_mat_clean)
```

208 sequences were removed from all samples.

Convert back to tibble and add sample index.

```{r}
Fungi_ASV <-
  as_tibble(Fungi_mat_clean) %>%
  bind_cols(select(Fungi_ASV, "Index"), .)
```

Remove control samples from ASV table.

```{r}
blank_index <-
  Meta %>%
  filter(!Type %in% c("active", "passive")) %>%
  pull(Index)

Fungi_ASV <-
  Fungi_ASV %>%
  filter(!Index %in% blank_index)
```

## Check for empty samples

Count reads per sample after cleaning.

```{r}
Sample_reads <-
  Fungi_ASV %>%
  pivot_longer(-Index) %>%
  group_by(Index) %>%
  summarise(value = sum(value)) %>%
  arrange(value) %>%
  left_join(Meta)

Sample_reads
```

Two samples have < 1000 reads: 1DD30 (Drone) and 1DB30_1 (Burkhart split). Remove these samples. All others have > 29000 reads.

```{r}
empty_samples <-
  Sample_reads %>%
  filter(value < 1000) %>%
  pull(Index)

Fungi_ASV <-
  Fungi_ASV %>%
  filter(!Index %in% empty_samples)
```

Create long format ASV table.

```{r}
Fungi_ASV_long <-
  pivot_longer(Fungi_ASV, -Index, names_to = "seq", values_to = "reads")
```

# Subset for target group

Filter taxonomy table to retain only sequences in cleaned ASV table.

```{r}
Fungi_tax <-
  Fungi_tax %>%
  filter(seq %in% colnames(Fungi_ASV))
```

Visualize reads and OTUs by kingdom and phylum.

```{r}
Fungi_ASV_long %>%
  group_by(seq) %>%
  summarise(reads = sum(reads)) %>%
  left_join(Fungi_tax) %>%
  group_by(kingdom, phylum) %>%
  summarise(n_OTUs = n(),
            reads = sum(reads)) %>%
  pivot_longer(one_of(c("n_OTUs", "reads"))) %>%
  ggplot(aes(x = phylum, y = value)) +
  facet_grid(kingdom~name, scales = "free", space = "free") +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_y_log10() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90))
```

The primers amplify more than fungi. Notably, many plant sequences (Tracheophyta) are present. For this analysis, retain only sequences assigned to kingdom Fungi with phylum-level identification.

```{r}
Fungi_seq_keep <-
  Fungi_tax %>%
  filter(kingdom == "Fungi") %>%
  filter(!is.na(phylum)) %>%
  pull(seq)
```

Count excluded OTUs.

```{r}
ncol(Fungi_ASV[,-1]) - length(Fungi_seq_keep)
```

This excludes 6000 OTUs.

Compute proportion of reads and sequences retained.

```{r}
Fungi_ASV_long %>%
  group_by(seq) %>%
  summarise(reads = sum(reads)) %>%
  left_join(Fungi_tax) %>%
  mutate(cat = case_when(kingdom == "Fungi" & !is.na(phylum) ~ "Fungi",
                         kingdom == "Fungi" & is.na(phylum) ~ "na_Fungi",
                         kingdom != "Fungi" ~ kingdom)) %>%
  group_by(cat) %>%
  summarise(reads = sum(reads),
            n_seq = n()) %>%
  mutate(reads_prct = signif(reads/sum(reads)*100, 2),
         seq_prct = signif(n_seq/sum(n_seq)*100, 2))
```

This keeps 59% of sequences accounting for 78% of reads. The largest excluded group is Plantae (9.5% of sequences, 20% of reads).

Create cleaned ASV table with fungi only.

```{r}
Fungi_ASV_clean <-
  Fungi_ASV %>%
  select(one_of("Index", Fungi_seq_keep))

Fungi_ASV_clean_long <-
  pivot_longer(Fungi_ASV_clean, -Index, names_to = "seq", values_to = "reads")
```

# Prevalence

Plot prevalence distribution (number of samples per OTU).

```{r}
Fungi_ASV_clean_long %>%
  filter(reads > 0) %>%
  group_by(seq) %>%
  summarise(prev = n()) %>%
  ggplot(aes(x = prev)) +
  geom_histogram(binwidth = 5, fill = NA, color = "darkred") +
  theme_minimal() +
  labs(x = "Prevalence (number of samples)", y = "Number of OTUs")
```

# Species richness

Count OTUs by phylum in cleaned dataset.

```{r}
Fungi_ASV_clean_long %>%
  left_join(Fungi_tax) %>%
  select(seq, phylum) %>%
  distinct() %>%
  group_by(phylum) %>%
  summarise(n = n())
```

Compute rarefaction curves on cleaned dataset.

```{r}
rar_curve <-
  Fungi_ASV_clean %>%
  select(-Index) %>%
  as.matrix() %>%
  `rownames<-`(Fungi_ASV_clean$Index) %>%
  rarecurve(x = ., step = 1000, tidy = TRUE)

rar_curve %>%
  left_join(Meta, by = c("Site" = "Index")) %>%
  filter(grepl("active|passive", Type)) %>%
  ggplot(aes(x = Sample, y = Species, group = Site, colour = duration)) +
  geom_line() +
  facet_wrap(~Sampler) +
  theme_minimal() +
  labs(x = "Sequencing depth", y = "Observed richness")
```

Compute observed richness per sample.

```{r}
S_df <-
  Fungi_ASV_clean_long %>%
  group_by(Index) %>%
  filter(reads > 0) %>%
  summarise(S = n()) %>%
  left_join(Meta) %>%
  filter(Sampler != "Drone") %>%
  mutate(Sampler = factor(Sampler, levels = c("Kärcher", "Coriolis", "Sass", "Hepa",
                                               "Burkhart", "Electrostatic", "MWAC",
                                               "WSL_Filter"))) %>%
  mutate(taxon = "Fungi")
```

Plot richness by sampler, duration, and site.

```{r}
S_df %>%
  ggplot(aes(y = S, x = duration, colour = Sites)) +
  facet_grid(~Sampler, scales = "free_x") +
  geom_line(aes(group = Sites), linewidth = 0.2, colour = "grey") +
  geom_point() +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
        legend.position = "bottom") +
  labs(x = "Duration", y = "Observed richness (S)")
```

## Formal statistical model

Fit generalized linear model for richness.

```{r}
S_df_model <-
  S_df %>%
  filter(!is.na(S))

richness_glm <- glm(S ~ Sampler * duration + Sites,
                    data = S_df_model,
                    family = poisson(link = "log"))

summary(richness_glm)
```

Extract predicted richness and confidence intervals.

```{r}
richness_pred <-
  S_df_model %>%
  mutate(pred_S = predict(richness_glm, type = "response"),
         pred_SE = predict(richness_glm, se.fit = TRUE, type = "response")$se.fit,
         pred_lower = pred_S - 1.96 * pred_SE,
         pred_upper = pred_S + 1.96 * pred_SE)

richness_pred %>%
  select(Sampler, duration, Sites, S, pred_S, pred_lower, pred_upper) %>%
  head(20)
```

## Comparison with Burkhart

Compute fold-differences relative to Burkhart at 5 hours.

```{r}
S_df %>%
  filter(duration != "30 min") %>%
  select(Sites, Sampler, S) %>%
  pivot_wider(names_from = Sampler, values_from = S) %>%
  mutate(across(!one_of("Sites"), ~.x/Burkhart)) %>%
  pivot_longer(-Sites) %>%
  group_by(name) %>%
  summarise(mean = round(mean(value), 1), sd = round(sd(value), 1))
```

## Export richness data

Export richness for cross-group figures.

```{r}
write_tsv(S_df, here("Data", "Fungi", "out_intermediate", "Fungi_richness.tsv"))
```

# Faith's phylogenetic diversity

Placeholder for phylogenetic diversity analysis. Requires phylogenetic tree from FastTree (not yet built).

```{r}
if(exists("Fungi_tree")) {

  # Prepare community matrix
  comm_mat <-
    Fungi_ASV_clean %>%
    data.frame() %>%
    `rownames<-`(.$Index) %>%
    select(-Index) %>%
    as.matrix()

  # Compute Faith's PD
  pd_result <- picante::pd(comm_mat, Fungi_tree, include.root = FALSE)

  # Add to S_df
  S_df <-
    S_df %>%
    left_join(
      pd_result %>%
        tibble::rownames_to_column("Index") %>%
        select(Index, PD)
    )

  # Plot PD by sampler, duration, and site
  S_df %>%
    ggplot(aes(y = PD, x = duration, colour = Sites)) +
    facet_grid(~Sampler, scales = "free_x") +
    geom_line(aes(group = Sites), linewidth = 0.2, colour = "grey") +
    geom_point() +
    theme_minimal() +
    scale_color_brewer(palette = "Set1") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.position = "bottom") +
    labs(x = "Duration", y = "Faith's PD")

  # Export updated richness with PD
  write_tsv(S_df, here("Data", "Fungi", "out_intermediate", "Fungi_richness.tsv"))

} else {
  message("Phylogenetic tree not available. Skipping Faith's PD analysis.")
}
```

# Richness vs air volume

Compute air volume for active samplers.

```{r}
flow_rate <-
  Meta %>%
  filter(Type == "active" & Sampler != "Drone") %>%
  select(Sampler, duration) %>%
  distinct() %>%
  mutate(flow = case_when(
    Sampler == "Hepa" ~ 60,
    Sampler == "Kärcher" ~ 3000,
    Sampler == "Sass" ~ 300,
    Sampler == "Coriolis" ~ 300,
    Sampler == "Electrostatic" ~ 10,
    Sampler == "Burkhart" ~ 16.5
  )) %>%
  mutate(air_vol = case_when(
    duration == "30 min" ~ flow * 30,
    duration == "5 hours" ~ flow * 5 * 60
  ))
```

Plot richness vs air volume with linear model.

```{r}
S_df %>%
  left_join(flow_rate) %>%
  filter(Type == "active" & Sampler != "Drone") %>%
  ggplot(aes(x = air_vol, y = S, colour = Sampler, shape = duration)) +
  geom_point() +
  geom_smooth(aes(group = 1), method = "lm", colour = "darkblue", linewidth = 0.5) +
  scale_x_log10() +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  labs(x = "Air volume (L)", y = "Observed richness (S)")
```

Fit linear model for richness vs log air volume.

```{r}
S_vol_df <-
  S_df %>%
  left_join(flow_rate) %>%
  filter(Type == "active" & Sampler != "Drone")

vol_lm <- lm(S ~ log10(air_vol), data = S_vol_df)
summary(vol_lm)
```

# Sampling completeness

Compute fraction of pooled 5h reference community captured per sampler × duration. Pool all samplers at each site for 5h as reference.

```{r}
ref_community <-
  Fungi_ASV_clean_long %>%
  left_join(Meta) %>%
  filter(duration == "5 hours" | duration == "2 weeks") %>%
  group_by(Sites, seq) %>%
  summarise(present = sum(reads) > 0) %>%
  group_by(Sites) %>%
  summarise(S_reference = sum(present))

completeness_df <-
  Fungi_ASV_clean_long %>%
  left_join(Meta) %>%
  filter(Sampler != "Drone") %>%
  group_by(Sites, Sampler, duration, seq) %>%
  summarise(reads = sum(reads)) %>%
  group_by(Sites, Sampler, duration) %>%
  summarise(S_sampler = sum(reads > 0)) %>%
  left_join(ref_community) %>%
  mutate(fraction_detected = S_sampler / S_reference) %>%
  mutate(taxon = "Fungi")
```

Plot completeness by sampler and duration.

```{r}
completeness_df %>%
  ggplot(aes(x = Sampler, y = fraction_detected, fill = duration)) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal() +
  labs(x = "Sampler", y = "Fraction of reference community detected", fill = "Duration")
```

Export completeness data.

```{r}
write_tsv(completeness_df, here("Data", "Fungi", "out_intermediate", "Fungi_completeness.tsv"))
```

# Richness correlation across sites

Compute correlation matrix of richness across sampler × duration combinations.

```{r}
Cor_mat <-
  S_df %>%
  left_join(flow_rate) %>%
  mutate(Sampler_dur = paste(Sampler, duration)) %>%
  arrange(Type, duration, desc(flow)) %>%
  select(Sites, Sampler_dur, S) %>%
  pivot_wider(names_from = Sampler_dur, values_from = S) %>%
  select(-Sites) %>%
  cor(method = "pearson", use = "pairwise.complete.obs")
```

Plot correlation matrix.

```{r}
Cor_mat %>%
  corrplot(type = "lower", diag = FALSE, tl.col = "black", tl.cex = 0.8,
           method = "ellipse", na.label = "square", na.label.col = "white")
```

# Richness by taxonomic subgroup

Test whether sampler effects differ between Ascomycota and Basidiomycota.

```{r}
S_subgroup_df <-
  Fungi_ASV_clean_long %>%
  left_join(Fungi_tax) %>%
  filter(phylum %in% c("Ascomycota", "Basidiomycota")) %>%
  left_join(Meta) %>%
  filter(Sampler != "Drone") %>%
  group_by(Index, phylum) %>%
  filter(reads > 0) %>%
  summarise(S = n()) %>%
  left_join(Meta) %>%
  mutate(Sampler = factor(Sampler, levels = c("Kärcher", "Coriolis", "Sass", "Hepa",
                                               "Burkhart", "Electrostatic", "MWAC",
                                               "WSL_Filter")))
```

Plot richness by phylum.

```{r}
S_subgroup_df %>%
  ggplot(aes(y = S, x = duration, colour = Sites)) +
  facet_grid(phylum ~ Sampler, scales = "free_y") +
  geom_line(aes(group = Sites), linewidth = 0.2, colour = "grey") +
  geom_point() +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
        legend.position = "bottom") +
  labs(x = "Duration", y = "Observed richness (S)")
```

Fit GLM for each phylum separately.

```{r}
ascomycota_glm <- glm(S ~ Sampler * duration + Sites,
                      data = filter(S_subgroup_df, phylum == "Ascomycota"),
                      family = poisson(link = "log"))
summary(ascomycota_glm)

basidiomycota_glm <- glm(S ~ Sampler * duration + Sites,
                         data = filter(S_subgroup_df, phylum == "Basidiomycota"),
                         family = poisson(link = "log"))
summary(basidiomycota_glm)
```

Analyze Agaricomycetes specifically (fruiting body fungi).

```{r}
S_agaricomycetes_df <-
  Fungi_ASV_clean_long %>%
  left_join(Fungi_tax) %>%
  filter(class == "Agaricomycetes") %>%
  left_join(Meta) %>%
  filter(Sampler != "Drone") %>%
  group_by(Index) %>%
  filter(reads > 0) %>%
  summarise(S = n()) %>%
  left_join(Meta) %>%
  mutate(Sampler = factor(Sampler, levels = c("Kärcher", "Coriolis", "Sass", "Hepa",
                                               "Burkhart", "Electrostatic", "MWAC",
                                               "WSL_Filter")))
```

Plot Agaricomycetes richness.

```{r}
S_agaricomycetes_df %>%
  ggplot(aes(y = S, x = duration, colour = Sites)) +
  facet_grid(~ Sampler, scales = "free_x") +
  geom_line(aes(group = Sites), linewidth = 0.2, colour = "grey") +
  geom_point() +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
        legend.position = "bottom") +
  labs(x = "Duration", y = "Observed richness (Agaricomycetes)")
```

# Nestedness vs turnover

Use betapart to decompose Sørensen dissimilarity into nestedness and turnover components. Compare sampler pairs averaged across sites.

```{r}
comm_mat <-
  Fungi_ASV_clean %>%
  data.frame() %>%
  `rownames<-`(.$Index) %>%
  select(-Index) %>%
  as.matrix()

beta_result <- betapart::beta.pair(comm_mat, index.family = "sorensen")
```

Extract pairwise nestedness and turnover matrices.

```{r}
beta_sor_mat <- as.matrix(beta_result$beta.sor)
beta_sim_mat <- as.matrix(beta_result$beta.sim)
beta_sne_mat <- as.matrix(beta_result$beta.sne)
```

Create metadata for sampler × duration combinations.

```{r}
sample_meta <-
  Meta %>%
  filter(Index %in% Fungi_ASV_clean$Index) %>%
  filter(Sampler != "Drone") %>%
  mutate(Sampler_dur = paste(Sampler, duration))
```

Compute mean beta diversity components across sites for each sampler pair.

```{r}
beta_df_list <- list()

for(i in 1:(nrow(beta_sor_mat)-1)) {
  for(j in (i+1):nrow(beta_sor_mat)) {

    meta1 <- sample_meta[i, ]
    meta2 <- sample_meta[j, ]

    if(meta1$Sites == meta2$Sites & meta1$Sampler != meta2$Sampler) {
      beta_df_list[[length(beta_df_list) + 1]] <- tibble(
        Sampler1 = meta1$Sampler_dur,
        Sampler2 = meta2$Sampler_dur,
        Sites = meta1$Sites,
        beta_sor = beta_sor_mat[i, j],
        beta_sim = beta_sim_mat[i, j],
        beta_sne = beta_sne_mat[i, j]
      )
    }
  }
}

beta_df <- bind_rows(beta_df_list)
```

Compute mean across sites for each sampler pair.

```{r}
beta_summary <-
  beta_df %>%
  group_by(Sampler1, Sampler2) %>%
  summarise(
    beta_sor = mean(beta_sor),
    beta_sim = mean(beta_sim),
    beta_sne = mean(beta_sne)
  ) %>%
  mutate(taxon = "Fungi")
```

Plot nestedness vs turnover.

```{r}
beta_summary %>%
  select(Sampler1, Sampler2, beta_sim, beta_sne) %>%
  pivot_longer(c(beta_sim, beta_sne), names_to = "component", values_to = "value") %>%
  ggplot(aes(x = paste(Sampler1, Sampler2, sep = " vs "), y = value, fill = component)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_minimal() +
  labs(x = "Sampler pair", y = "Beta diversity component", fill = "Component") +
  scale_fill_brewer(palette = "Set1", labels = c("Nestedness", "Turnover"))
```

Export betapart results.

```{r}
write_tsv(beta_summary, here("Data", "Fungi", "out_intermediate", "Fungi_nestedness_turnover.tsv"))
```

# Rank-abundance correlation

Compute Spearman correlation on relative abundance across shared OTUs between sampler pairs.

```{r}
rel_abund_mat <-
  Fungi_ASV_clean %>%
  data.frame() %>%
  `rownames<-`(.$Index) %>%
  select(-Index) %>%
  as.matrix()

rel_abund_mat <- sweep(rel_abund_mat, 1, rowSums(rel_abund_mat), "/")
```

Compute rank-abundance correlation for each sampler pair at same site.

```{r}
rank_cor_list <- list()

for(i in 1:(nrow(rel_abund_mat)-1)) {
  for(j in (i+1):nrow(rel_abund_mat)) {

    meta1 <- sample_meta[i, ]
    meta2 <- sample_meta[j, ]

    if(meta1$Sites == meta2$Sites) {

      shared_otus <- (rel_abund_mat[i, ] > 0) & (rel_abund_mat[j, ] > 0)

      if(sum(shared_otus) > 2) {
        cor_test <- cor.test(rel_abund_mat[i, shared_otus],
                            rel_abund_mat[j, shared_otus],
                            method = "spearman")

        rank_cor_list[[length(rank_cor_list) + 1]] <- tibble(
          Sampler1 = meta1$Sampler_dur,
          Sampler2 = meta2$Sampler_dur,
          Sites = meta1$Sites,
          rho = cor_test$estimate,
          p_value = cor_test$p.value
        )
      }
    }
  }
}

rank_cor_df <- bind_rows(rank_cor_list)
```

Compute mean correlation across sites.

```{r}
rank_cor_summary <-
  rank_cor_df %>%
  group_by(Sampler1, Sampler2) %>%
  summarise(
    rho = mean(rho, na.rm = TRUE),
    p_value = mean(p_value, na.rm = TRUE)
  ) %>%
  mutate(taxon = "Fungi")
```

Create correlation matrix for plotting.

```{r}
rank_cor_mat <-
  rank_cor_summary %>%
  select(Sampler1, Sampler2, rho) %>%
  bind_rows(
    rank_cor_summary %>%
      select(Sampler1 = Sampler2, Sampler2 = Sampler1, rho)
  ) %>%
  pivot_wider(names_from = Sampler2, values_from = rho) %>%
  data.frame() %>%
  `rownames<-`(.$Sampler1) %>%
  select(-Sampler1) %>%
  as.matrix()

diag(rank_cor_mat) <- 1
```

Plot correlation matrix.

```{r}
rank_cor_mat %>%
  corrplot(type = "lower", diag = FALSE, tl.col = "black", tl.cex = 0.8,
           method = "color", na.label = "square", na.label.col = "white")
```

Export rank correlation results.

```{r}
write_tsv(rank_cor_summary, here("Data", "Fungi", "out_intermediate", "Fungi_rank_correlation.tsv"))
```

# Venn diagram

Compare detection overlap for 4 main active samplers at 5 hours.

```{r}
sampler_venn <-
  Fungi_ASV_clean_long %>%
  left_join(Meta) %>%
  filter(Sampler %in% c("Kärcher", "Sass", "Coriolis", "Hepa")) %>%
  filter(duration == "5 hours") %>%
  group_by(Sampler, seq) %>%
  summarise(reads = sum(reads)) %>%
  mutate(present = reads > 0) %>%
  select(-reads) %>%
  pivot_wider(names_from = Sampler, values_from = present) %>%
  select(-seq) %>%
  euler(shape = "ellipse")

eulerr_options(quantities = list(fontsize = 10, col = "black"))

plot(sampler_venn, legend = TRUE, quantities = TRUE)
```

# iNEXT sensitivity analysis

Use coverage-based rarefaction and extrapolation to confirm sampler ranking matches observed richness.

```{r}
inext_data <-
  Fungi_ASV_clean %>%
  left_join(Meta) %>%
  filter(Sampler != "Drone") %>%
  filter(duration == "5 hours" | duration == "2 weeks") %>%
  mutate(Sampler_dur = paste(Sampler, duration)) %>%
  select(Sampler_dur, -Index, -Sites, -Sampler, -duration) %>%
  group_by(Sampler_dur) %>%
  summarise(across(everything(), sum)) %>%
  data.frame() %>%
  `rownames<-`(.$Sampler_dur) %>%
  select(-Sampler_dur) %>%
  t()

inext_result <- iNEXT(inext_data, q = 0, datatype = "abundance", endpoint = max(colSums(inext_data)) * 2)
```

Plot iNEXT rarefaction curves.

```{r}
ggiNEXT(inext_result, type = 1, color.var = "Assemblage") +
  theme_minimal() +
  labs(title = "Coverage-based rarefaction (q=0)",
       x = "Number of individuals",
       y = "Species richness")
```

Extract asymptotic richness estimates.

```{r}
inext_result$AsyEst %>%
  filter(Diversity == "Species richness") %>%
  select(Assemblage, Observed, Estimator, s.e., LCL, UCL)
```

# Red list detection

Summarize detection rates of Swiss macrofungi checklist by sampler.

```{r}
swiss_redlist <-
  read_tsv(here("Data", "Fungi", "in_files", "redlist_taxonomy.tsv"))

STATUS <- c("LC", "NT", "VU", "EN", "CR", "EW", "EX", "DD")
Full_Name <- c("Least Concern", "Near Threatened", "Vulnerable", "Endangered",
               "Critically Endangered", "Extinct in the Wild", "Extinct", "Data Deficient")
Rank <- c(1, 2, 3, 4, 5, 6, 7, 0)

IUCN_Red_List_Status <- data.frame(STATUS, Full_Name, Rank)
```

Match detected species to red list.

```{r}
swiss_redlist_filt <-
  swiss_redlist %>%
  select(accepted_name, SPECIES, STATUS) %>%
  filter(!is.na(accepted_name)) %>%
  left_join(IUCN_Red_List_Status) %>%
  group_by(accepted_name) %>%
  filter(Rank == min(Rank)) %>%
  dplyr::rename(species = accepted_name) %>%
  group_by(species) %>%
  dplyr::slice(1)

Fungi_ASV_clean_long_rl <-
  Fungi_ASV_clean_long %>%
  left_join(Fungi_tax) %>%
  left_join(swiss_redlist_filt)
```

Compute detection rates by sampler.

```{r}
redlist_detection <-
  Fungi_ASV_clean_long_rl %>%
  left_join(Meta) %>%
  filter(Sampler != "Drone") %>%
  filter(duration == "5 hours" | duration == "2 weeks") %>%
  filter(!is.na(STATUS)) %>%
  group_by(Sampler, species) %>%
  summarise(detected = sum(reads) > 0) %>%
  group_by(Sampler) %>%
  summarise(n_species = sum(detected)) %>%
  mutate(total_checklist = length(unique(swiss_redlist_filt$species))) %>%
  mutate(detection_rate = n_species / total_checklist * 100)

redlist_detection
```

Plot detection rates.

```{r}
redlist_detection %>%
  ggplot(aes(x = Sampler, y = detection_rate)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  theme_minimal() +
  labs(x = "Sampler", y = "% of Swiss macrofungi checklist detected")
```

# Export results

Ensure all required exports are written.

```{r}
# Richness (already exported above)
if(!file.exists(here("Data", "Fungi", "out_intermediate", "Fungi_richness.tsv"))) {
  write_tsv(S_df, here("Data", "Fungi", "out_intermediate", "Fungi_richness.tsv"))
}

# Contamination (already exported above)
if(!file.exists(here("Data", "Fungi", "out_intermediate", "Fungi_contamination.tsv"))) {
  write_tsv(cont_summary, here("Data", "Fungi", "out_intermediate", "Fungi_contamination.tsv"))
}

# Completeness (already exported above)
if(!file.exists(here("Data", "Fungi", "out_intermediate", "Fungi_completeness.tsv"))) {
  write_tsv(completeness_df, here("Data", "Fungi", "out_intermediate", "Fungi_completeness.tsv"))
}

# Nestedness/turnover (already exported above)
if(!file.exists(here("Data", "Fungi", "out_intermediate", "Fungi_nestedness_turnover.tsv"))) {
  write_tsv(beta_summary, here("Data", "Fungi", "out_intermediate", "Fungi_nestedness_turnover.tsv"))
}

# Rank correlation (already exported above)
if(!file.exists(here("Data", "Fungi", "out_intermediate", "Fungi_rank_correlation.tsv"))) {
  write_tsv(rank_cor_summary, here("Data", "Fungi", "out_intermediate", "Fungi_rank_correlation.tsv"))
}
```

# Session info

Record R session information for reproducibility.

```{r}
sessionInfo()
```
