---
title: "Fungi data analysis"
output: html_notebook
---

# Setup

Load required libraries for analysis.

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(forcats)
library(here)
library(readr)
library(ggplot2)
library(vegan)
library(corrplot)
library(Biostrings)
library(decontam)
library(eulerr)
library(betapart)
library(ape)
library(picante)
library(iNEXT)
library(glmmTMB)
library(DHARMa)
```

Define the taxonomic group for this analysis.

```{r}
primer <- "Fungi"
```

Load metadata from the standardized file.

```{r}
Meta <- read_tsv(here("Data", "Meta_data.tsv"))
```

# Load data

Load ASV count table, taxonomy, representative sequences, and DNA concentrations.

```{r}
Fungi_ASV <- read_tsv(here("Data", primer, "out_final", "Fungi_ASV_glom.txt")) %>%
  mutate(Sample = gsub(".+?(NGS.+)", "\\1", Sample)) %>%
  dplyr::rename(Index = Sample)

Fungi_tax <- read_tsv(here("Data", primer, "out_final", "Fungi_taxa_80_glom.txt"))

Fungi_seq <- readDNAStringSet(here("Data", primer, "out_final", paste(primer, "Seq_glom", sep = "_")))

DNAC <- read_tsv(here("Data", "DNA_conc_air_sampler_comp.txt"))
```

Load phylogenetic tree (placeholder until trees are built).

```{r}
# Fungi_tree <- ape::read.tree(here("Data", "Fungi", "out_final", "Fungi_tree.nwk"))
```

# Fragment length distribution

Compute sequence lengths for all representative sequences.

```{r}
Fungi_len <-
  Fungi_seq %>%
  width() %>%
  tibble(len = .) %>%
  mutate(seq = names(Fungi_seq))
```

Plot histogram of sequence lengths.

```{r}
Fungi_len %>%
  group_by(len) %>%
  summarise(n = n()) %>%
  ggplot(aes(x = len, y = n)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(x = "Sequence length (bp)", y = "Number of ASVs")
```

Compute reads per ASV and plot against sequence length.

```{r}
Fungi_len_reads <-
  Fungi_ASV %>%
  pivot_longer(-Index, names_to = "seq") %>%
  group_by(seq) %>%
  summarise(prev = sum(value > 0),
            reads = sum(value)) %>%
  na.omit() %>%
  filter(reads < quantile(reads, 0.9)) %>%
  left_join(Fungi_len) %>%
  mutate(reads_o_prev = reads/prev)

Fungi_len_reads %>%
  ggplot(aes(x = len, y = reads_o_prev)) +
  geom_point(shape = 21, alpha = 0.5, size = 3) +
  theme_minimal() +
  labs(x = "Sequence length (bp)", y = "Reads per prevalence")
```

# Rarefaction (pre-cleaning)

Define major phyla to examine rarefaction curves by taxonomic group.

```{r}
major_phyla <- c("Ascomycota", "Basidiomycota", "Tracheophyta", "Other")
```

Compute rarefaction curves for each major phylum.

```{r}
rc_df <- map_df(major_phyla, function(ph) {

  if(ph == "Other"){
    seqs <- Fungi_tax %>%
      filter(!phylum %in% major_phyla) %>%
      pull(seq)
  } else{
    seqs <- Fungi_tax %>%
      filter(phylum == ph) %>%
      pull(seq)
  }

  asv_sub <- Fungi_ASV %>%
    select(Index, any_of(seqs)) %>%
    data.frame()

  rownames(asv_sub) <- asv_sub$Index
  asv_mat <- asv_sub[, -1] %>% as.matrix()

  rarecurve(
    asv_mat,
    step = 1000,
    label = FALSE,
    tidy = TRUE
  ) %>%
    dplyr::rename(Index = Site) %>%
    mutate(phylum = ph)
})

rc_plot_df <- rc_df %>%
  left_join(Meta, by = "Index")
```

Plot rarefaction curves faceted by phylum.

```{r}
ggplot(rc_plot_df, aes(x = Sample, y = Species, color = duration, group = Index)) +
  geom_line() +
  facet_wrap(~phylum, scales = "free_y", ncol = 2) +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Set1") +
  labs(x = "Sequencing depth", y = "ASV richness")
```

Plot rarefaction curves faceted by sampler.

```{r}
ggplot(rc_plot_df, aes(x = Sample, y = Species, color = phylum, group = paste(Index, phylum))) +
  geom_line() +
  facet_wrap(~Sampler, scales = "free_y") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Set1") +
  labs(x = "Sequencing depth", y = "ASV richness")
```

Sequencing depth varied between samples but reached saturation in all cases for all target taxa.

# Clean data

## Known sample issues

Sample notes from lab records:
- 9F2control2 got contaminated during DNA extraction (cannot be used as blank)
- 1DK5: Filter dropped on floor, only buffer kept
- 1DD30: No liquid left after 1st lysis
- 1CD_: No liquid left after 1st lysis
- 1DB30: Only small amount of liquid, half transferred
- 8RH5: Some buffer missing, reduced beads
- Plate1: Possible pipetting error column 1/2
- Well E3: Very little liquid

## Index hopping

Check for reads in unused indices (should be absent but can occur due to index hopping).

```{r}
index_hop <-
  Fungi_ASV %>%
  filter(!Index %in% Meta$Index) %>%
  pivot_longer(-Index) %>%
  filter(value > 0)

index_hop
```

Five unused indices have reads (maximum 2 reads per index).

```{r}
Fungi_ASV %>%
  select(unique(index_hop$name)) %>%
  colSums()
```

All reads are from highly abundant ASVs, indicating low-level index hopping. Remove unused indices from ASV table.

```{r}
Fungi_ASV <-
  Fungi_ASV %>%
  filter(Index %in%
           {filter(Meta, Sample_names != "BLANK8") %>%
             pull(Index)
             }
         )
```

## Check controls

### Decontam (prevalence-based)

Define which samples are blanks. Field controls are set to NA (excluded) as they contain genuine environmental sequences.

```{r}
Blanks <-
  Meta %>%
  filter(Index %in% Fungi_ASV$Index) %>%
  select(Index, Sample_names, Type) %>%
  filter(Sample_names != "BLANK8") %>%
  mutate(is_neg = case_when(Type %in% c("active", "passive") ~ FALSE,
                            Type %in% c("extr_control", "PCR_control") ~ TRUE,
                            Type == "field_control" ~ NA)) %>%
  arrange(Index)

all(Fungi_ASV$Index == Blanks$Index)
```

Run decontam with prevalence method.

```{r}
contamdf.prev <- isContaminant(as.matrix(Fungi_ASV[,-1]),
                               neg = Blanks$is_neg,
                               method = "prevalence")

n_contaminants <- sum(contamdf.prev$p < 0.1, na.rm = TRUE)
```

Decontam (prevalence method, p < 0.1) identifies `r n_contaminants` contaminants. This is expected for air eDNA where contaminants are also environmental organisms, making the prevalence signal weak.

### Visualize control contamination

Compute mean and max abundance ratios between controls and samples.

```{r}
cont_ratio <-
  Fungi_ASV %>%
  pivot_longer(-Index, names_to = "seq") %>%
  left_join(Meta) %>%
  select(Index, seq, Sample_names, Type, value) %>%
  mutate(Type = case_when(Type %in% c("active", "passive") ~ "Sample",
                          Type %in% c("extr_control", "PCR_control") ~ "Lab",
                          Type == "field_control" ~ "Field")) %>%
  group_by(Type, seq) %>%
  summarise(mean_val = round(mean(value), 2),
            max_val = max(value)) %>%
  pivot_longer(one_of(c("mean_val", "max_val"))) %>%
  pivot_wider(names_from = Type, values_from = value) %>%
  filter(Lab > 0 | Field > 0) %>%
  pivot_longer(!one_of("seq", "Sample", "name"), names_to = "Type") %>%
  mutate(ratio = round(value/Sample, 3))

Field_only <-
  cont_ratio %>%
  filter(Type == "Field") %>%
  filter(value > 0)

Lab_only <-
  cont_ratio %>%
  filter(Type == "Lab") %>%
  filter(value > 0)
```

### Lab controls

Blank8 got contaminated by a sample and is excluded. Count unique sequences in lab controls.

```{r}
Lab_only$seq %>% unique() %>% length()
```

Count sequences with ratio > 1 (more abundant in controls than samples).

```{r}
Lab_only %>%
  filter(name == "mean_val") %>%
  filter(ratio > 1) %>%
  nrow()
```

Summarize lab contaminant taxonomy.

```{r}
Lab_only %>%
  filter(name == "mean_val") %>%
  left_join(Fungi_tax, by = "seq") %>%
  mutate(contaminant = ratio > 1) %>%
  group_by(phylum, class, contaminant) %>%
  summarise(
    n_OTUs = n(),
    total_reads = sum(value),
    .groups = "drop"
  ) %>%
  arrange(desc(contaminant), desc(n_OTUs))
```

There are 28 sequences present across extraction blanks and PCR negatives. Six are more abundant in controls than samples (clear contaminants). The remaining are likely cross-contamination from true OTUs. Expected contaminant taxa include molds (Aspergillus, Penicillium, Cladosporium) which are omnipresent lab and air contaminants, skin-associated fungi (Malassezia) which are classical human contaminants, and soil fungi (Mortierella) which may indicate sample handling contamination.

### Field controls

Count unique sequences in field controls.

```{r}
Field_only$seq %>% unique() %>% length()
```

Count sequences with ratio > 1.

```{r}
Field_only %>%
  filter(name == "mean_val") %>%
  filter(ratio > 1) %>%
  nrow()
```

Count sequences with ratio > 10.

```{r}
Field_only %>%
  filter(name == "mean_val") %>%
  filter(ratio > 10) %>%
  nrow()
```

There are 542 sequences present across 14 field blanks, with 223 more abundant in controls than samples (119 with ratio > 10).

Create field control dataset for visualization.

```{r}
Field_cont <-
  Fungi_ASV %>%
  pivot_longer(-Index, names_to = "seq") %>%
  filter(seq %in% unique(Field_only$seq)) %>%
  left_join(Meta) %>%
  filter(Type == "field_control") %>%
  filter(value > 0) %>%
  left_join(
    {filter(cont_ratio, Type == "Field",
            name == "mean_val") %>%
        select(seq, ratio)
        }
  )
```

Plot contamination levels by sampler.

```{r}
Field_cont %>%
  mutate(ratio_cat = ifelse(ratio > 1, "> 1", "< 1")) %>%
  ggplot(aes(x = Sampler, y = value, colour = ratio_cat)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.6)) +
  scale_y_log10() +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  coord_flip() +
  labs(x = "Sampler", y = "Reads in field control", colour = "Ratio")
```

Plot number of contaminant OTUs by sampler.

```{r}
Field_cont %>%
  mutate(ratio_cat = ifelse(ratio > 1, "> 1", "< 1")) %>%
  group_by(ratio_cat, Sampler) %>%
  summarise(n = n()) %>%
  ggplot(aes(x = Sampler, y = n, colour = ratio_cat)) +
  geom_point(position = position_dodge(width = 0.4), size = 3) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  coord_flip() +
  labs(x = "Sampler", y = "Number of OTUs", colour = "Ratio")
```

Count total OTUs per sampler in field controls.

```{r}
Field_cont %>%
  group_by(Sampler) %>%
  summarise(n = n())
```

Summarize contamination by ratio categories.

```{r}
Field_cont %>%
  mutate(ratio_cat = case_when(
    ratio < 1 ~ "< 1",
    ratio > 1 & ratio < 10 ~ "> 1",
    ratio > 10 ~ "> 10")) %>%
  group_by(ratio_cat, Sampler) %>%
  summarise(n = n()) %>%
  pivot_wider(names_from = ratio_cat, values_from = n, values_fill = 0)
```

There are significant differences in contamination levels across samplers. Passive methods have very low detection (6-7 OTUs). Burkhart, Coriolis, and Sass have 4-19 OTUs. Kärcher, Electrostatic, HEPA, and Drone have 89-310 OTUs. For Kärcher, most OTUs have ratio < 1 (likely environmental not contaminant). Drone, Electrostatic, and HEPA show high ratios indicating true contamination.

Summarize field contaminant taxonomy by sampler.

```{r}
Field_cont %>%
  filter(ratio > 1) %>%
  left_join(Fungi_tax, by = "seq") %>%
  group_by(Sampler, phylum, class) %>%
  summarise(n_OTUs = n(), .groups = "drop") %>%
  arrange(Sampler, desc(n_OTUs))
```

Field control contaminants show sampler-specific patterns. High-flow active samplers (Kärcher, HEPA, Electrostatic) show contamination dominated by common air and surface fungi. The taxonomic composition reflects handling contamination and ambient laboratory air rather than the target environmental signal.

### Export contamination summary

Create contamination summary table for cross-group comparison.

```{r}
cont_summary_lab <-
  Lab_only %>%
  filter(name == "mean_val") %>%
  summarise(
    n_OTUs_lab = n(),
    n_ratio_gt1_lab = sum(ratio > 1),
    total_reads_lab = sum(value)
  )

cont_summary_field <-
  Field_cont %>%
  group_by(Sampler) %>%
  summarise(
    n_OTUs_field = n(),
    n_ratio_gt1 = sum(ratio > 1),
    n_ratio_gt10 = sum(ratio > 10),
    total_reads_field = sum(value)
  ) %>%
  mutate(taxon = "Fungi")

cont_summary <-
  cont_summary_field %>%
  bind_cols(cont_summary_lab) %>%
  select(taxon, Sampler, n_OTUs_field, n_ratio_gt1, n_ratio_gt10,
         total_reads_field, n_OTUs_lab, n_ratio_gt1_lab)
```

Export contamination summary to intermediate output folder.

```{r}
write_tsv(cont_summary, here("Data", "Fungi", "out_intermediate", "Fungi_contamination.tsv"))
```

## Exclude contaminants

Remove reads from contaminant OTUs using 2× max subtraction method. For lab controls: subtract from all samples. For field controls: only subtract if ratio > 1.

```{r}
Remove_reads_lab <-
  Lab_only %>%
  filter(name == "max_val") %>%
  select(seq, value)

Remove_reads_field <-
  Field_only %>%
  group_by(seq) %>%
  filter(min(ratio) > 1) %>%
  ungroup() %>%
  filter(name == "max_val") %>%
  select(seq, value)

Remove_reads <-
  bind_rows(Remove_reads_lab,
            Remove_reads_field) %>%
  group_by(seq) %>%
  filter(value == max(value)) %>%
  mutate(seq_remove = ceiling(value * 2))

Fungi_mat <- as.matrix(Fungi_ASV[,-1])
Remove_mat <- Fungi_mat
Remove_mat[] <- 0
Remove_mat[, Remove_reads$seq] <- rep(Remove_reads$seq_remove, each = nrow(Remove_mat))

Fungi_mat_clean <- Fungi_mat - Remove_mat
Fungi_mat_clean[Fungi_mat_clean < 0] <- 0
```

Remove empty OTUs after contamination subtraction.

```{r}
Fungi_mat_clean <- Fungi_mat_clean[, colSums(Fungi_mat_clean) > 0]
```

Count OTUs removed.

```{r}
ncol(Fungi_mat) - ncol(Fungi_mat_clean)
```

208 sequences were removed from all samples.

Convert back to tibble and add sample index.

```{r}
Fungi_ASV <-
  as_tibble(Fungi_mat_clean) %>%
  bind_cols(select(Fungi_ASV, "Index"), .)
```

Remove control samples from ASV table.

```{r}
blank_index <-
  Meta %>%
  filter(!Type %in% c("active", "passive")) %>%
  pull(Index)

Fungi_ASV <-
  Fungi_ASV %>%
  filter(!Index %in% blank_index)
```

## Check for empty samples

Count reads per sample after cleaning.

```{r}
Sample_reads <-
  Fungi_ASV %>%
  pivot_longer(-Index) %>%
  group_by(Index) %>%
  summarise(value = sum(value)) %>%
  arrange(value) %>%
  left_join(Meta)

Sample_reads
```

Two samples have < 1000 reads: 1DD30 (Drone) and 1DB30_1 (Burkhart split). Remove these samples. All others have > 29000 reads.

```{r}
empty_samples <-
  Sample_reads %>%
  filter(value < 1000) %>%
  pull(Index)

Fungi_ASV <-
  Fungi_ASV %>%
  filter(!Index %in% empty_samples)
```

Create long format ASV table.

```{r}
Fungi_ASV_long <-
  pivot_longer(Fungi_ASV, -Index, names_to = "seq", values_to = "reads")
```

# Subset for target group

Filter taxonomy table to retain only sequences in cleaned ASV table.

```{r}
Fungi_tax <-
  Fungi_tax %>%
  filter(seq %in% colnames(Fungi_ASV))
```

Visualize reads and OTUs by kingdom and phylum.

```{r}
Fungi_ASV_long %>%
  group_by(seq) %>%
  summarise(reads = sum(reads)) %>%
  left_join(Fungi_tax) %>%
  group_by(kingdom, phylum) %>%
  summarise(n_OTUs = n(),
            reads = sum(reads)) %>%
  pivot_longer(one_of(c("n_OTUs", "reads"))) %>%
  ggplot(aes(x = phylum, y = value)) +
  facet_grid(kingdom~name, scales = "free", space = "free") +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_y_log10() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90))
```

The primers amplify more than fungi. Notably, many plant sequences (Tracheophyta) are present. For this analysis, retain only sequences assigned to kingdom Fungi with phylum-level identification.

```{r}
Fungi_seq_keep <-
  Fungi_tax %>%
  filter(kingdom == "Fungi") %>%
  filter(!is.na(phylum)) %>%
  pull(seq)
```

Count excluded OTUs.

```{r}
ncol(Fungi_ASV[,-1]) - length(Fungi_seq_keep)
```

This excludes 6013 OTUs.

Compute proportion of reads and sequences retained.

```{r}
Fungi_ASV_long %>%
  group_by(seq) %>%
  summarise(reads = sum(reads)) %>%
  left_join(Fungi_tax) %>%
  mutate(cat = case_when(kingdom == "Fungi" & !is.na(phylum) ~ "Fungi",
                         kingdom == "Fungi" & is.na(phylum) ~ "na_Fungi",
                         kingdom != "Fungi" ~ kingdom)) %>%
  group_by(cat) %>%
  summarise(reads = sum(reads),
            n_seq = n()) %>%
  mutate(reads_prct = signif(reads/sum(reads)*100, 2),
         seq_prct = signif(n_seq/sum(n_seq)*100, 2))
```

This keeps 59% of sequences accounting for 78% of reads. The largest excluded group is Plantae (9.5% of sequences, 20% of reads).

Create cleaned ASV table with fungi only.

```{r}
Fungi_ASV_clean <-
  Fungi_ASV %>%
  select(one_of("Index", Fungi_seq_keep))

Fungi_ASV_clean_long <-
  pivot_longer(Fungi_ASV_clean, -Index, names_to = "seq", values_to = "reads")
```

# Prevalence

Plot prevalence distribution (number of samples per OTU).

```{r}
Fungi_ASV_clean_long %>%
  filter(reads > 0) %>%
  group_by(seq) %>%
  summarise(prev = n()) %>%
  ggplot(aes(x = prev)) +
  geom_histogram(binwidth = 5, fill = NA, color = "darkred") +
  theme_minimal() +
  labs(x = "Prevalence (number of samples)", y = "Number of OTUs")
```

# Species richness

Count OTUs by phylum in cleaned dataset.

```{r}
Fungi_ASV_clean_long %>%
  left_join(Fungi_tax) %>%
  select(seq, phylum) %>%
  distinct() %>%
  group_by(phylum) %>%
  summarise(n = n())
```

Compute rarefaction curves on cleaned dataset.

```{r}
rar_curve <-
  Fungi_ASV_clean %>%
  select(-Index) %>%
  as.matrix() %>%
  `rownames<-`(Fungi_ASV_clean$Index) %>%
  rarecurve(x = ., step = 1000, tidy = TRUE)

rar_curve %>%
  left_join(Meta, by = c("Site" = "Index")) %>%
  filter(grepl("active|passive", Type)) %>%
  ggplot(aes(x = Sample, y = Species, group = Site, colour = duration)) +
  geom_line() +
  facet_wrap(~Sampler) +
  theme_minimal() +
  labs(x = "Sequencing depth", y = "Observed richness")
```

Compute observed richness per sample.

```{r}
S_df <-
  Fungi_ASV_clean_long %>%
  group_by(Index) %>%
  filter(reads > 0) %>%
  summarise(S = n()) %>%
  left_join(Meta) %>%
  filter(Sampler != "Drone") %>%
  mutate(Sampler = factor(Sampler, levels = c("Kärcher", "Coriolis", "Sass", "Hepa",
                                               "Burkhart", "Electrostatic", "MWAC",
                                               "WSL_Filter"))) %>%
  mutate(taxon = "Fungi")
```

Plot richness by sampler, duration, and site.

```{r}
S_df %>%
  ggplot(aes(y = S, x = duration, colour = Sites)) +
  facet_grid(~Sampler, scales = "free_x") +
  geom_line(aes(group = Sites), linewidth = 0.2, colour = "grey") +
  geom_point() +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
        legend.position = "bottom") +
  labs(x = "Duration", y = "Observed richness (S)")
```

## Formal statistical model

Sites are a blocking factor (random sample of locations). We fit a random intercept per site. Fit main effects model (no interaction) since passive samplers only have one duration level, making the full interaction uninterpretable.

Overdispersion is ubiquitous in metabarcoding richness data, so we use negative binomial family directly.

```{r}
S_df_model <-
  S_df %>%
  filter(!is.na(S))

richness_glmm <- glmmTMB(S ~ Sampler + duration + (1|Sites),
                         data = S_df_model,
                         family = nbinom2)

summary(richness_glmm)
```

Check residuals to verify model adequacy.

```{r}
sim_res <- simulateResiduals(richness_glmm, n = 1000)
plot(sim_res)
```

Extract predicted richness and confidence intervals.

```{r}
# Get predictions and standard errors
pred_obj <- predict(richness_glmm, se.fit = TRUE, type = "response")

richness_pred <-
  S_df_model %>%
  mutate(
    pred_S = pred_obj$fit,
    pred_SE = pred_obj$se.fit,
    pred_lower = pred_S - 1.96 * pred_SE,
    pred_upper = pred_S + 1.96 * pred_SE
  )

richness_pred %>%
  select(Sampler, duration, Sites, S, pred_S, pred_lower, pred_upper) %>%
  head(20)
```

Plot observed richness with model predictions overlaid.

```{r}
ggplot(richness_pred, aes(x = duration, y = S)) +
  facet_grid(~Sampler, scales = "free_x") +
  # Observed data
  geom_line(aes(group = Sites, colour = Sites), linewidth = 0.2, colour = "grey") +
  geom_point(aes(colour = Sites)) +
  # Model predictions
  geom_ribbon(aes(y = pred_S, ymin = pred_lower, ymax = pred_upper, group = 1),
              alpha = 0.2, fill = "darkblue") +
  geom_line(aes(y = pred_S, group = 1), colour = "darkblue", linewidth = 1) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
        legend.position = "bottom") +
  labs(x = "Duration", y = "Observed richness (S)")
```

## Comparison with Burkhart

Compute fold-differences relative to Burkhart at 5 hours.

```{r}
S_df %>%
  filter(duration != "30 min") %>%
  select(Sites, Sampler, S) %>%
  pivot_wider(names_from = Sampler, values_from = S) %>%
  mutate(across(!one_of("Sites"), ~.x/Burkhart)) %>%
  pivot_longer(-Sites) %>%
  group_by(name) %>%
  summarise(mean = round(mean(value), 1), sd = round(sd(value), 1))
```

## Export richness data

Export richness for cross-group figures.

```{r}
write_tsv(S_df, here("Data", "Fungi", "out_intermediate", "Fungi_richness.tsv"))
```

# Faith's phylogenetic diversity

Placeholder for phylogenetic diversity analysis. Requires phylogenetic tree from FastTree (not yet built).

```{r}
if(exists("Fungi_tree")) {

  # Prepare community matrix
  comm_mat <-
    Fungi_ASV_clean %>%
    data.frame() %>%
    `rownames<-`(.$Index) %>%
    select(-Index) %>%
    as.matrix()

  # Compute Faith's PD
  pd_result <- picante::pd(comm_mat, Fungi_tree, include.root = FALSE)

  # Add to S_df
  S_df <-
    S_df %>%
    left_join(
      pd_result %>%
        tibble::rownames_to_column("Index") %>%
        select(Index, PD)
    )

  # Plot PD by sampler, duration, and site
  S_df %>%
    ggplot(aes(y = PD, x = duration, colour = Sites)) +
    facet_grid(~Sampler, scales = "free_x") +
    geom_line(aes(group = Sites), linewidth = 0.2, colour = "grey") +
    geom_point() +
    theme_minimal() +
    scale_color_brewer(palette = "Set1") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.position = "bottom") +
    labs(x = "Duration", y = "Faith's PD")

  # Export updated richness with PD
  write_tsv(S_df, here("Data", "Fungi", "out_intermediate", "Fungi_richness.tsv"))

} else {
  message("Phylogenetic tree not available. Skipping Faith's PD analysis.")
}
```

# Richness vs air volume

Compute air volume for active samplers.

```{r}
flow_rate <-
  Meta %>%
  filter(Type == "active" & Sampler != "Drone") %>%
  select(Sampler, duration) %>%
  distinct() %>%
  mutate(flow = case_when(
    Sampler == "Hepa" ~ 60,
    Sampler == "Kärcher" ~ 3000,
    Sampler == "Sass" ~ 300,
    Sampler == "Coriolis" ~ 300,
    Sampler == "Electrostatic" ~ 10,
    Sampler == "Burkhart" ~ 16.5
  )) %>%
  mutate(air_vol = case_when(
    duration == "30 min" ~ flow * 30,
    duration == "5 hours" ~ flow * 5 * 60
  ))
```

Plot richness vs air volume with linear model.

```{r}
S_df %>%
  left_join(flow_rate) %>%
  filter(Type == "active" & Sampler != "Drone") %>%
  ggplot(aes(x = air_vol, y = S, colour = Sampler, shape = duration)) +
  geom_point() +
  geom_smooth(aes(group = 1), method = "lm", colour = "darkblue", linewidth = 0.5) +
  scale_x_log10() +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  labs(x = "Air volume (L)", y = "Observed richness (S)")
```

Fit linear model for richness vs log air volume.

```{r}
S_vol_df <-
  S_df %>%
  left_join(flow_rate) %>%
  filter(Type == "active" & Sampler != "Drone")

vol_lm <- lm(S ~ log10(air_vol), data = S_vol_df)
summary(vol_lm)
```

# Sampling completeness

Assess what fraction of the site community each sampler detects. Reference community = ALL species detected at a site across all samplers and durations (pooled). This answers: "Of all species we know are present, what fraction does this sampler × duration detect?"

Build reference community per site (pool all samplers × durations).

```{r}
ref_community_species <-
  Fungi_ASV_clean_long %>%
  left_join(Meta) %>%
  filter(Sampler != "Drone") %>%
  filter(reads > 0) %>%
  group_by(Sites, seq) %>%
  summarise(in_reference = TRUE)

ref_community_richness <-
  ref_community_species %>%
  group_by(Sites) %>%
  summarise(S_reference = n())
```

Compute species-level completeness (fraction of reference species detected).

```{r}
completeness_df <-
  Fungi_ASV_clean_long %>%
  left_join(Meta) %>%
  filter(Sampler != "Drone") %>%
  filter(reads > 0) %>%
  group_by(Sites, Sampler, duration) %>%
  summarise(S_sampler = n_distinct(seq), .groups = "drop") %>%
  left_join(ref_community_richness) %>%
  mutate(
    fraction_detected = S_sampler / S_reference,
    taxon = "Fungi"
  )
```

Plot completeness by sampler and duration.

```{r}
completeness_df %>%
  ggplot(aes(x = Sampler, y = fraction_detected, fill = duration)) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal() +
  labs(x = "Sampler", y = "Fraction of site community detected", fill = "Duration") +
  ylim(0, 1)
```

Export completeness data.

```{r}
write_tsv(completeness_df, here("Data", "Fungi", "out_intermediate", "Fungi_completeness.tsv"))
```

# Richness correlation across sites

Compute correlation matrix of richness across sampler × duration combinations.

```{r}
Cor_mat <-
  S_df %>%
  left_join(flow_rate) %>%
  mutate(Sampler_dur = paste(Sampler, duration)) %>%
  arrange(Type, duration, desc(flow)) %>%
  select(Sites, Sampler_dur, S) %>%
  pivot_wider(names_from = Sampler_dur, values_from = S) %>%
  select(-Sites) %>%
  cor(method = "pearson", use = "pairwise.complete.obs")
```

Plot correlation matrix.

```{r}
Cor_mat %>%
  corrplot(type = "lower", diag = FALSE, tl.col = "black", tl.cex = 0.8,
           method = "ellipse", na.label = "square", na.label.col = "white")
```

# Richness by taxonomic subgroup

Test whether sampler effects differ between Ascomycota and Basidiomycota.

```{r}
S_subgroup_df <-
  Fungi_ASV_clean_long %>%
  left_join(Fungi_tax) %>%
  filter(phylum %in% c("Ascomycota", "Basidiomycota")) %>%
  left_join(Meta) %>%
  filter(Sampler != "Drone") %>%
  group_by(Index, phylum) %>%
  filter(reads > 0) %>%
  summarise(S = n()) %>%
  left_join(Meta) %>%
  mutate(Sampler = factor(Sampler, levels = c("Kärcher", "Coriolis", "Sass", "Hepa",
                                               "Burkhart", "Electrostatic", "MWAC",
                                               "WSL_Filter")))
```

Plot richness by phylum.

```{r}
S_subgroup_df %>%
  ggplot(aes(y = S, x = duration, colour = Sites)) +
  facet_grid(phylum ~ Sampler, scales = "free_y") +
  geom_line(aes(group = Sites), linewidth = 0.2, colour = "grey") +
  geom_point() +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
        legend.position = "bottom") +
  labs(x = "Duration", y = "Observed richness (S)")
```

Fit GLMM for each phylum separately (main effects only, negative binomial family).

```{r}
ascomycota_glmm <- glmmTMB(S ~ Sampler + duration + (1|Sites),
                           data = filter(S_subgroup_df, phylum == "Ascomycota"),
                           family = nbinom2)
summary(ascomycota_glmm)

basidiomycota_glmm <- glmmTMB(S ~ Sampler + duration + (1|Sites),
                              data = filter(S_subgroup_df, phylum == "Basidiomycota"),
                              family = nbinom2)
summary(basidiomycota_glmm)
```

Analyze Agaricomycetes specifically (fruiting body fungi).

```{r}
S_agaricomycetes_df <-
  Fungi_ASV_clean_long %>%
  left_join(Fungi_tax) %>%
  filter(class == "Agaricomycetes") %>%
  left_join(Meta) %>%
  filter(Sampler != "Drone") %>%
  group_by(Index) %>%
  filter(reads > 0) %>%
  summarise(S = n()) %>%
  left_join(Meta) %>%
  mutate(Sampler = factor(Sampler, levels = c("Kärcher", "Coriolis", "Sass", "Hepa",
                                               "Burkhart", "Electrostatic", "MWAC",
                                               "WSL_Filter")))
```

Plot Agaricomycetes richness.

```{r}
S_agaricomycetes_df %>%
  ggplot(aes(y = S, x = duration, colour = Sites)) +
  facet_grid(~ Sampler, scales = "free_x") +
  geom_line(aes(group = Sites), linewidth = 0.2, colour = "grey") +
  geom_point() +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
        legend.position = "bottom") +
  labs(x = "Duration", y = "Observed richness (Agaricomycetes)")
```

# Nestedness vs turnover

Use betapart to decompose Sørensen dissimilarity into nestedness and turnover components. Compare sampler pairs averaged across sites.

Build community matrix and sample metadata from the same filtered set to ensure row alignment.

```{r}
non_drone_idx <-
  Meta %>%
  filter(Index %in% Fungi_ASV_clean$Index) %>%
  filter(Sampler != "Drone") %>%
  pull(Index)

comm_mat <-
  Fungi_ASV_clean %>%
  filter(Index %in% non_drone_idx) %>%
  data.frame() %>%
  `rownames<-`(.$Index) %>%
  select(-Index) %>%
  as.matrix()

sample_meta <-
  tibble(Index = rownames(comm_mat)) %>%
  left_join(Meta) %>%
  mutate(Sampler_dur = paste(Sampler, duration))
```

Compute pairwise beta diversity.

```{r}
beta_result <- betapart::beta.pair(comm_mat, index.family = "sorensen")
```

Extract pairwise nestedness and turnover matrices.

```{r}
beta_sor_mat <- as.matrix(beta_result$beta.sor)
beta_sim_mat <- as.matrix(beta_result$beta.sim)
beta_sne_mat <- as.matrix(beta_result$beta.sne)
```

Compute mean beta diversity components across sites for each sampler pair.

```{r}
beta_df_list <- list()

for(i in 1:(nrow(beta_sor_mat)-1)) {
  for(j in (i+1):nrow(beta_sor_mat)) {

    meta1 <- sample_meta[i, ]
    meta2 <- sample_meta[j, ]

    if(meta1$Sites == meta2$Sites & meta1$Sampler != meta2$Sampler) {
      beta_df_list[[length(beta_df_list) + 1]] <- tibble(
        Sampler1 = meta1$Sampler,
        Sampler2 = meta2$Sampler,
        duration1 = meta1$duration,
        duration2 = meta2$duration,
        Sites = meta1$Sites,
        beta_sor = beta_sor_mat[i, j],
        beta_sim = beta_sim_mat[i, j],
        beta_sne = beta_sne_mat[i, j]
      )
    }
  }
}

beta_df <- bind_rows(beta_df_list)
```

Compute mean across sites for each sampler pair.

```{r}
beta_summary <-
  beta_df %>%
  group_by(Sampler1, Sampler2, duration1, duration2) %>%
  summarise(
    beta_sor = mean(beta_sor),
    beta_sim = mean(beta_sim),
    beta_sne = mean(beta_sne),
    .groups = "drop"
  ) %>%
  mutate(taxon = "Fungi")
```

Plot nestedness vs turnover.

```{r}
beta_summary %>%
  mutate(comparison = paste(Sampler1, duration1, "vs", Sampler2, duration2)) %>%
  select(comparison, beta_sim, beta_sne) %>%
  pivot_longer(c(beta_sim, beta_sne), names_to = "component", values_to = "value") %>%
  ggplot(aes(x = comparison, y = value, fill = component)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_minimal() +
  labs(x = "Sampler pair", y = "Beta diversity component", fill = "Component") +
  scale_fill_brewer(palette = "Set1", labels = c("Turnover", "Nestedness"))
```

Export betapart results.

```{r}
write_tsv(beta_summary, here("Data", "Fungi", "out_intermediate", "Fungi_nestedness_turnover.tsv"))
```

# Rank-abundance correlation

Compute Spearman correlation on relative abundance across shared OTUs between sampler pairs. Use the same filtered index set as betapart to ensure alignment.

```{r}
rel_abund_mat <-
  Fungi_ASV_clean %>%
  filter(Index %in% non_drone_idx) %>%
  data.frame() %>%
  `rownames<-`(.$Index) %>%
  select(-Index) %>%
  as.matrix()

rel_abund_mat <- sweep(rel_abund_mat, 1, rowSums(rel_abund_mat), "/")
```

Compute rank-abundance correlation for each sampler pair at same site.

```{r}
rank_cor_list <- list()

for(i in 1:(nrow(rel_abund_mat)-1)) {
  for(j in (i+1):nrow(rel_abund_mat)) {

    meta1 <- sample_meta[i, ]
    meta2 <- sample_meta[j, ]

    if(meta1$Sites == meta2$Sites) {

      shared_otus <- (rel_abund_mat[i, ] > 0) & (rel_abund_mat[j, ] > 0)

      if(sum(shared_otus) > 2) {
        cor_test <- cor.test(rel_abund_mat[i, shared_otus],
                            rel_abund_mat[j, shared_otus],
                            method = "spearman")

        rank_cor_list[[length(rank_cor_list) + 1]] <- tibble(
          Sampler1 = meta1$Sampler,
          Sampler2 = meta2$Sampler,
          duration1 = meta1$duration,
          duration2 = meta2$duration,
          Sites = meta1$Sites,
          rho = cor_test$estimate,
          p_value = cor_test$p.value
        )
      }
    }
  }
}

rank_cor_df <- bind_rows(rank_cor_list)
```

Compute mean correlation across sites.

```{r}
rank_cor_summary <-
  rank_cor_df %>%
  group_by(Sampler1, Sampler2, duration1, duration2) %>%
  summarise(
    rho = mean(rho, na.rm = TRUE),
    p_value = mean(p_value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(taxon = "Fungi")
```

Create correlation matrix for plotting.

```{r}
rank_cor_mat <-
  rank_cor_summary %>%
  mutate(
    Sampler_dur1 = paste(Sampler1, duration1),
    Sampler_dur2 = paste(Sampler2, duration2)
  ) %>%
  select(Sampler_dur1, Sampler_dur2, rho) %>%
  bind_rows(
    rank_cor_summary %>%
      mutate(
        Sampler_dur1 = paste(Sampler1, duration1),
        Sampler_dur2 = paste(Sampler2, duration2)
      ) %>%
      select(Sampler_dur1 = Sampler_dur2, Sampler_dur2 = Sampler_dur1, rho)
  ) %>%
  pivot_wider(names_from = Sampler_dur2, values_from = rho) %>%
  data.frame() %>%
  `rownames<-`(.$Sampler_dur1) %>%
  select(-Sampler_dur1) %>%
  as.matrix()

diag(rank_cor_mat) <- 1
```

Plot correlation matrix.

```{r}
rank_cor_mat %>%
  corrplot(type = "lower", diag = FALSE, tl.col = "black", tl.cex = 0.8,
           method = "color", na.label = "square", na.label.col = "white")
```

Export rank correlation results.

```{r}
write_tsv(rank_cor_summary, here("Data", "Fungi", "out_intermediate", "Fungi_rank_correlation.tsv"))
```

# Venn diagram

Compare detection overlap for 4 main active samplers at 5 hours.

```{r}
sampler_venn <-
  Fungi_ASV_clean_long %>%
  left_join(Meta) %>%
  filter(Sampler %in% c("Kärcher", "Sass", "Coriolis", "Hepa")) %>%
  filter(duration == "5 hours") %>%
  group_by(Sampler, seq) %>%
  summarise(reads = sum(reads)) %>%
  mutate(present = reads > 0) %>%
  select(-reads) %>%
  pivot_wider(names_from = Sampler, values_from = present) %>%
  select(-seq) %>%
  euler(shape = "ellipse")

eulerr_options(quantities = list(fontsize = 10, col = "black"))

plot(sampler_venn, legend = TRUE, quantities = TRUE)
```

# Richness estimation sensitivity

Use coverage-based richness estimation at 95% coverage to confirm sampler rankings match observed richness. Use `nboot = 0` to skip confidence intervals for speed on 100+ samples. Consistent coverage-based approach throughout the analysis.

```{r}
# Prepare per-sample abundance data
sample_abundance <-
  Fungi_ASV_clean_long %>%
  left_join(Meta) %>%
  filter(Sampler != "Drone") %>%
  filter(reads > 0) %>%
  group_by(Index) %>%
  group_split()

# Coverage-based estimation at 95% per sample (nboot = 0 for speed)
sensitivity_df <-
  map_df(sample_abundance, function(samp) {
    idx <- unique(samp$Index)
    abund_vec <- samp$reads

    est <- tryCatch(
      estimateD(abund_vec,
                datatype = "abundance",
                base = "coverage",
                level = 0.95,
                nboot = 0) %>%
        filter(Order.q == 0),
      error = function(e) NULL
    )

    if(is.null(est)) return(NULL)

    tibble(
      Index = idx,
      S_coverage95 = est$qD
    )
  }) %>%
  left_join(Meta) %>%
  left_join(S_df %>% select(Index, S))
```

Plot observed vs coverage-standardized richness.

```{r}
sensitivity_df %>%
  filter(!is.na(S_coverage95)) %>%
  pivot_longer(c(S, S_coverage95), names_to = "metric", values_to = "richness") %>%
  mutate(metric = ifelse(metric == "S", "Observed", "Coverage-standardized (95%)")) %>%
  ggplot(aes(y = richness, x = duration, colour = Sites)) +
  facet_grid(metric ~ Sampler, scales = "free_y") +
  geom_line(aes(group = paste(Sites, metric)), linewidth = 0.2, colour = "grey") +
  geom_point() +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
        legend.position = "bottom") +
  labs(x = "Duration", y = "Species richness")
```

Coverage-standardized richness at 95% coverage confirms that sampler rankings are robust to differences in sequencing depth. The relative performance of samplers remains consistent under coverage standardization.

# Red list detection

Summarize detection rates of Swiss macrofungi checklist by sampler.

```{r}
swiss_redlist <-
  read_tsv(here("Data", "Fungi", "in_files", "redlist_taxonomy.tsv"))

STATUS <- c("LC", "NT", "VU", "EN", "CR", "EW", "EX", "DD")
Full_Name <- c("Least Concern", "Near Threatened", "Vulnerable", "Endangered",
               "Critically Endangered", "Extinct in the Wild", "Extinct", "Data Deficient")
Rank <- c(1, 2, 3, 4, 5, 6, 7, 0)

IUCN_Red_List_Status <- data.frame(STATUS, Full_Name, Rank)
```

Match detected species to red list.

```{r}
swiss_redlist_filt <-
  swiss_redlist %>%
  select(accepted_name, SPECIES, STATUS) %>%
  filter(!is.na(accepted_name)) %>%
  left_join(IUCN_Red_List_Status) %>%
  group_by(accepted_name) %>%
  filter(Rank == min(Rank)) %>%
  dplyr::rename(species = accepted_name) %>%
  group_by(species) %>%
  dplyr::slice(1)

Fungi_ASV_clean_long_rl <-
  Fungi_ASV_clean_long %>%
  left_join(Fungi_tax) %>%
  left_join(swiss_redlist_filt)
```

Compute detection rates by sampler.

```{r}
redlist_detection <-
  Fungi_ASV_clean_long_rl %>%
  left_join(Meta) %>%
  filter(Sampler != "Drone") %>%
  filter(duration == "5 hours" | duration == "2 weeks") %>%
  filter(!is.na(STATUS)) %>%
  group_by(Sampler, species) %>%
  summarise(detected = sum(reads) > 0) %>%
  group_by(Sampler) %>%
  summarise(n_species = sum(detected)) %>%
  mutate(total_checklist = length(unique(swiss_redlist_filt$species))) %>%
  mutate(detection_rate = n_species / total_checklist * 100)

redlist_detection
```

Plot detection rates.

```{r}
redlist_detection %>%
  ggplot(aes(x = Sampler, y = detection_rate)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  theme_minimal() +
  labs(x = "Sampler", y = "% of Swiss macrofungi checklist detected")
```


# Session info

Record R session information for reproducibility.

```{r}
sessionInfo()
```
