---
title: "Sample_indexing"
author: "Fabian Roger"
date: "`r Sys.Date()`"
output: html_notebook
---

Loading libraries

```{r, message=FALSE}

library(dplyr)
library(tidyr)
library(readxl)
library(readr)
library(Biostrings)
library(here)
library(clipr)
library(here)
library(ggplot2)

#helper functions for sequence analyis
source(here("Scripts", "Seq_analysis_helper.R"))
```

```{r}
sessionInfo()
```

# filter Ns and PhiX
we use bbduk to filter reads containing N's and remaining PhiX

[LINK](https://github.com/BioInfoTools/BBMap/blob/master/docs/guides/BBDukGuide.txt)

setting up variables:

The demultiplexed reads are on an external harddrive. 

```{r}
#path to sequences
path_demulti_ext <- "/Volumes/B_physalus/23_Air_sampler_test/Sequencing_data/o34396_Aviti_2023-10-22_AV009"

#path to filtered sequences
path_filtered <- paste(path_demulti_ext, "filtered", sep = "/")
dir.create(path_filtered)

#get files names
dm_files <- list.files(path_demulti_ext, full.names = F)
dm_files <- dm_files[grepl("fastq.gz", dm_files)]

forward_raw <- dm_files[grepl("_R1.fastq.gz", dm_files)]
reverse_raw <- dm_files[grepl("_R2.fastq.gz", dm_files)]

#number of cores to use 
Cores <- 8
```

bbduk

the default parameter for bbduk are: 

minlength=10 (no length filtering)
minavgquality=0  (no Quality filtering based on Phread scores)


```{r}
dir.create(here("Data", "AVITI", "bbduk_stats"))

for (i in seq_along(forward_raw)) {
  
  message(paste("processing file", i, "out of", length(forward_raw), sep=" "))

  system2("/Applications/bbmap/bbduk.sh",
          args = c(
            paste(" in1=", paste(path_demulti_ext , '/', forward_raw[i], sep = ''), sep = ""),
            paste(" in2=", paste(path_demulti_ext , '/', reverse_raw[i], sep = ''), sep = ""),
            paste(" out=", paste(path_filtered , '/', forward_raw[i], sep = ''), sep = ""),
            paste(" out2=", paste(path_filtered , '/', reverse_raw[i], sep = ''), sep = ""),
            "maxns=0",
            "ref=/Applications/bbmap/resources/phix174_ill.ref.fa.gz k=20 hdist=2",
            paste("threads=", Cores, sep = "")
          ),
          stderr = gsub("My Drive", "My\ Drive", (paste(paste(here("Data", "AVITI", "bbduk_stats"),
                     paste("bbduk", gsub(".fastq.gz", "", i), ".txt", sep = ""), sep = "/"))))
          )
}
```


```{r}

bbduk_files <- list.files(here("Data", "AVITI", "bbduk_stats"), full.names = TRUE)

file_order <- as.numeric(gsub(".+?(\\d+).txt", "\\1", bbduk_files))

bbduk_stat = lapply(bbduk_files, function(x){
  temp <- read_lines(x)
  out <- parse_bbduk_output(temp)
} ) %>% bind_rows() %>% 
  mutate(Index = gsub("_R1.fastq.gz", "", forward_raw[file_order])) %>% 
  relocate(Index, .before = NULL) %>% 
  mutate(across(ends_with("reads"), function(x) x/2))  #bbduk counts both pairs of paired reads

bbduk_stat %>% 
  write_tsv(here("Data", "AVITI", "bbduk_stat.txt"))

bbduk_stat <- read_tsv(here("Data", "AVITI", "bbduk_stat.txt"))
```

total reads
```{r}
bbduk_stat %>% 
  pivot_longer(-Index) %>% 
  group_by(name) %>% 
  summarise(mean = mean(value),
            sum = sum(value),
            min = min(value),
            max = max(value),
            sd = sd(value))
```


# demultiplexing by primer

We pooled four different primer pairs

1) COI Primer for Insects 
2) ITS Primer for Fungi 
3) 16S Primer for Bacteria 
4) 16S Primer for Vertebrates 
5) ITS Primer for Plants

I will use cutadapt to sort the reads by primer into separate files

I use [cutadapt](https://cutadapt.readthedocs.io/en/stable/guide.html) to sort the reads by primer into different files.

Citation:

Marcel Martin. Cutadapt removes adapter sequences from high-throughput
sequencing reads. EMBnet.Journal, 17(1):10-12, May 2011.
http://dx.doi.org/10.14806/ej.17.1.200

setting up variables:

```{r}
#minimum length for sequences after trimming (shorter sequences are discarded)
minlength <- 100

#number of cores to use 
Cores <- 8
```


-g specifies that the primer adaptor is at the 5' end of the read.

> 5’ adapters preceed the sequence of interest


# Insects

COI-ArtF12_P5 : GCHCCHGAYATRGCHTTYCCNC
COI-ArtR22xls_P7: GRTCAAARAADGAWGTATTWARATTWCG


```{r}
Primer <- list()
Primer$Frwd <- "GCHCCHGAYATRGCHTTYCCNC" 
Primer$Rev <- "GRTCAAARAADGAWGTATTWARATTWCG" 

Primer$Frwd_rc <- as.character(reverseComplement(DNAStringSet(Primer$Frwd)))
Primer$Rev_rc <- as.character(reverseComplement(DNAStringSet(Primer$Rev)))
```

```{r}
#create directories for primer pair
Primer_path <- here("Data", "AVITI", "Insects")
dir.create(Primer_path)
```


cutadapt version

```{r}
system(paste("source ~/cutadapt-venv/bin/activate", "&&", 'cutadapt --version '))
```


This is cutadapt 4.8

```{r}

cutadapt_out <- vector(mode = "list", length = length(forward_raw))
names(cutadapt_out) <- gsub("(.+?)_R\\d.+", "\\1", forward_raw)

#name of virtual environment in which cutadapt is installed

for (i in seq_along(forward_raw)) {
  
  message(paste("processing file", i, "out of", length(forward_raw), sep=" "))

# will sort in separate file below
temp <- 
system(
paste("source ~/cutadapt-venv/bin/activate ", "&&",
      'cutadapt -a ', #replace by your path to cutadapt
      '"', Primer$Frwd,';required;min_overlap=12','"','...', Primer$Rev_rc,
      ' -A ', '"',Primer$Rev,';required;min_overlap=12','"', '...', Primer$Frwd_rc,
      ' --discard-untrimmed',
      ' --minimum-length=', minlength, #minimum length for sequences after trimming
      ' --cores=', Cores, # Number of cores to use
      ' -o ', shQuote(paste(Primer_path, '/', forward_raw[i], sep = '')),
      ' -p ', shQuote(paste(Primer_path, '/', reverse_raw[i], sep = '')),
      paste(' ',path_filtered , '/', forward_raw[i], sep = ''),
      paste(' ', path_filtered, '/', reverse_raw[i], sep = ''),
      ' --report=minimal',
      sep = ''
      ), intern = TRUE
)

df <- read.table(text = temp)[2,]
colnames(df) <- read.table(text = temp)[1,]
cutadapt_out[[i]] <- df
}

cutadapt_out_df <- bind_rows(cutadapt_out, .id = "Index")

cutadapt_out_df$Marker <- "Insects"

cutadapt_out_df %>% 
  write_tsv(here("Data", "AVITI", "cutadapt_stats_inscects.txt"))

cutadapt_out_df
```
## read lengths

we check the read length of the trimmed reads 

```{r}
readsR1 <-  ShortRead::readFastq(paste(Primer_path, '/', forward_raw[1], sep = '')) %>% 
  ShortRead::sread() %>% 
  width() %>% 
  table() %>% 
  as.data.frame() %>%
  dplyr::rename(Var1 = 1) %>% 
  mutate(dir = "Forward")

readsR2 <-  ShortRead::readFastq(paste(Primer_path, '/', reverse_raw[1], sep = '')) %>% 
  ShortRead::sread() %>% 
  width() %>% 
  table() %>% 
  as.data.frame() %>%
  dplyr::rename(Var1 = 1) %>% 
  mutate(dir = "Reverse")

bind_rows(readsR1, readsR2) %>% 
  group_by(dir) %>% 
  mutate(across(one_of("Var1", "Freq"), function(x) as.numeric(as.character(x)))) %>% 
  arrange(desc(Freq)) %>% 
  mutate(prct = Freq / sum(Freq)*100) %>% 
  mutate(rank = 1:n()) %>% 
  mutate(Label = case_when( rank < 6 ~ paste(Var1, "(",round(prct,1), " %",")", sep = ""),
                   TRUE ~ NA_character_)) %>% 
    ggplot(aes(x = Var1, y = Freq))+
  geom_line(colour = "darkgrey", alpha = 0.5)+
  geom_text_repel(aes(label = Label))+
  geom_point(data = ~filter(.x, !is.na(Label)), colour = "red", size = 0.5)+
  facet_wrap(~dir)+
  theme_bw()+
  scale_x_continuous(limits = c(0,500))
```


# Vertebrates

For vertebrates I used primers targeting the mitochondrial vertebrate 16S gene
The forward Primer has two versions, one targeting birds specifically

mam1 :   GGTTGGGGTGACCTCGGA
mam1_b : GGTTGGGGCGACCTTGGA
                 Y     Y
ve16S : AATCGTTGAACAAACGAAC


```{r}
Primer <- list()
Primer$Frwd <- "GGTTGGGGYGACCTYGGA" 
Primer$Rev <- "AATCGTTGAACAAACGAAC" 

Primer$Frwd_rc <- as.character(reverseComplement(DNAStringSet(Primer$Frwd)))
Primer$Rev_rc <- as.character(reverseComplement(DNAStringSet(Primer$Rev)))
```

```{r}
#create directories for primer pair
Primer_path <- here("Data", "AVITI",  "Vertebrates")
dir.create(Primer_path)
```


```{r}

cutadapt_out <- vector(mode = "list", length = length(forward_raw))
names(cutadapt_out) <- gsub("(.+?)_R\\d.+", "\\1", forward_raw)

for (i in seq_along(forward_raw)) {
  
  message(paste("processing file", i, "out of", length(forward_raw), sep=" "))

# will sort in separate file below
temp <- 
system(
paste("source ~/cutadapt-venv/bin/activate ", "&&",
      'cutadapt -a ', #replace by your path to cutadapt
      '"', Primer$Frwd,';required;min_overlap=12','"','...', Primer$Rev_rc,
      ' -A ', '"',Primer$Rev,';required;min_overlap=12','"', '...', Primer$Frwd_rc,
      ' --discard-untrimmed',
      ' --minimum-length=', minlength, #minimum length for sequences after trimming
      ' --cores=', Cores, # Number of cores to use
      ' -o ', shQuote(paste(Primer_path, '/', forward_raw[i], sep = '')),
      ' -p ', shQuote(paste(Primer_path, '/', reverse_raw[i], sep = '')),
      paste(' ',path_filtered , '/', forward_raw[i], sep = ''),
      paste(' ', path_filtered, '/', reverse_raw[i], sep = ''),
      ' --report=minimal',
      sep = ''
      ), intern = TRUE
)

df <- read.table(text = temp)[2,]
colnames(df) <- read.table(text = temp)[1,]
cutadapt_out[[i]] <- df
}

cutadapt_out_df <- bind_rows(cutadapt_out, .id = "Index")

cutadapt_out_df$Marker <- "Vertebrates"

cutadapt_out_df %>% 
  write_tsv(here("Data", "AVITI", "cutadapt_stats_vertebrates.txt"))

cutadapt_out_df
```

## read lengths

we check the read length of the trimmed reads 

```{r}
readsR1 <-  ShortRead::readFastq(paste(Primer_path, '/', forward_raw[1], sep = '')) %>% 
  ShortRead::sread() %>% 
  width() %>% 
  table() %>% 
  as.data.frame() %>%
  dplyr::rename(Var1 = 1) %>% 
  mutate(dir = "Forward")

readsR2 <-  ShortRead::readFastq(paste(Primer_path, '/', reverse_raw[1], sep = '')) %>% 
  ShortRead::sread() %>% 
  width() %>% 
  table() %>% 
  as.data.frame() %>%
  dplyr::rename(Var1 = 1) %>% 
  mutate(dir = "Reverse")

bind_rows(readsR1, readsR2) %>% 
  group_by(dir) %>% 
  mutate(across(one_of("Var1", "Freq"), function(x) as.numeric(as.character(x)))) %>% 
  arrange(desc(Freq)) %>% 
  mutate(prct = Freq / sum(Freq)*100) %>% 
  mutate(rank = 1:n()) %>% 
  mutate(Label = case_when( rank < 6 ~ paste(Var1, "(",round(prct,1), " %",")", sep = ""),
                   TRUE ~ NA_character_)) %>% 
    ggplot(aes(x = Var1, y = Freq))+
  geom_line(colour = "darkgrey", alpha = 0.5)+
  geom_text_repel(aes(label = Label))+
  geom_point(data = ~filter(.x, !is.na(Label)), colour = "red", size = 0.5)+
  facet_wrap(~dir)+
  theme_bw()+
  scale_x_continuous(limits = c(0,500))
```

# Bacteria

515F : GTGYCAGCMGCCGCGGTAA
806R : GGACTACNVGGGTWTCTAAT


```{r}
Primer <- list()
Primer$Frwd <- "GTGYCAGCMGCCGCGGTAA" 
Primer$Rev <- "GGACTACNVGGGTWTCTAAT" 

Primer$Frwd_rc <- as.character(reverseComplement(DNAStringSet(Primer$Frwd)))
Primer$Rev_rc <- as.character(reverseComplement(DNAStringSet(Primer$Rev)))
```

```{r}
#create directories for primer pair
Primer_path <- here("Data", "AVITI",  "Bacteria")
dir.create(Primer_path)
```


```{r}

cutadapt_out <- vector(mode = "list", length = length(forward_raw))
names(cutadapt_out) <- gsub("(.+?)_R\\d.+", "\\1", forward_raw)

for (i in seq_along(forward_raw)) {
  
  message(paste("processing file", i, "out of", length(forward_raw), sep=" "))

# will sort in separate file below
temp <- 
system(
paste("source ~/cutadapt-venv/bin/activate ", "&&",
      'cutadapt -a ', #replace by your path to cutadapt
      '"', Primer$Frwd,';required;min_overlap=12','"','...', Primer$Rev_rc,
      ' -A ', '"',Primer$Rev,';required;min_overlap=12','"', '...', Primer$Frwd_rc,
      ' --discard-untrimmed',
      ' --minimum-length=', minlength, #minimum length for sequences after trimming
      ' --cores=', Cores, # Number of cores to use
      ' -o ', shQuote(paste(Primer_path, '/', forward_raw[i], sep = '')),
      ' -p ', shQuote(paste(Primer_path, '/', reverse_raw[i], sep = '')),
      paste(' ',path_filtered , '/', forward_raw[i], sep = ''),
      paste(' ', path_filtered, '/', reverse_raw[i], sep = ''),
      ' --report=minimal',
      sep = ''
      ), intern = TRUE
)

df <- read.table(text = temp)[2,]
colnames(df) <- read.table(text = temp)[1,]
cutadapt_out[[i]] <- df
}

cutadapt_out_df <- bind_rows(cutadapt_out, .id = "Index")

cutadapt_out_df$Marker <- "Bacteria"

cutadapt_out_df %>% 
  write_tsv(here("Data", "AVITI", "cutadapt_stats_bacteria.txt"))

cutadapt_out_df
```
## read lengths

we check the read length of the trimmed reads 

```{r}
readsR1 <-  ShortRead::readFastq(paste(Primer_path, '/', forward_raw[1], sep = '')) %>% 
  ShortRead::sread() %>% 
  width() %>% 
  table() %>% 
  as.data.frame() %>%
  dplyr::rename(Var1 = 1) %>% 
  mutate(dir = "Forward")

readsR2 <-  ShortRead::readFastq(paste(Primer_path, '/', reverse_raw[1], sep = '')) %>% 
  ShortRead::sread() %>% 
  width() %>% 
  table() %>% 
  as.data.frame() %>%
  dplyr::rename(Var1 = 1) %>% 
  mutate(dir = "Reverse")

bind_rows(readsR1, readsR2) %>% 
  group_by(dir) %>% 
  mutate(across(one_of("Var1", "Freq"), function(x) as.numeric(as.character(x)))) %>% 
  arrange(desc(Freq)) %>% 
  mutate(prct = Freq / sum(Freq)*100) %>% 
  mutate(rank = 1:n()) %>% 
  mutate(Label = case_when( rank < 6 ~ paste(Var1, "(",round(prct,1), " %",")", sep = ""),
                   TRUE ~ NA_character_)) %>% 
    ggplot(aes(x = Var1, y = Freq))+
  geom_line(colour = "darkgrey", alpha = 0.5)+
  geom_text_repel(aes(label = Label))+
  geom_point(data = ~filter(.x, !is.na(Label)), colour = "red", size = 0.5)+
  facet_wrap(~dir)+
  theme_bw()+
  scale_x_continuous(limits = c(0,500))
```
# Fungi

```{r}
ITS_primer <- read_xlsx(here("Data", "Fungal_primer_Schlegel.xlsx"))
```

we sequence wit a mix of primers, we specify the adapters in separate files:

```{r}
ITS_fwrd <- 
ITS_primer %>% 
  filter(Primer_dir == "Fwrd") %>% 
  pull(`primer sequence`) %>% 
  unique() %>% 
  DNAStringSet()

fwrd_search <- 
paste(ITS_fwrd, ";required;min_overlap=12","...",reverseComplement(ITS_fwrd), sep = "")

paste(">primer", 1:length(fwrd_search), sep = "_") %>% 
  rbind(., fwrd_search) %>% 
  c() %>% 
  write_lines(here("Data", "ITS_Primer_fwrd.fasta"))


ITS_rev <- 
ITS_primer %>% 
  filter(Primer_dir == "Rev") %>% 
  pull(`primer sequence`) %>% 
  unique() %>% 
  DNAStringSet()

rev_search <- 
paste(ITS_rev, ";required;min_overlap=12","...",reverseComplement(ITS_rev), sep = "")

paste(">primer", 1:length(rev_search), sep = "_") %>% 
  rbind(., rev_search) %>% 
  c() %>% 
  write_lines(here("Data", "ITS_Primer_rev.fasta"))
```

```{r}
#create directories for primer pair
Primer_path <- here("Data", "AVITI",  "Fungi")
dir.create(Primer_path)
```


```{r}

cutadapt_out <- vector(mode = "list", length = length(forward_raw))
names(cutadapt_out) <- gsub("(.+?)_R\\d.+", "\\1", forward_raw)

for (i in seq_along(forward_raw)) {
  
  message(paste("processing file", i, "out of", length(forward_raw), sep=" "))

# will sort in separate file below
temp <- 
system(
paste("source ~/cutadapt-venv/bin/activate ", "&&",
      'cutadapt -a ', #replace by your path to cutadapt
      '"', 
      'file:',here("Data", "ITS_Primer_fwrd.fasta"),
      '"',
      ' -A ', 
      '"',
      'file:', here("Data", "ITS_Primer_rev.fasta"),
      '"', 
      ' --discard-untrimmed',
      ' --minimum-length=', minlength, #minimum length for sequences after trimming
      ' --cores=', Cores, # Number of cores to use
      ' -o ', shQuote(paste(Primer_path, '/', forward_raw[i], sep = '')),
      ' -p ', shQuote(paste(Primer_path, '/', reverse_raw[i], sep = '')),
      paste(' ',path_filtered , '/', forward_raw[i], sep = ''),
      paste(' ', path_filtered, '/', reverse_raw[i], sep = ''),
      ' --report=minimal',
      sep = ''
      ), intern = TRUE
)

df <- read.table(text = temp)[2,]
colnames(df) <- read.table(text = temp)[1,]
cutadapt_out[[i]] <- df
}

cutadapt_out_df <- bind_rows(cutadapt_out, .id = "Index")

cutadapt_out_df$Marker <- "Fungi"

cutadapt_out_df %>% 
  write_tsv(here("Data", "AVITI", "cutadapt_stats_fungi.txt"))

cutadapt_out_df
```
## read lengths

we check the read length of the trimmed reads 

```{r}
readsR1 <-  ShortRead::readFastq(paste(Primer_path, '/', forward_raw[2], sep = '')) %>% 
  ShortRead::sread() %>% 
  width() %>% 
  table() %>% 
  as.data.frame() %>%
  dplyr::rename(Var1 = 1) %>% 
  mutate(dir = "Forward")

readsR2 <-  ShortRead::readFastq(paste(Primer_path, '/', reverse_raw[2], sep = '')) %>% 
  ShortRead::sread() %>% 
  width() %>% 
  table() %>% 
  as.data.frame() %>%
  dplyr::rename(Var1 = 1) %>% 
  mutate(dir = "Reverse")

bind_rows(readsR1, readsR2) %>% 
  group_by(dir) %>% 
  mutate(across(one_of("Var1", "Freq"), function(x) as.numeric(as.character(x)))) %>% 
  arrange(desc(Freq)) %>% 
  mutate(prct = Freq / sum(Freq)*100) %>% 
  mutate(rank = 1:n()) %>% 
  mutate(Label = case_when( rank < 6 ~ paste(Var1, "(",round(prct,1), " %",")", sep = ""),
                   TRUE ~ NA_character_)) %>% 
    ggplot(aes(x = Var1, y = Freq))+
  geom_line(colour = "darkgrey", alpha = 0.5)+
  geom_text_repel(aes(label = Label))+
  geom_point(data = ~filter(.x, !is.na(Label)), colour = "red", size = 0.5)+
  facet_wrap(~dir)+
  theme_bw()+
  scale_x_continuous(limits = c(0,500))
```

# Plants

ITS2F : ATGCGATACTTGGTGTGAAT
ITS2R : GACGCTTCTCCAGACTACAAT


```{r}
Primer <- list()
Primer$Frwd <- "ATGCGATACTTGGTGTGAAT" 
Primer$Rev <- "GACGCTTCTCCAGACTACAAT" 

Primer$Frwd_rc <- as.character(reverseComplement(DNAStringSet(Primer$Frwd)))
Primer$Rev_rc <- as.character(reverseComplement(DNAStringSet(Primer$Rev)))
```

```{r}
#create directories for primer pair
Primer_path <- here("Data", "AVITI",  "Plants")
dir.create(Primer_path)
```


```{r}

cutadapt_out <- vector(mode = "list", length = length(forward_raw))
names(cutadapt_out) <- gsub("(.+?)_R\\d.+", "\\1", forward_raw)

for (i in seq_along(forward_raw)) {
  
  message(paste("processing file", i, "out of", length(forward_raw), sep=" "))

# will sort in separate file below
temp <- 
system(
paste("source ~/cutadapt-venv/bin/activate ", "&&",
      'cutadapt -a ', #replace by your path to cutadapt
      '"', Primer$Frwd,';required;min_overlap=12','"','...', Primer$Rev_rc,
      ' -A ', '"',Primer$Rev,';required;min_overlap=12','"', '...', Primer$Frwd_rc,
      ' --discard-untrimmed',
      ' --minimum-length=', minlength, #minimum length for sequences after trimming
      ' --cores=', Cores, # Number of cores to use
      ' -o ', shQuote(paste(Primer_path, '/', forward_raw[i], sep = '')),
      ' -p ', shQuote(paste(Primer_path, '/', reverse_raw[i], sep = '')),
      paste(' ',path_filtered , '/', forward_raw[i], sep = ''),
      paste(' ', path_filtered, '/', reverse_raw[i], sep = ''),
      ' --report=minimal',
      sep = ''
      ), intern = TRUE
)

df <- read.table(text = temp)[2,]
colnames(df) <- read.table(text = temp)[1,]
cutadapt_out[[i]] <- df
}

cutadapt_out_df <- bind_rows(cutadapt_out, .id = "Index")

cutadapt_out_df$Marker <- "Plants"

cutadapt_out_df %>% 
  write_tsv(here("Data", "AVITI", "cutadapt_stats_plants.txt"))

cutadapt_out_df
```
## read lengths

we check the read length of the trimmed reads 

```{r}
readsR1 <-  ShortRead::readFastq(paste(Primer_path, '/', forward_raw[1], sep = '')) %>% 
  ShortRead::sread() %>% 
  width() %>% 
  table() %>% 
  as.data.frame() %>%
  dplyr::rename(Var1 = 1) %>% 
  mutate(dir = "Forward")

readsR2 <-  ShortRead::readFastq(paste(Primer_path, '/', reverse_raw[1], sep = '')) %>% 
  ShortRead::sread() %>% 
  width() %>% 
  table() %>% 
  as.data.frame() %>%
  dplyr::rename(Var1 = 1) %>% 
  mutate(dir = "Reverse")

bind_rows(readsR1, readsR2) %>% 
  group_by(dir) %>% 
  mutate(across(one_of("Var1", "Freq"), function(x) as.numeric(as.character(x)))) %>% 
  arrange(desc(Freq)) %>% 
  mutate(prct = Freq / sum(Freq)*100) %>% 
  mutate(rank = 1:n()) %>% 
  mutate(Label = case_when( rank < 6 ~ paste(Var1, "(",round(prct,1), " %",")", sep = ""),
                   TRUE ~ NA_character_)) %>% 
    ggplot(aes(x = Var1, y = Freq))+
  geom_line(colour = "darkgrey", alpha = 0.5)+
  geom_text_repel(aes(label = Label))+
  geom_point(data = ~filter(.x, !is.na(Label)), colour = "red", size = 0.5)+
  facet_wrap(~dir)+
  theme_bw()+
  scale_x_continuous(limits = c(0,500))
```

# distribution of reads

```{r}
gURL <- "https://docs.google.com/spreadsheets/d/1GKWzD0q683oH3I_i5ueulbf4P4Q-9-a9YPbdyUZyF2A/edit#gid=847966271"

#load metadata
Meta <- read_sheet(gURL, sheet = "Sample_data_corr") %>% 
  select(Sample_names, Sites, Sampler, duration, Type)

# get plate layouts
Plates_Ins_Vert_Plant <- read_sheet(gURL, sheet = "Plates", range = "A1:M21", col_types = "c") %>% 
  dplyr::rename(row = 1) %>% 
  filter(row %in% LETTERS[1:8]) %>% 
  mutate(Plate = rep(c("Plate_1", "Plate_2"), each = 8)) %>% 
  pivot_longer(!one_of(c("row", "Plate")), names_to = "col", values_to = "Ins_Vert_Plant_sample") %>% 
  select(Plate, row, col, Ins_Vert_Plant_sample) %>% 
  filter(!is.na(Ins_Vert_Plant_sample)) 

Plates_Bac <- read_sheet(gURL, sheet = "Plates", range = "A48:M66", col_types = "c") %>% 
  dplyr::rename(row = 1) %>% 
  filter(row %in% LETTERS[1:8]) %>% 
  mutate(Plate = rep(c("Plate_1", "Plate_2"), each = 8)) %>% 
  pivot_longer(!one_of(c("row", "Plate")), names_to = "col", values_to = "Bac_sample") %>% 
  select(Plate, row, col, Bac_sample) %>% 
  filter(!is.na(Bac_sample)) 

Plates_Fungi <- read_sheet(gURL, sheet = "Plates", range = "A26:M44", col_types = "c") %>% 
  dplyr::rename(row = 1) %>% 
  filter(row %in% LETTERS[1:8]) %>% 
  mutate(Plate = rep(c("Plate_1", "Plate_2"), each = 8)) %>% 
  pivot_longer(!one_of(c("row", "Plate")), names_to = "col", values_to = "Fungi_sample") %>% 
  select(Plate, row, col, Fungi_sample) %>% 
  filter(!is.na(Fungi_sample)) 

# get plate layout of barcodes 
shini_barcodes <- 
  read_xlsx(here("Documents", "Shini_barcodes_4_plates.xlsx")) %>% 
  mutate(Plate = rep(rep(paste("Plate", 1:4, sep = "_"), each = 96),2)) %>% 
  filter(Plate %in% c("Plate_1", "Plate_2")) %>% 
  dplyr::rename(Index = 1) %>% 
  select(Index, Plate) %>% 
  filter(grepl("_i5$", Index)) %>% 
  mutate(Index = gsub("(.+?)_i5", "\\1", Index)) %>% 
  group_by(Plate) %>% 
  mutate(row = rep(LETTERS[1:8], each = 12)) %>% 
  mutate(col = as.character(rep(1:12, 8)))

Index_meta <- 
  shini_barcodes %>% 
  left_join(Plates_Ins_Vert_Plant) %>% 
  left_join(Plates_Bac) %>% 
  left_join(Plates_Fungi) %>% 
  ungroup() %>% 
  select(-Plate, -row, -col)
```


```{r}
ca_insect <- read_tsv(here("Data", "AVITI", "cutadapt_stats_inscects.txt"))

ca_vert <- read_tsv(here("Data", "AVITI", "cutadapt_stats_vertebrates.txt"))

ca_bac <- read_tsv(here("Data", "AVITI", "cutadapt_stats_bacteria.txt")) 

ca_fungi <- read_tsv(here("Data", "AVITI", "cutadapt_stats_fungi.txt"))

ca_plants <- read_tsv(here("Data", "AVITI", "cutadapt_stats_plants.txt"))

bbduk_stat <- read_tsv(here("Data", "AVITI", "bbduk_stat.txt"))

read_stat <- 
list(ca_insect, ca_vert, ca_bac, ca_fungi, ca_plants) %>% 
  bind_rows() %>% 
  select(Index, out_reads, Marker) %>% 
  pivot_wider(names_from = Marker, values_from = out_reads) %>% 
 # left_join(select(ca_bac, Index, in_reads, too_short)) %>% 
  left_join(select(bbduk_stat, Index, Input_reads, Total_removed_reads, Result_reads),.) %>% 
  mutate(accounted = (Insects + Vertebrates + Bacteria + Fungi + Plants + Total_removed_reads)) %>% 
  mutate(missing_prct = 100-round(accounted/Input_reads*100,1)) %>% 
  mutate(Index = gsub(".+?-(.+)", "\\1", Index)) %>% 
  left_join(Index_meta)


write_tsv(read_stat, here("Data", "AVITI", "readStat.txt"))

read_stat <- read_tsv(here("Data", "AVITI", "readStat.txt"))
read_stat %>% 
  filter(Input_reads > 10000) %>% 
  arrange(desc(missing_prct))

```

##index leaking

We did not use all indices from Plate 1 and 2, but we demultiplexed using all indices. We can look at the unused indices to assess if there was index leaking, i.e. if reads got assigned to those samples even though they were not present in the data.

Across 19 unused barcodes, Zero reads where assigned to the unused barcodes and passed cutadapt

```{r}
read_stat %>% 
  filter(is.na(Ins_Vert_Plant_sample) & is.na(Bac_sample) & is.na(Fungi_sample))
```

```{r}
read_stat <- read_tsv(here("Data", "AVITI", "readStat.txt"))
```

# Plate visualisation

##Bacteria
```{r}
Plates_Bac %>% 
  left_join(shini_barcodes) %>% 
  left_join(select(read_stat, Index, Bacteria)) %>% 
  mutate(col = factor(col, levels = as.character(1:12))) %>% 
  mutate(row = factor(row, levels = LETTERS[8:1])) %>% 
  ggplot(aes(x = col, y = row, fill = Bacteria))+
  geom_tile()+
  geom_text(aes(label = Bac_sample), size = 2, colour = "white")+
  facet_grid(Plate~.)+
  scale_y_discrete(limits = rev(levels(row)))+
  scale_fill_viridis_c()
```

##Fungi
```{r}
Plates_Fungi %>% 
  left_join(shini_barcodes) %>% 
  left_join(select(read_stat, Index, Fungi)) %>% 
  mutate(col = factor(col, levels = as.character(1:12))) %>% 
  mutate(row = factor(row, levels = LETTERS[8:1])) %>% 
  ggplot(aes(x = col, y = row, fill = Fungi))+
  geom_tile()+
  geom_text(aes(label = Fungi_sample), size = 2, colour = "white")+
  facet_grid(Plate~.)+
  scale_y_discrete(limits = rev(levels(row)))+
  scale_fill_viridis_c()
```
##Insects
```{r}
Plates_Ins_Vert_Plant %>% 
  left_join(shini_barcodes) %>% 
  left_join(select(read_stat, Index, Insects)) %>% 
  mutate(col = factor(col, levels = as.character(1:12))) %>% 
  mutate(row = factor(row, levels = LETTERS[8:1])) %>% 
  ggplot(aes(x = col, y = row, fill = Insects))+
  geom_tile()+
  geom_text(aes(label = Ins_Vert_Plant_sample), size = 2, colour = "white")+
  facet_grid(Plate~.)+
  scale_y_discrete(limits = rev(levels(row)))+
  scale_fill_viridis_c()
```

##Vertebrates
```{r}
Plates_Ins_Vert_Plant %>% 
  left_join(shini_barcodes) %>% 
  left_join(select(read_stat, Index, Vertebrates)) %>% 
  mutate(col = factor(col, levels = as.character(1:12))) %>% 
  mutate(row = factor(row, levels = LETTERS[8:1])) %>% 
  ggplot(aes(x = col, y = row, fill = Vertebrates))+
  geom_tile()+
  geom_text(aes(label = Ins_Vert_Plant_sample), size = 2, colour = "white")+
  facet_grid(Plate~.)+
  scale_y_discrete(limits = rev(levels(row)))+
  scale_fill_viridis_c()
```

##Plants
```{r}
Plates_Ins_Vert_Plant %>% 
  left_join(shini_barcodes) %>% 
  left_join(select(read_stat, Index, Plants)) %>% 
  filter(Ins_Vert_Plant_sample != "5LK30") %>% #this sample has 3 times as many reads as the others (for plants)
  mutate(col = factor(col, levels = as.character(1:12))) %>% 
  mutate(row = factor(row, levels = LETTERS[8:1])) %>% 
  ggplot(aes(x = col, y = row, fill = Plants))+
  geom_tile()+
  geom_text(aes(label = Ins_Vert_Plant_sample), size = 2, colour = "white")+
  facet_grid(Plate~.)+
  scale_y_discrete(limits = rev(levels(row)))+
  scale_fill_viridis_c()
```
