---
title: "Meta_data"
output: html_notebook
---

```{r}
library(tidyverse)
library(lubridate)
library(googlesheets4)
```

```{r}

googlesheet_url <- "https://docs.google.com/spreadsheets/d/1GKWzD0q683oH3I_i5ueulbf4P4Q-9-a9YPbdyUZyF2A/edit#gid=1771030448"

meta <- read_sheet(googlesheet_url, sheet = "Sample_names")
```


Table of Sampler codes

```{r}

Sampler <- tibble(
  Sampler = c("Hepa","Burkhart","Kärcher","Coriolis",
              "Electrostatic","Sass","Drone",
              "WSL_Filter","MWAC"),
  Sampler_Code = c("H","B","K","C","E","S","D","F","M")
)

```
Table of Site codes

```{r}
Sites <- tibble(
  Sites = c("Duebendorf","Talent","Coruz","Haslibach",
            "Leugene","Gaebelbach","Weierbach","Riedgraben")
) %>% 
  mutate(Site_Code = substr(Sites, 1, 1))
```

Table of Date codes

```{r}
Dates <- tibble(
  Date_Code = as.character(c(1:9)),
  Dates = c("05.07.2022","06.07.2022","07.07.2022","08.07.2022",
            "09.07.2022","10.07.2022","11.07.2022","12.07.2022",
            "31.07.2022")) %>% 
  mutate(Dates = dmy(Dates))
```

correct sample names

9H5F2 --> 9HF2
9C3M2 --> 9CM2
9H5M2 --> 9HM2


```{r}

meta <- 
  meta %>% 
  mutate(Sample_names = case_when(Sample_names == "9H5F2" ~ "9HF2",
                                  Sample_names == "9C3M2" ~ "9CM2",
                                  Sample_names == "9H5M2" ~ "9HM2",
                                  TRUE~Sample_names,
                                  ))
  
```


```{r}
meta <- 
meta %>% 
  mutate(Date_Code = substr(Sample_names, 1,1)) %>% 
  left_join(Dates) %>% 
  mutate(Site_Code = substr(Sample_names, 2,2)) %>% 
  left_join(Sites) %>% 
  mutate(Sampler_Code = substr(Sample_names, 3,3)) %>% 
  left_join(Sampler) %>% 
  mutate(duration = case_when(grepl("30$", Sample_names) ~ "30 min",
                              grepl("5$", Sample_names) ~ "5 hours",
                              grepl("M|F", Sample_names) ~ "2 weeks",
                              grepl("control", Sample_names) ~ NA_character_,
                              TRUE ~ NA_character_)) %>% 
  mutate(duration = case_when(Sampler == "Drone" & 
                                grepl("CD[1-6]$", Sample_names) ~ "10 min",
                              TRUE ~ duration)) %>% 
  mutate(Type = case_when(Sampler_Code %in% c("F", "M") ~ "passive",
                          grepl("control", Sample_names) ~ "field_control",
                          grepl("B[1-2]", Sample_names) ~ "field_control",
                          TRUE ~ "active")) %>% 
  arrange(Dates, Sampler, duration)
```

update tube and buffer volume

[1] "Burkhart"      "Coriolis"    "Hepa"         
 [6] "Kärcher"       "Sass"          "MWAC"          "WSL_Filter"    NA             

```{r}
meta <- 
meta %>% 
  mutate(Material = 
           case_when(Material = Sampler %in% c("Electrostatic") ~ "Swab",
            Sampler %in% c("Drone") ~ "Swab_and_grease",
            Sampler %in% c("Burkhart") ~ "none",
            Sampler %in% c("Coriolis") ~ "PC_filter_0.2_25mm",
            Sampler %in% c("Hepa") ~ "Hepa_filter_90mm",
            Sampler %in% c("Kärcher") ~ "Sterivex_0.2",
            Sampler %in% c("Sass") ~ "SASS_filer",
            Sampler %in% c("MWAC") ~ "PC_filter_0.2_25mm",
            Sampler %in% c("WSL_Filter") ~ "Glass_Fiber_90mm"
            )) %>%
  mutate(Tube = 
           case_when(Material = Sampler %in% c("Electrostatic") ~ "BeadRupt_2ml",
            Sampler %in% c("Drone") ~ "BeadRupt_2ml",
            Sampler %in% c("Burkhart") ~ "Eppi_1.5",
            Sampler %in% c("Coriolis") ~ "BeadRupt_2ml",
            Sampler %in% c("Hepa") ~ "BeadRupt_7ml",
            Sampler %in% c("Kärcher") ~ "Sterivex_in_Falcon",
            Sampler %in% c("Sass") ~ "BeadRupt_7ml",
            Sampler %in% c("MWAC") ~ "BeadRupt_2ml",
            Sampler %in% c("WSL_Filter") ~ "BeadRupt_7ml"
            )) %>% 
  mutate(Buffer_vol = 
           case_when(Material = Sampler %in% c("Electrostatic") ~ "2",
            Sampler %in% c("Drone") ~ "2",
            Sampler %in% c("Burkhart") ~ "1.5",
            Sampler %in% c("Coriolis") ~ "2",
            Sampler %in% c("Hepa") ~ "3",
            Sampler %in% c("Kärcher") ~ "2",
            Sampler %in% c("Sass") ~ "3",
            Sampler %in% c("MWAC") ~ "2",
            Sampler %in% c("WSL_Filter") ~ "3"
            ))
```

```{r}
# meta %>%
#   select(Sample_names, Sites, Sampler, duration, Type, Tube, Material, Buffer_vol, Dates, Site_Code, Sampler_Code, Date_Code) %>%
# write_sheet(., googlesheet_url, sheet = "Sample_data")
```


#Plate layout

As we already edited the metadata in the Google sheet (e.g. added comments during extraction) we download it. 

```{r}
Meta <- read_sheet(googlesheet_url, sheet = "Sample_data")

Plates <- read_sheet(googlesheet_url, sheet = "Plates", col_types = "c")

Plate_meta <- 
Plates %>% 
  filter(!is.na(Plate_1) & Plate_1 != "Plate_2") %>% 
  mutate(Plate = rep(c("Plate_1", "Plate_2"), each = 8)) %>% 
  rename(row = "Plate_1") %>% 
  pivot_longer(!one_of(c("Plate", "row")), names_to = "col", values_to = "Sample_names") %>% 
  mutate(well = paste(row, col, sep = "_")) %>% 
  select(Sample_names, Plate, well) %>% 
  filter(!is.na(Sample_names)) #%>% 
 # mutate(Sample_names = case_when(Plate == "Plate_1" & well == "H_10" ~ "??",
#                                  TRUE ~ Sample_names))
```


I checked the tubes in the cleanlab to see if can know which sample is which. Both are clearly marked 
1DB30, but one is in a clear tube with a lot of liquid and one is an opaque tube with barely any liquid. 
There is also a comment about this in the metadata. 

I name tube with less liquid 1DB30_1.

```{r}

Plate_meta %>% 
  filter(grepl("1DB30", Sample_names))
  
```


The extraction blanks are not part of the metadate yet
```{r}
Plate_meta %>% 
  filter(!Sample_names %in% Meta$Sample_names)
```

```{r}
Meta %>% filter(!Sample_names %in% Plate_meta$Sample_names)
```
Sample 9F2control2 is excluded (see comment), it's normal it's not part of the meta data

4R isn't possible as R is Riedgraben, which we visited on Date 8.

8RS30 and 8RH30 already exist in the metadata and on the PLates, I must have added them when preparing the plates. 

I remove  4RS30 and 4RH30 from the metadata

```{r}
Meta <- 
  Meta %>% 
  filter(!Sample_names %in% c("4RS30", "4RH30"))
```


```{r}

Meta <- 
Meta %>% 
  full_join(Plate_meta) %>% 
  mutate(Sampler = case_when(grepl("BLANK", Sample_names) ~ "Extraction Blanks",
                          TRUE ~ Sampler)) %>% 
  mutate(Type = case_when(grepl("BLANK", Sample_names) ~ "field_control",
                          TRUE ~ Type))

```


```{r}
Meta %>%
 write_sheet(., googlesheet_url, sheet = "Sample_data_corr")
```

