---
title: "Nextflex"
output: html_notebook
---

```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(Biostrings)
library(here)
```

```{r}
Nextflex <- read_xlsx(here("Data", "NEXTFLEX-UDI-Barcodes-1536-indexes_10NT.xlsx"))
```

#i7

1536 unique i7 indicies
```{r}
Nextflex$`I7 Index` %>%
  unique() %>% 
  length()
```

minimum hamming distance is 3 bps

--> maybe worth filtering out those that have 
```{r}
Nextflex$`I7 Index` %>% 
  stringDist(method = "hamming") %>% 
  table()
```

#i5

1536 unique i5 indicies
```{r}
Nextflex$`I5 Index` %>%
  unique() %>% 
  length()
```

minimum hamming distance is 3 bps

--> maybe worth filtering out those that have 
```{r}
Nextflex$`I5 Index` %>% 
  stringDist(method = "hamming") %>% 
  table()
```


the I5 and I7 indicies are completely unique
```{r}
sum(Nextflex$`I5 Index` %in% Nextflex$`I7 Index`)
```

two i7 indicies happen to be the reverse compl of two i5 indicies:

```{r}
sum(Nextflex$`I7 Index` %in% Nextflex$`I5 Index reverse complement`)
```


the combined indices have at least 6 bps difference (maybe worth filtering out pairs that have <10 bp difference)
```{r}
ind_comb <- paste(Nextflex$`I7 Index`, Nextflex$`I5 Index`, sep = "")

ind_comb %>% 
  stringDist(method = "hamming") %>% 
  table()

```



attempt at optimising the indices but not sure it works great. 

Seems to work to find a subset of index combinations that has a minimum hamming distance of 8 (1360), 9 (975) or 10 (513).

Might not be the most efficient algorithm. 
```{r}
min_ham <- 9

ind_comb_sel <- ind_comb

#  initialize temp with high value
maxtemp <- 20
i = 1

while(maxtemp > 1 & i <= 100){
  
  #hamming sitance between all index pairs
  distM <- 
    ind_comb_sel %>% 
    stringDist(method = "hamming") %>% 
    as.matrix()
  
  #which indices have a hamming distance of less than min_ham bp
  distM10_list <- 
    apply(distM, 1, function(x){
    unique(which(x < min_ham))
  })
  
  #how many indicies with such a hmmaing distance are there (for each index)
  temp <- sort(unlist(lapply(distM10_list, length)))
  
  #maximum amount of 
  maxtemp = max(temp)
  
  #exclude 10 index pairs with highest amount of other index pairs with hamming distance < 10
  excl <- temp[(length(temp)-10):length(temp)]
  ind_comb_sel <- ind_comb_sel[-c(as.numeric(names(excl)))]
  
  #print status message
  message(paste("itteration", i, "maxtemp =", maxtemp, sep = " "))
  
  i = i+1
}
```


```{r}
ind_comb_sel %>% 
  stringDist(method = "hamming") %>% 
  table()
```


```{r}
length(ind_comb_sel)
```

```{r}

Nextflex_9 <- 
Nextflex %>% 
  mutate(ind_comb = ind_comb) %>% 
  filter(ind_comb %in% ind_comb_sel)
  
```

