---
title: "R Notebook"
output: tax_glom_level
---
```{r}
library(dplyr)
library(Biostrings)
library(here)
library(readr)
library(tidyr)
library(purrr)
library(data.table)
library(rgbif)
library(job)
library(eulerr)
```
#Functions

define function to check against gbif backbone in chunks
```{r}
# Function to split the dataframe into chunks
split_dataframe <- function(df, chunk_size) {
  split(df, ceiling(seq_along(1:nrow(df)) / chunk_size))
}

# Combined function to process each chunk with retries
process_chunk_with_retry <- function(chunk, max_retries = 5, wait_seconds = 10) {
  attempt <- 1
  success <- FALSE
  result <- NULL
  
  while (attempt <= max_retries && !success) {
    tryCatch({
      # Attempt to process the chunk
      result <- name_backbone_checklist(name_data = chunk)
      success <- TRUE  # If successful, set success to TRUE
    }, error = function(e) {
      message("Error processing chunk on attempt ", attempt, ": ", e$message)
      attempt <- attempt + 1  # Increment the attempt counter
      if (attempt > max_retries) {
        stop("Failed to process chunk after ", max_retries, " attempts.")
      }
      Sys.sleep(wait_seconds)  # Wait before retrying
    })
  }
  
  return(result)
}

# Main function to process dataframe in chunks and re-merge
process_dataframe_in_chunks <- function(df, chunk_size, save_path) {
  chunks <- split_dataframe(df, chunk_size)
  total_chunks <- length(chunks)
  chunk_numbers <- seq_along(chunks)
  
  results <- map2(chunks, chunk_numbers, function(chunk, chunk_number) {
    save_file <- file.path(save_path, paste0("chunk_", chunk_number, ".rds"))
    
    # Check if the chunk has already been processed
    if (file.exists(save_file)) {
      message("Chunk ", chunk_number, " already processed. Loading from file.")
      return(readRDS(save_file))
    } else {
      message("Processing chunk ", chunk_number, " of ", total_chunks)
      
      # Process the chunk with retries
      result <- process_chunk_with_retry(chunk)
      
      # Adjust verbatim_index if needed
      result$verbatim_index <- (chunk_number - 1) * chunk_size + result$verbatim_index
      
      # Save the result
      saveRDS(result, save_file)
      return(result)
    }
  })
  
  combined_results <- bind_rows(results)
  return(combined_results)
}
```


#rdp

```{r}
rdp <- readDNAStringSet(here("Data", "trainset19_072023.rdp", "trainset19_072023.rdp.fasta"))

tax_rdp <- read_table(here("Data", "trainset19_072023.rdp", "trainset19_072023.rdp.tax"), col_names = F)
colnames(tax_rdp) <- c("ID", "tax")

tax_rdp <- 
tax_rdp %>% 
  tidyr::separate_wider_delim(tax, delim = ";", 
                              names = c("d", "p", "c", "o", "f", "g"), 
                              too_many = "drop", 
                              too_few = "align_start")

#structure
"KU904771.1;tax=p:Chlorophyta,c:Ulvophyceae,o:Cladophorales,f:Cladophoraceae,g:Basicladia,s:unknown Basicladia"

tax_rdp_sintax <- 
tax_rdp %>%
  pivot_longer(-ID) %>% 
  mutate(rank = paste(name, value, sep = ":")) %>% 
  group_by(ID) %>% 
  summarise(rank = paste0(rank, collapse = ",")) %>% 
  mutate(Sintax = paste(ID, ";tax=", rank, sep = ""))

seq_order <- gsub("(.+?)\\s.+", "\\1", names(rdp))

names(rdp) <- tax_rdp_sintax[match(seq_order, tax_rdp_sintax$ID),]$Sintax

writeXStringSet(rdp, here("Data", "trainset19_072023.rdp", "trainset19_072023.rdp_sintax.fasta"))
```


#Unite
```{r}
refDB <- readDNAStringSet(here("Data", primer, "in_files", "sh_general_release_dynamic_s_all_29.11.2022.fasta"))

names_refDB <- names(refDB)

names_refDB_sintax <- 
strsplit(names_refDB, "\\|") %>% 
  lapply(function(x) paste(x[2], x[5], sep = ";tax=")) %>% 
  unlist() %>% 
  gsub("__", ":", .)

names_refDB_sintax <- 
names_refDB_sintax %>% 
  gsub(";", ",", .) %>% 
  gsub(",tax", ";tax",. )


names(refDB) <- names_refDB_sintax

refDB %>% 
writeXStringSet(here("Data", primer, "in_files", "sh_general_release_dynamic_s_all_29.11.2022.sintax.fasta"))
```


##get taxonomy
```{r}
Unite <- readDNAStringSet(here("Data", "Fungi", "in_files", "sh_general_release_dynamic_s_all_29.11.2022.sintax.fasta"))

Unite_tax <- 
  tibble(seqname = names(Unite)) %>% 
  mutate(
    tax = str_extract(seqname, "tax=.*"),
    tax = str_remove(tax, "^tax="),
    tax = str_remove(tax, "[;,]?$")
  ) %>% 
  mutate(tax = strsplit(tax, ",")) %>% 
  unnest(tax) %>% 
  group_by(seqname) %>% 
  mutate(taxonomy_name = gsub("^\\w:(.+?);*$", "\\1", tax),
         taxonomy_level = gsub("(^\\w):.+", "\\1", tax)) %>% 
  left_join(tax_match_df)

Unite_tax_w <- 
Unite_tax %>% 
  select(seqname, taxonomy_name, taxon) %>% 
  pivot_wider(names_from = taxon, values_from = taxonomy_name)


write_tsv(Unite_tax_w, here("Data", "Fungi", "in_files", "Unite_taxonomy.txt"))
Unite_tax_w <- read_tsv(here("Data", "Fungi", "in_files", "Unite_taxonomy.txt"))
```

## rgbif
```{r}
Unite_gbif_in <- 
  Unite_tax_w %>% 
  ungroup %>% 
  select(-seqname) %>% 
  distinct() %>% 
  mutate(name = species) %>% 
  mutate(name = gsub("_", " ", name))

# Define the chunk size
chunk_size <- 2000

# Define the path to save intermediate results
save_path <- here("Data", "Fungi", "in_files")

Unite_gbif_out <- process_dataframe_in_chunks(Unite_gbif_in, chunk_size, save_path)

#check if everything has been processed
Unite_gbif_in %>% 
  ungroup() %>% 
  mutate(index = 1:n()) %>% 
  filter(!index %in% Unite_gbif_out$verbatim_index) %>% 
  nrow()

# Delete chunk files
chunk_files <- list.files(save_path, pattern = "\\.rds$", full.names = TRUE)
file.remove(chunk_files)

write_tsv(Unite_gbif_out, here("Data", "Fungi", "in_files", "Unite_gbif_out.txt"))

Unite_gbif_out <- read_tsv(here("Data", "Fungi", "in_files", "Unite_gbif_out.txt"))
```

###Merge gbif taxonomy to Unite taxonomy
```{r}

Unite_tax_newname <- 
Unite_gbif_out %>% 
  select(starts_with("verbatim"), kingdom, phylum, class, order, family, genus, species) %>% 
  dplyr::rename_with( ~paste("gbif",., sep = "_"), .cols = !starts_with("verbatim")) %>% 
  dplyr::rename_with( ~gsub("verbatim_", "",.), .cols = starts_with("verbatim")) %>% 
  dplyr::rename(species = name) %>% 
  mutate(species = gsub(" ", "_", species)) %>% 
  select(-index) %>% 
  distinct() %>% 
  left_join(Unite_tax_w,.) 

```

###make Sintax compatible sequence names
```{r}

Unite_tax_newseqname <- 
Unite_tax_newname %>%
  ungroup() %>% 
  select(seqname, starts_with("gbif")) %>% 
  mutate(ID_short = gsub("^(.+?);.+", "\\1", seqname)) %>% 
  pivot_longer(!one_of(c("seqname", "ID_short")), names_to = "taxlev", values_to = "tax") %>% 
  mutate(taxlev = gsub("gbif_(\\w).+", "\\1", taxlev)) %>% 
  mutate(taxp = paste(taxlev, tax, sep = ":")) %>% 
  filter(!is.na(tax)) %>% 
  select(seqname, ID_short, taxp) %>% 
  group_by(seqname, ID_short) %>% 
  summarise(taxs = paste(taxp, collapse = ",")) %>% 
  ungroup() %>% 
  mutate(newSeqname = paste(ID_short, ";tax=",taxs,";", sep = ""))

write_tsv(Unite_tax_newseqname, here("Data", "Fungi", "in_files", "Unite_tax_newseqname.txt"))
Unite_tax_newseqname <- read_tsv(here("Data", "Fungi", "in_files", "Unite_tax_newseqname.txt"))
```

###rename sequences
```{r}
Unite_gbif <- Unite

#reorder sequences to match dataframe
Unite_gbif <- Unite_gbif[match(Unite_tax_newseqname$seqname, names(Unite_gbif))]

#rename seqeunces
names(Unite_gbif) <- Unite_tax_newseqname$newSeqname

#export
Unite_gbif <- readDNAStringSet(here("Data", "Fungi", "in_files", "Unite_db_sintax_gbif.fa"))

#make version with only sequences with at least genus assignment
Unite_gbif[grepl("g:", names(Unite_gbif))] %>% 
  writeXStringSet(., here("Data", "Fungi", "in_files", "Unite_db_sintax_gbif_genus.fa"))
```

```{r}
Unite_sename_tax <- 
Unite_tax_newseqname %>% 
  ungroup() %>% 
  select(seqname, newSeqname) %>%
  left_join(select(Unite_tax_newname, seqname, starts_with("gbif"))) %>% 
  rename_all(~gsub("gbif_", "", .x))

write_tsv(Unite_sename_tax, here("Data", "Fungi", "in_files", "Unite_gbif_tax.txt"))
Unite_seqname_tax <- read_tsv(here("Data", "Fungi", "in_files", "Unite_gbif_tax.txt"))

Unite_seqname_tax %>%
  filter(kingdom == "Plantae") %>% 
  group_by(genus, species) %>% summarise(n = n())
  pull(species) %>% unique %>% length
```


#ITS Plants Quaresma et al 

https://www.nature.com/articles/s41597-024-02962-5

```{r}
its_global <- readDNAStringSet(here("Data", "Plants", "in_files", "Plant_ITS_db",  "its2.global.2023-01-17.curated.tax.mc.add.fa"))
```

## extract taxonomy

get taxonomy dataframe for all seqeunces in ITS refDB
```{r}

its_global_tax <- tibble(seqname = names(its_global)) %>% 
  mutate(
    TaxonomyString = str_extract(seqname, "tax=.*"),
    TaxonomyString = str_remove(TaxonomyString, "^tax="),
    TaxonomyString = str_remove(TaxonomyString, "[;,]?$")
  )

# Split the taxonomy string into key-value pairs
its_global_tax <- its_global_tax %>%
  mutate(
    TaxaPairs = str_split(TaxonomyString, ",")
  )

# Function to process key-value pairs
process_taxa_pairs <- function(taxa_pairs) {
  # Split each key-value pair by colon
  pairs_list <- str_split(taxa_pairs, ":")
  
  # Create a named vector
  taxa_list <- sapply(pairs_list, function(x) {
    key <- x[1]
    value <- ifelse(length(x) > 1, x[2], NA)
    names(value) <- key
    return(value)
  })
  
  return(taxa_list)
}

# Apply the function to each row and expand the taxonomy
its_global_tax <- its_global_tax %>%
  mutate(
    TaxaList = lapply(TaxaPairs, process_taxa_pairs)
  ) %>%
  unnest_wider(TaxaList)

its_global_tax <- 
  its_global_tax %>% 
  select(seqname, k,p,c,o,f,g,s) %>% 
  pivot_longer(-seqname, names_to = "taxonomy_level", values_to = "taxa") %>% 
  left_join(tax_match_df) %>% 
  select(-taxonomy_level) %>% 
  pivot_wider(names_from = taxon, values_from = taxa)

#export taxonomy table
write_tsv(its_global_tax,here("Data", "Plants", "in_files", "Plant_ITS_db", "its_global_tax.txt"))
```



## rgbif
```{r}

its_global_gbif_in <- 
  its_global_tax %>% 
  select(-seqname) %>% 
  distinct() %>% 
  mutate(name = gsub("_", " ", species))

# Define the chunk size
chunk_size <- 2000

# Define the path to save intermediate results
save_path <- here("Data", "Plants", "in_files", "Plant_ITS_db")

its_global_gbif_out <- process_dataframe_in_chunks(its_global_gbif_in, chunk_size, save_path)

retry_df <- 
its_global_gbif_in %>% 
  ungroup() %>% 
  mutate(index = 1:n()) %>% 
  filter(!index %in% its_global_gbif_out$verbatim_index)

retry_out <- process_dataframe_in_chunks(retry_df, chunk_size, save_path)

its_global_gbif_out <- bind_rows(its_global_gbif_out, retry_out)

write_tsv(its_global_gbif_out,here("Data", "Plants", "in_files", "Plant_ITS_db", "its_global_gbif.txt"))

# Delete chunk files
chunk_files <- list.files(save_path, pattern = "\\.rds$", full.names = TRUE)
file.remove(chunk_files)



```

Merge gbif taxonomy to its_global_tax taxonomy
```{r}

its_global_gbif_out <- read_tsv(here("Data", "Plants", "in_files", "Plant_ITS_db", "its_global_gbif.txt"))

its_global_newname <- 
  its_global_gbif_in %>%
  rename_with(~ paste("verbatim", ., sep = "_"), everything()) %>% 
  left_join(its_global_gbif_out) %>% 
  select(starts_with("verbatim"), kingdom, phylum, class, order, family, genus, species) %>% 
  dplyr::rename_with( ~paste("gbif",., sep = "_"), .cols = !starts_with("verbatim")) %>% 
  select(-verbatim_name, -verbatim_index) %>% 
  dplyr::rename_with( ~gsub("verbatim_(\\w+)", "\\1",.), .cols = starts_with("verbatim")) %>% 
  distinct() %>% 
  left_join(its_global_tax,.) %>% 
  mutate(ID = gsub("(.+?);tax=.+", "\\1", seqname))

#check that ID's are unique
length(its_global_newname$ID) == unique(length(its_global_newname$ID))
```

replace Plant ITS taxonomy

example name
```{r}
names(its_global)[1]
```


```{r}

its_global_newseqname <- 
its_global_newname %>% 
  select(seqname, ID, starts_with("gbif")) %>% 
  mutate(gbif_species = gsub(" ", "_", gbif_species)) %>% #no spaces in sequence names for vsearch
  pivot_longer(!one_of("ID", "seqname"), names_to = "taxlev", values_to = "tax") %>% 
  mutate(taxlev = gsub("gbif_(\\w).+", "\\1", taxlev)) %>% 
  mutate(taxp = paste(taxlev, tax, sep = ":")) %>% 
  select(seqname, ID, taxp) %>% 
  group_by(seqname, ID) %>% 
  summarise(taxs = paste(taxp, collapse = ",")) %>% 
  mutate(newSeqname = paste(ID, ";tax=",taxs,";", sep = ""))

its_global_gbif <- its_global

#reorder sequences to match dataframe
its_global_gbif <- its_global_gbif[match(its_global_newseqname$ID, gsub("(.+?);.+", "\\1", names(its_global_gbif)))]

#rename seqeunces
names(its_global_gbif) <- its_global_newseqname$newSeqname

#export
writeXStringSet(its_global_gbif, here("Data", "Plants", "in_files", "Plant_ITS_db", "its_global_gbif.fa"))

#version with only sequences with species level assignment
its_global_gbif_species <- its_global_gbif[!grepl(",s:NA", names(its_global_gbif))]

#export
writeXStringSet(its_global_gbif_species, here("Data", "Plants", "in_files", "Plant_ITS_db", "its_global_gbif_sp.fa"))
```

export seqname tax df
```{r}
its_global_seqname_tax <- 
its_global_newseqname %>% 
  ungroup() %>% 
  select(ID, newSeqname) %>% 
  left_join(select(its_global_newname, ID, starts_with("gbif"))) %>% 
  rename_all(~gsub("gbif_", "", .x)) %>% 
  select(-ID)

write_tsv(its_global_seqname_tax, here("Data", "Plants", "in_files", "Plant_ITS_db", "its_global_gbif_tax.txt"))

its_global_seqname_tax <- read_tsv(here("Data", "Plants", "in_files", "Plant_ITS_db", "its_global_gbif_tax.txt"))


its_global_seqname_tax %>% 
  group_by(genus, species) %>% 
  summarise(n = n())
```

#Midori COI

1.Leray M, Knowlton N, Machida RJ. In press. MIDORI2: A collection of quality controlled, preformatted, and regularly updated reference databases for taxonomic assignment of eukaryotic mitochondrial sequences. Environmental DNA n/a. (doi:10.1002/edn3.303)

https://www.reference-midori.info/download/Databases/GenBank261_2024-06-15/SINTAX/uniq/MIDORI2_UNIQ_NUC_GB261_CO1_SINTAX.fasta.gz

```{r}
Midori <- readDNAStringSet(here("Data", "Insects", "in_files",  "MIDORI2_UNIQ_NUC_GB261_CO1_SINTAX.fasta"))
```

## extract taxonomy

get taxonomy dataframe for all seqeunces in refDB
```{r}

Midori_tax <-
  tibble(seqname = names(Midori)) %>%
#ä  dplyr::slice(1:10) %>% 
  mutate(
    TaxonomyString = str_extract(seqname, "tax=.*"),
    TaxonomyString = str_remove(TaxonomyString, "^tax="),
    TaxonomyString = str_remove(TaxonomyString, "[;,]?$")
  ) %>% 
  mutate(TaxonomyString = gsub("_\\d+", "", TaxonomyString))

# Split the taxonomy string into key-value pairs
Midori_tax <- Midori_tax %>%
  mutate(
    TaxaPairs = str_split(TaxonomyString, ",")
  )

# Function to process key-value pairs
process_taxa_pairs <- function(taxa_pairs) {
  # Split each key-value pair by colon
  pairs_list <- str_split(taxa_pairs, ":")
  
  # Create a named vector
  taxa_list <- sapply(pairs_list, function(x) {
    key <- x[1]
    value <- ifelse(length(x) > 1, x[2], NA)
    names(value) <- key
    return(value)
  })
  
  return(taxa_list)
}

# Apply the function to each row and expand the taxonomy
Midori_tax <- Midori_tax %>%
  mutate(
    TaxaList = lapply(TaxaPairs, process_taxa_pairs)
  ) %>%
  unnest_wider(TaxaList)

Midori_tax <- 
  Midori_tax %>% 
  select(seqname, k,p,c,o,f,g,s) %>% 
  pivot_longer(-seqname, names_to = "taxonomy_level", values_to = "taxa") %>% 
  left_join(tax_match_df) %>% 
  select(-taxonomy_level) %>% 
  pivot_wider(names_from = taxon, values_from = taxa)

#export taxonomy table
write_tsv(Midori_tax,here("Data", "Plants", "in_files", "Plant_ITS_db", "Midori_tax.txt"))
```



## rgbif
```{r}

Midori_gbif_in <- 
  Midori_tax %>% 
  select(-seqname) %>% 
  distinct() %>% 
  mutate(name = gsub("_", " ", species))

# Define the chunk size
chunk_size <- 2000

# Define the path to save intermediate results
save_path <- here("Data", "Insects", "in_files")

Midori_gbif_out <- process_dataframe_in_chunks(Midori_gbif_in, chunk_size, save_path)

retry_df <- 
Midori_gbif_in %>% 
  ungroup() %>% 
  mutate(index = 1:n()) %>% 
  filter(!index %in% Midori_gbif_out$verbatim_index)

retry_out <- process_dataframe_in_chunks(retry_df, chunk_size, save_path)

Midori_gbif_out <- bind_rows(Midori_gbif_out, retry_out)

write_tsv(Midori_gbif_out,here("Data", "Insects", "in_files", "Midori_gbif.txt"))

# Delete chunk files
chunk_files <- list.files(save_path, pattern = "\\.rds$", full.names = TRUE)
file.remove(chunk_files)



```

Merge gbif taxonomy to Midori_tax taxonomy
```{r}

Midori_gbif_out <- read_tsv(here("Data", "Insects", "in_files", "Midori_gbif.txt"))

Midori_newname <- 
  Midori_gbif_in %>%
  rename_with(~ paste("verbatim", ., sep = "_"), everything()) %>% 
  left_join(Midori_gbif_out) %>% 
  select(starts_with("verbatim"), kingdom, phylum, class, order, family, genus, species) %>% 
  dplyr::rename_with( ~paste("gbif",., sep = "_"), .cols = !starts_with("verbatim")) %>% 
  select(-verbatim_name, -verbatim_index) %>% 
  dplyr::rename_with( ~gsub("verbatim_(\\w+)", "\\1",.), .cols = starts_with("verbatim")) %>% 
  distinct() %>% 
  left_join(Midori_tax,.) %>% 
  mutate(ID = gsub("(.+?);tax=.+", "\\1", seqname))

#check that ID's are unique
length(Midori_newname$ID) == unique(length(Midori_newname$ID))
```

replace Midori taxonomy

example name
```{r}
names(Midori)[1]
```


```{r}

Midori_newseqname <- 
Midori_newname %>% 
  select(seqname, ID, starts_with("gbif")) %>% 
  mutate(gbif_species = gsub(" ", "_", gbif_species)) %>% #no spaces in sequence names for vsearch
  pivot_longer(!one_of("ID", "seqname"), names_to = "taxlev", values_to = "tax") %>% 
  mutate(taxlev = gsub("gbif_(\\w).+", "\\1", taxlev)) %>% 
  mutate(taxp = paste(taxlev, tax, sep = ":")) %>% 
  select(seqname, ID, taxp) %>% 
  group_by(seqname, ID) %>% 
  summarise(taxs = paste(taxp, collapse = ",")) %>% 
  mutate(newSeqname = paste(ID, ";tax=",taxs,";", sep = ""))

Midori_gbif <- Midori

#reorder sequences to match dataframe
Midori_gbif <- Midori_gbif[match(Midori_newseqname$ID, gsub("(.+?);.+", "\\1", names(Midori_gbif)))]

#rename seqeunces
names(Midori_gbif) <- Midori_newseqname$newSeqname

#export
writeXStringSet(Midori_gbif, here("Data", "Plants", "in_files", "Plant_ITS_db", "Midori_gbif.fa"))

#version with only sequences at least genus level assignment
Midori_gbif_species <- Midori_gbif[!grepl(",g:NA", names(Midori_gbif))]

#export
writeXStringSet(Midori_gbif_species, here("Data", "Insects", "in_files", "Midori_gbif_sp.fa"))
```

export seqname tax df
```{r}
Midori_seqname_tax <- 
Midori_newseqname %>% 
  ungroup() %>% 
  select(ID, newSeqname) %>% 
  left_join(select(Midori_newname, ID, starts_with("gbif"))) %>% 
  rename_all(~gsub("gbif_", "", .x)) %>% 
  select(-ID)

write_tsv(Midori_seqname_tax, here("Data", "Insects", "in_files", "Midori_gbif_tax.txt"))
Midori_seqname_tax <- read_tsv(here("Data", "Insects", "in_files", "Midori_gbif_tax.txt"))

Midori_seqname_tax %>% filter(phylum == "Chordata") %>% group_by(genus, species) %>% summarise(n = n())
```


#Midori 16S

1.Leray M, Knowlton N, Machida RJ. In press. MIDORI2: A collection of quality controlled, preformatted, and regularly updated reference databases for taxonomic assignment of eukaryotic mitochondrial sequences. Environmental DNA n/a. (doi:10.1002/edn3.303)

https://www.reference-midori.info/download/Databases/GenBank261_2024-06-15/SINTAX/uniq/MIDORI2_UNIQ_NUC_GB261_lrRNA_SINTAX.fasta.gz

```{r}
Midori_lr <- readDNAStringSet(here("Data", "Vertebrates", "in_files",  "MIDORI2_UNIQ_NUC_GB261_lrRNA_SINTAX.fasta"))
```

## extract taxonomy

get taxonomy dataframe for all seqeunces in refDB
```{r}

Midori_lr_tax <-
  tibble(seqname = names(Midori_lr)) %>%
 #dplyr::slice(1:10) %>% 
  mutate(
    TaxonomyString = str_extract(seqname, "tax=.*"),
    TaxonomyString = str_remove(TaxonomyString, "^tax="),
    TaxonomyString = str_remove(TaxonomyString, "[;,]?$")
  ) %>% 
  mutate(TaxonomyString = gsub("_\\d+", "", TaxonomyString))

# Split the taxonomy string into key-value pairs
Midori_lr_tax <- Midori_lr_tax %>%
  mutate(
    TaxaPairs = str_split(TaxonomyString, ",")
  )

# Function to process key-value pairs
process_taxa_pairs <- function(taxa_pairs) {
  # Split each key-value pair by colon
  pairs_list <- str_split(taxa_pairs, ":")
  
  # Create a named vector
  taxa_list <- sapply(pairs_list, function(x) {
    key <- x[1]
    value <- ifelse(length(x) > 1, x[2], NA)
    names(value) <- key
    return(value)
  })
  
  return(taxa_list)
}

# Apply the function to each row and expand the taxonomy
Midori_lr_tax <- Midori_lr_tax %>%
  mutate(
    TaxaList = lapply(TaxaPairs, process_taxa_pairs)
  ) %>%
  unnest_wider(TaxaList)

Midori_lr_tax <- 
  Midori_lr_tax %>% 
  select(seqname, k,p,c,o,f,g,s) %>% 
  pivot_longer(-seqname, names_to = "taxonomy_level", values_to = "taxa") %>% 
  left_join(tax_match_df) %>% 
  select(-taxonomy_level) %>% 
  pivot_wider(names_from = taxon, values_from = taxa)

#export taxonomy table
write_tsv(Midori_lr_tax,here("Data", "Vertebrates", "in_files", "Midori_lr_tax.txt"))
```



## rgbif
```{r}

Midori_lr_gbif_in <- 
  Midori_lr_tax %>% 
  select(-seqname) %>% 
  distinct() %>% 
  mutate(name = gsub("_", " ", species))

# Define the chunk size
chunk_size <- 2000

# Define the path to save intermediate results
save_path <- here("Data", "Vertebrates", "in_files")

Midori_lr_gbif_out <- process_dataframe_in_chunks(Midori_lr_gbif_in, chunk_size, save_path)

retry_df <- 
Midori_lr_gbif_in %>% 
  ungroup() %>% 
  mutate(index = 1:n()) %>% 
  filter(!index %in% Midori_lr_gbif_out$verbatim_index)

retry_out <- process_dataframe_in_chunks(retry_df, chunk_size, save_path)

Midori_lr_gbif_out <- bind_rows(Midori_lr_gbif_out, retry_out)

write_tsv(Midori_lr_gbif_out,here("Data", "Vertebrates", "in_files", "Midori_lr_gbif.txt"))

# Delete chunk files
chunk_files <- list.files(save_path, pattern = "\\.rds$", full.names = TRUE)
file.remove(chunk_files)



```

Merge gbif taxonomy to Midori_lr_tax taxonomy
```{r}

Midori_lr_gbif_out <- read_tsv(here("Data", "Vertebrates", "in_files", "Midori_lr_gbif.txt"))

Midori_lr_newname <- 
  Midori_lr_gbif_in %>%
  rename_with(~ paste("verbatim", ., sep = "_"), everything()) %>% 
  left_join(Midori_lr_gbif_out) %>% 
  select(starts_with("verbatim"), kingdom, phylum, class, order, family, genus, species) %>% 
  dplyr::rename_with( ~paste("gbif",., sep = "_"), .cols = !starts_with("verbatim")) %>% 
  select(-verbatim_name, -verbatim_index) %>% 
  dplyr::rename_with( ~gsub("verbatim_(\\w+)", "\\1",.), .cols = starts_with("verbatim")) %>% 
  distinct() %>% 
  left_join(Midori_lr_tax,.) %>% 
  mutate(ID = gsub("(.+?);tax=.+", "\\1", seqname))

#check that ID's are unique
length(Midori_lr_newname$ID) == unique(length(Midori_lr_newname$ID))
```

replace Midori_lr taxonomy

example name
```{r}
names(Midori_lr)[1]
```


```{r}

Midori_lr_newseqname <- 
Midori_lr_newname %>% 
  select(seqname, ID, starts_with("gbif")) %>% 
  mutate(gbif_species = gsub(" ", "_", gbif_species)) %>% #no spaces in sequence names for vsearch
  pivot_longer(!one_of("ID", "seqname"), names_to = "taxlev", values_to = "tax") %>% 
  mutate(taxlev = gsub("gbif_(\\w).+", "\\1", taxlev)) %>% 
  mutate(taxp = paste(taxlev, tax, sep = ":")) %>% 
  select(seqname, ID, taxp) %>% 
  group_by(seqname, ID) %>% 
  summarise(taxs = paste(taxp, collapse = ",")) %>% 
  mutate(newSeqname = paste(ID, ";tax=",taxs,";", sep = ""))

Midori_lr_gbif <- Midori_lr

#reorder sequences to match dataframe
Midori_lr_gbif <- Midori_lr_gbif[match(Midori_lr_newseqname$ID, gsub("(.+?);.+", "\\1", names(Midori_lr_gbif)))]

#rename seqeunces
names(Midori_lr_gbif) <- Midori_lr_newseqname$newSeqname

#export
writeXStringSet(Midori_lr_gbif, here("Data", "Vertebrates", "in_files", "Midori_lr_gbif.fa"))

#version with only sequences with species level assignment
Midori_lr_gbif_species <- Midori_lr_gbif[!grepl(",s:NA", names(Midori_lr_gbif))]

#export
writeXStringSet(Midori_lr_gbif_species, here("Data", "Vertebrates", "in_files", "Midori_lr_gbif_sp.fa"))
```

export seqname tax df
```{r}
Midori_lr_seqname_tax <- 
Midori_lr_newseqname %>% 
  ungroup() %>% 
  select(ID, newSeqname) %>% 
  left_join(select(Midori_lr_newname, ID, starts_with("gbif"))) %>% 
  rename_all(~gsub("gbif_", "", .x)) %>% 
  select(-ID)

write_tsv(Midori_lr_seqname_tax, here("Data", "Vertebrates", "in_files", "Midori_lr_gbif_tax.txt"))
Midori_lr_seqname_tax <- read_tsv(here("Data", "Vertebrates", "in_files", "Midori_lr_gbif_tax.txt"))

Arth_16S <- Midori_lr_seqname_tax %>% filter(phylum == "Chordata") %>% group_by(genus, species) %>%  summarise(n = n())
Arth_COI <- Midori_seqname_tax %>% filter(phylum == "Chordata") %>% group_by(genus, species) %>% summarise(n = n())

bind_rows(Arth_16S, Arth_COI) %>% select(-n) %>% distinct()
```

##BOLD Data

```{r}

# Function to process FASTA and load into SQLite
process_fasta_to_db <- function(filepath, db_name, chunk_size = 1000) {
  # Create SQLite connection
  conn <- dbConnect(SQLite(), dbname = db_name)
  
  # Initialize database with a table
  dbExecute(conn, "CREATE TABLE fasta_data (
    sample_id TEXT,
    marker TEXT,
    country TEXT,
    kingdom TEXT,
    phylum TEXT,
    class TEXT,
    order_col TEXT,
    family TEXT,
    subfamily TEXT,
    genus TEXT,
    species TEXT,
    subspecies TEXT,
    sequence TEXT
  );")
  
  # Initialize counters
  skip_lines <- 0
  
  repeat {
    #progress
    message(paste("reading lines" , skip_lines, "to", skip_lines+chunk_size, "...", sep = " "))
    
    # Read chunk
    chunk <- readDNAStringSet(filepath, format = "fasta", skip = skip_lines, nrec = chunk_size)
    
    # Break if no sequences left
    if (length(chunk) == 0) break
    
    # Extract headers and sequences
    headers <- names(chunk)
    sequences <- as.character(chunk)
    
    # Parse metadata from headers
    metadata <- str_split_fixed(headers, "\\|", n = 4)
    taxonomy <- str_split_fixed(metadata[, 4], ",", n = 9)
    
    # Create a data frame
    fasta_data <- data.frame(
      sample_id = metadata[, 1],
      marker = metadata[, 2],
      country = metadata[, 3],
      kingdom = taxonomy[, 1],
      phylum = taxonomy[, 2],
      class = taxonomy[, 3],
      order_col = taxonomy[, 4],
      family = taxonomy[, 5],
      subfamily = taxonomy[, 6],
      genus = taxonomy[, 7],
      species = taxonomy[, 8],
      subspecies = taxonomy[, 9],
      sequence = sequences,
      stringsAsFactors = FALSE
    )
    
    # Write data to the database
    dbWriteTable(conn, "fasta_data", fasta_data, append = TRUE, row.names = FALSE)
    
    # Update skip_lines
    skip_lines <- skip_lines + chunk_size
  }
  
  # Close connection
  dbDisconnect(conn)
  
  return(db_name)
}
```

```{r}
# Process the FASTA file and populate the database
db_name <- here("Data", "Insects", "in_files", "BOLD_Public.sqlite")
fasta_file <- here("Data", "Insects", "in_files", "BOLD_Public.27-Sep-2024.fasta")
process_fasta_to_db(fasta_file, db_name, chunk_size = 1e6)
```

split into taxonomy and sequence table
```{r}

# Connect to the database
db_name <- here("Data", "Insects", "in_files", "BOLD_Public.sqlite")
conn <- dbConnect(RSQLite::SQLite(), dbname = db_name)

# Step 1: Create the taxonomy table with a primary key
dbExecute(conn, "DROP TABLE IF EXISTS taxonomy;")
dbExecute(conn, "
  CREATE TABLE taxonomy (
    taxonomy_id INTEGER PRIMARY KEY AUTOINCREMENT,
    kingdom TEXT,
    phylum TEXT,
    class TEXT,
    order_col TEXT,
    family TEXT,
    subfamily TEXT,
    genus TEXT,
    species TEXT
  );
")

# Insert distinct taxonomy records into the taxonomy table
dbExecute(conn, "
  INSERT INTO taxonomy (kingdom, phylum, class, order_col, family, subfamily, genus, species)
  SELECT DISTINCT kingdom, phylum, class, order_col, family, subfamily, genus, species
  FROM fasta_data;
")

# Step 2: Create the sequences table with a reference to taxonomy_id
dbExecute(conn, "DROP TABLE IF EXISTS sequences;")
dbExecute(conn, "
  CREATE TABLE sequences AS
  SELECT 
    f.sample_id,
    f.marker,
    f.country,
    t.taxonomy_id,
    f.sequence
  FROM fasta_data f
  INNER JOIN taxonomy t
  ON f.kingdom = t.kingdom
     AND f.phylum = t.phylum
     AND f.class = t.class
     AND f.order_col = t.order_col
     AND f.family = t.family
     AND f.subfamily = t.subfamily
     AND f.genus = t.genus
     AND f.species = t.species;
")

```

get gbif taxonomy
```{r}
db_name <- here("Data", "Insects", "in_files", "BOLD_Public.sqlite")
conn <- dbConnect(RSQLite::SQLite(), dbname = db_name)

taxonomy_tbl <- tbl(conn, "taxonomy")

Bold_tax_gbif <- collect(taxonomy_tbl)

Bold_tax_gbif <- 
  Bold_tax_gbif %>% 
  dplyr::rename(order = order_col) %>%
  mutate(across(-taxonomy_id, ~gsub("^None$", NA, .x)))

# Define the chunk size
chunk_size <- 2000

# Define the path to save intermediate results
save_path <- here("Data", "Insects", "in_files")

bold_gbif_out <- process_dataframe_in_chunks(Bold_tax_gbif, chunk_size, save_path)

write_tsv(bold_gbif_out, here("Data", "Insects", "in_files", "Bold_gbif.txt"))

# Delete chunk files
chunk_files <- list.files(save_path, pattern = "\\.rds$", full.names = TRUE)
file.remove(chunk_files)

```

make new database with gbif taxonomy
```{r}
bold_gbif_out <- 
  dplyr::rename(bold_gbif_out, verbatim_species = verbatim_name) %>% 
  select(starts_with("verbatim"), kingdom, phylum, class, order, family, genus, species) %>% 
  select(-verbatim_index) %>% 
  distinct()

gbif_taxonomy <- 
  Bold_tax_gbif %>%
  rename_with(~ paste("verbatim", ., sep = "_"), -taxonomy_id) %>% 
  left_join(bold_gbif_out) %>% 
  select(taxonomy_id, kingdom, phylum, class, order, family, genus, species) 

#add sintax taxonomy string 
gbif_taxonomy <- 
gbif_taxonomy %>% 
 mutate(species = gsub(" ", "_", species)) %>% #no spaces in sequence names for vsearch
  pivot_longer(!one_of("taxonomy_id"), names_to = "taxlev", values_to = "tax") %>% 
  mutate(taxlev = gsub("(\\w).+", "\\1", taxlev)) %>% 
  mutate(taxp = paste(taxlev, tax, sep = ":")) %>% 
  select(taxonomy_id, taxp) %>% 
  group_by(taxonomy_id) %>% 
  summarise(sinatx_tax = paste(taxp, collapse = ",")) %>% 
  left_join(gbif_taxonomy, .)

# Create sequences table by writing directly to the database
dbExecute(conn, "DROP TABLE IF EXISTS gbif_taxonomy;") 

# write gbif taxonomy table
dbWriteTable(conn, "gbif_taxonomy", gbif_taxonomy, row.names = FALSE)
```

```{r}
# 1. Load the sequence table as a dplyr table
sequence_tbl <- tbl(conn, "sequences")

# 2. Load the GBIF taxonomy table as a dplyr table
gbif_tax_tbl <- tbl(conn, "gbif_taxonomy")

# 3. Join the two tables on tax_id
joined_tbl <- sequence_tbl %>%
  inner_join(gbif_tax_tbl, by = "taxonomy_id")

# 4. Filter for the marker of interest
filtered_tbl <- joined_tbl %>%
  filter(marker == "COI-5P")

# Separate sequences with and without species-level assignments
sequences_with_species <- filtered_tbl %>%
  filter(!is.na(species))

sequences_without_species <- filtered_tbl %>%
  filter(is.na(species))

# Sample max 10 sequences per species for those with species-level assignments
sampled_with_species <- sequences_with_species %>%
  group_by(species) %>%
  slice_sample(n = 10, replace = FALSE) %>%
  ungroup()

# Combine both datasets
final_sequences_tbl <- sampled_with_species %>%
  union_all(sequences_without_species)

# 5. Create the new sequence name
result_tbl <- final_sequences_tbl %>%
  mutate(sequence_name = paste0(sample_id, ";tax=", sinatx_tax, sep = "")) %>%
  select(sequence_name, sequence) %>% 
  compute()

# 6. Write the sequences to a FASTA file in chunks
output_fasta <- here("Data", "Insects", "in_files", "Bold_gbif_COI.fasta")

# Remove existing FASTA file if it exists
if (file.exists(output_fasta)) file.remove(output_fasta)


# Loop through the database table in chunks

# Define chunk size
chunk_size <- 1000000

# Step 3: Initialize pagination
offset <- 0

repeat {
  # progress: 
  message(paste("processing lines", offset, "to", offset + chunk_size, "..."))
  
  # Use dplyr to slice a chunk from the computed table
  chunk <- 
    result_tbl %>% 
    filter(row_number() > offset & row_number() <= (offset + chunk_size)) %>%
    collect()
  
  # Break the loop if no more rows are fetched
  if (nrow(chunk) == 0) break
  
  # Convert the chunk into DNAStringSet
  dna_set <- chunk$sequence
  names(dna_set) <- chunk$sequence_name
  dna_set <- DNAStringSet(dna_set)
  
  # Append the chunk to the FASTA file
  writeXStringSet(dna_set, filepath = output_fasta, append = TRUE)
  
  # Increment the offset
  offset <- offset + chunk_size
}
```


```{r}
# Connect to the database
conn <- dbConnect(SQLite(), dbname = db_name)

# Use dplyr to interact with the database
fasta_tbl <- tbl(conn, "fasta_data")

# Example 1: Count total records
fasta_tbl %>%
  summarise(total_records = n()) %>%
  collect()

# the marker we want is COI-5P
fasta_tbl %>% group_by(marker) %>% summarise(n = n()) %>% collect()

Taxa_unique <- 
fasta_tbl %>% 
  filter(marker == "COI-5P") %>% 
  group_by(kingdom,phylum,class,order_col,family,subfamily,genus,species) %>% 
  summarise(n =n()) %>% 
  collect()
```



## Swiss Vertebrates

GBIF.org (13 November 2024) GBIF Occurrence Download  https://doi.org/10.15468/dl.ujk7mf

Basis of record: Observation Machine; observation Human; observation
Country or area: Switzerland
Scientific name: Chordata

```{r}
SwissVert_gbif <- read_tsv(here("Data", "Vertebrates", "in_files", "0012786-241107131044228.csv"))

SwissVert_gbif <- 
SwissVert_gbif %>% 
  select(phylum, class, order, family, genus, species, iucnRedListCategory, numberOfOccurrences) 

SwissVert_gbif <-  
SwissVert_gbif %>% 
  group_by(genus) %>% 
  filter(
        !is.na(species) | all(is.na(species))
      ) %>% 
  group_by(phylum, class, order, family, genus, species) %>% 
  reframe(iucnRedListCategory = unique(iucnRedListCategory[!is.na(iucnRedListCategory)]),
            numberOfOccurrences = sum(numberOfOccurrences)) %>% 
  mutate(inGbif = TRUE) %>% 
  group_by(genus, species) %>% 
  mutate(iucnRedListCategory = case_when(n() > 1 ~ NA_character_, 
                                         TRUE ~ iucnRedListCategory)) %>% 
  distinct()


write_tsv(SwissVert_gbif, here("Data", "Vertebrates", "in_files", "SwissVert_gbif.txt"))
```







##Unite_Swiss_Plants

We want a curated reference databases which contains only the species known to occur in Switzerland

As not all species from InfoFlora are present in either Unite or PlantITS, we attempt to search for the sequences on NCBI & BOLD

### bold & ncbi
```{r}
InfoFlora_sp <- 
species_venn_df %>% 
  filter(InfoFlora & !Unite & !plantITS) %>% pull(species)

InfoFlora_bold <- 
  InfoFlora_sp %>% 
  split(., ceiling(seq_along(.)/200)) %>% 
  lapply(., function(x) bold_seqspec(taxon = x)) %>% 
  bind_rows()

InfoFlora_bold_seq <- 
InfoFlora_bold %>% 
  filter(markercode == "ITS2") %>% 
  select(sequenceID, species_name, nucleotides) %>% 
  dplyr::rename(ID = sequenceID) %>% 
  distinct()

InfoFlora_ncbi <- 
  InfoFlora_sp %>% 
  split(., ceiling(seq_along(.)/50)) %>% 
  lapply(., function(x) {refdb_import_NCBI(
    query = paste("(", paste0(
      x, "[Organism]", collapse = " OR "),
      ")",
      " AND ITS2[All Fields] AND gene[All Fields]",
      sep = ""))})

InfoFlora_ncbi_seq <- 
InfoFlora_ncbi %>% 
  bind_rows() %>% 
  select(id, species, sequence) %>% 
  dplyr::rename(ID = id) %>% 
  distinct()

IF_eu_df <- 
tibble(species = InfoFlora_sp) %>% 
  left_join(InfoFlora_ncbi_seq) %>% 
  select(-ID) %>% 
  dplyr::rename(NCBI = sequence) %>% 
  mutate(NCBI = !is.na(NCBI)) %>% 
  distinct() %>% 
  left_join(dplyr::rename(InfoFlora_bold_seq, species = species_name)) %>% 
  select(-ID) %>% 
  dplyr::rename(BOLD = nucleotides) %>% 
  mutate(BOLD = !is.na(BOLD)) %>% 
  distinct() %>% 
  mutate(InfoFlora = TRUE) %>% 
  mutate(Beaver = species %in% Beaver_plants_gbif$species) 

IF_eu <- 
  IF_eu_df %>% 
  select(-species) %>% 
  euler(shape = "ellipse") 

plot(IF_eu, legend = TRUE, quantities = TRUE)
```


```{r}
IF_eu_df %>% filter(Beaver & !NCBI & !BOLD)
```


## SwissPlants
get species from each database
```{r}
Unite_ch <- 
species_venn_df %>% 
  filter(InfoFlora & Unite) %>% 
  pull(species)

PlantITS_ch_u <- 
species_venn_df %>% 
  filter(InfoFlora & plantITS & !Unite) %>% 
  pull(species)

Bold_NCBI_u <- 
  IF_eu_df %>% 
  filter(BOLD | NCBI)
```

##get sequences

for species with more than 10 reference sequences in the database I keep the 10 longest sequences

###Unite

```{r}
Unite_seqs <- 
Unite_tax_newname %>%
  ungroup() %>% 
  left_join(tibble(ID = names(Unite), len = width(Unite))) %>% 
  filter(gbif_species %in% Unite_ch) %>% 
  group_by(gbif_species, SH) %>% 
  arrange(species, SH, desc(len)) %>% 
  dplyr::slice(min(10, n())) %>% 
  pull(ID)
```

```{r}
Unite_sub <- Unite[Unite_seqs]

Unite_sub_names <- 
Unite_tax_newseqname %>%
  ungroup() %>% 
  dplyr::slice(match(names(Unite_sub),ID)) %>% 
  pull(newSeqname)

names(Unite_sub) <- Unite_sub_names
```

###PlantITS
```{r}
PlantITS_seq <- 
  PlantITS_tax_newname %>% 
  ungroup() %>% 
  left_join(tibble(Seqname = names(plantITS), len = width(plantITS))) %>% 
  filter(gbif_species %in% PlantITS_ch_u) %>% 
  group_by(gbif_species) %>% 
  arrange(species, desc(len)) %>% 
  dplyr::slice(min(10, n())) %>% 
  pull(Seqname)

PlantITS_sub <- plantITS[PlantITS_seq]

PlantITS_sub_names <- 
PlantITS_newseqname %>%
  ungroup() %>% 
  dplyr::slice(match(names(PlantITS_sub),Seqname)) %>% 
  pull(newSeqname)

names(PlantITS_sub) <- PlantITS_sub_names
```

###ncbi & bold
```{r}

ncbi_bold_seq <- 
InfoFlora_bold_seq %>% 
  dplyr::rename(sequence = nucleotides, species = species_name) %>% 
  mutate(sequence = gsub("-", "", sequence)) %>% 
  mutate(ID = as.character(ID)) %>% 
  bind_rows(InfoFlora_ncbi_seq) %>% 
  distinct() %>% 
  mutate(len = nchar(sequence)) %>% 
  group_by(species) %>% 
  arrange(species, desc(len)) %>% 
  dplyr::slice(min(10, n()))
 
ncbi_bold_seq_names <- 
  ncbi_bold_seq %>% 
  select(ID, species) %>% 
  name_backbone_checklist() %>% 
  mutate(ID = ncbi_bold_seq$ID[.$verbatim_index]) %>% 
  select(ID, kingdom, phylum, class, order, family, genus, species) %>% 
  filter(!is.na(species)) %>% 
  filter(species %in% InfoFlora_gbif$species) %>% 
  pivot_longer(-ID, names_to = "taxlev", values_to = "tax") %>% 
  mutate(taxlev = gsub("(\\w).+", "\\1", taxlev)) %>% 
  mutate(taxp = paste(taxlev, tax, sep = ":")) %>% 
  select(ID, taxp) %>% 
  group_by(ID) %>% 
  summarise(taxs = paste(taxp, collapse = ",")) %>% 
  mutate(newSeqname = paste(ID, ";tax=",taxs,";", sep = ""))

ncbi_bold_seq <- 
  ncbi_bold_seq %>% 
  left_join(ncbi_bold_seq_names) %>% 
  filter(!is.na(newSeqname))


ncbi_bold_ch <- DNAStringSet(ncbi_bold_seq$sequence)
names(ncbi_bold_ch) <- ncbi_bold_seq$newSeqname
```

##join

```{r}
Plants_CH <- 
c(Unite_sub, PlantITS_sub, ncbi_bold_ch)

#ITSx chops off the seqname after a space, so we need to remove them
names(Plants_CH) <- gsub("\\s", "_", names(Plants_CH))

writeXStringSet(Plants_CH, here("Data", "Plants_CH_refDB.fasta"), width = max(width(Plants_CH)))
```

##ITSx
```{r}
cores = 6

#copy files to downloads because of path problems (space in path)
system2(command = "cp", 
        args = c(shQuote(here("Data", "Plants_CH_refDB.fasta")),
                 "~/Downloads/"))

#create tmp output directory 
dir.create("~/Downloads/ITSX_out")

system(paste("source ~/.bash_profile && /Applications/ITSx_1.1.3/ITSx",
             "ITSx -i ",
             "~/Downloads/Plants_CH_refDB.fasta ",
             "-o ~/Downloads/ITSX_out/ITSx ", 
             "-t all ",
             "--allow_single_domain '1e-5' ",
             "--complement F ",
             "--save_regions ITS2 ",
             "--cpu ", 
             cores
             ))

#move results
system2(command = "mv", 
        args = c("~/Downloads/ITSX_out",
                 shQuote(here("Data"))))

#clean up
file.remove("~/Downloads/Plants_CH_refDB.fasta")
```

read in sequences with ITS2 domains

```{r}
ITSx_pos <- read_tsv(here("Data", "ITSX_out", "ITSx.positions.txt"), col_names = F) %>% 
  `colnames<-`(c("newSeqname", "length", "SSU", "ITS1", "S58", "ITS2", "LSU", "tab")) %>% 
  select(-tab)

#4 sequences have no ITS2 (they most stop after ITS1)
ITSx_pos %>% 
  filter(grepl("Not found", ITS2)) %>% 
  nrow()

ITS2_pos <- 
ITSx_pos %>% 
  mutate(ITS2 = gsub("ITS2: ", "", ITS2)) %>% 
  tidyr::separate(ITS2, into = c("ITS2_start", "ITS2_end"), sep = "-") %>% 
  mutate(across(starts_with("ITS"), as.numeric)) %>% 
  mutate(ITS2_len = ITS2_end - ITS2_start) %>% 
  mutate(newSeqname = gsub("_", " ", newSeqname)) %>% 
  filter(newSeqname != "--END--") #not sure why these arise
```


```{r}
names(Plants_CH) <- gsub("_", " ", names(Plants_CH))

#check if all sequence names are found
ITS2_pos$newSeqname[which(!ITS2_pos$newSeqname %in% names(Plants_CH))]

#check if sequences from Plants_CH are missing
sum(!names(Plants_CH) %in% ITS2_pos$newSeqname)

#199 Sequences are missing. These are in the file ITSx_no_detections.fasta 
readDNAStringSet(here("Data", "ITSX_out", "ITSx_no_detections.fasta")) %>% 
  length()

```

```{r}
Plants_CH_ITS <- Plants_CH

#exclude sequences were ITS2 was not found
ITS_pos_clean <- 
ITS2_pos %>% 
  filter(!is.na(ITS2_len))

#get cumulativve size distribution
ITS_pos_clean %>% 
  group_by(ITS2_len) %>% 
  summarise(n_len = n()) %>% 
  arrange(ITS2_len) %>% 
  mutate(cum_prct = cumsum(n_len/sum(n_len))) %>% 
  ggplot(aes(x = ITS2_len,y = cum_prct))+
  geom_line()+
  scale_x_continuous(limits = c(0,1000))+
  geom_vline(xintercept = c(50, 500))

#we filter sequences <50 and above 500 bps
ITS_pos_clean <- 
  ITS_pos_clean %>% 
  filter(ITS2_len > 50) %>% 
  filter(ITS2_len < 500)
```


```{r}
#subset DB and make sure sequence order matches
Plants_CH_ITS <- Plants_CH_ITS[ITS_pos_clean$newSeqname]

all(names(Plants_CH_ITS) == ITS_pos_clean$newSeqname)

#subset sequences for only ITS2
Plants_CH_ITS <- subseq(Plants_CH_ITS, ITS_pos_clean$ITS2_start, ITS_pos_clean$ITS2_end)

#export
writeXStringSet(seq_ITS2, here("Data", "Plants_CH_refDB_ITSx.fasta"), format = "fasta")
```

```{r}
P_Frwd <- "ATGCGATACTTGGTGTGAAT" 
P_Rev <- "GACGCTTCTCCAGACTACAAT" %>% DNAString() %>% reverseComplement() %>% as.character()

Plant_seqs <- 
Unite_tax_newname %>% 
  left_join(distinct(select(Unite_tax_gbif_out, kingdom, phylum))) %>% 
  filter(kingdom == "Plantae") %>% 
  left_join(Unite_tax_newseqname) %>% 
  filter(ID_short %in% fullSeq) %>% 
  pull(var = ID_short) %>% 
  unique()

Unite_full_gbif_P <- Unite[which(gsub("(.+?)\\|.+", "\\1", names(Unite)) %in% Plant_seqs)]

c(DNAStringSet(P_Frwd),Unite[1:3]) %>% AlignSeqs() %>% BrowseSeqs()
  

vmatchPattern(P_Frwd, Unite_full_gbif_P[1:100], max.mismatch=3, fixed = FALSE)
as.numeric(min(end(vmatchPattern(P_Rev, Unite_full_gbif_P[1:100], max.mismatch=3, fixed = FALSE))))

revT <- vcountPattern(P_Rev, Unite_full_gbif_P, max.mismatch = 3, fixed = FALSE) > 0
Unite_frwd <- subseq(Unite_full_gbif_P[revT], start = as.numeric(start(vmatchPattern(P_Frwd, Unite_full_gbif_P[revT], max.mismatch = 3, fixed = FALSE))), end = Inf)

fullSeq <- 
ITSx_pos %>% 
  filter(!grepl("Not found", LSU)) %>% 
  filter(!grepl("Not found", S58)) %>% 
  pull(Seq) %>% 
  gsub("(.+?);.+", "\\1", .)


c(DNAStringSet(P_Rev), Unite_full_gbif_P[7:11]) %>% AlignSeqs() %>% BrowseSeqs()

```




##ITSx
```{r}
cores = 6

#copy files to downloads because of path problems (space in path)
system2(command = "cp", 
        args = c(shQuote(here("Data", "Plant_ITS_db", "its2.global.2023-01-17.curated.tax.mc.add.fa")),
                 "~/Downloads/"))

#create tmp output directory 
dir.create("~/Downloads/ITSX_out")

system(paste("source ~/.bash_profile && /Applications/ITSx_1.1.3/ITSx",
             "ITSx -i ",
             "~/Downloads/its2.global.2023-01-17.curated.tax.mc.add.fa ",
             "-o ~/Downloads/ITSX_out/ITSx ", 
             "-t all ",
             "--allow_single_domain '1e-5' ",
             "--complement F ",
             "--save_regions ITS2 ",
             "--cpu ", 
             cores
             ))

#move results
system2(command = "mv", 
        args = c("~/Downloads/ITSX_out",
                 shQuote(here("Data", "Plant_ITS_db"))))

#clean up
file.remove("~/Downloads/its2.global.2023-01-17.curated.tax.mc.add.fa")
```

read in sequences with ITS2 domains

```{r}
ITSx_pos <- read_tsv(here("Data", "Plant_ITS_db", "ITSX_out", "ITSx.positions.txt"), col_names = F) %>% 
  `colnames<-`(c("newSeqname", "length", "SSU", "ITS1", "S58", "ITS2", "LSU", "tab")) %>% 
  select(-tab)

#892 sequences have no ITS2 (they most stop after ITS1)
ITSx_pos %>% 
  filter(grepl("Not found", ITS2)) %>% 
  nrow()

ITS2_pos <- 
ITSx_pos %>% 
  filter(!grepl("Not found", ITS2)) %>% 
  mutate(ITS2 = gsub("ITS2: ", "", ITS2)) %>%
  tidyr::separate(ITS2, into = c("ITS2_start", "ITS2_end"), sep = "-") %>%
  mutate(across(starts_with("ITS"), as.numeric)) %>%
  mutate(ITS2_len = ITS2_end - ITS2_start) %>% 
  mutate(ID = gsub("(.+?);tax=.+", "\\1", newSeqname)) %>%
  dplyr::rename(ITSx_newseqname = newSeqname) %>% 
  left_join(select(ungroup(its_global_newseqname), ID, newSeqname)) %>% 
  filter(!is.na(newSeqname)) %>% 
  select(-ITSx_newseqname) %>% 
  dplyr::relocate(newSeqname, .before = NULL)
  
```


```{r}
ITS_global_gbif <- readDNAStringSet(here("Data", "Plant_ITS_db", "its_global_gbif.fa"))

#check if all sequence names are found 
Missing_seq <- which(!ITS2_pos$newSeqname %in% names(ITS_global_gbif))

#check if sequences from ITS_global are missing
sum(!names(ITS_global_gbif) %in% ITS2_pos$newSeqname)

#16272 Sequences are missing. Most of these are in the file ITSx_no_detections.fasta 
readDNAStringSet(here("Data", "Plant_ITS_db", "ITSX_out", "ITSx_no_detections.fasta")) %>% 
  length()

```

```{r}

#get cumulative size distribution
ITS2_pos %>% 
  group_by(ITS2_len) %>% 
  summarise(n_len = n()) %>% 
  arrange(ITS2_len) %>% 
  mutate(cum_prct = cumsum(n_len/sum(n_len))) %>% 
  ggplot(aes(x = ITS2_len,y = cum_prct))+
  geom_line()+
  scale_x_continuous(limits = c(0,1000))+
  geom_vline(xintercept = c(50, 500))

#we filter sequences <100 and above 300 bps
ITS2_pos_f <- 
  ITS2_pos %>% 
  filter(ITS2_len >= 100) %>% 
  filter(ITS2_len <= 500)

nrow(ITS2_pos) - nrow(ITS2_pos_f)
```


```{r}
#subset DB and make sure sequence order matches
ITS_global_ITS2 <- ITS_global_gbif[ITS2_pos_f$newSeqname]

all(names(ITS_global_ITS2) == ITS2_pos_f$newSeqname)

#subset sequences for only ITS2
ITS_global_ITS2 <- subseq(ITS_global_ITS2, ITS2_pos_f$ITS2_start, ITS2_pos_f$ITS2_end)

#export
writeXStringSet(ITS_global_ITS2, here("Data", "Plant_ITS_db", "ITS_global_ITS2.fasta"), format = "fasta")

#read
ITS_global_ITS2 <- readDNAStringSet(here("Data", "Plant_ITS_db", "ITS_global_ITS2.fasta"))
```

```{r}
system2( "/Applications/vsearch/bin/vsearch",
         args= c("--cluster_fast", 
                 shQuote(here("Data", "Plant_ITS_db", "ITS_global_ITS2.fasta")),
                 "--id 1.0",
                 "--uc",
                 shQuote(here("Data", "Plant_ITS_db", "ITS_global_clusters.uc")),
                 "--threads 6"
                 )
)
```


```{r}
ID_seq <- read_tsv(here("Data", "Plant_ITS_db", "ITS_global_clusters.uc"))

colnames(ID_seq) <- c(
  "Type", "ClusterNumber", "SequenceLength", "PercentageIdentity",
  "Strand", "Unused1", "Unused2", "Unused3", "QueryLabel", "TargetLabel"
)

# Extract the sequence ID 
ID_seq <- ID_seq %>%
  mutate(
    ID = str_extract(QueryLabel, "^[^;]+"),
  )

its_global_ID_to_tax <- 
  its_global_newname %>% 
  select(ID, starts_with("gbif")) %>% 
  dplyr::rename_all(~gsub("gbif_", "", .))

# Join the taxonomy 
ID_seq <- ID_seq %>%
  left_join(its_global_ID_to_tax) 
```

```{r}


# Assume 'ID_seq' is your dataframe with columns:
# Type, ClusterNumber, ID, species, genus, family

# Ensure we have unique combinations of ClusterNumber, ID, and species
cluster_species <- ID_seq %>%
  select(ClusterNumber, ID, species, genus, family) %>%
  distinct()

# Get the clusters each species appears in
species_clusters <- cluster_species %>%
  select(ClusterNumber, species) %>%
  distinct()

# Create a mapping of clusters to species
cluster_to_species <- species_clusters

# For each species, find all other species in the same clusters
species_cooccurrence <- species_clusters %>%
  inner_join(cluster_to_species, by = "ClusterNumber",  relationship = "many-to-many") %>%
  filter(species.x != species.y) %>%  # Exclude self-matches
  group_by(species.x) %>%
  summarise(
    CooccurringSpecies = paste(unique(species.y), collapse = "|"),
    CooccurringSpeciesCount = n_distinct(species.y)
  ) %>%
  dplyr::rename(species = species.x)

# Get all unique species
all_species <- cluster_species %>% select(species) %>% distinct()

# Left join to include species with no co-occurrences
species_cooccurrence_full <- all_species %>%
  left_join(species_cooccurrence, by = "species") %>%
  mutate(
    CooccurringSpecies = ifelse(is.na(CooccurringSpecies), "", CooccurringSpecies),
    CooccurringSpeciesCount = ifelse(is.na(CooccurringSpeciesCount), 0L, CooccurringSpeciesCount)
  )

```


check for Swiss species
```{r}
InfoFlora_check <- 
  InfoFlora_gbif %>% 
  distinct() %>% 
  mutate(in_ref_db = species %in% ID_seq$species) %>% 
  left_join(species_cooccurrence_full) %>% 
  filter(CooccurringSpecies != "") %>% 
  separate_rows(CooccurringSpecies, sep = "\\|") %>%
mutate(
    IsInSpeciesList = CooccurringSpecies %in% species
  ) %>% 
  group_by(species)  %>% 
  summarise(
    CooccurringSpeciesCount = unique(CooccurringSpeciesCount),
    AnyMatchingSpecies = any(IsInSpeciesList),
    cooccurrence_with_list = paste(unique(CooccurringSpecies[IsInSpeciesList]), collapse = "|"),
    CooccurringSpecies = paste(CooccurringSpecies, collapse = "|"),
    .groups = 'drop'
  )
  
InfoFlora_check %>% 
  filter(AnyMatchingSpecies) %>% select(species, cooccurrence_with_list) %>% View()
```

```{r}
Beaver_check <- 
Beaver_plants_gbif %>% 
  select(-verbatim_name) %>% 
  distinct() %>% 
  mutate(in_ref_db = species %in% ID_seq$species) %>% 
  left_join(species_cooccurrence_full) %>% 
  filter(CooccurringSpecies != "") %>% 
  separate_rows(CooccurringSpecies, sep = "\\|") %>%
mutate(
    IsInSpeciesList = CooccurringSpecies %in% species
  ) %>% 
  group_by(species)  %>% 
  summarise(
    CooccurringSpeciesCount = unique(CooccurringSpeciesCount),
    AnyMatchingSpecies = any(IsInSpeciesList),
    cooccurrence_with_list = paste(unique(CooccurringSpecies[IsInSpeciesList]), collapse = "|"),
    CooccurringSpecies = paste(CooccurringSpecies, collapse = "|"),
    .groups = 'drop'
  )
  
mutate(
    cooccurrence_with_list = ifelse(is.na(cooccurrence_with_list), "", cooccurrence_with_list),
    AnyMatchingSpecies = ifelse(is.na(AnyMatchingSpecies), FALSE, AnyMatchingSpecies)
  )
  
Beaver_check %>% 
  filter(AnyMatchingSpecies) %>% select(species, cooccurrence_with_list)
  filter(cooccurence_beaver)
```




#comapre refDBs

```{r}
Beaver_plants <- read_xlsx(here("Data", "data_vegetation.xlsx"), sheet = "pivot_data")

Beaver_plant_table <- 
  select(Beaver_plants, c(1,57:438)) %>% 
  group_by(Site) %>% 
  mutate(across(everything(), ~ifelse(is.na(.), 0, .))) %>% 
  summarise(across(everything(), sum))
  

Beaver_plants_tax <- tibble(species = colnames(Beaver_plant_table)[-1])

Beaver_plants_gbif <- 
rgbif::name_backbone_checklist(Beaver_plants_tax, kingdom = "Plantae")

Beaver_plants_gbif <- 
Beaver_plants_gbif %>% 
  select(verbatim_name, kingdom, phylum, class, order, family, genus, species)
```

```{r}
InfoFlora <- read_xlsx(here("Data", "Checklist_2017_simple_version_20230503.xlsx"), skip = 5)

InfoFlora <- 
  InfoFlora %>% 
  mutate(name = `Nom du taxon`,
         family = Famille)

InfoFlora_gbif <- 
  name_backbone_checklist(InfoFlora)

InfoFlora_gbif <-
  InfoFlora_gbif %>% 
  select(phylum, class, order, family, genus, species) %>% 
  distinct()

```
I complete the check-list with occurrence data from GBIF

GBIF.org (25 September 2024) GBIF Occurrence Download  https://doi.org/10.15468/dl.8emdev

```{r}
Swiss_gbif <- read_tsv(here("Data", "0030105-240906103802322.csv"))

Swiss_gbif <- 
Swiss_gbif %>% 
  select(phylum, class, order, family, genus, species, iucnRedListCategory, numberOfOccurrences) 

Swiss_gbif <-  
Swiss_gbif %>% 
  group_by(genus) %>% 
  filter(
        !is.na(species) | all(is.na(species))
      ) %>% 
  group_by(phylum, class, order, family, genus, species) %>% 
  reframe(iucnRedListCategory = unique(iucnRedListCategory[!is.na(iucnRedListCategory)]),
            numberOfOccurrences = sum(numberOfOccurrences)) %>% 
  mutate(inGbif = TRUE) %>% 
  group_by(genus, species) %>% 
  mutate(iucnRedListCategory = case_when(n() > 1 ~ NA_character_, 
                                         TRUE ~ iucnRedListCategory)) %>% 
  distinct()

Swiss_gbif %>% 
  group_by(species) %>% 
  filter(n() > 1) %>% 
  filter(!is.na(species))

#InfoFlora_gbif <- 
InfoFlora_gbif %>% full_join(Swiss_gbif)

write_tsv(InfoFlora_gbif, here("Data", "InfoFlora_gbif.txt"))
```


```{r}

PlantITS_tax_gbif_out <- read_tsv(here("Data", "PLANiTS_29-03-2020", "PlantITS_tax_gbif.txt"))

species_venn_df <- 
  Unite_tax_gbif_out %>% 
  filter(kingdom == "Plantae") %>% 
  select(species) %>% 
  mutate(DB = "Unite") %>% 
  bind_rows(.,tibble(species = PlantITS_tax_gbif_out$species, DB = "plantITS")) %>% 
  bind_rows(.,tibble(species = InfoFlora_gbif$species, DB = "InfoFlora")) %>% 
  bind_rows(.,tibble(species = Beaver_plants_gbif$species, DB = "Beaver")) %>% 
  filter(!is.na(species)) %>% 
  distinct() %>%
  mutate(species2 = species) %>% 
  pivot_wider(names_from = DB, values_from = species2) %>% 
  mutate_at(-1, ~!is.na(.)) 

species_venn <- 
  species_venn_df %>% 
  filter(InfoFlora | Beaver) %>% 
  select(-InfoFlora) %>% 
  select(-species) %>% 
  euler(shape = "ellipse") 

plot(species_venn, legend = TRUE, quantities = TRUE)
```

```{r}
species_venn$original.values
```
There are 30 species in the beaver dataset which are not in either Unite nor plantITS refDB

There are 4 species which are only in plantITS

```{r}
genus_venn <- 
  Unite_tax_gbif_out %>% 
  filter(kingdom == "Plantae") %>% 
  select(genus) %>% 
  mutate(DB = "Unite") %>% 
  bind_rows(.,tibble(genus = PlantITS_tax_gbif_out$genus, DB = "plantITS")) %>% 
  bind_rows(.,tibble(genus = InfoFlora_gbif$genus, DB = "InfoFlora")) %>% 
  bind_rows(.,tibble(genus = Beaver_plants_gbif$genus, DB = "Beaver")) %>% 
  filter(!is.na(genus)) %>% 
  distinct() %>% 
  mutate(genus2 = genus) %>% 
  pivot_wider(names_from = DB, values_from = genus2) %>% 
  mutate_at(-1, ~!is.na(.)) %>% 
  select(-genus) %>% 
 # filter(InfoFlora | Beaver) %>% 
  euler(shape = "ellipse") 

plot(genus_venn, legend = TRUE, quantities = TRUE)
```

```{r}
genus_venn$original.values %>% `[`(grepl("Beaver", names(.)))
```
all genera are in refDB, 3 only in PlantITS