---
title: "Plant ITS Analysis"
author: "Fabian Roger"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, message = FALSE}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(readr)
library(here)
library(stringr)
library(ggplot2)
library(ShortRead)
library(dada2)
library(ggrepel)
library(phyloseq)


# Load helper functions
source(here("Scripts", "Seq_analysis_helper.R"))
```

```{r}
sessionInfo()
```

# Set-Up
This script follows the [DADA2 ITS Tutorial (1.8)](https://benjjneb.github.io/dada2/ITS_workflow.html).
The data are demultiplexed (see `Demultiplexing.Rmd`)
The original article can be found [here](http://rdcu.be/ipGh)

The analysis proceeds through read filtering, error learning, denoising, merging, chimera removal, taxonomic assignment, and clustering.


## Setup Directories and Paths
Set up paths for input and output data and create necessary directories.

```{r}
primer <- "Plants"
trimmed_path <- here("Data", primer, "demultiplexed")
filter_path <- here("Data", primer, "filtered")
dir.create(filter_path, showWarnings = FALSE)
dir.create(here("Data", primer, "out_intermediate"))
dir.create(here("Data", primer, "out_final"))
```

## Filter and Prepare Reads
Prepare and sort forward and reverse read files. Extract sample names for downstream processing.

```{r, "sort fwrd and rev", cache=TRUE}
# List forward and reverse read files
fnFs <- sort(list.files(trimmed_path, pattern = "_R1.fastq.gz", full.names = TRUE))
fnRs <- sort(list.files(trimmed_path, pattern = "_R2.fastq.gz", full.names = TRUE))

# Extract sample names from filenames
sample.names <- sapply(fnFs, function(x) gsub(paste(".+", primer, "/demultiplexed/(.+)_R1.fastq.gz", sep = ""), "\\1", x), USE.NAMES = FALSE)
```

## Check Read Lengths
Check read length distributions to ensure read quality.

```{r}
# Check read length distributions for quality
readsR1 <- ShortRead::readFastq(fnFs[1]) %>%
  ShortRead::sread() %>%
  width() %>%
  table() %>%
  as.data.frame() %>%
  dplyr::rename(Length = 1) %>%
  mutate(Direction = "Forward")

readsR2 <- ShortRead::readFastq(fnRs[1]) %>%
  ShortRead::sread() %>%
  width() %>%
  table() %>%
  as.data.frame() %>%
  dplyr::rename(Length = 1) %>%
  mutate(Direction = "Reverse")

bind_rows(readsR1, readsR2) %>%
  mutate(Length = as.numeric(as.character(Length))) %>% 
  ggplot(aes(x = Length, y = Freq, color = Direction)) +
  geom_line(alpha = 0.5) +
  theme_bw() +
  facet_wrap(~Direction) +
  scale_x_continuous(limits = c(0, 500))
```

## Quality Profiles
Visualize the quality profile of the forward and reverse reads to determine trimming parameters.

```{r, "plot quality of reads", cache=TRUE}
# plot one forward and one reverse read in the rapport
plotQualityProfile(fnFs[1])+
  scale_y_continuous(limits = c(0,45))+
  scale_x_continuous(breaks = seq(0,300, 20))+
  geom_hline(yintercept = 30, colour = "grey", linetype = "dashed")+
  geom_hline(yintercept = 20, colour = "grey", linetype = "dashed")+
  labs(title = paste(sample.names[1], "foward", sep = " - " ))

plotQualityProfile(fnRs[1])+
  scale_y_continuous(limits = c(0,45))+
  scale_x_continuous(breaks = seq(0,300, 20))+
  geom_hline(yintercept = 30, colour = "grey", linetype = "dashed")+
  geom_hline(yintercept = 20, colour = "grey", linetype = "dashed")+
  labs(title = paste(sample.names[1], "reverse", sep = " - "))
```

The quality of the forwards reads is very high, throughout the full length of the read.  The quality of the reverse read is >= Q30 up to ~200 bps

## Merge Read Lengths
Merge paired reads to assess overlap and read length distribution for optimal trimming decisions.

```{r}
# Merge the first file's forward and reverse reads using vsearch for length distribution analysis
 system(
  paste(
    "/Applications/vsearch/bin/vsearch --fastq_mergepairs ", 
    gsub("My Drive", "My\\\\ Drive", fnFs[1]), 
    " --reverse ", gsub("My Drive", "My\\\\ Drive", fnRs[1]), 
    " --fastaout", gsub("My Drive", "My\\\\ Drive", here("Data", primer, "test_merge.fasta")), 
    "--fastq_allowmergestagger",
    "--fastq_qmax 45"
    ), intern = TRUE
  )

# Read the merged sequences into R
file_merged <- readDNAStringSet(here("Data", primer, "test_merge.fasta"))

# Delete the temporary merged file after use
file.remove(here("Data", primer, "test_merge.fasta"))

# Create a table of sequence lengths and plot the distribution
table(width(file_merged)) %>%
  as.data.frame() %>%
  mutate(across(c("Var1", "Freq"), function(x) as.numeric(as.character(x)))) %>%  # Ensure columns are numeric
  arrange(desc(Freq)) %>%  # Sort by frequency
  mutate(
    prct = Freq / sum(Freq) * 100,  # Calculate the percentage of each length
    rank = 1:n(),  # Create a rank column
    Label = case_when(  # Add labels for the top 5 most frequent lengths
      rank < 6 ~ paste(Var1, "(", round(prct, 1), " %)", sep = ""),
      TRUE ~ NA_character_
    )
  ) %>%
  ggplot(aes(x = Var1, y = Freq)) +
  geom_line(colour = "darkgrey", alpha = 0.5) +  # Line plot for length distribution
  geom_text_repel(aes(label = Label)) +  # Annotate top labels
  geom_point(data = ~filter(.x, !is.na(Label)), colour = "red", size = 0.5) +  # Highlight labeled points
  theme_bw() +
  scale_x_continuous(limits = c(0, 500))  # Limit x-axis range
```

We sequenced with 2 x 300 bp.
the merged fragments are 471 bp long

The trimmed forward and reverse reads are 279 and 273 bps long, respectively.

Thus we have a minimum overlap of `276+275-471` = 80 bps.

given the lower quality of the reverse reads, I trimm them down to 230, leaving an overlap of 35 bps.

## Set Read Parameters
Set trimming parameters based on observed read quality and overlap.

```{r}
# Set parameters for fragment and read lengths
frag_len <- 471
fwrd_len <- 276
rev_len <- 275
fwrd_trim <- 276
rev_trim <- 230

# Create a plot showing the trimmed regions and overlap
ggplot(data = tibble(x = c(1, frag_len), y = c(1, 5)), aes(x = x, y = y)) +
  geom_blank() +  # Base blank plot to set up the axes
  geom_segment(x = 0, xend = frag_len, y = 2, yend = 2) +  # Full fragment line
  geom_segment(x = 0, xend = fwrd_len, y = 3, yend = 3) +  # Full forward read
  geom_segment(x = frag_len - rev_len, xend = frag_len, y = 3.5, yend = 3.5) +  # Full reverse read
  geom_segment(x = fwrd_trim, xend = fwrd_len, y = 3, yend = 3,  # Trimmed forward segment
               colour = "red", size = 2, alpha = 0.2) +
  geom_segment(x = frag_len - rev_trim, xend = frag_len - rev_len,  # Trimmed reverse segment
               y = 3.5, yend = 3.5, colour = "red", size = 3, alpha = 0.2) +
  geom_segment(x = frag_len - rev_trim, xend = fwrd_trim, y = 2,  # Overlap region
               yend = 2, colour = "green", size = 3, alpha = 0.2) +
  geom_text(x = ((frag_len - rev_trim) + fwrd_trim) / 2, y = 2.25,  # Label showing overlap length
            label = paste(as.character(fwrd_trim + rev_trim - frag_len), "bp"),
            colour = "blue", size = 5) +
  theme_void() +  # Removes Fungikground and axes for clarity
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +  # Adjust x-axis text orientation
  scale_x_continuous(breaks = seq(0, frag_len, 12))  # X-axis ticks
```

The filtering parameters we’ll use are stringent as the quality is very high and we have a lot of reads. 
I allow 2 expected errors on the reverse read as we cannot trim to much because of the long amplicon size. 

+ maxN=0 (DADA2 requires no Ns)
+ truncQ=2 
+ maxEE=c(1,2)

The `maxEE` parameter sets the maximum number of “expected errors” allowed in a read. 

We use the `fastqPairedFilter` function to jointly filter the forward and reverse reads.

#DADA2 Pipeline

## Filter and Trim Reads
Filter and trim the reads based on quality, expected errors, and length settings. Write filtered read statistics to file.

```{r}
# Create file paths for the filtered forward and reverse reads, stored in the filtered/ subdirectory
filtFs <- file.path(here("Data", primer, "filtered"), paste0(sample.names, "_F_filt.fastq.gz"))  # Path for forward filtered reads
filtRs <- file.path(here("Data", primer, "filtered"), paste0(sample.names, "_R_filt.fastq.gz"))  # Path for reverse filtered reads

# Name the paths
names(filtFs) <- sample.names
names(filtRs) <- sample.names
```

```{r}
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, 
                     truncLen = c(fwrd_trim, rev_trim),
                     maxN=0,
                     maxEE=c(1,1),
                     truncQ=2,
                     rm.phix=TRUE,
                     compress=TRUE, multithread=6) 
out_df <- 
out %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = "Sample_name") %>% 
  mutate(Sample_name = gsub("_R1.+", "", Sample_name)) %>% 
  mutate(prct = signif(reads.out / reads.in * 100, 2)) %>% 
  mutate(Sample_name = gsub("(.+?)_S.+.fastq.gz", "\\1", Sample_name))

write_tsv(out_df, here("Data", primer, "out_intermediate", paste(primer, "ee_filter_stats.txt", sep = "_")))

out_df 
```

## Learn Errors
Estimate error rates from the filtered reads to inform the denoising step.

```{r}
# remove files whith no reads left after filtering
files_passed_ee <- 
  which(out[,2] > 0) 
  
filtFs <- filtFs[files_passed_ee]
filtRs <- filtRs[files_passed_ee]
```

```{r}
errF <- learnErrors(
  filtFs,
  multithread = TRUE,
  nbases = 1e9,
  errorEstimationFunction = loessErrfun_mod4,
  verbose = TRUE
)

errR <- learnErrors(
  filtRs,
  multithread = TRUE,
  nbases = 1e9,
  errorEstimationFunction = loessErrfun_mod4,
  verbose = TRUE
)
```

```{r, cache=TRUE}
plotErrors(errF, nominalQ=TRUE)

plotErrors(errR, nominalQ=TRUE)
```

## Denoise
Apply DADA2 denoising to forward and reverse reads to identify unique sequences.

```{r}
filt <- list.files(filter_path, full.names = T)

filtFs <- filt[grepl("*_F_filt.fastq.gz", filt)]
names(filtFs) <- gsub(".+filtered/(.+?)_F.+", "\\1", filtFs)

filtRs <- filt[grepl("*_R_filt.fastq.gz", filt)]
names(filtRs) <- gsub(".+filtered/(.+?)_F.+", "\\1", filtRs)
```

```{r}
dadaFs <- dada(filtFs, err=errF, multithread=TRUE)
dadaRs <- dada(filtRs, err=errR, multithread=TRUE)
```

```{r}

denoise_stat <- 
sapply(dadaFs, function(x) sum(getUniques((x)))) %>% 
  data.frame(denoised = .) %>% 
  tibble::rownames_to_column(var = "Sample_name") %>% 
  left_join(select(out_df, Sample_name, reads.out)) %>% 
  mutate(prct = round(denoised / reads.out*100, 2))
  
write_tsv(denoise_stat, here("Data", primer, "out_intermediate", paste(primer, "denoise_stat.txt", sep = "_")))

denoise_stat
```

## Merge
Merge the forward and reverse reads and create a summary of merged reads for each sample.

```{r}

filtFs <- filtFs[names(dadaFs)]
filtRs <- filtRs[names(dadaRs)]

mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose=TRUE)

merge_stat <- 
lapply(mergers, function(x) {sum(x$abundance)}) %>% 
  as_tibble() %>% 
  pivot_longer(everything(),
    names_to = "Sample_name",
               values_to = "merged.out") %>% 
  left_join(select(out_df, Sample_name, reads.out)) %>% 
  mutate(prct = round(merged.out/reads.out * 100, 2))

merge_stat %>% 
write_tsv(here("Data", primer, "out_intermediate", paste(primer, "merge_stat.txt", sep = "_")))

merge_stat

```

## Construct Sequence Table
Generate the sequence table from the merged reads.

```{r, "Seqeunce table"}
seqtab <- makeSequenceTable(mergers)
dim(seqtab)
```

## Chimera Removal
Remove chimeric sequences from the sequence table and save chimera statistics to file.

```{r, "Chimera removal", cache=TRUE}

seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread=TRUE, verbose=TRUE)

write_rds(seqtab.nochim, here("Data", primer, "out_intermediate", paste(primer, "seqtab.nochim.rds", sep = "_")))
seqtab.nochim <- read_rds(here("Data", primer, "out_intermediate", paste(primer, "seqtab.nochim.rds", sep = "_")))

dim(seqtab.nochim)
```

```{r}
chimera_stat <- 
tibble(Sample_name = rownames(seqtab),
       before = rowSums(seqtab),
       after = rowSums(seqtab.nochim),
       prct = round(after/before*100,2))

chimera_stat %>% 
write_tsv(here("Data", primer,  "out_intermediate", paste(primer, "Chimera_stat.txt", sep = "_"))) 

chimera_stat
```

percentage of Chimeric reads

```{r}
(1-(sum(seqtab.nochim)/sum(seqtab)))*100
```

percentage of Chimeric ASVs

```{r}
(1-(ncol(seqtab.nochim)/ncol(seqtab)))*100
```

## Length Filtering
Visualize the length distribution of ASVs to confirm expected fragment size.

```{r}
data.frame(table(nchar(colnames(seqtab.nochim)))) %>% 
  mutate(across(c("Var1", "Freq"), function(x) as.numeric(as.character(x)))) %>%  # Ensure columns are numeric
  arrange(desc(Freq)) %>%  # Sort by frequency
  mutate(
    prct = Freq / sum(Freq) * 100,  # Calculate the percentage of each length
    rank = 1:n(),  # Create a rank column
    Label = case_when(  # Add labels for the top 5 most frequent lengths
      rank < 6 ~ paste(Var1, "(", round(prct, 1), " %)", sep = ""),
      TRUE ~ NA_character_
    )
  ) %>%
  ggplot(aes(x = Var1, y = Freq)) +
  geom_line(colour = "darkgrey", alpha = 0.5) +  # Line plot for length distribution
  geom_text_repel(aes(label = Label)) +  # Annotate top labels
  geom_point(data = ~filter(.x, !is.na(Label)), colour = "red", size = 0.5) +  # Highlight labeled points
  theme_bw() +
  scale_x_continuous(limits = c(300, 500))  # Limit x-axis range
```

We have a large lengths variation as expected for ITS2. We do not apply lengths filtering.

```{r}
# sum(nchar(colnames(seqtab.nochim)) > 257)
# sum(nchar(colnames(seqtab.nochim)) > 257)/ncol(seqtab.nochim)*100
# 
# seqtab.nochim <- seqtab.nochim[,nchar(colnames(seqtab.nochim)) <= 257]
# dim(seqtab.nochim)
```


## Export Raw ASV and Fasta Files
Export the non-chimeric ASV sequences to a FASTA file and save the ASV abundance table as a TSV.


```{r}
# Convert the columns of the sequence table (ASVs) to a DNAStringSet object for export
seqs <- DNAStringSet(colnames(seqtab.nochim))

# Assign sequence names in the format: seq_001, seq_002, etc.
names(seqs) <- paste("seq", 
                     formatC(1:length(seqs), flag = "0", width = nchar(length(seqs))), 
                     sep = "_")

# Export the ASVs as a FASTA file, prefaced with the primer name
writeXStringSet(seqs, here("Data", primer, "out_intermediate", paste(primer, "ASVs.fasta", sep = "_")), format = "fasta")

# Prepare the ASV table for export as a TSV
ASV <- seqtab.nochim %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = "Sample") %>% 
  `colnames<-`(c("Sample", names(seqs)))

# Export the ASV table as a TSV file, prefaced with the primer name
write_tsv(ASV, here("Data", primer, "out_intermediate", paste(primer, "ASVs.txt", sep = "_")))

# Code to read the TSV (if needed for future processing)
# ASV <- read_tsv(here("Data", primer, "out_intermediate", paste(primer, "ASVs.txt", sep = "_")))
```

As the reference databases for Plants are very complete for Swiss plants, we do not apply any clustering but will rather merge the ASVs by taxonomic assignment


# ITSx
This section runs the ITSx tool to extract the ITS2 region from the ASV sequences

```{r}
cores = 6

# Copy ASV FASTA file to the Downloads directory to avoid path issues
system2(command = "cp", 
        args = c(shQuote(here("Data", primer, "out_intermediate", paste(primer, "ASVs.fasta", sep = "_"))),
                 "~/Downloads/"))

# Create a temporary output directory for ITSx results
dir.create("~/Downloads/ITSX_out")

# Run ITSx for ITS2 region extraction
system(paste("source ~/.bash_profile && /Applications/ITSx_1.1.3/ITSx",
             "ITSx -i ",
             paste("~/Downloads/", paste(primer, "ASVs.fasta ", sep = "_"), sep = ""),
             "-o ~/Downloads/ITSX_out/ITSx ", 
             "-t all ",
             "--allow_single_domain '1e-5' ",
             "--complement F ",
             "--save_regions ITS2 ",
             "--cpu ", 
             cores
             ))

# Move ITSx results back to the project directory
system2(command = "mv", 
        args = c("~/Downloads/ITSX_out",
                 shQuote(here("Data", primer, "out_intermediate"))))

# Clean up: Remove temporary ASV FASTA file from Downloads
file.remove(paste("~/Downloads/", paste(primer, "ASVs.fasta", sep = "_"), sep = ""))
```


read in sequences with ITS2 domains
```{r}
# Read the ITS2 region sequences from the ITSx output
seq_ITS2 <- readDNAStringSet(here("Data", primer, "out_intermediate", "ITSX_out", "ITSx.ITS2.fasta"))

# foramt sequence names to match input
names(seq_ITS2) <- gsub("(seq_\\d+)\\|.+", "\\1", names(seq_ITS2))

# Check how many original sequences are not in the ITS2 set
length(which(!names(seqs) %in% names(seq_ITS2)))

# Export the ITS2 sequences as a new FASTA file, prefaced with the primer name
writeXStringSet(seq_ITS2, here("Data", primer, "out_intermediate", paste(primer, "ASVs_ITS2.fasta", sep = "_")), format = "fasta")

# Uncomment this line if needed for reading the ITS2 sequences back in the future
# seq_ITS2 <- readDNAStringSet(here("Data", primer, "out_intermediate", paste(primer, "ASVs_ITS2.fasta", sep = "_")))
```

# Taxonomy Assignment

## blast
I use the a curated Plant ITS database for taxonomic assignment. See reference below. I harmonized the taxonomy against the gbif backbone. I use blast for exact matching. 

https://doi.org/10.15156/BIO/2959341


```{r}

# Build an index of the reference database
system2( "/Applications/vsearch/bin/vsearch",
         args= c("--makeudb_usearch", 
                 shQuote(here("Data", primer, "in_files", "Plant_ITS_db", "its_global_gbif_sp.fa")),
                 "--output",
                 shQuote(here("Data", primer, "in_files", "Plant_ITS_db", "its_global_gbif_sp.udb"))
                 ))
 
```

optimising 
```{r}
# sample 100 sequence at random 
test <- readDNAStringSet(here("Data", primer, "out_intermediate", "Plants_ASVs_ITS2.fasta"))
set.seed(7890)
test <- test[sample(1:length(test), 1000)]
writeXStringSet(test, here("Data", primer, "out_intermediate", "Plants_ASVs_ITS2_sample.fasta"))


library(lubridate)

# Create a data frame to store results
results_df <- data.frame(
  maxrejects = integer(),
  run_time = numeric(),
  num_hits = integer(),
  median_hits = integer(),
  stringsAsFactors = FALSE
)

# Iterate over different maxrejects values
for (MR in c(2, 4, 8, 16, 32, 64, 128, 256, 512)) {
  
  # Start timing
  start_time <- Sys.time()
  
  # Run the vsearch command
  system2("/Applications/vsearch/bin/vsearch",
          args = c("--usearch_global", 
                   shQuote(here("Data", primer, "out_intermediate", "Plants_ASVs_ITS2_sample.fasta")),
                   "--db",
                   shQuote(here("Data", primer, "in_files", "Plant_ITS_db", "its_global_gbif_sp.udb")),
                   "--id .90",
                   "--maxaccepts 0",
                   paste("--maxrejects", MR),
                   "--blast6out",
                   shQuote(here("Data", primer, "out_intermediate", paste("ITS2_Plants_exact_MR", MR, ".txt", sep = ""))),
                   "--threads 6"
          ))
  
  # End timing
  end_time <- Sys.time()
  run_time <- as.numeric(difftime(end_time, start_time, units = "secs"))
  
  # Read the output file
  blast_output_path <- here("Data", primer, "out_intermediate", paste("ITS2_Plants_exact_MR", MR, ".txt", sep = ""))

  blast_output <- read_tsv(blast_output_path, col_names = c(
      "query_id", "subject_id", "%identity",
      "alignment_length", "mismatches", "gap_opens",
      "q_start", "q_end", "s_start", "s_end", 
      "evalue", "bit_score"))
    
    # Store the total number of hits
    num_hits <- nrow(blast_output)
    
    # Store the median number of hits
    median_hits <- blast_output %>% 
      group_by(query_id) %>% 
      summarise(n = n()) %>% 
      pull(n) %>% 
      median()
  
  # Add results to the data frame
  results_df <- results_df %>% 
    add_row(maxrejects = MR, run_time = run_time, num_hits = num_hits, median_hits = median_hits)
}

write_csv(results_df, here("Data", primer, "out_intermediate", "blast_parameter_search_results_id90.csv"))
```


```{r}
results_df95 <- read_csv(here("Data", primer, "out_intermediate", "blast_parameter_search_results_id95.csv"))

results_df90 <- read_csv(here("Data", primer, "out_intermediate", "blast_parameter_search_results_id90.csv"))

results_df95 %>% 
  mutate(id = "95") %>% 
  bind_rows(mutate(results_df90, id = "90")) %>% 
  pivot_longer(one_of("median_hits", "num_hits", "run_time"), names_to = "stat") %>% 
  ggplot(aes(x = maxrejects, y = value, colour = id))+
  geom_point()+
  facet_wrap(~stat, nrow = 1, scales = "free_y")

```
 We set maxrejects to 128 and maxaccepts to all to make sure we do not miss potential matches which could lead to incorrect assignments in the LCA taxonomy. 

```{r}
# Perform exact matching
system2( "/Applications/vsearch/bin/vsearch",
         args= c("--usearch_global", 
                 shQuote(here("Data", primer, "out_intermediate", "Plants_ASVs_ITS2.fasta")),
                 "--db",
                 shQuote(here("Data", primer, "in_files", "Plant_ITS_db", "its_global_gbif_sp.udb")),
                 "--id .95",
                 "--maxaccepts 0",
                 "--maxrejects 128",
                 "--blast6out",
                 shQuote(here("Data", primer, "out_intermediate", "ITS2_Plants_exact.txt")),
                 "--threads 6"
                 ))
```

```{r}
#read in seqname taxonomy
seqname_to_tax <- read_tsv(here("Data", "Plants", "in_files", "Plant_ITS_db", "its_global_gbif_tax.txt"))

#read in vsearch output
ITS_exact <- read_tsv(here("Data", primer, "out_intermediate", "ITS2_Plants_exact.txt"), col_names = F)

colnames(ITS_exact) <- c("query_id","subject_id","%identity","alignment_length",
                         "mismatches","gap_opens","q_start","q_end","s_start",
                         "s_end","evalue","bit_score")

ITS_exact %>% 
  select(1:3) %>% 
  left_join(seqname_to_tax, by = c("subject_id" = "newSeqname")) %>% 
  select(-subject_id) %>% 
  distinct() %>% 
  arrange(query_id) %>% 
  group_by(query_id) %>%
  filter(!is.na(species)) %>% 
  filter(n()>1)
```

filter with Swiss occurence list
```{r}
InfoFlora_gbif <- read_tsv(here("Data", primer, "in_files", "InfoFlora_gbif.txt")) %>% 
  mutate(in_reflist = TRUE)

ITS_exact_sel <- 
ITS_exact %>% 
  select(1:3) %>% 
  left_join(seqname_to_tax, by = c("subject_id" = "newSeqname")) %>% 
  select(-subject_id) %>% 
  distinct() %>% 
  left_join(InfoFlora_gbif) %>% 
  mutate(in_reflist = ifelse(is.na(in_reflist), FALSE, in_reflist))

# Function to apply the heuristic to each group
process_query <- function(data, key) {
  # data: dataframe for the current group (query_id)
  # key: list containing the group key (query_id)
  
  # Step 1: Check if any matches occur in Switzerland
  if (any(data$in_reflist == TRUE)) {
    # Keep only matches occurring in Switzerland
    data <- data %>% filter(in_reflist == TRUE)
  }
  # If no matches occur in Switzerland, keep all matches
  
  # Step 2: Keep matches with the highest %identity
  max_identity <- max(data$`%identity`)
  data <- data %>% filter(`%identity` == max_identity)
  
  # Step 3: Resolve ties with consensus taxonomy
  if (nrow(data) == 1) {
    # Only one best match remains
    return(data)
  } else {
    # Multiple matches with equal %identity
    # Create consensus taxonomy
    taxonomic_levels <- c("kingdom", "phylum", "class", "order", "family", "genus", "species")
    consensus <- data[1, ]  # Start with the first match
    
    for (level in taxonomic_levels) {
      level_values <- data[[level]]
      # Remove NAs
      non_na_values <- level_values[!is.na(level_values)]
      # Unique non-NA values
      unique_values <- unique(non_na_values)
      
      if (length(unique_values) == 1) {
        # All non-NA values are the same
        consensus[[level]] <- unique_values
      } else {
        # Disagreement at this level
        consensus[[level]] <- NA
      }
    }
    
    # Update `in_reflist` field
    consensus$in_reflist <- any(data$in_reflist == TRUE)
    # Update `%identity` field
    consensus$`%identity` <- max_identity
    
    return(consensus)
  }
}

# Apply the function to each group
ITS_exact_sel <- ITS_exact_sel %>%
  group_by(query_id) %>%
  group_modify(~ process_query(.x, .y)) %>%
  ungroup()

#export
ITS_exact_sel %>% 
  dplyr::rename(prct_id = '%identity',
                seq = query_id) %>% 
  write_tsv(here("Data", primer, "out_intermediate", paste(primer, "Tax_blast_LCA.txt", sep = "_")))
  


ITS_exact_sel %>% 
  ungroup %>% 
  dplyr::rename(prct_id = '%identity') %>% 
  arrange(prct_id) %>% 
  select(-query_id) %>% 
  group_by_all() %>% 
  summarise(n= n()) %>% 
  group_by(prct_id) %>% 
  summarise(n = n()) %>% 
  arrange(desc(prct_id)) %>% 
  mutate(cum_prct = round(cumsum(n) / sum(n) * 100,2)) %>% 
  ggplot(aes(y = prct_id, x = cum_prct))+
  geom_line()
```

##Sintax

what sequences have not been found? 
```{r}
unassigne_tax <- seq_ITS2[!names(seq_ITS2) %in% ITS_exact_sel$query_id]

writeXStringSet(unassigne_tax, here("Data", primer, "out_intermediate", paste(primer, "ASVs_ITS2_unassigned.fasta", sep = "_")))
```

I probabilistically assign the sequences that haven't been found against a curated version of the Unite DB utax_reference_dataset_all_04.04.2024.fasta


```{r}
system2( "/Applications/vsearch/bin/vsearch",
         args= c("--sintax", 
                 shQuote(here("Data", primer, "out_intermediate",
                              paste(primer, "ASVs_ITS2_unassigned.fasta", sep = "_"))),
                 paste("--db", shQuote(here("Data", "Plants", "in_files",
                                            "Unite_db_sintax_gbif_genus.fa")), sep = " "),
                 "--sintax_cutoff 0.1",
                 "-strand both",
                 "--tabbedout", 
                 shQuote(here("Data", primer, "out_intermediate", 
                              paste(primer, "ITS2_Plants_sintax.txt", sep = "_")))
                 )
)
```

```{r}
Sintax_taxa <- read_tsv(here("Data", primer, "out_intermediate", 
                              paste(primer, "ITS2_Plants_sintax.txt", sep = "_")), 
                        col_names = FALSE) %>% 
  select(1:2)

colnames(Sintax_taxa) <- c("seq", "tax")

#get unassigned
Sintax_na <- Sintax_taxa %>% filter(is.na(tax)) %>% select(seq)

Sintax_prep <- 
  Sintax_taxa %>% 
  filter(!is.na(tax)) %>% 
  mutate(tax = strsplit(tax, ",")) %>% 
  unnest(tax) %>% 
  group_by(seq) %>% 
  dplyr::slice(-1) %>% 
  mutate(taxonomy_name = gsub("\\w:(.+)\\(.+", "\\1", tax),
         taxonomy_level = gsub("(\\w):.+", "\\1", tax),
         taxonomy_conf = gsub(".+\\((.+)\\)", "\\1", tax)) %>% 
 left_join(tax_match_df)

#get table with all assignments (even the unlikely ones)
Sintax_taxa_table_full <- 
 Sintax_prep %>% 
  select(-tax, -taxonomy_conf, -taxonomy_level) %>% 
  pivot_wider(names_from = taxon, values_from = taxonomy_name) %>% 
  select("seq", "phylum", "class", "order", "family", "genus", "species") %>% 
  bind_rows(Sintax_na) %>% 
  arrange(seq)

#get table with probabilities for each assignment
Sintax_taxa_table_prob <- 
  Sintax_prep %>% 
  select(-tax, -taxonomy_name, -taxonomy_level) %>% 
  pivot_wider(names_from = taxon, values_from = taxonomy_conf) %>% 
  select("seq", "phylum", "class", "order", "family", "genus", "species") %>% 
  bind_rows(Sintax_na) %>% 
  arrange(seq)
```


```{r}
Sintax_taxa_80 <- Sintax_taxa_table_full
Sintax_taxa_80[,colnames(Sintax_taxa_80) %in% colnames(Sintax_taxa_table_prob)][Sintax_taxa_table_prob < 0.8] <- NA_character_ 

Sintax_taxa_80 %>% mutate(in_reflist = case_when(is.na(species) ~ NA,
                                                     TRUE ~ species %in% InfoFlora_gbif$species))
```

```{r}
Sintax_taxa_80 %>% 
  group_by(phylum) %>% 
 # filter(phylum == "Tracheophyta") %>%
  #group_by(family) %>%
  summarise(n = n())
```

##Join Taxonomy
### comment

Not sure what to do here. What are the sequences that don't have a 95% match with blast? Are they missing from the database? (True for fungi - but they should be excluded anyway, but what about the 110 plants?)

For now I only keep the sequences that have a blast assignment

## Merge and Aggregate ASV by Taxonomic Level
Use phyloseq to merge ASVs with the same species assignment.


```{r}
#define taxonomic aggregation level
tax_glom_level <- "species"

#read the blast LCA taxonomy
# replace missing species with genus sp. to agglomerate at genus level if species level is missing
Plant_blast_tax <- read_tsv(here("Data", primer, "out_intermediate", 
                           paste(primer, "Tax_blast_LCA.txt", sep = "_"))) %>% 
  mutate(species = case_when(
    is.na(species) & !is.na(genus) ~ paste(genus, "sp."),
    TRUE ~ species)) 

# Read the ASV abundance table
ASV <- read_tsv(here("Data", primer, "out_intermediate", paste(primer, "ASVs.txt", sep = "_")))

# exclude ASV sequences not asisgned in blast
ASV <- ASV %>% select(one_of("Sample", Plant_blast_tax$seq))

# Prepare ASV matrix and set sample names as row names
ASV_ps <- ASV[, -1]  # Remove the first column (sample identifiers)
rownames(ASV_ps) <- ASV$Sample

# Filter and prepare the taxonomic table 
Plant_blast_tax_ps <-
  Plant_blast_tax %>%
  filter(!is.na(!!sym(tax_glom_level))) %>%
  select(-seq, -prct_id, -in_reflist) %>%
  as.matrix()

# Set sequence names as row names in the taxonomic matrix
rownames(Plant_blast_tax_ps) <- filter(Plant_blast_tax, !is.na(!!sym(tax_glom_level)))$seq

# Create a phyloseq object using the ASV and taxonomic data
ps_obj <- phyloseq(otu_table(ASV_ps, taxa_are_rows = FALSE),
                   tax_table(Plant_blast_tax_ps))

# Aggregate ASVs at the selected level
ps_glom <- tax_glom(ps_obj, taxrank = tax_glom_level, NArm = FALSE)

# Convert the aggregated taxonomic table to a data frame
Plant_blast_tax_glom <- as.data.frame(tax_table(ps_glom)) %>%
  tibble::rownames_to_column("seq")

# Convert the aggregated ASV table to a data frame
ASW_glom <- as.data.frame(otu_table(ps_glom)) %>%
  tibble::rownames_to_column("Sample")

# Merge and adjust columns in the final ASV table
ASW_glom_j <- ASV %>%
  select(!one_of(rownames(Plant_blast_tax_ps))) %>%
  left_join(ASW_glom, .)

# Sort columns alphabetically for consistency
ASW_glom_j <- ASW_glom_j[sort(colnames(ASW_glom_j))]

# Combine the original taxonomic table with the merged table
Plant_blast_tax_glom_j <- Plant_blast_tax %>%
  select(-prct_id, -in_reflist) %>% 
  filter(!seq %in% rownames(Plant_blast_tax_ps)) %>%
  bind_rows(Plant_blast_tax_glom, .) %>%
  arrange(seq)

# Prepare a sequence table with just the representative sequences
seqs_glom <- seqs[Plant_blast_tax_glom_j$seq]

dim(ASV)
dim(ASW_glom_j)
```
# Plant in fungal primer

read in fungal data.

I read in the raw data to apply the same pipeline as to the Plant data above
```{r}
Fungi_tax <- read_tsv(here("Data", "Fungi", "out_intermediate", paste("Fungi", "Sintax_80.txt", sep = "_")))

Fungi_ASW <- read.table(here("Data", "Fungi", "out_intermediate", paste("Fungi", "ASV_swarm.txt", sep = "_")))

Fungi_seqs <-readDNAStringSet(here("Data", "Fungi", "out_intermediate", paste("Fungi", "ASV_swarm.fasta", sep = "_")))
```

subset for Plant species
```{r}
#get plant sequences
Plant_seqs <- Fungi_tax %>% filter(kingdom == "Plantae") %>% pull(seq)

#export plant sequences
Fungi_seqs[Plant_seqs] %>% 
  writeXStringSet(here("Data", primer, "out_intermediate", paste(primer, "ASV_fungi.fasta", sep = "_")))

```

## blast

```{r}
# Perform exact matching
system2( "/Applications/vsearch/bin/vsearch",
         args= c("--usearch_global", 
                 shQuote(here("Data", primer, "out_intermediate", 
                              paste(primer, "ASV_fungi.fasta", sep = "_"))),
                 "--db",
                 shQuote(here("Data", primer, "in_files", "Plant_ITS_db", "its_global_gbif_sp.udb")),
                 "--id .95",
                 "--maxaccepts 0",
                 "--maxrejects 128",
                 "--blast6out",
                 shQuote(here("Data", primer, "out_intermediate", "Plants_Fungi_exact.txt")),
                 "--threads 6"
                 ))
```

```{r}
#read in vsearch output
Plants_Fungi_exact <- read_tsv(here("Data", primer, "out_intermediate", "Plants_Fungi_exact.txt"), col_names = F)

colnames(Plants_Fungi_exact) <- c("query_id","subject_id","%identity","alignment_length",
                         "mismatches","gap_opens","q_start","q_end","s_start",
                         "s_end","evalue","bit_score")
```

assign best match with Swiss occurence list
```{r}

Plants_Fungi_sel <- 
Plants_Fungi_exact %>% 
  select(1:3) %>% 
  left_join(seqname_to_tax, by = c("subject_id" = "newSeqname")) %>% 
  select(-subject_id) %>% 
  distinct() %>% 
  left_join(InfoFlora_gbif) %>% 
  mutate(in_reflist = ifelse(is.na(in_reflist), FALSE, in_reflist))


# Apply selection function to each group
Plants_Fungi_sel <- Plants_Fungi_sel %>%
  group_by(query_id) %>%
  group_modify(~ process_query(.x, .y)) %>%
  ungroup()

#export
Plants_Fungi_sel %>% 
  dplyr::rename(prct_id = '%identity',
                seq = query_id) %>% 
  write_tsv(here("Data", primer, "out_intermediate", paste(primer, "Fungi_Tax_blast_LCA.txt", sep = "_")))


Plants_Fungi_sel %>% 
  ungroup %>% 
  dplyr::rename(prct_id = '%identity') %>% 
  arrange(prct_id) %>% 
  select(-query_id) %>% 
  group_by_all() %>% 
  summarise(n= n()) %>% 
  group_by(prct_id) %>% 
  summarise(n = n()) %>% 
  arrange(desc(prct_id)) %>% 
  mutate(cum_prct = round(cumsum(n) / sum(n) * 100,2)) %>% 
  ggplot(aes(y = prct_id, x = cum_prct))+
  geom_line()
```

```{r}
#define taxonomic aggregation level
tax_glom_level <- "species"

#read the blast LCA taxonomy
# replace missing species with genus sp. to agglomerate at genus level if species level is missing
Fungi_blast_tax <- read_tsv(here("Data", primer, "out_intermediate", paste(primer, "Fungi_Tax_blast_LCA.txt", sep = "_"))) %>% 
  mutate(species = case_when(
    is.na(species) & !is.na(genus) ~ paste(genus, "sp."),
    TRUE ~ species)) 

# Read the ASV abundance table
ASV <- Fungi_ASW

# exclude ASV sequences not asisgned in blast
ASV <- ASV %>% select(one_of("Sample", Fungi_blast_tax$seq))

# Prepare ASV matrix and set sample names as row names
ASV_ps <- ASV[, -1]  # Remove the first column (sample identifiers)
rownames(ASV_ps) <- ASV$Sample

# Filter and prepare the taxonomic table 
Fungi_blast_tax_ps <-
  Fungi_blast_tax %>%
  filter(!is.na(!!sym(tax_glom_level))) %>%
  select(-seq, -prct_id, -in_reflist) %>%
  as.matrix()

# Set sequence names as row names in the taxonomic matrix
rownames(Fungi_blast_tax_ps) <- filter(Fungi_blast_tax, !is.na(!!sym(tax_glom_level)))$seq

# Create a phyloseq object using the ASV and taxonomic data
ps_obj <- phyloseq(otu_table(ASV_ps, taxa_are_rows = FALSE),
                   tax_table(Fungi_blast_tax_ps))

# Aggregate ASVs at the selected level
ps_glom <- tax_glom(ps_obj, taxrank = tax_glom_level, NArm = FALSE)

# Convert the aggregated taxonomic table to a data frame
Fungi_blast_tax_glom <- as.data.frame(tax_table(ps_glom)) %>%
  tibble::rownames_to_column("seq")

# Convert the aggregated ASV table to a data frame
Fungi_ASW_glom <- as.data.frame(otu_table(ps_glom)) %>%
  tibble::rownames_to_column("Sample")

# Merge and adjust columns in the final ASV table
Fungi_ASW_glom_j <- ASV %>%
  select(!one_of(rownames(Fungi_blast_tax_ps))) %>%
  left_join(Fungi_ASW_glom, .)

# Sort columns alphabetically for consistency
Fungi_ASW_glom_j <- Fungi_ASW_glom_j[sort(colnames(Fungi_ASW_glom_j))]

# Combine the original taxonomic table with the merged table
Fungi_blast_tax_glom_j <- Fungi_blast_tax %>%
  select(-prct_id, -in_reflist) %>% 
  filter(!seq %in% rownames(Fungi_blast_tax_ps)) %>%
  bind_rows(Fungi_blast_tax_glom, .) %>%
  arrange(seq)
```

# Export Final Aggregated Data
Export the aggregated ASV and taxonomic data tables

```{r}
# Export the processed taxonomic table with genus-level aggregation
write_tsv(Plant_blast_tax_glom_j, here("Data", primer, "out_final", paste(primer, "taxa_blast_glom.txt", sep = "_")))

write_tsv(Fungi_blast_tax_glom_j, here("Data", primer, "out_final", paste(primer, "taxa_blast_glom_fungi.txt", sep = "_")))

# Export the aggregated ASV table
write_tsv(ASW_glom_j, here("Data", primer, "out_final", paste(primer, "ASV_glom.txt", sep = "_")))

write_tsv(Fungi_ASW_glom_j, here("Data", primer, "out_final", paste(primer, "ASV_glom_fungi.txt", sep = "_")))

# Export the subset sequence table
#writeXStringSet(seqs_glom, here("Data", primer, "out_final", paste(primer, "Seq_glom", sep = "_")))
```

```{r}
library(eulerr)

species_venn_df <- 
  Plant_blast_tax_glom_j %>% 
  filter(kingdom == "Plantae") %>% 
  select(species) %>% 
  mutate(DB = "Plant_ITS") %>% 
  bind_rows(.,tibble(species = Fungi_blast_tax_glom_j$species, DB = "Fungi_ITS")) %>% 
  filter(!is.na(species)) %>% 
#  filter(!grepl("sp.", species)) %>% 
  distinct() %>%
  mutate(species2 = species) %>% 
  pivot_wider(names_from = DB, values_from = species2) %>% 
  mutate_at(-1, ~!is.na(.)) 

species_venn <- 
  species_venn_df %>% 
  select(-species) %>% 
  euler(shape = "ellipse") 

plot(species_venn, legend = TRUE, quantities = TRUE ,main = "species detection")
```
```{r}

genus_venn_df <- 
  Plant_blast_tax_glom_j %>% 
  filter(kingdom == "Plantae") %>% 
  select(genus) %>% 
  mutate(DB = "Plant_ITS") %>% 
  bind_rows(.,tibble(genus = Fungi_blast_tax_glom_j$genus, DB = "Fungi_ITS")) %>% 
  filter(!is.na(genus)) %>% 
#  filter(!grepl("sp.", genus)) %>% 
  distinct() %>%
  mutate(genus2 = genus) %>% 
  pivot_wider(names_from = DB, values_from = genus2) %>% 
  mutate_at(-1, ~!is.na(.)) 

genus_venn <- 
  genus_venn_df %>% 
  select(-genus) %>% 
  euler(shape = "ellipse") 

plot(genus_venn, legend = TRUE, quantities = TRUE ,main = "genus detection")
```
```{r}
Fungi_tax

Fungi_ASW %>% 
  pivot_longer(-Sample, names_to = "seq", values_to = "reads") %>% 
  group_by(seq) %>% 
  summarise(reads = sum(reads)) %>% 
  left_join(Fungi_tax) %>% 
  group_by(phylum) %>% 
  summarise(reads = sum(reads)) %>% 
  mutate(reads_prct = round(reads / sum(reads) *100,2))
```

