---
title: "Fungi ITS Analysis"
author: "Fabian Roger"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, message = FALSE}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(readr)
library(here)
library(stringr)
library(ggplot2)
library(ShortRead)
library(dada2)
library(ggrepel)
library(phyloseq)


# Load helper functions
source(here("Scripts", "Seq_analysis_helper.R"))
```

```{r}
sessionInfo()
```

# Set-Up
This script follows the [DADA2 ITS Tutorial (1.8)](https://benjjneb.github.io/dada2/ITS_workflow.html).
The data are demultiplexed (see `Demultiplexing.Rmd`)
The original article can be found [here](http://rdcu.be/ipGh)

The analysis proceeds through read filtering, error learning, denoising, merging, chimera removal, taxonomic assignment, and clustering.


## Setup Directories and Paths
Set up paths for input and output data and create necessary directories.

```{r}
primer <- "Fungi"
trimmed_path <- here("Data", primer, "demultiplexed")
filter_path <- here("Data", primer, "filtered")
dir.create(filter_path, showWarnings = FALSE)
dir.create(here("Data", primer, "out_intermediate"))
dir.create(here("Data", primer, "out_final"))
```

## Filter and Prepare Reads
Prepare and sort forward and reverse read files. Extract sample names for downstream processing.

```{r, "sort fwrd and rev", cache=TRUE}
# List forward and reverse read files
fnFs <- sort(list.files(trimmed_path, pattern = "_R1.fastq.gz", full.names = TRUE))
fnRs <- sort(list.files(trimmed_path, pattern = "_R2.fastq.gz", full.names = TRUE))

# Extract sample names from filenames
sample.names <- sapply(fnFs, function(x) gsub(paste(".+", primer, "/demultiplexed/(.+)_R1.fastq.gz", sep = ""), "\\1", x), USE.NAMES = FALSE)
```

## Check Read Lengths
Check read length distributions to ensure read quality.

```{r}
# Check read length distributions for quality
readsR1 <- ShortRead::readFastq(fnFs[1]) %>%
  ShortRead::sread() %>%
  width() %>%
  table() %>%
  as.data.frame() %>%
  dplyr::rename(Length = 1) %>%
  mutate(Direction = "Forward")

readsR2 <- ShortRead::readFastq(fnRs[1]) %>%
  ShortRead::sread() %>%
  width() %>%
  table() %>%
  as.data.frame() %>%
  dplyr::rename(Length = 1) %>%
  mutate(Direction = "Reverse")

bind_rows(readsR1, readsR2) %>%
  mutate(Length = as.numeric(as.character(Length))) %>% 
  ggplot(aes(x = Length, y = Freq, color = Direction)) +
  geom_line(alpha = 0.5) +
  theme_bw() +
  facet_wrap(~Direction) +
  scale_x_continuous(limits = c(0, 500))
```

## Quality Profiles
Visualize the quality profile of the forward and reverse reads to determine trimming parameters.

```{r, "plot quality of reads", cache=TRUE}
# plot one forward and one reverse read in the rapport
plotQualityProfile(fnFs[1])+
  scale_y_continuous(limits = c(0,45))+
  scale_x_continuous(breaks = seq(0,300, 20))+
  geom_hline(yintercept = 30, colour = "grey", linetype = "dashed")+
  geom_hline(yintercept = 20, colour = "grey", linetype = "dashed")+
  labs(title = paste(sample.names[1], "foward", sep = " - " ))

plotQualityProfile(fnRs[1])+
  scale_y_continuous(limits = c(0,45))+
  scale_x_continuous(breaks = seq(0,300, 20))+
  geom_hline(yintercept = 30, colour = "grey", linetype = "dashed")+
  geom_hline(yintercept = 20, colour = "grey", linetype = "dashed")+
  labs(title = paste(sample.names[1], "reverse", sep = " - "))
```

The quality of the forwards reads is very high, throughout the full length of the read.  The quality of reverse read is >= Q35 up to ~200 bps.

## Merge Read Lengths
Merge paired reads to assess overlap and read length distribution for optimal trimming decisions.

```{r}
# Merge the first file's forward and reverse reads using vsearch for length distribution analysis
 system(
  paste(
    "/Applications/vsearch/bin/vsearch --fastq_mergepairs ", 
    gsub("My Drive", "My\\\\ Drive", fnFs[1]), 
    " --reverse ", gsub("My Drive", "My\\\\ Drive", fnRs[1]), 
    " --fastaout", gsub("My Drive", "My\\\\ Drive", here("Data", primer, "test_merge.fasta")), 
    "--fastq_allowmergestagger",
    "--fastq_qmax 45"
    ), intern = TRUE
  )

# Read the merged sequences into R
file_merged <- readDNAStringSet(here("Data", primer, "test_merge.fasta"))

# Delete the temporary merged file after use
file.remove(here("Data", primer, "test_merge.fasta"))

# Create a table of sequence lengths and plot the distribution
table(width(file_merged)) %>%
  as.data.frame() %>%
  mutate(across(c("Var1", "Freq"), function(x) as.numeric(as.character(x)))) %>%  # Ensure columns are numeric
  arrange(desc(Freq)) %>%  # Sort by frequency
  mutate(
    prct = Freq / sum(Freq) * 100,  # Calculate the percentage of each length
    rank = 1:n(),  # Create a rank column
    Label = case_when(  # Add labels for the top 5 most frequent lengths
      rank < 6 ~ paste(Var1, "(", round(prct, 1), " %)", sep = ""),
      TRUE ~ NA_character_
    )
  ) %>%
  ggplot(aes(x = Var1, y = Freq)) +
  geom_line(colour = "darkgrey", alpha = 0.5) +  # Line plot for length distribution
  geom_text_repel(aes(label = Label)) +  # Annotate top labels
  geom_point(data = ~filter(.x, !is.na(Label)), colour = "red", size = 0.5) +  # Highlight labeled points
  theme_bw() +
  scale_x_continuous(limits = c(0, 500))  # Limit x-axis range
```

We sequenced with 2 x 300 bp.
the largest merged fragments are 387 bps long

The trimmed forward and reverse reads are 278 and 275 bps long, respectively.

Thus we have a minimum overlap of `278+275-387` = 166 bps.

given the lower quality of the reverse reads, I trimm them down to 200, leaving an overlap of 83 bps.

## Set Read Parameters
Set trimming parameters based on observed read quality and overlap.

```{r}
# Set parameters for fragment and read lengths
frag_len <- 387
fwrd_len <- 278
rev_len <- 275
fwrd_trim <- 270
rev_trim <- 200

# Create a plot showing the trimmed regions and overlap
ggplot(data = tibble(x = c(1, frag_len), y = c(1, 5)), aes(x = x, y = y)) +
  geom_blank() +  # Base blank plot to set up the axes
  geom_segment(x = 0, xend = frag_len, y = 2, yend = 2) +  # Full fragment line
  geom_segment(x = 0, xend = fwrd_len, y = 3, yend = 3) +  # Full forward read
  geom_segment(x = frag_len - rev_len, xend = frag_len, y = 3.5, yend = 3.5) +  # Full reverse read
  geom_segment(x = fwrd_trim, xend = fwrd_len, y = 3, yend = 3,  # Trimmed forward segment
               colour = "red", size = 2, alpha = 0.2) +
  geom_segment(x = frag_len - rev_trim, xend = frag_len - rev_len,  # Trimmed reverse segment
               y = 3.5, yend = 3.5, colour = "red", size = 3, alpha = 0.2) +
  geom_segment(x = frag_len - rev_trim, xend = fwrd_trim, y = 2,  # Overlap region
               yend = 2, colour = "green", size = 3, alpha = 0.2) +
  geom_text(x = ((frag_len - rev_trim) + fwrd_trim) / 2, y = 2.25,  # Label showing overlap length
            label = paste(as.character(fwrd_trim + rev_trim - frag_len), "bp"),
            colour = "blue", size = 5) +
  theme_void() +  # Removes Fungikground and axes for clarity
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +  # Adjust x-axis text orientation
  scale_x_continuous(breaks = seq(0, frag_len, 12))  # X-axis ticks
```

The filtering parameters we’ll use are stringent as the quality is very high and we have a lot of reads. 

+ maxN=0 (DADA2 requires no Ns)
+ truncQ=2 
+ maxEE=c(1,2)

The `maxEE` parameter sets the maximum number of “expected errors” allowed in a read. 

We use the `fastqPairedFilter` function to jointly filter the forward and reverse reads.

#DADA2 Pipeline

## Filter and Trim Reads
Filter and trim the reads based on quality, expected errors, and length settings. Write filtered read statistics to file.

```{r}
# Create file paths for the filtered forward and reverse reads, stored in the filtered/ subdirectory
filtFs <- file.path(here("Data", primer, "filtered"), paste0(sample.names, "_F_filt.fastq.gz"))  # Path for forward filtered reads
filtRs <- file.path(here("Data", primer, "filtered"), paste0(sample.names, "_R_filt.fastq.gz"))  # Path for reverse filtered reads

# Name the paths
names(filtFs) <- sample.names
names(filtRs) <- sample.names
```

```{r}
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, 
                     truncLen = c(fwrd_trim, rev_trim),
                     maxN=0,
                     maxEE=c(1,1),
                     truncQ=2,
                     rm.phix=TRUE,
                     compress=TRUE, multithread=6) 
out_df <- 
out %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = "Sample_name") %>% 
  mutate(Sample_name = gsub("_R1.+", "", Sample_name)) %>% 
  mutate(prct = signif(reads.out / reads.in * 100, 2)) %>% 
  mutate(Sample_name = gsub("(.+?)_S.+.fastq.gz", "\\1", Sample_name))

write_tsv(out_df, here("Data", primer, "out_intermediate", paste(primer, "ee_filter_stats.txt", sep = "_")))

out_df 
```

## Learn Errors
Estimate error rates from the filtered reads to inform the denoising step.

```{r}
# remove files whith no reads left after filtering
files_passed_ee <- 
  which(out[,2] > 0) 
  
filtFs <- filtFs[files_passed_ee]
filtRs <- filtRs[files_passed_ee]
```

```{r}
errF <- learnErrors(
  filtFs,
  multithread = TRUE,
  nbases = 1e9,
  errorEstimationFunction = loessErrfun_mod4,
  verbose = TRUE
)

errR <- learnErrors(
  filtRs,
  multithread = TRUE,
  nbases = 1e9,
  errorEstimationFunction = loessErrfun_mod4,
  verbose = TRUE
)
```

```{r, cache=TRUE}
plotErrors(errF, nominalQ=TRUE)

plotErrors(errR, nominalQ=TRUE)
```

## Denoise
Apply DADA2 denoising to forward and reverse reads to identify unique sequences.

```{r}
filt <- list.files(filter_path, full.names = T)

filtFs <- filt[grepl("*_F_filt.fastq.gz", filt)]
names(filtFs) <- gsub(".+filtered/(.+?)_F.+", "\\1", filtFs)

filtRs <- filt[grepl("*_R_filt.fastq.gz", filt)]
names(filtRs) <- gsub(".+filtered/(.+?)_F.+", "\\1", filtRs)
```

```{r}
dadaFs <- dada(filtFs, err=errF, multithread=TRUE)
dadaRs <- dada(filtRs, err=errR, multithread=TRUE)
```

```{r}

denoise_stat <- 
sapply(dadaFs, function(x) sum(getUniques((x)))) %>% 
  data.frame(denoised = .) %>% 
  tibble::rownames_to_column(var = "Sample_name") %>% 
  left_join(select(out_df, Sample_name, reads.out)) %>% 
  mutate(prct = round(denoised / reads.out*100, 2))
  
write_tsv(denoise_stat, here("Data", primer, "out_intermediate", paste(primer, "denoise_stat.txt", sep = "_")))

denoise_stat
```

## Merge
Merge the forward and reverse reads and create a summary of merged reads for each sample.

```{r}

filtFs <- filtFs[names(dadaFs)]
filtRs <- filtRs[names(dadaRs)]

mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose=TRUE)

merge_stat <- 
lapply(mergers, function(x) {sum(x$abundance)}) %>% 
  as_tibble() %>% 
  pivot_longer(everything(),
    names_to = "Sample_name",
               values_to = "merged.out") %>% 
  left_join(select(out_df, Sample_name, reads.out)) %>% 
  mutate(prct = round(merged.out/reads.out * 100, 2))

merge_stat %>% 
write_tsv(here("Data", primer, "out_intermediate", paste(primer, "merge_stat.txt", sep = "_")))

merge_stat

```

## Construct Sequence Table
Generate the sequence table from the merged reads.

```{r, "Seqeunce table"}
seqtab <- makeSequenceTable(mergers)
dim(seqtab)
```

## Chimera Removal
Remove chimeric sequences from the sequence table and save chimera statistics to file.

```{r, "Chimera removal", cache=TRUE}

seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread=TRUE, verbose=TRUE)

write_rds(seqtab.nochim, here("Data", primer, "out_intermediate", paste(primer, "seqtab.nochim.rds", sep = "_")))
seqtab.nochim <- read_rds(here("Data", primer, "out_intermediate", paste(primer, "seqtab.nochim.rds", sep = "_")))

dim(seqtab.nochim)
```

```{r}
chimera_stat <- 
tibble(Sample_name = rownames(seqtab),
       before = rowSums(seqtab),
       after = rowSums(seqtab.nochim),
       prct = round(after/before*100,2))

chimera_stat %>% 
write_tsv(here("Data", primer,  "out_intermediate", paste(primer, "Chimera_stat.txt", sep = "_"))) 

chimera_stat
```

percentage of Chimeric reads

```{r}
(1-(sum(seqtab.nochim)/sum(seqtab)))*100
```

percentage of Chimeric ASVs

```{r}
(1-(ncol(seqtab.nochim)/ncol(seqtab)))*100
```

## Length Filtering
Visualize the length distribution of ASVs to confirm expected fragment size.

```{r}
data.frame(table(nchar(colnames(seqtab.nochim)))) %>% 
  mutate(across(c("Var1", "Freq"), function(x) as.numeric(as.character(x)))) %>%  # Ensure columns are numeric
  arrange(desc(Freq)) %>%  # Sort by frequency
  mutate(
    prct = Freq / sum(Freq) * 100,  # Calculate the percentage of each length
    rank = 1:n(),  # Create a rank column
    Label = case_when(  # Add labels for the top 5 most frequent lengths
      rank < 6 ~ paste(Var1, "(", round(prct, 1), " %)", sep = ""),
      TRUE ~ NA_character_
    )
  ) %>%
  ggplot(aes(x = Var1, y = Freq)) +
  geom_line(colour = "darkgrey", alpha = 0.5) +  # Line plot for length distribution
  geom_text_repel(aes(label = Label)) +  # Annotate top labels
  geom_point(data = ~filter(.x, !is.na(Label)), colour = "red", size = 0.5) +  # Highlight labeled points
  theme_bw() +
  scale_x_continuous(limits = c(200, 400))  # Limit x-axis range
```

We have a large lengths variation as expected for ITS2. We do not apply lengths filtering.

```{r}
# sum(nchar(colnames(seqtab.nochim)) > 257)
# sum(nchar(colnames(seqtab.nochim)) > 257)/ncol(seqtab.nochim)*100
# 
# seqtab.nochim <- seqtab.nochim[,nchar(colnames(seqtab.nochim)) <= 257]
# dim(seqtab.nochim)
```


## Export Raw ASV and Fasta Files
Export the non-chimeric ASV sequences to a FASTA file and save the ASV abundance table as a TSV.


```{r}
# Convert the columns of the sequence table (ASVs) to a DNAStringSet object for export
seqs <- DNAStringSet(colnames(seqtab.nochim))

# Assign sequence names in the format: seq_001, seq_002, etc.
names(seqs) <- paste("seq", 
                     formatC(1:length(seqs), flag = "0", width = nchar(length(seqs))), 
                     sep = "_")

# Export the ASVs as a FASTA file, prefaced with the primer name
writeXStringSet(seqs, here("Data", primer, "out_intermediate", paste(primer, "ASVs.fasta", sep = "_")), format = "fasta")

# Prepare the ASV table for export as a TSV
ASV <- seqtab.nochim %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = "Sample") %>% 
  `colnames<-`(c("Sample", names(seqs)))

# Export the ASV table as a TSV file, prefaced with the primer name
write_tsv(ASV, here("Data", primer, "out_intermediate", paste(primer, "ASVs.txt", sep = "_")))

# Code to read the TSV (if needed for future processing)
# ASV <- read_tsv(here("Data", primer, "out_intermediate", paste(primer, "ASVs.txt", sep = "_")))
```


# Swarm Clustering

We use swarm for single linkage clustering of the ASVs. The method is described here:

> Mahé F, Rognes T, Quince C, Vargas C de, Dunthorn M. 2015 Swarm v2: highly-scalable and high-resolution amplicon clustering. PeerJ 3, e1420. (doi:10.7717/peerj.1420)

The github is [here](https://github.com/torognes/swarm)


Swarm needs a fasta file with size annotation as input. It just separates the size by a "_"

```{r}
seqs_size <- 
  colSums(ASV[,-1]) %>% 
  data.frame(size = .) %>% 
  tibble::rownames_to_column(var = "seq") %>% 
  mutate(new_name = gsub("_", "", seq)) %>% 
  mutate(new_name = paste(new_name, size, sep = "_"))

#make sure sequences are in correct order
seqs <- seqs[seqs_size$seq]

#rename
names(seqs) <- seqs_size$new_name

# Export ASV sequences for Swarm clustering input
writeXStringSet(seqs, here("Data", primer, "out_intermediate", paste(primer, "ASVs_swarm_in.fasta", sep = "_")), format = "fasta")

```


## Run Swarm Clustering
Run Swarm with the processed ASV input to generate clusters and identify sequence centroids.

```{r}
# Run Swarm for single-linkage clustering of ASVs
tmp <- system2("/Applications/swarm-3.1.0/bin/swarm",
        args = c("-d 1", "-f", "-t 8", "-w",
                 shQuote(here("Data", primer, "out_intermediate", 
                              paste(primer, "ASVs_swarm_out.fasta", sep = "_"))), 
                 paste("-u", shQuote(here("Data", primer, "out_intermediate", 
                                          paste(primer, "ASVs_swarm_cluster.txt", sep = "_"))), sep = " "),
                 shQuote(here("Data", primer, "out_intermediate", paste(primer, "ASVs_swarm_in.fasta", sep = "_")))),
        stdout = TRUE)
```
## Process Swarm output

The ITS_ASVs_swarm_cluster.txt file tells us which sequences have been clustered

column 09 gives the original sequence name
column 10 gives the centroid sequence to which it has been clustered
column 01 tells us if the sequence is itself a centroid sequence ("S") or clustered to a centroid ("H")
column 04 gives the % identity with the centroid (here between 100% and 99%)

```{r}
# Read the Swarm clustering result file
Cluster_ID <- read_delim(here("Data", primer, "out_intermediate", paste(primer, "ASVs_swarm_cluster.txt", sep = "_")), 
                         delim = "\t", col_names = FALSE)

# Filter only sequences that have been clustered and rename columns
Cluster_ID <- Cluster_ID %>%
  dplyr::rename(seq = X9, Type = X1, match = X10, pident = X4) %>% 
  select(seq, Type, match, pident) %>% 
  filter(Type == "H") %>%  # Keep only sequences that were clustered to a centroid
  arrange(match)

# Rename sequences to match the sequence table format
Cluster_ID <- Cluster_ID %>%
  mutate(across(c("seq", "match"), function(x) {
    gsub("(seq)(\\d+)_\\d+", "\\1_\\2", x)
  }))

# Split by cluster centroid for processing
Cluster_list <- split(Cluster_ID, Cluster_ID$match)

# Sum abundance of clustered sequences
ASV_list <- lapply(Cluster_list, function(x) {
  S_seq <- unique(x$match)
  C_seqs <- x$seq
  ASV[, S_seq] <- rowSums(ASV[, c(S_seq, C_seqs)])
})

# Create a new ASV table with only the centroid sequences
ASV_sw <- ASV
ASV_sw[, names(ASV_list)] <- bind_rows(ASV_list)
ASV_sw <- ASV_sw[, !colnames(ASV_sw) %in% Cluster_ID$seq]  # Remove clustered sequences

# Export the updated ASV table
ASV_sw %>% 
  write.table(., here("Data", primer, "out_intermediate", paste(primer, "ASV_swarm.txt", sep = "_")))

# Display the distribution of percent identities for clustered sequences
Cluster_ID %>%
  filter(Type == "H") %>%
  pull(pident) %>%
  as.numeric() %>%
  table()
```

## Update Fasta with Centroids
```{r}
# Read the FASTA file output from Swarm clustering
seqs <- readDNAStringSet(here("Data", primer, "out_intermediate", paste(primer, "ASVs_swarm_out.fasta", sep = "_")))

# Rename the sequences to match the ASV table format
names(seqs) <- gsub("(seq)(\\d+)_\\d+", "\\1_\\2", names(seqs))

# Check if any sequences in the FASTA file are missing in the ASV table
missing_count <- sum(!names(seqs) %in% colnames(ASV_sw[, -1]))
missing_count  # Output the count of missing sequences

# Export the updated FASTA file, prefaced with the primer name
writeXStringSet(seqs, here("Data", primer, "out_intermediate", paste(primer, "ASV_swarm.fasta", sep = "_")))
```

read_files
```{r}
# Read the final FASTA file of clustered ASV centroids
seqs <- readDNAStringSet(here("Data", primer, "out_intermediate", paste(primer, "ASV_swarm.fasta", sep = "_")))
```

# ITSx
This section runs the ITSx tool to extract the ITS2 region from the ASV sequences

```{r}
cores = 6

# Copy ASV FASTA file to the Downloads directory to avoid path issues
system2(command = "cp", 
        args = c(shQuote(here("Data", primer, "out_intermediate", paste(primer, "ASV_swarm.fasta", sep = "_"))),
                 "~/Downloads/"))

# Create a temporary output directory for ITSx results
dir.create("~/Downloads/ITSX_out")

# Run ITSx for ITS2 region extraction
system(paste("source ~/.bash_profile && /Applications/ITSx_1.1.3/ITSx",
             "ITSx -i ",
             paste("~/Downloads/", paste(primer, "ASV_swarm.fasta ", sep = "_"), sep = ""),
             "-o ~/Downloads/ITSX_out/ITSx ", 
             "-t all ",
             "--allow_single_domain '1e-5' ",
             "--complement F ",
             "--save_regions ITS2 ",
             "--cpu ", 
             cores
             ))

# Move ITSx results back to the project directory
system2(command = "mv", 
        args = c("~/Downloads/ITSX_out",
                 shQuote(here("Data", primer, "out_intermediate"))))

# Clean up: Remove temporary ASV FASTA file from Downloads
file.remove(paste("~/Downloads/", paste(primer, "ASV_swarm.fasta", sep = "_"), sep = ""))
```


read in sequences with ITS2 domains
```{r}
# Read the ITS2 region sequences from the ITSx output
seq_ITS2 <- readDNAStringSet(here("Data", primer, "out_intermediate", "ITSX_out", "ITSx.ITS2.fasta"))

# foramt sequence names to match input
names(seq_ITS2) <- gsub("(seq_\\d+)\\|.+", "\\1", names(seq_ITS2))

# Check how many original sequences are not in the ITS2 set
length(which(!names(seqs) %in% names(seq_ITS2)))

# Export the ITS2 sequences as a new FASTA file, prefaced with the primer name
writeXStringSet(seq_ITS2, here("Data", primer, "out_intermediate", paste(primer, "ASVs_ITS2.fasta", sep = "_")), format = "fasta")

# Uncomment this line if needed for reading the ITS2 sequences back in the future
# seq_ITS2 <- readDNAStringSet(here("Data", primer, "out_intermediate", paste(primer, "ASVs_ITS2.fasta", sep = "_")))
```

# Taxonomy Assignment

https://unite.ut.ee/repository.php

Abarenkov, Kessy; Zirk, Allan; Piirmann, Timo; Pöhönen, Raivo; Ivanov, Filipp; Nilsson, R. Henrik; Kõljalg, Urmas (2022): UNITE general FASTA release for eukaryotes 2. Version 16.10.2022. UNITE Community. https://doi.org/10.15156/BIO/2483914

Includes global and 3% distance singletons. 

I updated the taxonomy in the reference database to match the Gbif backbone. 

I exclude reference sequences with no genus level assignment. 

(see Prepare_RefDB.Rmd)

## Run Sintax for Taxonomic Assignment
Use vsearch to perform taxonomic classification

```{r}
# Run vsearch for taxonomic assignment using the Sintax algorithm
system2("/Applications/vsearch/bin/vsearch",
        args = c("--sintax", 
                 shQuote(here("Data", primer, "out_intermediate", 
                              paste(primer, "ASVs_ITS2.fasta", sep = "_"))),
                 paste("--db", shQuote(here("Data", "Fungi", "in_files",
                                            "Unite_db_sintax_gbif_genus.fa")), sep = " "),
                 "--sintax_cutoff 0.1",  # Set confidence threshold for taxonomic assignment
                 "-strand both",  # Analyze both strands
                 "--tabbedout", 
                 shQuote(here("Data", primer, "out_intermediate", paste(primer, "tax_raw.txt", sep = "_")))
                 )
)

```

## Prepare and Format Taxonomic Data
Format the Sintax output into comprehensive and confidence-filtered taxonomic tables.

```{r}
# Read the raw taxonomic output from the Sintax algorithm
Sintax_taxa <- read_tsv(here("Data", primer, "out_intermediate", paste(primer, "tax_raw.txt", sep = "_")), 
                        col_names = FALSE) %>% 
  select(1:2)  # Select only the sequence and taxonomy columns

# Rename columns 
colnames(Sintax_taxa) <- c("seq", "tax")

# Extract unassigned sequences for reference
Sintax_na <- Sintax_taxa %>% filter(is.na(tax)) %>% select(seq)

# Prepare and format the taxonomic data
Sintax_prep <- Sintax_taxa %>% 
  filter(!is.na(tax)) %>% 
  mutate(tax = strsplit(tax, ",")) %>%  # Split taxonomic data into levels
  unnest(tax) %>% 
  group_by(seq) %>% 
  mutate(taxonomy_name = gsub("\\w:(.+)\\(.+", "\\1", tax),  # Extract taxonomic names
         taxonomy_level = gsub("(\\w):.+", "\\1", tax),  # Extract taxonomy levels
         taxonomy_conf = gsub(".+\\((.+)\\)", "\\1", tax)) %>%  # Extract confidence scores
  left_join(tax_match_df)

# Create a table with all taxonomic assignments, even low-confidence ones
Sintax_taxa_table_full <- Sintax_prep %>% 
  select(-tax, -taxonomy_conf, -taxonomy_level) %>% 
  pivot_wider(names_from = taxon, values_from = taxonomy_name) %>% 
  bind_rows(Sintax_na) %>% 
  arrange(seq)

# Create a table with confidence scores for each taxonomic level
Sintax_taxa_table_prob <- Sintax_prep %>% 
  select(-tax, -taxonomy_name, -taxonomy_level) %>% 
  pivot_wider(names_from = taxon, values_from = taxonomy_conf) %>% 
  bind_rows(Sintax_na) %>% 
  arrange(seq)
```


## Export Taxonomic Tables

```{r}
# Export the full taxonomic assignment table 
write_tsv(Sintax_taxa_table_full, here("Data", primer, "out_intermediate", paste(primer, "Sintax_tax.txt", sep = "_")))

# Export the table with confidence scores 
write_tsv(Sintax_taxa_table_prob, here("Data", primer, "out_intermediate", paste(primer, "Sintax_prob.txt", sep = "_")))

# Read the files Fungi in
Sintax_taxa_table_full <- read_tsv(here("Data", primer, "out_intermediate", paste(primer, "Sintax_tax.txt", sep = "_")))
Sintax_taxa_table_prob <- read_tsv(here("Data", primer, "out_intermediate", paste(primer, "Sintax_prob.txt", sep = "_")))
```


```{r}
# Filter the taxonomic assignment table to retain only those with a confidence score >= 0.8
Fungi_taxa_80 <- Sintax_taxa_table_full
Fungi_taxa_80[, colnames(Fungi_taxa_80) %in% colnames(Sintax_taxa_table_prob)][Sintax_taxa_table_prob < 0.8] <- NA_character_

# Export the filtered taxonomic assignment table
write_tsv(Fungi_taxa_80, here("Data", primer, "out_intermediate", paste(primer, "Sintax_80.txt", sep = "_")))

# Read the filtered taxonomic table for further processing or analysis
Fungi_taxa_80 <- read_tsv(here("Data", primer, "out_intermediate", paste(primer, "Sintax_80.txt", sep = "_")))
```
## Lulu 

```{r generate_match_list, echo=TRUE}
# Generate the match list using VSEARCH
system(paste0("/Applications/vsearch/bin/vsearch --usearch_global ",
              shQuote(here("Data", primer, "out_intermediate", 
                              paste(primer, "ASVs_ITS2.fasta", sep = "_"))), 
             " --db ", 
             shQuote(here("Data", primer, "out_intermediate", 
                              paste(primer, "ASVs_ITS2.fasta", sep = "_"))), 
             " --self --id 0.84 --iddef 1 --maxaccepts 0 --query_cov .9 --maxhits 10 --userout ", 
             shQuote(here("Data", primer, "out_intermediate", 
                              paste(primer, "match_list.txt", sep = "_"))), 
             " --userfields query+target+id"))

```

```{r read_data, echo=TRUE}
# Read the match list
matchlist <- read_tsv(here("Data", primer, "out_intermediate", 
                              paste(primer, "match_list.txt", sep = "_")), col_names = FALSE)

# Rename columns to match the expected column names
colnames(matchlist) <- c("OTUid", "hit", "match")

# read in the ASV table and exclude sequences excluded 
ASV_sw <- read.table(here("Data", primer, "out_intermediate", paste(primer, "ASV_swarm.txt", sep = "_")))
```


```{r}
ASV_sw <- select(ASV_sw, one_of("Sample", matchlist$OTUid))

rownames(ASV_sw) <- ASV_sw$Sample
ASV_sw$Sample <- NULL
ASV_sw_t <- t(ASV_sw)

lulu_out <- lulu_optimized(ASV_sw_t, matchlist, 
                           minimum_ratio_type = "min", 
                           minimum_ratio = 1, 
                           minimum_match = 95, 
                           minimum_relative_cooccurence = 0.95)

ASV_lulu <- 
lulu_out$curated_table %>% 
  t() %>% 
  data.frame() %>% 
  tibble::rownames_to_column("Sample")

lulu_match <- 
lulu_out$otu_map %>% 
  filter(curated != "parent") %>% 
  arrange(parent_id) %>% 
  mutate(match = paste("pair", 1:n())) %>% 
  pivot_longer(one_of("OTUid", "parent_id"), values_to = "seq") %>%
  left_join(Fungi_taxa_80) 

lulu_tax_match <- 
  lulu_match %>%  
  group_by(match) %>% 
  mutate_at(8:14, function(x){length(unique(x)) == 1}) %>% 
  pivot_wider(names_from = name, values_from = seq)

no_match_phylum <- 
lulu_tax_match %>% 
  filter(!phylum) %>% 
  pull(match) %>% 
  unique()

lulu_tax_match %>% 
  filter(match %in% no_match_phylum)

lulu_tax_match %>% 
  ungroup() %>% 
  select(6:12) %>% 
  pivot_longer(everything()) %>% 
  group_by(name, value) %>% 
  summarise(n = n()) %>% 
  mutate(name = factor(name, levels = c("kingdom", "phylum", "class", "order", "family", "genus", "species"))) %>% 
  ggplot(aes(x = name, y = n, fill = value)) +
  geom_bar(stat = "identity", position = "stack")
  

```

## Merge and Aggregate ASV by Taxonomic Level
Use phyloseq to merge ASVs with the same species assignment.


```{r}
#define taxonomic aggregation level
tax_glom_level <- "species"

# Read the ASV abundance table after Swarm clustering
ASV_sw <- read.table(here("Data", primer, "out_intermediate", paste(primer, "ASV_swarm.txt", sep = "_")))

# Prepare ASV matrix and set sample names as row names
ASV_ps <- ASV_sw[, -1]  # Remove the first column (sample identifiers)
rownames(ASV_ps) <- ASV_sw$Sample

# Filter and prepare the taxonomic table 
Fungi_taxa_80_ps <- Fungi_taxa_80 %>%
  filter(!is.na(!!sym(tax_glom_level))) %>%
  select(-seq) %>%
  as.matrix()

# Set sequence names as row names in the taxonomic matrix
rownames(Fungi_taxa_80_ps) <- filter(Fungi_taxa_80, !is.na(!!sym(tax_glom_level)))$seq

# Create a phyloseq object using the ASV and taxonomic data
ps_obj <- phyloseq(otu_table(ASV_ps, taxa_are_rows = FALSE),
                   tax_table(Fungi_taxa_80_ps))

# Aggregate ASVs at the selected level
ps_glom <- tax_glom(ps_obj, taxrank = tax_glom_level, NArm = FALSE)

# Convert the aggregated taxonomic table to a data frame
Fungi_taxa_80_glom <- as.data.frame(tax_table(ps_glom)) %>%
  tibble::rownames_to_column("seq")

# Convert the aggregated ASV table to a data frame
ASW_glom <- as.data.frame(otu_table(ps_glom)) %>%
  tibble::rownames_to_column("Sample")

# Merge and adjust columns in the final ASV table
ASW_glom_j <- ASV_sw %>%
  select(!one_of(rownames(Fungi_taxa_80_ps))) %>%
  left_join(ASW_glom, .)

# Sort columns alphabetically for consistency
ASW_glom_j <- ASW_glom_j[sort(colnames(ASW_glom_j))]

# Combine the original taxonomic table with the merged table
Fungi_taxa_80_glom_j <- Fungi_taxa_80 %>%
  filter(!seq %in% rownames(Fungi_taxa_80_ps)) %>%
  bind_rows(Fungi_taxa_80_glom, .) %>%
  arrange(seq)

# Prepare a sequence table with just the representative sequences
seqs_glom <- seqs[Fungi_taxa_80_glom_j$seq]

dim(ASV_sw)
dim(ASW_glom_j)
```


# Export Final Aggregated Data
Export the aggregated ASV and taxonomic data tables

```{r}
# Export the processed taxonomic table with genus-level aggregation
write_tsv(Fungi_taxa_80_glom_j, here("Data", primer, "out_final", paste(primer, "taxa_80_glom.txt", sep = "_")))

# Export the aggregated ASV table
write_tsv(ASW_glom_j, here("Data", primer, "out_final", paste(primer, "ASV_glom.txt", sep = "_")))

# Export the subset sequence table
writeXStringSet(seqs_glom, here("Data", primer, "out_final", paste(primer, "Seq_glom", sep = "_")))
```
