---
title: "Vertebrate 16S Analysis"
author: "Fabian Roger"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, message = FALSE}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(readr)
library(here)
library(stringr)
library(ggplot2)
library(ShortRead)
library(dada2)
library(ggrepel)
library(phyloseq)


# Load helper functions
source(here("Scripts", "Seq_analysis_helper.R"))
```

```{r}
sessionInfo()
```

# Set-Up
This script follows the [DADA2 Pipeline Tutorial (1.8)](https://benjjneb.github.io/dada2/tutorial_1_8.html).
The data are demultiplexed (see `Demultiplexing.Rmd`)
The original article can be found [here](http://rdcu.be/ipGh)

The analysis proceeds through read filtering, error learning, denoising, merging, chimera removal, taxonomic assignment, and clustering.


## Setup Directories and Paths
Set up paths for input and output data and create necessary directories.

```{r}
primer <- "Vertebrates"
trimmed_path <- here("Data", primer, "demultiplexed")
filter_path <- here("Data", primer, "filtered")
dir.create(filter_path, showWarnings = FALSE)
dir.create(here("Data", primer, "out_intermediate"))
dir.create(here("Data", primer, "out_final"))
```

## Filter and Prepare Reads
Prepare and sort forward and reverse read files. Extract sample names for downstream processing.

```{r, "sort fwrd and rev", cache=TRUE}
# List forward and reverse read files
fnFs <- sort(list.files(trimmed_path, pattern = "_R1.fastq.gz", full.names = TRUE))
fnRs <- sort(list.files(trimmed_path, pattern = "_R2.fastq.gz", full.names = TRUE))

# Extract sample names from filenames
sample.names <- sapply(fnFs, function(x) gsub(paste(".+", primer, "/demultiplexed/(.+)_R1.fastq.gz", sep = ""), "\\1", x), USE.NAMES = FALSE)
```

## Check Read Lengths
Check read length distributions to ensure read quality.

```{r}
# Check read length distributions for quality
readsR1 <- ShortRead::readFastq(fnFs[1]) %>%
  ShortRead::sread() %>%
  width() %>%
  table() %>%
  as.data.frame() %>%
  dplyr::rename(Length = 1) %>%
  mutate(Direction = "Forward")

readsR2 <- ShortRead::readFastq(fnRs[1]) %>%
  ShortRead::sread() %>%
  width() %>%
  table() %>%
  as.data.frame() %>%
  dplyr::rename(Length = 1) %>%
  mutate(Direction = "Reverse")

bind_rows(readsR1, readsR2) %>%
  mutate(Length = as.numeric(as.character(Length))) %>% 
  ggplot(aes(x = Length, y = Freq, color = Direction)) +
  geom_line(alpha = 0.5) +
  theme_bw() +
  facet_wrap(~Direction) +
  scale_x_continuous(limits = c(0, 500))
```

## Quality Profiles
Visualize the quality profile of the forward and reverse reads to determine trimming parameters.

```{r, "plot quality of reads", cache=TRUE}
# plot one forward and one reverse read in the rapport
plotQualityProfile(fnFs[1])+
  scale_y_continuous(limits = c(0,45))+
  scale_x_continuous(breaks = seq(0,300, 20))+
  geom_hline(yintercept = 30, colour = "grey", linetype = "dashed")+
  geom_hline(yintercept = 20, colour = "grey", linetype = "dashed")+
  labs(title = paste(sample.names[1], "foward", sep = " - " ))

plotQualityProfile(fnRs[1])+
  scale_y_continuous(limits = c(0,45))+
  scale_x_continuous(breaks = seq(0,300, 20))+
  geom_hline(yintercept = 30, colour = "grey", linetype = "dashed")+
  geom_hline(yintercept = 20, colour = "grey", linetype = "dashed")+
  labs(title = paste(sample.names[1], "reverse", sep = " - "))
```

The quality of the forwards reads is exceptionally high, staying at \>=Q40 over the full length. The quality of the reverse read is also equally high over the trimmed lengths with \>= Q40.


## Merge Read Lengths
Merge paired reads to assess overlap and read length distribution for optimal trimming decisions.

```{r}
# Merge the first file's forward and reverse reads using vsearch for length distribution analysis
 system(
  paste(
    "/Applications/vsearch/bin/vsearch --fastq_mergepairs ", 
    gsub("My Drive", "My\\\\ Drive", fnFs[1]), 
    " --reverse ", gsub("My Drive", "My\\\\ Drive", fnRs[1]), 
    " --fastaout", gsub("My Drive", "My\\\\ Drive", here("Data", primer, "test_merge.fasta")), 
    "--fastq_allowmergestagger",
    "--fastq_qmax 45"
    ), intern = TRUE
  )

# Read the merged sequences into R
file_merged <- readDNAStringSet(here("Data", primer, "test_merge.fasta"))

# Delete the temporary merged file after use
file.remove(here("Data", primer, "test_merge.fasta"))

# Create a table of sequence lengths and plot the distribution
table(width(file_merged)) %>%
  as.data.frame() %>%
  mutate(across(c("Var1", "Freq"), function(x) as.numeric(as.character(x)))) %>%  # Ensure columns are numeric
  arrange(desc(Freq)) %>%  # Sort by frequency
  mutate(
    prct = Freq / sum(Freq) * 100,  # Calculate the percentage of each length
    rank = 1:n(),  # Create a rank column
    Label = case_when(  # Add labels for the top 5 most frequent lengths
      rank < 6 ~ paste(Var1, "(", round(prct, 1), " %)", sep = ""),
      TRUE ~ NA_character_
    )
  ) %>%
  ggplot(aes(x = Var1, y = Freq)) +
  geom_line(colour = "darkgrey", alpha = 0.5) +  # Line plot for length distribution
  geom_text_repel(aes(label = Label)) +  # Annotate top labels
  geom_point(data = ~filter(.x, !is.na(Label)), colour = "red", size = 0.5) +  # Highlight labeled points
  theme_bw() +
  scale_x_continuous(limits = c(0, 500))  # Limit x-axis range
```

the merged fragments are 204 bp long

The trimmed forward and reverse reads are 204 and 204 bs long, respectively.

Thus we have a minimum overlap of `204+204-204` = 204 bps, i.e a full overlap.

given the exceptional quality of the trimmed pairs, we do not trim the reads.

## Set Read Parameters
Set trimming parameters based on observed read quality and overlap.

```{r}
# Set parameters for fragment and read lengths
frag_len <- 204
fwrd_len <- 204
rev_len <- 204
fwrd_trim <- 204
rev_trim <- 204   

# Create a plot showing the trimmed regions and overlap
ggplot(data = tibble(x = c(1, frag_len), y = c(1, 5)), aes(x = x, y = y)) +
  geom_blank() +  # Base blank plot to set up the axes
  geom_segment(x = 0, xend = frag_len, y = 2, yend = 2) +  # Full fragment line
  geom_segment(x = 0, xend = fwrd_len, y = 3, yend = 3) +  # Full forward read
  geom_segment(x = frag_len - rev_len, xend = frag_len, y = 3.5, yend = 3.5) +  # Full reverse read
  geom_segment(x = fwrd_trim, xend = fwrd_len, y = 3, yend = 3,  # Trimmed forward segment
               colour = "red", size = 2, alpha = 0.2) +
  geom_segment(x = frag_len - rev_trim, xend = frag_len - rev_len,  # Trimmed reverse segment
               y = 3.5, yend = 3.5, colour = "red", size = 3, alpha = 0.2) +
  geom_segment(x = frag_len - rev_trim, xend = fwrd_trim, y = 2,  # Overlap region
               yend = 2, colour = "green", size = 3, alpha = 0.2) +
  geom_text(x = ((frag_len - rev_trim) + fwrd_trim) / 2, y = 2.25,  # Label showing overlap length
            label = paste(as.character(fwrd_trim + rev_trim - frag_len), "bp"),
            colour = "blue", size = 5) +
  theme_void() +  # Removes background and axes for clarity
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +  # Adjust x-axis text orientation
  scale_x_continuous(breaks = seq(0, frag_len, 12))  # X-axis ticks
```

The filtering parameters we’ll use are stringent as the quality is very high and we have a lot of reads. 

+ maxN=0 (DADA2 requires no Ns)
+ truncQ=2 
+ maxEE=c(1,1)

The `maxEE` parameter sets the maximum number of “expected errors” allowed in a read. 

We use the `fastqPairedFilter` function to jointly filter the forward and reverse reads.

#DADA2 Pipeline

## Filter and Trim Reads
Filter and trim the reads based on quality, expected errors, and length settings. Write filtered read statistics to file.

```{r}
# Create file paths for the filtered forward and reverse reads, stored in the filtered/ subdirectory
filtFs <- file.path(here("Data", primer, "filtered"), paste0(sample.names, "_F_filt.fastq.gz"))  # Path for forward filtered reads
filtRs <- file.path(here("Data", primer, "filtered"), paste0(sample.names, "_R_filt.fastq.gz"))  # Path for reverse filtered reads

# Name the paths
names(filtFs) <- sample.names
names(filtRs) <- sample.names
```

```{r}
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, 
                     truncLen = c(fwrd_trim, rev_trim),
                     maxN=0,
                     maxEE=c(1,1),
                     truncQ=2,
                     rm.phix=TRUE,
                     compress=TRUE, multithread=6) 
out_df <- 
out %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = "Sample_name") %>% 
  mutate(Sample_name = gsub("_R1.+", "", Sample_name)) %>% 
  mutate(prct = signif(reads.out / reads.in * 100, 2)) %>% 
  mutate(Sample_name = gsub("(.+?)_S.+.fastq.gz", "\\1", Sample_name))

write_tsv(out_df, here("Data", primer, "out_intermediate", paste(primer, "ee_filter_stats.txt", sep = "_")))

out_df 
```

## Learn Errors
Estimate error rates from the filtered reads to inform the denoising step.

```{r}
# remove files with no reads left after filtering
files_passed_ee <- 
  which(out[,2] > 0) 
  
filtFs <- filtFs[files_passed_ee]
filtRs <- filtRs[files_passed_ee]
```

```{r}
errF <- learnErrors(
  filtFs,
  multithread = TRUE,
  nbases = 1e9,
  errorEstimationFunction = loessErrfun_mod4,
  verbose = TRUE
)

errR <- learnErrors(
  filtRs,
  multithread = TRUE,
  nbases = 1e9,
  errorEstimationFunction = loessErrfun_mod4,
  verbose = TRUE
)
```

```{r, cache=TRUE}
plotErrors(errF, nominalQ=TRUE)

plotErrors(errR, nominalQ=TRUE)
```

## Denoise
Apply DADA2 denoising to forward and reverse reads to identify unique sequences.

```{r}
filt <- list.files(filter_path, full.names = T)

filtFs <- filt[grepl("*_F_filt.fastq.gz", filt)]
names(filtFs) <- gsub(".+filtered/(.+?)_F.+", "\\1", filtFs)

filtRs <- filt[grepl("*_R_filt.fastq.gz", filt)]
names(filtRs) <- gsub(".+filtered/(.+?)_F.+", "\\1", filtRs)
```

```{r}
dadaFs <- dada(filtFs, err=errF, multithread=TRUE)
dadaRs <- dada(filtRs, err=errR, multithread=TRUE)
```

```{r}

denoise_stat <- 
sapply(dadaFs, function(x) sum(getUniques((x)))) %>% 
  data.frame(denoised = .) %>% 
  tibble::rownames_to_column(var = "Sample_name") %>% 
  left_join(select(out_df, Sample_name, reads.out)) %>% 
  mutate(prct = round(denoised / reads.out*100, 2))
  
write_tsv(denoise_stat, here("Data", primer, "out_intermediate", paste(primer, "denoise_stat.txt", sep = "_")))

denoise_stat
```

## Merge
Merge the forward and reverse reads and create a summary of merged reads for each sample.

```{r}

filtFs <- filtFs[names(dadaFs)]
filtRs <- filtRs[names(dadaRs)]

mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose=TRUE)

merge_stat <- 
lapply(mergers, function(x) {sum(x$abundance)}) %>% 
  as_tibble() %>% 
  pivot_longer(everything(),
    names_to = "Sample_name",
               values_to = "merged.out") %>% 
  left_join(select(out_df, Sample_name, reads.out)) %>% 
  mutate(prct = round(merged.out/reads.out * 100, 2))

merge_stat %>% 
write_tsv(here("Data", primer, "out_intermediate", paste(primer, "merge_stat.txt", sep = "_")))

merge_stat

```

## Construct Sequence Table
Generate the sequence table from the merged reads.

```{r, "Seqeunce table"}
seqtab <- makeSequenceTable(mergers)
dim(seqtab)
```

## Chimera Removal
Remove chimeric sequences from the sequence table and save chimera statistics to file.

```{r, "Chimera removal", cache=TRUE}

seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread=TRUE, verbose=TRUE)

write_rds(seqtab.nochim, here("Data", primer, "out_intermediate", paste(primer, "seqtab.nochim.rds", sep = "_")))

seqtab.nochim <- read_rds(here("Data", primer, "out_intermediate", paste(primer, "seqtab.nochim.rds", sep = "_")))

dim(seqtab.nochim)
```

```{r}
chimera_stat <- 
tibble(Sample_name = rownames(seqtab),
       before = rowSums(seqtab),
       after = rowSums(seqtab.nochim),
       prct = round(after/before*100,2))

chimera_stat %>% 
write_tsv(here("Data", primer,  "out_intermediate", paste(primer, "Chimera_stat.txt", sep = "_"))) 

chimera_stat
```

percentage of Chimeric reads

```{r}
(1-(sum(seqtab.nochim)/sum(seqtab)))*100
```

percentage of Chimeric ASVs

```{r}
(1-(ncol(seqtab.nochim)/ncol(seqtab)))*100
```

## Length Filtering
Visualize the length distribution of ASVs to confirm expected fragment size.

```{r}
data.frame(table(nchar(colnames(seqtab.nochim)))) %>% 
  mutate(across(c("Var1", "Freq"), function(x) as.numeric(as.character(x)))) %>%  # Ensure columns are numeric
  arrange(desc(Freq)) %>%  # Sort by frequency
  mutate(
    prct = Freq / sum(Freq) * 100,  # Calculate the percentage of each length
    rank = 1:n(),  # Create a rank column
    Label = case_when(  # Add labels for the top 5 most frequent lengths
      rank < 6 ~ paste(Var1, "(", round(prct, 1), " %)", sep = ""),
      TRUE ~ NA_character_
    )
  ) %>%
  ggplot(aes(x = Var1, y = Freq)) +
  geom_line(colour = "darkgrey", alpha = 0.5) +  # Line plot for length distribution
  geom_text_repel(aes(label = Label)) +  # Annotate top labels
  geom_point(data = ~filter(.x, !is.na(Label)), colour = "red", size = 0.5) +  # Highlight labeled points
  theme_bw() #+
  scale_x_continuous(limits = c(100, 300))  # Limit x-axis range
```
~50% of the reads are <204 bps long, the remaining reads mostly differ in the length of 1-3 triplets. I filter reads < 328 and > 362 bps
```{r}
sum(nchar(colnames(seqtab.nochim)) < 213)/ncol(seqtab.nochim)*100

seqtab.nochim <- seqtab.nochim[,nchar(colnames(seqtab.nochim)) > 328 & nchar(colnames(seqtab.nochim)) < 362]

dim(seqtab.nochim)
```


## Export Raw ASV and Fasta Files
Export the non-chimeric ASV sequences to a FASTA file and save the ASV abundance table as a TSV.


```{r}
# Convert the columns of the sequence table (ASVs) to a DNAStringSet object for export
seqs <- DNAStringSet(colnames(seqtab.nochim))

# Assign sequence names in the format: seq_001, seq_002, etc.
names(seqs) <- paste("seq", 
                     formatC(1:length(seqs), flag = "0", width = nchar(length(seqs))), 
                     sep = "_")

# Export the ASVs as a FASTA file, prefaced with the primer name
writeXStringSet(seqs, here("Data", primer, "out_intermediate", paste(primer, "ASVs.fasta", sep = "_")), format = "fasta")

# Prepare the ASV table for export as a TSV
ASV <- seqtab.nochim %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = "Sample") %>% 
  `colnames<-`(c("Sample", names(seqs)))

# Export the ASV table as a TSV file, prefaced with the primer name
write_tsv(ASV, here("Data", primer, "out_intermediate", paste(primer, "ASVs.txt", sep = "_")))

# Code to read the TSV (if needed for future processing)
# ASV <- read_tsv(here("Data", primer, "out_intermediate", paste(primer, "ASVs.txt", sep = "_")))
```


# Taxonomy Assignment

## blast
I use the a curated 16S database for taxonomic assignment. See reference below. I harmonized the taxonomy against the gbif backbone. I use blast for exact matching. 

https://doi.org/10.15156/BIO/2959341


```{r}

# Build an index of the reference database
system2( "/Applications/vsearch/bin/vsearch",
         args= c("--makeudb_usearch", 
                 shQuote(here("Data", primer, "in_files", "Midori_lr_gbif_sp.fa")),
                 "--output",
                 shQuote(here("Data", primer, "in_files", "Midori_lr_gbif_sp.udb"))
                 ))
 
```

We set maxrejects to 128 and maxaccepts to all to make sure we do not miss potential matches which could lead to incorrect assignments in the LCA taxonomy. 

```{r}
# Perform exact matching
system2( "/Applications/vsearch/bin/vsearch",
         args= c("--usearch_global", 
                 shQuote(here("Data", primer, "out_intermediate", 
                              paste(primer, "ASVs.fasta", sep = "_"))),
                 "--db",
                 shQuote(here("Data", primer, "in_files", "Midori_lr_gbif_sp.udb")),
                 "--id .95",
                 "--maxaccepts 0",
                 "--maxrejects 128",
                 "--blast6out",
                 shQuote(here("Data", primer, "out_intermediate", 
                              paste(primer, "taxblast_exact.txt", sep = "_"))),
                 "--threads 6"
                 ))
```

```{r}
#read in seqname taxonomy
seqname_to_tax <- read_tsv(here("Data", "Vertebrates", "in_files", "Midori_lr_gbif_tax.txt"))

#read in vsearch output
Vert_exact <- read_tsv(here("Data", primer, "out_intermediate",
                            paste(primer, "taxblast_exact.txt", sep = "_")), col_names = F)

colnames(Vert_exact) <- c("query_id","subject_id","%identity","alignment_length",
                         "mismatches","gap_opens","q_start","q_end","s_start",
                         "s_end","evalue","bit_score")

Vert_exact %>% 
  select(1:3) %>% 
  left_join(seqname_to_tax, by = c("subject_id" = "newSeqname")) %>% 
  select(-subject_id) %>% 
  distinct() %>% 
  arrange(query_id) %>% 
  group_by(query_id) %>%
  filter(!is.na(species)) %>% 
  filter(n()>1)
```

filter with Swiss occurence list
```{r}
SwissVert_gbif <- read_tsv(here("Data", primer, "in_files", "SwissVert_gbif.txt")) %>% 
  mutate(in_reflist = TRUE)

Vert_exact_sel <- 
Vert_exact %>% 
  select(1:3) %>% 
  left_join(seqname_to_tax, by = c("subject_id" = "newSeqname")) %>% 
  select(-subject_id) %>% 
  distinct() %>% 
  left_join(SwissVert_gbif) %>% 
  mutate(in_reflist = ifelse(is.na(in_reflist), FALSE, in_reflist))

# Function to apply the heuristic to each group
process_query <- function(data, key) {
  # data: dataframe for the current group (query_id)
  # key: list containing the group key (query_id)
  
  # Step 1: Check if any matches occur in reference database
  if (any(data$in_reflist == TRUE)) {
    # Keep only matches occurring in reference database
    data <- data %>% filter(in_reflist == TRUE)
  }
  # If no matches occur in reference database, keep all matches
  
  # Step 2: Keep matches with the highest %identity
  max_identity <- max(data$`%identity`)
  
  data <- data %>% filter(`%identity` == max_identity)
  
  # Step 3: Resolve ties with consensus taxonomy
  if (nrow(data) == 1) {
    # Only one best match remains
    return(data)
  } else {
    # Multiple matches with equal %identity
    # Create consensus taxonomy
    taxonomic_levels <- c("kingdom", "phylum", "class", "order", "family", "genus", "species")
    consensus <- data[1, ]  # Start with the first match
    
    for (level in taxonomic_levels) {
      level_values <- data[[level]]
      # Remove NAs
      non_na_values <- level_values[!is.na(level_values)]
      # Unique non-NA values
      unique_values <- unique(non_na_values)
      
      if (length(unique_values) == 1) {
        # All non-NA values are the same
        consensus[[level]] <- unique_values
      } else {
        # Disagreement at this level
        consensus[[level]] <- NA
      }
    }
    
    # Update `in_reflist` field
    consensus$in_reflist <- any(data$in_reflist == TRUE)
    # Update `%identity` field
    consensus$`%identity` <- max_identity
    
    return(consensus)
  }
}

# Apply the function to each group
Vert_exact_sel <- Vert_exact_sel %>%
  group_by(query_id) %>%
  group_modify(~ process_query(.x, .y)) %>%
  ungroup()

#export
Vert_exact_sel %>% 
  dplyr::rename(prct_id = '%identity',
                seq = query_id) %>% 
  write_tsv(here("Data", primer, "out_intermediate", paste(primer, "Tax_blast_LCA.txt", sep = "_")))
```


## Merge and Aggregate ASV by Taxonomic Level
Use phyloseq to merge ASVs with the same species assignment.


```{r}
#define taxonomic aggregation level
tax_glom_level <- "species"

#read the blast LCA taxonomy
# replace missing species with genus sp. to agglomerate at genus level if species level is missing
Vert_blast_tax <- read_tsv(here("Data", primer, "out_intermediate", 
                           paste(primer, "Tax_blast_LCA.txt", sep = "_"))) %>% 
  mutate(species = case_when(
    is.na(species) & !is.na(genus) ~ paste(genus, "sp."),
    TRUE ~ species)) %>% 
  select(!one_of(c("iucnRedListCategory", "numberOfOccurrences", "inGbif", "in_reflist")))

# Read the ASV abundance table
ASV <- read_tsv(here("Data", primer, "out_intermediate", paste(primer, "ASVs.txt", sep = "_")))

# exclude ASV sequences not asisgned in blast
ASV <- ASV %>% select(one_of("Sample", Vert_blast_tax$seq))

# Prepare ASV matrix and set sample names as row names
ASV_ps <- ASV[, -1]  # Remove the first column (sample identifiers)
rownames(ASV_ps) <- ASV$Sample

# Filter and prepare the taxonomic table 
Vert_blast_tax_ps <-
  Vert_blast_tax %>%
  filter(!is.na(!!sym(tax_glom_level))) %>%
  select(-seq, -prct_id) %>%
  as.matrix()

# Set sequence names as row names in the taxonomic matrix
rownames(Vert_blast_tax_ps) <- filter(Vert_blast_tax, !is.na(!!sym(tax_glom_level)))$seq

# Create a phyloseq object using the ASV and taxonomic data
ps_obj <- phyloseq(otu_table(ASV_ps, taxa_are_rows = FALSE),
                   tax_table(Vert_blast_tax_ps))

# Aggregate ASVs at the selected level
ps_glom <- tax_glom(ps_obj, taxrank = tax_glom_level, NArm = FALSE)

# Convert the aggregated taxonomic table to a data frame
Vert_blast_tax_glom <- as.data.frame(tax_table(ps_glom)) %>%
  tibble::rownames_to_column("seq")

# Convert the aggregated ASV table to a data frame
ASW_glom <- as.data.frame(otu_table(ps_glom)) %>%
  tibble::rownames_to_column("Sample")

# Merge and adjust columns in the final ASV table
ASW_glom_j <- ASV %>%
  select(!one_of(rownames(Vert_blast_tax_ps))) %>%
  left_join(ASW_glom, .)

# Sort columns alphabetically for consistency
ASW_glom_j <- ASW_glom_j[sort(colnames(ASW_glom_j))]

# Combine the original taxonomic table with the merged table
Vert_blast_tax_glom_j <- Vert_blast_tax %>%
  select(-prct_id, ) %>% 
  filter(!seq %in% rownames(Vert_blast_tax_ps)) %>%
  bind_rows(Vert_blast_tax_glom, .) %>%
  arrange(seq)

# Prepare a sequence table with just the representative sequences
seqs_glom <- seqs[Vert_blast_tax_glom_j$seq]

dim(ASV)
dim(ASW_glom_j)
```
# Export Final Aggregated Data
Export the aggregated ASV and taxonomic data tables

```{r}
# Export the processed taxonomic table with genus-level aggregation
write_tsv(Vert_blast_tax_glom_j, here("Data", primer, "out_final", paste(primer, "taxa_blast_glom.txt", sep = "_")))

# Export the aggregated ASV table
write_tsv(ASW_glom_j, here("Data", primer, "out_final", paste(primer, "ASV_glom.txt", sep = "_")))

# Export the subset sequence table
writeXStringSet(seqs_glom, here("Data", primer, "out_final", paste(primer, "Seq_glom", sep = "_")))
```